%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

% 
% 
%\chapter{Preface}
% this suppresses an explicit chapter number for the preface.
\chapter*{Preface}%\normalsize
  \addcontentsline{toc}{chapter}{Preface}

This document is based on my lecture notes for the Fall 2012, University of Toronto Advanced Classical Optics course (PHY485H1F), taught by Prof.\ Joseph H. Thywissen.  

%My thanks to Professor Thywissen for teaching this course...

Official course description:

This course builds on a student's knowledge of basic electromagnetic theory by focusing attention on light including elementary aspects of the propagation of optical beams and their interaction with matter.  We examine light polarization, coherence, interference and diffraction as we move towards a description of lasers within a semiclassical picture in which the fields are treated classically and matter is treated quantum mechanically.  In between we discuss Gaussian beam modes and their relation to optical resonators as well as fibre and slab waveguides.

In the course of the lectures the books \citep{hecht1998hecht}, \citep{fowles1989introduction}, and \citep{born1980principles} are referred to, but none were required texts for the course.

This document will probably end up containing a few things

\begin{itemize}
\item Plain old lecture notes.

\item Personal notes exploring details that were not clear to me from the lectures, or from the texts associated with the lecture material.

\item Assigned problems.

\item Some worked problems attempted as course prep, for fun, or for exam preparation.

\item Links to Mathematica workbooks associated with course content.

\end{itemize}

Peeter Joot  \quad peeter.joot@gmail.com 
