%
% Copyright © 2015 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%
%\chapter{Preface}
% this suppresses an explicit chapter number for the preface.
\chapter*{Preface}%\normalsize
  \addcontentsline{toc}{chapter}{Preface}

This document was produced while taking the Winter 2015, University of Toronto Graduate Quantum Mechanics course (PHY1520H), taught by Prof.\ Arun Paramekanti.

\paragraph{Course Syllabus}

This course will discuss the following topics in quantum mechanics (time permitting)

\begin{enumerate}
\item Basics - Postulates, Wavefunctions, Density matrices, Measurements
\item Time evolution - Schrodinger picture, Heisenberg picture, Interaction picture
\item Harmonic oscillator - Operator method, Wavefunctions, Coherent states
\item Particle in a magnetic field - Local gauge invariance, 2D Landau levels
\item Symmetries - Parity, Translations, Rotations, Time-reversal
\item Angular momentum, Spin, and Angular momentum addition
\item Time-independent perturbation theory
\item Time-dependent perturbation theory
\item Variation approach
\item Scattering theory
\item Dirac equation - one dimension
\item Path integrals
\end{enumerate}

% Texts include
% \begin{itemize}
	% \item [Required Text]
% J. J. Sakurai, Modern Quantum Mechanics, 2nd edition. Addison Wesley 1993.
	% \item [Optional Text]
% K. Gottfried and T.-M. Yan, Quantum Mechanics (Springer, New York, 2004), 2nd Edition.
% %	\item [Reference books]
% %M. Le Bellac, Quantum Physics (Cambridge University Press, New York, 2006).
% %C. Cohen-Tannoudji, B. Diu & F. Laloe, Quantum Mechanics, Vol. 1-2, Wiley-VCH.
% %R. Shankar, Principles of Quantum Mechanics, 2nd edition, Plenum Press
% %E. Merzbacher, Quantum Mechanics, 3rd edition, Wiley.
% \end{itemize}

\withproblemsetsMessage{
\textcolor{Maroon}{
\textit{THIS DOCUMENT IS REDACTED.  THE PROBLEM SET SOLUTIONS AND ASSOCIATED MATHEMATICA CODE IS NOT VISIBLE.  PLEASE EMAIL ME FOR THE FULL VERSION IF YOU ARE NOT TAKING PHY1520.}
}
}

\paragraph{This document contains:}

\begin{itemize}
\item Lecture notes.
\item Personal notes exploring auxiliary details.
\item Worked practice problems.

\ifthenelse{\boolean{redacted}}%
{%
\item Links to Mathematica notebooks associated with the course material and problems (but not problem sets).
}%
{
\item Assigned problems.%
\item Links to Mathematica notebooks associated with problems and course material.%
}
\end{itemize}

My thanks go to Professor Paramekanti for teaching this course, and to Nishant Bhatt for providing me with a copy his notes for lecture 18, which are incorporated herein.

Peeter Joot  \quad peeterjoot@protonmail.com
