%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%
%
%\chapter{Preface}
% this suppresses an explicit chapter number for the preface.
\chapter*{Preface}%\normalsize
  \addcontentsline{toc}{chapter}{Preface}

These are my personal lecture notes for the Spring 2012, University of Toronto Advanced Classical Mechanics course (PHY354H1S).  This class was taught by Prof.\ Erich Poppitz.  I audited the Wednesday lectures since the timing was convenient.  I took occasional notes, and did some of the problems.

The official description of this course was:

Symmetry and conservation laws, stability and instability, generalized co-ordinates, Hamilton's principle, Hamilton's equations, phase space, Liouville's theorem, canonical transformations, Poisson brackets, Noether's theorem.

The text for the course is \citep{landau1976mechanics}.  Before purchasing that I studied from \cite{goldstein1951cm}, also a great little book.

This document contains a few things

\begin{itemize}
\item A collection of miscellaneous notes and problems for my personal (independent) classical mechanics studies.

\item My notes for some of the lectures I attended (some of the earlier lectures I did not actively take notes for).

Typos and errors are probably mine (Peeter), and no claim nor attempt of spelling or grammar correctness will be made.  

These notes, when I took them, likely track along with the Professor's hand written notes very closely, since his lectures follow his notes very closely.

\item Some assigned problems from the PHY354 course, ungraded (not submitted since I did not actually take the course).

\item Miscellaneous worked problems from other sources.

\end{itemize}

Peeter Joot  \quad peeter.joot@gmail.com 
