%\input{../peeter_prologue_print.tex}
\input{../peeter_prologue_widescreen.tex}

\chapter{1D forced harmonic oscillator.  Quick solution of non-homogeneous problem.}
\label{chap:1dharmonicOsc}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2010/1dharmonicOsc.pdf}
\date{Feb 19, 2010}
\revisionInfo{1dharmonicOsc.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Motivation.}

In \cite{brown1954feynman} equation (25) we have a forced harmonic oscillator equation

\begin{align}\label{eqn:1dharmonicOsc:1}
m \ddot{x} + m \omega^2 x = \gamma(t).
\end{align}

The solution of this equation is provided, but for fun lets derive it.

\section{Guts}

Writing

\begin{align}\label{eqn:1dharmonicOsc:2}
\omega u = \dot{x},
\end{align}

we can rewrite the second order equation as a first order linear system 

\begin{align}\label{eqn:1dharmonicOsc:3}
\dot{u} + \omega x &= \gamma(t)/m \omega \\
\dot{x} - \omega u &= 0,
\end{align}

Or, with $X = (u, x)$, in matrix form

\begin{align}\label{eqn:1dharmonicOsc:4}
\dot{X} + \omega 
\begin{bmatrix}
0 & 1 \\
-1 & 0
\end{bmatrix}
X
&=
\begin{bmatrix}
\gamma(t)/m \omega \\
0
\end{bmatrix}.
\end{align}

The two by two matrix has the same properties as the complex imaginary, squaring to the identity matrix, so the equation to solve is now of the form

\begin{align}\label{eqn:1dharmonicOsc:5}
\dot{X} + \omega i X &= \Gamma.
\end{align}

The homogeneous part of the solution is just the matrix

\begin{align*}
X 
&= e^{-i \omega t} A \\
&= 
\left( 
\cos(\omega t) 
\begin{bmatrix}
1 & 0 \\
0 & 1
\end{bmatrix}
- 
\sin(\omega t)
\begin{bmatrix}
0 & 1 \\
-1 & 0
\end{bmatrix}
\right) A,
\end{align*}

where $A$ is a two by one column matrix of constants.  Assuming for the specific solution $X = e^{-i \omega t} A(t)$, and substuiting we have

\begin{align}\label{eqn:1dharmonicOsc:6}
e^{-i \omega t} \dot{A} = \Gamma(t).
\end{align}

This integrates directly, fixing the unknown column vector function $A(t)$

\begin{align}\label{eqn:1dharmonicOsc:7}
A(t) = A(0) + \int_0^t e^{i \omega \tau} \Gamma(\tau).
\end{align}

Thus the non-homogeneous solution takes the form

\begin{align}\label{eqn:1dharmonicOsc:8}
X = e^{-i \omega t} A(0) + \int_0^t e^{i \omega (\tau - t)} \Gamma(\tau).
\end{align}

Note that $A(0) = (\dot{x}_0/\omega, x_0)$.  Multiplying this out, and discarding all but the second row of the matrix product gives $x(t)$, and Feynman's equation (26) follows directly.

\EndArticle
%\EndNoBibArticle
