\input{../peeter_prologue.tex}

\chapter{Integrating the equation of motion for a one dimensional problem.}
\label{chap:1dpotentialIntegral}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2010/1dpotentialIntegral.pdf}
\date{Jan 1, 2010}
\revisionInfo{1dpotentialIntegral.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Motivation.}

While linear approximations, such as the small angle approximation for the pendum, are often used to understand the dynamics of non-linear systems, exact solutions may be possible in some cases.  Walk through the setup for such an exact solution.

\section{Guts}

The equation to consider solutions of has the form

\begin{align}
\label{eqn:1dpotentialIntegral:1}
\frac{d}{dt} \left( m \frac{dx}{dt} \right) = -\PD{x}{U(x)}.
\end{align}

We have an unpleasant mix of space and time derivatives, preventing any sort of antidifferentiation.  Assuming constant mass $m$, and employing the chain rule a refactoring in terms of velocities is possible.

\begin{align*}
\frac{d}{dt} \left( \frac{dx}{dt} \right) 
&= 
\frac{dx}{dt} \frac{d}{dx} \left( \frac{dx}{dt} \right)  \\
&= 
\inv{2} \frac{d}{dx} \left( \frac{dx}{dt} \right)^2  \\
\end{align*}

The one dimensional Newton's law \autoref{eqn:1dpotentialIntegral:1} now takes the form
\begin{align}
\label{eqn:1dpotentialIntegral:2}
\frac{d}{dx} \left( \frac{dx}{dt} \right)^2 &= -\frac{2}{m} \PD{x}{U(x)}.
\end{align}

This can now be antidifferentiated for

\begin{align}\label{eqn:1dpotentialIntegral:3}
\left( \frac{dx}{dt} \right)^2 &= \frac{2}{m} (E - U(x)).
\end{align}

NOTE: this is a dumb way to get here.  This is just the Hamiltonian, rearranged, so one could start with the physics, instead of the calculus to get to this point.

Taking roots and rearranging produces an implicit differential form $x$ in terms of time

\begin{align}\label{eqn:1dpotentialIntegral:4}
dt = \frac{dx}{\sqrt{ \frac{2}{m} (E - U(x)) } }.
\end{align}

One can concievably integrate this and invert to solve for position as a function of time, but substitution of a more specific potential is required to go further.

\begin{align}\label{eqn:1dpotentialIntegral:5}
t(x) = t(x_0) + \int_{y=x_0}^{x} \frac{dy}{\sqrt{ \frac{2}{m} (E - U(y)) } }.
\end{align}


%\EndArticle
\EndNoBibArticle
