%
% Copyright © 2015 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\makeproblem{Long thin wire dipoles}{advancedantenna:problemSet3:2}{ 

\makesubproblem{}{advancedantenna:problemSet3:2a}
On a single diagram, plot the polar patterns for 
\( l = 0.5 \lambda, l = 1.0 \lambda, l = 1.25 \lambda \) and \( l = 2.0 \lambda \) long thin wire dipole antennas.

\makesubproblem{}{advancedantenna:problemSet3:2b}
Use numerical integration to calculate the maximum directivity for each dipole.
Make a table with your results. Which length corresponds to the highest
directivity?

\makesubproblem{}{advancedantenna:problemSet3:2c}
Use numerical integration to calculate the radiation resisistance of the  \( l = 1.25 \lambda \) dipole. Do you expect this dipole to be capacitive or inductive?
} % makeproblem

\makeanswer{advancedantenna:problemSet3:2}{ 
\makeSubAnswer{}{advancedantenna:problemSet3:2a}

Assuming a \( \zcap \) oriented dipole, in the far field, the electric field is

\begin{dmath}\label{eqn:advancedantennaProblemSet3Problem2:20}
E_\theta \approx j \eta \frac{I_0 e^{-j k r}}{ 2 \pi r } 
\lr{ 
\frac{\cos\lr{ \frac{k l}{2} \cos\theta} - \cos\lr{\frac{kl}{2}}}{\sin\theta}
},
\end{dmath}

with the magnetic field given by \( H_\phi \approx E_\theta/\eta \), and an average Poynting vector of

\begin{dmath}\label{eqn:advancedantennaProblemSet3Problem2:40}
\BW_{\textrm{av}} = 
\rcap 
\eta \frac{\Abs{I_0}^2}{8 \pi^2 r^2}
\lr{ 
\frac{\cos\lr{ \frac{k l}{2} \cos\theta} - \cos\lr{\frac{kl}{2}}}{\sin\theta}
}^2.
\end{dmath}

Writing \( l = \alpha \lambda \), the radiation intensity \( U = r^2 W_{\textrm{av}} \) is

\begin{dmath}\label{eqn:advancedantennaProblemSet3Problem2:60}
U = 
\eta \frac{\Abs{I_0}^2}{8 \pi^2 r^2}
\lr{ 
\frac{\cos\lr{ \pi \alpha \cos\theta} - \cos\lr{ \pi \alpha }}{\sin\theta}
}^2.
\end{dmath}

\makeSubAnswer{}{advancedantenna:problemSet3:2b}

TODO.
\makeSubAnswer{}{advancedantenna:problemSet3:2c}

TODO.
}
