\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Lorentz transformation of an antisymmetric tensor.}
\label{chap:antisymmetricTensorTx}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/antisymmetricTensorTx.pdf}
\date{Jan 14, 2011}
\revisionInfo{antisymmetricTensorTx.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

I have an ancient copy of \cite{landau1980classical} from the library right now (mine is on order still) for my PHY450H1S course (relativistic electrodynamics).  Given the transformation rule for a first rank tensor

\begin{equation}\label{eqn:antisymmetricTensorTx:5}
A_{i} = \alpha_{im} A'_{m},
\end{equation}

they list the transformation rule for a second rank tensor as

\begin{equation}\label{eqn:antisymmetricTensorTx:10}
A_{ik} = \alpha_{im} \alpha_{kl} A'_{ml}.
\end{equation}

This isn't motivated in any way.  Let's compare to transformation of a bivector expressed in the Dirac basis, transformed by outermorphism.  That's specifically a transformation of a antisymmetric tensor (once expressed in components anyways), but should provide some intuition.

It is also worthwhile to note that there are some old fashioned notational quirks in this text (at least the old version that I have currently borrowed).  Specifically, they uses Latin indexes four vectors with Greek indexes for three vectors, completely opposite to what appears to be the current conventions.  They also don't use upper and lower indexes to keep track of bookkeeping.  I'll use the conventions I'm used to for now.

\section{Notation and use of Geometric Algebra herein.}

I'll use conventions from \cite{doran2003gap} using the Dirac basis, with a preference for index upper coordinates, and express a vector as

\begin{equation}\label{eqn:antisymmetricTensorTx:20}
x = x^\alpha \gamma_\alpha = x_\alpha \gamma^\alpha,
\end{equation}

Here the basis pairs $\{\gamma_\mu\}$ and $\{\gamma^\mu\}$ are reciprocal frames with $\gamma^\mu \cdot \gamma_\nu = {\delta^\mu}_\nu$.  I'll have no need for any specific metric convention here.

The dot and wedge products used will be defined in terms of their Clifford Algebra formulation

\begin{align}\label{eqn:antisymmetricTensorTx:25}
a \cdot b &= \inv{2} (a b + b a) \\
a \wedge b &= \inv{2} (a b - b a).
\end{align}

The dot product between two bivectors $A$, $B$ will also be used, defined as the scalar part of the product $AB$.  In particular the identity for extraction of that scalar component from the dot product of two wedge products will be required

\begin{equation}\label{eqn:antisymmetricTensorTx:26}
(a \wedge b ) \cdot (c \wedge d)
= ( a (b \cdot c) - b ( a \cdot c) ) \cdot d
= (a \cdot d) (b \cdot c) - (b \cdot d) ( a \cdot c) 
\end{equation}

\section{Transformation of the coordinates.}

Let's assume our transformation is linear, and we will denote its action on vectors as follows

\begin{equation}\label{eqn:antisymmetricTensorTx:30}
x' = L(x) = x^\alpha L( \gamma_\alpha).
\end{equation}

Extracting coordinates for the transformed coordinates (assuming a non-moving frame where the unit vectors on both sides are the same), we have after dotting with $\gamma^\mu$

\begin{equation}\label{eqn:antisymmetricTensorTx:40}
{x'}^\mu = \left( {x'}^\alpha \gamma_\alpha \right) \cdot \gamma^\mu
= x^\alpha \left( L( \gamma_\alpha) \cdot \gamma^\mu \right) 
\end{equation}

Now introduce a coordinate representation for the transformation $L$ 

\begin{equation}\label{eqn:antisymmetricTensorTx:50}
L( \gamma_\alpha) \cdot \gamma^\mu  = {L_\alpha}^\mu,
\end{equation}

so our transformation rule for the four vector coordinates becomes

\begin{equation}\label{eqn:antisymmetricTensorTx:55}
{x'}^\mu = x^\alpha {L_\alpha}^\mu.
\end{equation}

We are now ready to look at the transformation of a bivector (a quantity having a rank two antisymmetric tensor representation in coordinates), and see how the coordinates transform.

Let's transform by outermorphism of the transformed vector factors the bivector 

\begin{equation}\label{eqn:antisymmetricTensorTx:60}
c = a \wedge b \rightarrow a' \wedge b'.
\end{equation}

First we'll need the coordinate representation of the bivector before transformation.  We dot with $\gamma^\nu \wedge \gamma^\mu$ to pick up the desired term

\begin{align*}
(a \wedge b) \cdot (\gamma_\nu \wedge \gamma_\mu)
&=
a^\alpha b^\beta (\gamma_\alpha \wedge \gamma_\beta) \cdot (\gamma^\nu \wedge \gamma^\mu) \\
&=
a^\alpha b^\beta ( \gamma_\alpha {\delta_\beta}^\nu -\gamma_\beta {\delta_\alpha}^\nu ) \cdot \gamma^\mu \\
&=
a^\alpha b^\beta ( {\delta_\alpha}^\mu {\delta_\beta}^\nu -{\delta_\beta}^\mu {\delta_\alpha}^\nu ) \\
&=
a^\mu b^\nu 
-a^\nu b^\mu \\
\end{align*}

If we introduce a rank two tensor now, say

\begin{equation}\label{eqn:antisymmetricTensorTx:70}
T^{\mu\nu} = a^\mu b^\nu -a^\nu b^\mu,
\end{equation}

we recover our bivector with
\begin{equation}\label{eqn:antisymmetricTensorTx:80}
a \wedge b = \inv{2} T^{\alpha \beta} \gamma_\alpha \wedge \gamma_\beta.
\end{equation}

Now let's look at the coordinate representation of the transformed bivector.  It will also be helpful to make use of the identity that can be observed above from the initial coordinate extraction

\begin{equation}\label{eqn:antisymmetricTensorTx:90}
(\gamma_\alpha \wedge \gamma_\beta) \cdot (\gamma^\nu \wedge \gamma^\mu) = {\delta_\alpha}^\mu {\delta_\beta}^\nu -{\delta_\beta}^\mu {\delta_\alpha}^\nu 
\end{equation}

In coordinates our transformed bivector is

\begin{equation}\label{eqn:antisymmetricTensorTx:100}
a' \wedge b' =
a^\sigma {L_\sigma}^\alpha
b^\pi {L_\pi}^\beta
\gamma_\alpha \wedge \gamma_\beta,
\end{equation}

and we can proceed with the coordinate extraction by taking dot products with $\gamma^\nu \wedge \gamma^\mu$ as before.  This gives us

\begin{align*}
( a' \wedge b' ) \cdot (\gamma^\nu \wedge \gamma^\mu)
&=
a^\sigma {L_\sigma}^\alpha
b^\pi {L_\pi}^\beta
\gamma_\alpha \wedge \gamma_\beta \\
&=
a^\sigma {L_\sigma}^\alpha
b^\pi {L_\pi}^\beta
( {\delta_\alpha}^\mu {\delta_\beta}^\nu -{\delta_\beta}^\mu {\delta_\alpha}^\nu  ) \\
&=
a^\sigma {L_\sigma}^\mu
b^\pi {L_\pi}^\nu
-
a^\sigma {L_\sigma}^\nu
b^\pi {L_\pi}^\mu \\
&=
a^\sigma {L_\sigma}^\mu
b^\pi {L_\pi}^\nu
-
a^\pi {L_\pi}^\nu
b^\sigma {L_\sigma}^\mu \\
&=
(a^\sigma b^\pi - a^\pi b^\sigma) {L_\sigma}^\mu {L_\pi}^\nu
\\
&=
T^{\sigma \pi} {L_\sigma}^\mu {L_\pi}^\nu
\\
\end{align*}

We are able to conclude that the bivector coordinates transform as

\begin{equation}\label{eqn:antisymmetricTensorTx:200}
T^{\mu \nu} \rightarrow T^{\sigma \pi} {L_\sigma}^\mu {L_\pi}^\nu.
\end{equation}

Except for the lowering index differences this verifies the rule \ref{eqn:antisymmetricTensorTx:10} from the text.

It would be reasonable seeming to impose such a tensor transformation rule on any antisymmetric rank 2 tensor, and in the text this is also imposed as the rule for transformation of symmetric rank 2 tensors.  Do we have a simple example of a rank 2 symmetric tensor that can be expressed geometrically?  The only one that comes to mind off the top of my head is the electrodynamic stress tensor, which isn't exactly simple to work with.

\section{Lorentz transformation of the metric tensors.}

Following up on the previous thought, it is not hard to come up with an example of a symmetric tensor a whole lot simpler than the electrodynamic stress tensor.  The metric tensor is probably the simplest symmetric tensor, and we get that by considering the dot product of two vectors.  Taking the dot product of vectors $a$ and $b$ for example we have

\begin{equation}\label{eqn:antisymmetricTensorTx:300}
a \cdot b 
= a^\mu b^\nu \gamma_\mu \cdot \gamma_\nu
\end{equation}

From this, the metric tensors are defined as

\begin{align}\label{eqn:antisymmetricTensorTx:310}
\eta_{\mu\nu} &= \gamma_\mu \cdot \gamma_\nu \\
\eta^{\mu\nu} &= \gamma^\mu \cdot \gamma^\nu
\end{align}

These are both symmetric and diagonal, and in fact equal (regardless of whether one picks a $+,-,-,-$ or $-,+,+,+$ signature for the space).

Let's look at the transformation of the dot product, utilizing the transformation of the four vectors being dotted to do so.  By definition, when both vectors are equal, we have the (squared) spacetime interval, which based on the speed of light being constant, has been found to be an invariant under transformation.

\begin{equation}\label{eqn:antisymmetricTensorTx:320}
a' \cdot b'
= 
a^\mu b^\nu L(\gamma_\mu) \cdot L(\gamma_\nu)
\end{equation}

We note that, like any other vector, the image $L(\gamma_\mu)$ of the Lorentz transform of the vector $\gamma_\mu$ can be written as

\begin{equation}\label{eqn:antisymmetricTensorTx:330}
L(\gamma_\mu) = \left( L(\gamma_\mu) \cdot \gamma^\nu \right) \gamma_\nu
\end{equation}

Similarily we can write any vector in terms of the reciprocal frame
\begin{equation}\label{eqn:antisymmetricTensorTx:340}
\gamma_\nu = (\gamma_\nu \cdot \gamma_\mu) \gamma^\mu.
\end{equation}

The dot product factor is a component of the metric tensor

\begin{equation}\label{eqn:antisymmetricTensorTx:350}
\eta_{\nu \mu} = \gamma_\nu \cdot \gamma_\mu,
\end{equation}

so we see that the dot product transforms as

\begin{equation}\label{eqn:antisymmetricTensorTx:320b}
a' \cdot b' 
= a^\mu b^\nu 
( L(\gamma_\mu) \cdot \gamma^\alpha ) 
( L(\gamma_\nu) \cdot \gamma^\beta ) 
\gamma_\alpha
\cdot
\gamma_\beta
= a^\mu b^\nu 
{L_\mu}^\alpha
{L_\nu}^\beta
\eta_{\alpha \beta}
\end{equation}

In particular, for $a = b$ where we have the invariant interval defined by the condition $a^2 = {a'}^2$, we must have

\begin{equation}\label{eqn:antisymmetricTensorTx:360}
a^\mu a^\nu \eta_{\mu \nu}
= a^\mu a^\nu 
{L_\mu}^\alpha
{L_\nu}^\beta
\eta_{\alpha \beta}
\end{equation}

This implies that the symmetric metric tensor transforms as

\begin{equation}\label{eqn:antisymmetricTensorTx:370}
\eta_{\mu\nu}
=
{L_\mu}^\alpha
{L_\nu}^\beta
\eta_{\alpha \beta}
\end{equation}

Recall from \ref{eqn:antisymmetricTensorTx:200} that the coordinates representation of a bivector, an antisymmetric quantity transformed as

\begin{equation}\label{eqn:antisymmetricTensorTx:200b}
T^{\mu \nu} \rightarrow T^{\sigma \pi} {L_\sigma}^\mu {L_\pi}^\nu.
\end{equation}

This is a very similar transformation, but differs from the bivector case where our free indexes were upper indexes.  Suppose that we define an alternate set of coordinates for the Lorentz transformation.  Let

\begin{equation}\label{eqn:antisymmetricTensorTx:380}
{L^\mu}_\nu = L(\gamma^\mu) \cdot \gamma_\nu.
\end{equation}

This can be related to the previous coordinate matrix by
\begin{equation}\label{eqn:antisymmetricTensorTx:390}
{L^\mu}_\nu = \eta^{\mu \alpha } \eta_{\nu \beta } {L_\alpha}^\beta. 
\end{equation}

If we examine how the coordinates of $x^2$ transform in thier lower index representation we find

\begin{equation}\label{eqn:antisymmetricTensorTx:400}
{x'}^2 = x_\mu x_\nu {L^\mu}_\alpha {L^\nu}_\beta \eta^{\alpha \beta} = x^2 = x_\mu x_\nu \eta^{\mu \nu},
\end{equation}

and therefore find that the (upper index) metric tensor transforms as

\begin{equation}\label{eqn:antisymmetricTensorTx:410}
\eta^{\mu \nu} \rightarrow
\eta^{\alpha \beta}
{L^\mu}_\alpha {L^\nu}_\beta .
\end{equation}

Compared to $\ref{eqn:antisymmetricTensorTx:200b}$ we have almost the same structure of transformation.  Are these the same?  Does the notation I picked here introduce an apparent difference that does not actually exist?  We really want to know if we have the identity

\begin{equation}\label{eqn:antisymmetricTensorTx:420}
L(\gamma_\mu) \cdot \gamma^\nu
\stackrel{?}{=}
L(\gamma^\nu) \cdot \gamma_\mu,
\end{equation}

which given the notation selected would mean that ${L_\mu}^\nu = {L^\nu}_\mu$, and justify a notational simplification ${L_\mu}^\nu = {L^\nu}_\mu = L^\nu_\mu$.

\section{The inverse Lorentz transformation}

To answer this question, let's consider a specific example, an x-axis boost of rapidity $\alpha$.  For that our Lorentz transformation takes the following form

\begin{equation}\label{eqn:antisymmetricTensorTx:430}
L(x) = e^{-\sigma_1 \alpha/2} x e^{\sigma_1 \alpha/2},
\end{equation}

where $\sigma_k = \gamma_k \gamma_0$.  Since $\sigma_1$ anticommutes with $\gamma_0$ and $\gamma_1$, but commutes with $\gamma_2$ and $\gamma_3$, we have

\begin{equation}\label{eqn:antisymmetricTensorTx:440}
L(x) = (x^0 \gamma_0 + x^1 \gamma_1) e^{\sigma_1 \alpha} + x^2 \gamma_2 + x^3 \gamma_3,
\end{equation}

and after expansion this is
\begin{equation}\label{eqn:antisymmetricTensorTx:441}
L(x) = 
\gamma_0 ( x^0 \cosh \alpha - x^1 \sinh \alpha ) 
+\gamma_1 ( x^1 \cosh \alpha - x^0 \sinh \alpha )
+\gamma_2
+\gamma_3.
\end{equation}

Note that this is the first time a specific metric preference has been imposed, and $+,-,-,-$ has been used.

Observe that for the basis vectors themselves we have

\begin{equation}\label{eqn:antisymmetricTensorTx:450}
\begin{bmatrix}
L(\gamma_0) \\
L(\gamma_1) \\
L(\gamma_2) \\
L(\gamma_3)
\end{bmatrix}
=
\begin{bmatrix}
\gamma_0 \cosh \alpha - \gamma_1 \sinh \alpha \\
-\gamma_0 \sinh \alpha + \gamma_1 \cosh \alpha \\
\gamma_2 \\
\gamma_3
\end{bmatrix}
\end{equation}

Forming a matrix with $\mu$ indexing over rows and $\nu$ indexing over columns we have

\begin{equation}\label{eqn:antisymmetricTensorTx:460}
{L_\mu}^\nu =
\begin{bmatrix}
\cosh \alpha &- \sinh \alpha & 0 & 0 \\
-\sinh \alpha & \cosh \alpha & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1
\end{bmatrix}
\end{equation}

Performing the same expansion for ${L^\nu}_\mu$, again with $\mu$ indexing over rows, we have

\begin{equation}\label{eqn:antisymmetricTensorTx:470}
{L^\nu}_\mu =
\begin{bmatrix}
\cosh \alpha & \sinh \alpha & 0 & 0 \\
\sinh \alpha & \cosh \alpha & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1
\end{bmatrix}.
\end{equation}

This answers the question.  We cannot assume that ${L_\mu}^\nu = {L^\nu}_\mu$.  In fact, in this particular case, we have ${L^\nu}_\mu = ({L_\mu}^\nu)^{-1}$.  Is that a general condition?  Note that for the general case, we have to consider compounded transformations, where each can be a boost or rotation.

Let's come back to this, and temporarily consider the question of duality first.

\section{Duality in tensor form.}

Let's consider the subject of duality to antisymmetric forms.  Within a geometric algebra context our duality is provided by multiplication by the pseudoscalar for the space.

For instance in \R{3} the dual to a bivector is the familiar cross product

\begin{equation}\label{eqn:antisymmetricTensorTx:500}
\Ba \cross \Bb = -I (\Ba \wedge \Bb),
\end{equation}

where $I = \Be_1 \Be_2 \Be_3$.  In our spacetime context we use the pseudoscalar $I = \gamma_0 \gamma_1 \gamma_2 \gamma_3$.  Let's compute the coordinate representation of our vector, bivector, and trivector duals, which should compare with the tensor representation of the text.

\EndArticle
