\makeproblem{Diffusion}{basicStatMech:problemSet2:1}{ 

The usual diffusion equation for the probability density in one dimension is given by

\begin{equation}\label{eqn:basicStatMechProblemSet2Problem1:20}
\PD{t}{P}(x, t) = D \PDSq{x}{P}(x, t)
\end{equation}

where $D$ is the diffusion constant.  Define the Fourier components of the probability distribution via

\begin{subequations}
\begin{equation}\label{eqn:basicStatMechProblemSet2Problem1:40}
P(x, t) = \int_{-\infty}^\infty \frac{dk}{2 \pi} \tilde{P}(k, t) \exp\left( i k x \right)
\end{equation}
\begin{equation}\label{eqn:basicStatMechProblemSet2Problem1:60}
\tilde{P}(k, t) = \int_{-\infty}^\infty dx P(x, t) \exp\left( -i k x \right)
\end{equation}
\end{subequations}

This is useful since the diffusion equation is linear in the probability and each Fourier component will evolve independently.  Using this, solve the diffusion equation to obtain $P(k,t)$ in Fourier space given the initial $\tilde{P}(k,0)$.

\makesubproblem{Assuming an initial Gaussian profile}{basicStatMech:problemSet2:1a}

\begin{equation}\label{eqn:basicStatMechProblemSet2Problem1:80}
P(x, 0) = \inv{\sqrt{2 \pi \sigma^2}} \exp
\left(
-\frac{x^2}{2 \sigma^2}
\right),
\end{equation}

obtain the probability density $P(x,t)$ at a later time $t$.  (NB: Fourier transform, get the solution, transform back.)  Schematically plot the profile at the initial time and a later time.

\makesubproblem{A small modulation on top of a uniform value}{basicStatMech:problemSet2:1b}
Let the probability density be proportional to

\begin{equation}\label{eqn:basicStatMechProblemSet2Problem1:100}
\inv{L} + A \sin(k_0 x)
\end{equation}

at an initial time $t = 0$.  Assume this is in a box of large size $L$, but ignore boundary effects except to note that it will help to normalize the constant piece, assuming the oscillating piece integrates to zero.  Also note that we have
to assume $A < 1/L$ to ensure that the probability density is positive.  Obtain $P(x,t)$ at a later time $t$. Roughly how long does the modulation take to decay away? Schematically plot the profile at the initial time and a later time.
}

\makeanswer{basicStatMech:problemSet2:1}{ 

TODO.

\makeSubAnswer{XXX}{basicStatMech:problemSet2:1a}

TODO.

\makeSubAnswer{XXX}{basicStatMech:problemSet2:1b}

TODO.
}
