\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY354 Advanced Classical Mechanics.  Problem set 1.}
\label{chap:classicalMechanicsPs1}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/classicalMechanicsPs1.pdf}
\date{Jan 24, 2012}
\revisionInfo{classicalMechanicsPs1.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Problem 1.  Lorentz force Lagrangian.}
\subsection{Evaluate the Euler-Lagrange equations.}

This problem has two parts.   The first is to derive the Lorentz force equation

\begin{align}\label{eqn:classicalMechanicsPs1:n}
\BF &= q (\BE + \Bv \cross \BB) \\
\BE &= -\spacegrad \phi - \PD{t}{\BA} \\
\BB &= \spacegrad \cross \BA
\end{align}

using the Euler-Lagrange equations using the Lagrangian

\begin{equation}\label{eqn:classicalMechanicsPs1:n}
\LL = \inv{2} m \Bv^2 + q \Bv \cdot \BA - q \phi.
\end{equation}

In coordinates, employing summation convention, this Lagrangian is

\begin{equation}\label{eqn:classicalMechanicsPs1:n}
\LL = \inv{2} m \xdot_j \xdot_j + q \xdot_j A_j - q \phi.
\end{equation}

Taking derivatives

\begin{equation}\label{eqn:classicalMechanicsPs1:n}
\PD{\xdot_i}{\LL} = m \xdot_i + q A_i,
\end{equation}

\begin{align*}
\frac{d}{dt} \PD{\xdot_i}{\LL} 
&= 
m \ddot{x}_i 
+ q \PD{t}{A_i}
+ q \PD{x_j}{A_i} \frac{dx_j}{dt} \\
&=
m \ddot{x}_i 
+ q \PD{t}{A_i}
+ q \PD{x_j}{A_i} \xdot_j
\end{align*}

This must equal

\begin{equation}\label{eqn:classicalMechanicsPs1:n}
\PD{x_i}{\LL} = q \xdot_j \PD{x_i}{A_j} - q \PD{x_i}{\phi},
\end{equation}

So we have

\begin{align*}
m \ddot{x}_i 
&= 
-q \PD{t}{A_i}
- q \PD{x_j}{A_i} \xdot_j
+q \xdot_j \PD{x_i}{A_j} - q \PD{x_i}{\phi} \\
&=
-q \left( \PD{t}{A_i} - \PD{x_i}{\phi} \right)
+q v_j \left( \PD{x_i}{A_j} - \PD{x_j}{A_i} \right)
\end{align*}

The first term is just $E_i$.  If we expand out $(\Bv \cross \BB)_i$ we see that matches

\begin{align*}
(\Bv \cross \BB)_i
&=
v_a B_b \epsilon_{abi} \\
&=
v_a \partial_r A_s \epsilon_{rsb} \epsilon_{abi} \\
&=
v_a \partial_r A_s \delta_{rs}^{[ia]} \\
&=
v_a (\partial_i A_a - \partial_a A_i).
\end{align*}

A $a \rightarrow j$ substition, and comparision of this with the Euler-Lagrange result above completes the exersize.

\subsection{Show that the Lagrangian is gauge invariant.}

With a gauge transformation of the form

\begin{align}\label{eqn:classicalMechanicsPs1:n}
\phi &\rightarrow \phi + \PD{t}{\chi} \\
\BA &\rightarrow \BA - \spacegrad \chi,
\end{align}

show that the Lagrangian is invariant.

We really only have to show that 

\begin{equation}\label{eqn:classicalMechanicsPs1:n}
\Bv \cdot \BA - \phi
\end{equation}

is invariant.  Making the transformation we have

\begin{align*}
\Bv \cdot \BA - \phi
&\rightarrow 
v_j \left(A_j - \partial_j \chi \right) - \left(\phi + \PD{t}{\chi} \right) \\
&=
v_j A_j - \phi - v_j \partial_j \chi - \PD{t}{\chi} \\
&=
\Bv \cdot \BA - \phi
- \left( \frac{d x_j}{dt} \PD{x_j}{\chi} + \PD{t}{\chi} \right) \\
&=
\Bv \cdot \BA - \phi
- \frac{d \chi(\Bx, t)}{dt}.
\end{align*}

We see then that the Lagrangian transforms as

\begin{equation}\label{eqn:classicalMechanicsPs1:n}
\LL \rightarrow \LL + \frac{d}{dt}\left( -q \chi \right),
\end{equation}

and differs only by a total derivative.  With the lemma from the lecture, we see that this gauge transformation does not have any effect on the end result of applying the Euler-Lagrange equations.

\section{Problem 2.}

\section{Problem 3.}

\EndArticle
%\EndNoBibArticle
