\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Playing with complex notation for relativistic applications in a plane}
\label{chap:complexPlanarBoost}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/complexPlanarBoost.pdf}
\date{Mar 8, 2011}
\revisionInfo{complexPlanarBoost.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

In the recent electrodynamics midterm we had a question on circular motion.  This screamed for use of complex numbers to describe the spatial parts of the spacetime trajectories.

Let's play with this a bit.

\section{Our invariant.}

Suppose we describe our spacetime point as a paired time and complex number

\begin{equation}\label{eqn:complexPlanarBoost:n}
X = (ct, z).
\end{equation}

Our spacetime invariant interval in this form is thus

\begin{equation}\label{eqn:complexPlanarBoost:n}
X^2 \equiv (ct)^2 - \Abs{z}^2.
\end{equation}

Not much different than the usual coordinate representation of the spatial coordinates, except that we have a $\Abs{z}^2$ replacing the usual $\Bx^2$.

Taking the spacetime distance between $X$ and another point, say $\tilde{X} = ( c \tilde{t}, \tilde{z})$ motivates the inner product between two points in this representation

\begin{align*}
(X - \tilde{X})^2 
&= (ct - c \tilde{t})^2 - \Abs{z - \tilde{z}}^2 \\
&= (ct - c \tilde{t})^2 - (z - \tilde{z})(z^\conj - \tilde{z}^\conj) \\
&= (ct)^2 - 2 (ct) (c \tilde{t}) + (c \tilde{t})^2 
- \Abs{z}^2 - \Abs{\tilde{z}}^2 + (z \tilde{z}^\conj + z^\conj \tilde{z}) \\
&= X^2 + \tilde{X}^2 - 2 \left( (ct) (c \tilde{t}) - \inv{2}(z \tilde{z}^\conj + z^\conj \tilde{z}) \right) \\
\end{align*}

It's clear that it makes sense to define

\begin{equation}\label{eqn:complexPlanarBoost:n}
X \cdot \tilde{X} = (ct) (c \tilde{t}) - \Real (z \tilde{z}^\conj),
\end{equation}

consistent with our original starting point

\begin{equation}\label{eqn:complexPlanarBoost:n}
X^2 = X \cdot X.
\end{equation}

Let's also introduce a complex inner product

\begin{equation}\label{eqn:complexPlanarBoost:n}
\innerprod{z}{\tilde{z}} \equiv \inv{2} \left( z \tilde{z}^\conj + z^\conj \tilde{z}) \right) = \Real (z \tilde{z}^\conj).
\end{equation}

Our dot product can now be written

\begin{equation}\label{eqn:complexPlanarBoost:n}
X \cdot \tilde{X} = (ct) (c \tilde{t}) - \innerprod{z}{\tilde{z}}.
\end{equation}

\section{Change of basis.}

Our standard basis for our spatial components is $\{1, i\}$, but we are free to pick any other basis should we choose.  In particular, if we rotate our basis counterclockwise by $\theta$, our new basis, still orthonormal, is $\{ e^{i\theta}, i e^{i\theta} \}$.

In any orthonormal basis the coordinates of a point with respect to that basis are real, so just as we can write

\begin{equation}\label{eqn:complexPlanarBoost:n}
z = \innerprod{1}{z} + i \innerprod{i}{z},
\end{equation}

we can extract the coordinates in the rotated frame, also simply by taking inner products

\begin{equation}\label{eqn:complexPlanarBoost:n}
z = e^{i \theta} \innerprod{e^{i \theta}}{z} + i e^{i\theta} \innerprod{i e^{i\theta} }{z}.
\end{equation}

The values $\innerprod{e^{i \theta}}{z}$, and $\innerprod{i e^{i\theta} }{z}$ are the (real) coordinates of the point $z$ in this rotated basis.

This is enough that we can write the Lorentz boost immediately for a velocity $\vec{v} = c \beta e^{i\theta}$ at an arbitrary angle $\theta$ in the plane

\begin{equation}\label{eqn:complexPlanarBoost:n}
\begin{bmatrix}
ct' \\
\innerprod{e^{i\theta}}{z'} \\
\innerprod{i e^{i\theta}}{z'} 
\end{bmatrix}
=
\begin{bmatrix}
\gamma & -\gamma \beta & 0 \\
-\gamma \beta & \gamma & 0 \\
0 & 0 & 1
\end{bmatrix}
\begin{bmatrix}
ct \\
\innerprod{e^{i\theta}}{z} \\
\innerprod{i e^{i\theta}}{z} 
\end{bmatrix}
\end{equation}

\EndArticle
%\EndNoBibArticle
