%
% Copyright © 2013 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\input{../blogpost.tex}
\renewcommand{\basename}{condensedMatterLecture1}
\renewcommand{\dirname}{notes/phy487/}
\newcommand{\keywords}{Condensed matter physics, PHY487H1S}
\input{../peeter_prologue_print2.tex}

\beginArtNoToc
\generatetitle{PHY487H1S Condensed Matter Physics.  Lecture 1: Course overview.  Taught by Prof.\ Stephen Julian}
%\chapter{Course overview}
\label{chap:condensedMatterLecture1}
% classmates: Seth (eng sci, tall and lanky)

\section{Disclaimer}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{Course overview}

\begin{itemize}
\item Bonding and crystal structure: Why solids form, describing periodic order mathematically, diffraction.
\item Lattice vibrations: ``elementary excitation'' of a periodic array of atoms (periodicity allows a $10^{23}$ body problem to be solved).
\item Electrons in solids: Explaining this introduces a need for quantum mechanics.  
%Periodicity -> band structure -> insulators vs metals.
\item Electrical conduction in solids.  (metals and semiconductors).
\end{itemize}

\section{Chemical bonding in solids}

% Reading: L&L chapter 1.

\begin{itemize}
\item Different types of chemical bonds explain many of the differences between solids.
\item Differences in solids: hard/soft.  Example: Lithium, so soft that a pure sample will flow if set on a desk ; metal vs insulating, melting points
\end{itemize}

FIXME: Fig1: periodic table annoted with orbital filling notes.

\S 1.1 \underline{The periodic table}.  Elements are classified according to which type of orbital is being filled.

FIXME: Fig2: Hydrogenic atom (only one electron).  Hydrogen, or other super ionized material (example: iron with all but one electron observed in supernova spectra).

FIXME: Fig3: Two electron atom. eg: $\text{Ti}_{22}$.

Read: \S 1.1.

FIXME: Fig4.  s, m, l ranges.

\begin{itemize}
\item upper right hand of periodic table: covalent bonding
\item lower left hand of periodic table: metallic bonding
\item mixed left hand with right hand of periodic table: ionic bonding
\end{itemize}

\section{Covalent bonding}

Consider a pair of hydrogen nuclei sharing one electron.  \S 1.2 \citep{ibach2009solid} has a mathematical description (not examinable)

FIXME: Fig5

F6:  far apart

F7: close together.  bonding

F8: close together.  anti-bonding

F9: Chemistry diagram.

%\paragraph{Adding more electrons}

\EndArticle
