%
% Copyright © 2013 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\makeproblem{Analytic calculations of density of states and effective mass}{condensedMatter:problemSet9:2}{ 

The dispersion relation of 
a two-dimensional square lattice with lattice parameter $a$, 
with nearest-neighbour tight-binding integrals $A$ and $B$, is 

\begin{eqnarray*}
E(\Bk) \simeq E_s - A - 2B(\cos(k_xa) + \cos(k_ya)).
\end{eqnarray*}

\makesubproblem{}{condensedMatter:problemSet9:2a}
Expanding $E(k)$ around the bottom and the top of the band to second 
  order in $\Bk$ (i.e.\ approximating the bands as parabolic) 
  show that the density of states per unit area jumps from zero when $E$ is 
  outside the band, to a finite value as $E$ passes the bottom or top of the band. 
  What is the finite value?

\makesubproblem{}{condensedMatter:problemSet9:2b}
Consider now the analogous three-dimensional simple cubic system, 
  such that 
  \begin{eqnarray*}
    E(\Bk) \simeq E_s - A - 2B(\cos(k_xa) + \cos(k_ya) + \cos(k_za)).
  \end{eqnarray*}
Show that the density of states now grows continuously from zero, 
 and is proportional to  
 $\sqrt(E-E_b)$, and $\sqrt(E_t-E)$, as the bottom ($E_b$) and top ($E_t$) 
 of the band are crossed.  
Find the constants of proportionality. 

\makesubproblem{}{condensedMatter:problemSet9:2c}
For the three-dimensional cubic system of part (b), find the group 
  velocity $\Bv_g$ along the $k_x$ direction, as a function of $k_x$, 
  and sketch its behaviour from $k_x=0$ to $\pi/a$. 

\makesubproblem{}{condensedMatter:problemSet9:2d}
Using our definition of group velocity from class, and defining the 
  velocity effective mass $m^*$ by $v_g = \Hbar k/m^*$, 
  plot $m^*$ vs $k_x$ from 0 to $\pi/a$.
} % makeproblem

\makeanswer{condensedMatter:problemSet9:2}{ 
\makeSubAnswer{}{condensedMatter:problemSet9:2a}

TODO.
\makeSubAnswer{}{condensedMatter:problemSet9:2b}

TODO.
\makeSubAnswer{}{condensedMatter:problemSet9:2c}

TODO.
\makeSubAnswer{}{condensedMatter:problemSet9:2d}

TODO.
}

