%
% Copyright © 2013 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\input{../blogpost.tex}
\renewcommand{\basename}{condensedMatterLecture20}
\renewcommand{\dirname}{notes/phy487/}
\newcommand{\keywords}{Condensed matter physics, PHY487H1F}
\input{../peeter_prologue_print2.tex}

%\citep{harald2003solid} \S x.y
%\citep{ibach2009solid} \S x.y

%\usepackage{mhchem}
\usepackage{bm} % \EE
\usepackage[version=3]{mhchem}
\usepackage{units}
\newcommand{\nought}[0]{\circ}
%\newcommand{\EF}[0]{\epsilon_{\mathrm{F}}}
\newcommand{\EF}[0]{E_{\mathrm{F}}}
\newcommand{\kF}[0]{k_{\mathrm{F}}}
\newcommand{\vF}[0]{v_{\mathrm{F}}}

\beginArtNoToc
\generatetitle{PHY487H1F Condensed Matter Physics.  Lecture 20: Electric current (cont.).  Taught by Prof.\ Stephen Julian}
%\chapter{Electric current (cont.)}
\label{chap:condensedMatterLecture20}

%\section{Disclaimer}
%
%Peeter's lecture notes from class.  May not be entirely coherent.

\section{Electric current (cont.)}

$\EE \ne 0$ shifts the Fermi distribution so that so that $+\Bv_1 - \Bv$ no longer cancel.

F1

Current flows, unless the band is completely full, or empty.

Full band has $f(E, \BE) = F(E, 0)$, so that there is no current 

F2

An \dquoteAndIndex{insulator} has all bands either completely full or completely empty.

A \dquoteAndIndex{metal} is a solid with a Fermi surface (partly filled band(s)).

The $\EE$ field displaces the Fermi surface, but scattering restores equilibrium, limiting $\Bj$.

Note that a periodic lattice does \underline{not} cause scattering, it causes \underline{band structure}.

Scattering is due to \underline{departures} from periodicity, and is due to impurities and/or vacancies and lattice vibrations (phonons).

F3

A metric for this is called the resistivity, a temperature dependent effect

F4

\paragraph{Non-equilibrium $f(E, \EE)$}

F5

Electrons in the range (1) cannot scatter due to the Pauli exclusion principle, whereas those in the energy range (2) \underline{can} scatter from $+k$ to $-k$.  The net scattering is from $+\Bv$ to $-\Bv$.

For 3D see \citep{ibach2009solid} fig. 9.5, roughly:

F6: a displaced Fermi sphere.

\paragraph{Steady state} rate of scattering from $+\Bv$ to $-\Bv$ is the rate at which new $+\Bv$ carriers appear, where

\begin{dmath}\label{eqn:condensedMatterLecture20:n}
\Hbar \dot{\Bk} = - e \EE.
\end{dmath}

Introduce $\tau$ as the mean scattering time, and let

\begin{dmath}\label{eqn:condensedMatterLecture20:n}
\Delta k = \dot{k} \tau = -\frac{e \mathcal{E}}{\Hbar} \tau,
\end{dmath}

the amount by which the Fermi surface is displaced.

F7

Here 
In the shaded region, the electrons are inert, because the $-\Bv$ and $\Bv$ contributions cancel.  It's only the non-shaded portions of the overlapping spheres that we have to consider.  

\begin{dmath}\label{eqn:condensedMatterLecture20:n}
j_x 
= -\frac{e}{V} \sum_{k, \sigma} v_x(k, \sigma)
= -\frac{e}{\cancel{V}}  \frac{2 \cancel{V} }{(2 \pi)^3}
\int 
\mathLabelBox
{
\kF^2 \sin\theta d\theta d\phi 
}
{
area element at $\theta, $\phi$
}
\mathLabelBox
[
   labelstyle={below of=m\themathLableNode, below of=m\themathLableNode}
]
{
\Bigr{ -\frac{e \mathcal{E}}{\Hbar} \tau \cos\theta \Bigl} 
}
{
$\delta k$ at $\theta, \phi$
}
\times
\mathLabelBox
{
\vF \cos\theta
}
{
v_x
}
\approx 
\frac{e^2}{4 \pi^3} \tau \kF^2 \frac{\vF}{\Hbar} 
\mathLabelBox
{
\int_0^\pi \sin\theta \cos^2\theta d\theta
}
{$
-\int_0^\pi \cos^2\theta d \cos\theta = - \int_1^{-1} u^2 du = 2/3
$}
\mathLabelBox
[
   labelstyle={below of=m\themathLableNode, below of=m\themathLableNode}
]
{
2 \pi
}
{
$\phi$ integral
}
\mathcal{E}_x
\end{dmath}

With 

\begin{dmath}\label{eqn:condensedMatterLecture20:n}
\vF = \frac{\Hbar \kF}{m^\conj},
\end{dmath}

this is

\begin{dmath}\label{eqn:condensedMatterLecture20:n}
j_x \approx
\frac{e^2\tau}{m^\conj} \frac{\kF^3}{3 \pi^2} \mathcal{E}_x
\end{dmath}

\begin{dmath}\label{eqn:condensedMatterLecture20:n}
j_x = \frac{n e^2 \tau}{m^\conj} \mathcal{E}_x
\end{dmath}

or

\begin{dmath}\label{eqn:condensedMatterLecture20:n}
\myBoxed{
\sigma = \frac{n e^2 \tau}{m^\conj}
}
\end{dmath}

This is the \textAndIndex{Drude formula for conductivity}.

Note that Drude's derivation predated quantum mechanics.  He treated the electrons classically introducing a drift velocity

F8

where

\begin{dmath}\label{eqn:condensedMatterLecture20:n}
v_{\text{drift}} = \text{acceleration} \tau = -\frac{e \mathcal{E}}{m} \tau,
\end{dmath}

so that 

\begin{dmath}\label{eqn:condensedMatterLecture20:n}
\Bj = - e n v_{\text{drift}} \EE = \frac{ n e^2 \tau }{m} \EE.
\end{dmath}

%\EndArticle
\EndNoBibArticle
