\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY454H1S Continuum Mechanics.  Lecture 20: Singular pertubation theory.  Taught by Prof. K. Das.}
\label{chap:continuumL20}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumL20.pdf}
\date{Mar 28, 2012}
\gitRevisionInfo{continuumL20}
\keywords{Navier-Stokes, PHY454H1S} 

\beginArtWithToc
%\beginArtNoToc

\section{Disclaimer.}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{Review.}

\begin{enumerate}
\item 
\begin{equation}\label{eqn:continuumL20:10}
\epsilon \ddx{u} + u = x
\end{equation}

With 
\begin{align*}
u(0) &= 1 \\
\epsilon &\rightarrow 0
\end{align*}
\item
\begin{equation}\label{eqn:continuumL20:30}
\epsilon \frac{d^2u}{dy^2} + \frac{du}{dy} = 1
\end{equation}

\begin{align*}
u(0) &= 0 \\
u(1) &= 2 \\
0 < \epsilon \ll 1
\end{align*}
\end{enumerate}

Regular pertubation 

\begin{align*}
\epsilon \rightarrow 0
du &= dy \\
u &= y + C \\
u(1) &= 2 \\
c &= 1 \\
u &= y + 1
\end{align*}

\subsection{Case 1.  Exact solution}

\begin{equation}\label{eqn:continuumL20:50}
u = ( 1 + \epsilon ) e^{-x/\epsilon} + x - \epsilon
\end{equation}

FIXME: F1:

\begin{equation}\label{eqn:continuumL20:70}
0 < \epsilon \ll 1
\end{equation}

\begin{equation}\label{eqn:continuumL20:90}
\epsilon \rightarrow 0
\end{equation}

If $x = O(1)$

\begin{align*}
u &= 1 \times 0 + x \\
u &= x
\end{align*}

If $x = O(\epsilon)$ then we have to be more careful constructing an approximation.

FIXME: F2:

With $\epsilon = 0.01 \text{m}$, then we have

\begin{equation}\label{eqn:continuumL20:110}
e^{-x/\epsilon} \rightarrow 0
\end{equation}

\begin{align*}
x 
&= 0.05 \text{m} \\
&= 0.01 x 1 \text{m} \\
&= \epsilon O(1)
\end{align*}

When $x$ is very small

\begin{equation}\label{eqn:continuumL20:130}
e^{-x/\epsilon} \ne 0
\end{equation}

if $\epsilon \rightarrow 0$ and $x \rightarrow O(\epsilon)$

\begin{equation}\label{eqn:continuumL20:150}
e^{-x/\epsilon} \rightarrow e^{-\epsilon O(1) /\epsilon} \rightarrow e^{-O(1)}
\end{equation}

so 

\begin{equation}\label{eqn:continuumL20:170}
u = e^{-x/\epsilon}
\end{equation}

\section{Solution in the inner region.  First order example.}

When $x = O(\epsilon)$ define a new scale

\begin{equation}\label{eqn:continuumL20:190}
X = \frac{x}{\epsilon}
\end{equation}

\begin{equation}\label{eqn:continuumL20:210}
x = O(1)
\end{equation}

so that our DE takes the form

\begin{equation}\label{eqn:continuumL20:230}
\frac{du}{dX} + u = \epsilon X
\end{equation}

When $\epsilon \rightarrow 0$ we have

\begin{equation}\label{eqn:continuumL20:250}
\frac{du}{dX} + u \approx 0.
\end{equation}

We have solution

\begin{equation}\label{eqn:continuumL20:270}
\ln u = -X + ln C,
\end{equation}

or

\begin{equation}\label{eqn:continuumL20:290}
u \propto e^{-X} = e^{-x/\epsilon}.
\end{equation}

\paragraph{Question:} Couldn't we just Laplace transform.
\paragraph{Answer given:} We'd still get into trouble when we take $\epsilon \rightarrow 0$.  My comment: I don't think that's strictly true.  In an example like this where we have an exact solution, a Laplace transform technique should also yield that solution.  I think the real trouble will come when we attempt to incorporate the non-linear inertial terms of the Navier-Stokes equation.

\section{Second order example.}

Our exact solution is

\begin{equation}\label{eqn:continuumL20:310}
u = y + \frac{1 + e^{-y/\epsilon}}{1 - e^{1/\epsilon}}
\end{equation}

FIXME: show.

FIXME: graph.

when $\epsilon \rightarrow 0$ and $y = O(1)$ we have

\begin{equation}\label{eqn:continuumL20:330}
u = y + 1.
\end{equation}

when $y = O(\epsilon)$ and both $\epsilon$ and $y$ tend to zero, we have approximately

\begin{equation}\label{eqn:continuumL20:350}
u = 1 - e^{-y/\epsilon}
\end{equation}

FIXME: show.  This one isn't as obvious.

Consider the inner region and write

\begin{equation}\label{eqn:continuumL20:n}
Y = \frac{y}{\epsilon},
\end{equation}

when $y = O(\epsilon)$ and $Y = O(1)$ we have

\begin{equation}\label{eqn:continuumL20:n}
\cancel{\epsilon} \frac{d^2 u}{\cancel{\epsilon} \epsilon dY^2} + \inv{\epsilon} \frac{du}{dY} = 1
\end{equation}

or
\begin{equation}\label{eqn:continuumL20:n}
\frac{d^2 u}{dY^2} + \frac{du}{dY} = \epsilon.
\end{equation}

when $\epsilon \rightarrow 0$ we have approximately

\begin{equation}\label{eqn:continuumL20:n}
\frac{d^2 u}{dY^2} + \frac{du}{dY} = 0.
\epsilon.
\end{equation}

This we can solve in the normal way, computing the characteristic equation and so forth with an assumed solution of $u = e^{mY}$.

FIXME: show in detail:

We find
\begin{equation}\label{eqn:continuumL20:n}
m^2 + m = 0
\end{equation}

with solutions

\begin{equation}\label{eqn:continuumL20:n}
m = 0, -1
\end{equation}

Our solution therefore has the form

\begin{equation}\label{eqn:continuumL20:n}
u = A + B e^{-Y}
\end{equation}

at $Y = 0$ we have

\begin{equation}\label{eqn:continuumL20:n}
u = A + B = 0.
\end{equation}

so that 
\begin{equation}\label{eqn:continuumL20:n}
B = -A.
\end{equation}

and we find

\begin{equation}\label{eqn:continuumL20:n}
u = A (1 - e^{-Y}) = A( 1 - e^{-y/\epsilon} )
\end{equation}

In the outer region with $\epsilon \rightarrow 0$ and $y = O(1)$ we have

\begin{equation}\label{eqn:continuumL20:n}
\frac{du}{dy} = 1
\end{equation}

or 

\begin{equation}\label{eqn:continuumL20:n}
u = y + 1
\end{equation}

Taking both together we have something that looks like

FIXME: F4

%FIXME: Reading: \S XX from \cite{acheson1990elementary}

%\EndArticle
\EndNoBibArticle
