\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY454H1S Continuum Mechanics.  Lecture 16: Boundary Layers.  Taught by Prof. K. Das.}
\label{chap:continuumL16}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumL16.pdf}
\date{Mar 14, 2012}
\gitRevisionInfo{continuumL16}

\beginArtWithToc
%\beginArtNoToc

\section{Disclaimer.}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{Boundary Layers.}

Suppose we have an obsticle to fluid flow, as in

FIXME: F1

we have a couple conditions on fluid flow.

\begin{enumerate}
\item No fluid can cross the solid boundary
\item Due to viscosity the tangential velocity of the fluid should match the velocity of the solid boundary.
\end{enumerate}

In study of this type of flow, we can consider the flow separated into two portions, one is a flow that is largely viscous, and the other that is largely inertial.  This is depicted in 

FIXME: F2

We call the study of these two regions boundary layer flow.

\section{Unsteady rectilinear flow.}

Given

\begin{equation}\label{eqn:continuumL16:10}
\Bu = u(x, y, z, t) \xcap
\end{equation}

Our continuity equation (incompressibility assumption) is 

\begin{equation}\label{eqn:continuumL16:30}
\PD{x}{u} + 
\cancel{\PD{y}{v}} + 
\cancel{\PD{z}{w}} = 0.
\end{equation}

Our non-linear term of Navier-Stokes is then killed

\begin{equation}\label{eqn:continuumL16:50}
u \PD{x}{u} = 0
\end{equation}

so that NS 

\begin{subequations}
\begin{equation}\label{eqn:continuumL16:70}
\rho \PD{t}{u} = -\PD{x}{p} + 
\mu \left( \PDSq{y}{u} +\PDSq{z}{u} \right)
\end{equation}
\begin{equation}\label{eqn:continuumL16:90}
0 = -\PD{y}{p} 
\end{equation}
\begin{equation}\label{eqn:continuumL16:110}
0 = -\PD{z}{p} 
\end{equation}
\end{subequations}

Taking x partial of \ref{eqn:continuumL16:70}, we have

\begin{equation}\label{eqn:continuumL16:130}
\rho \PD{t}{} \cancel{\PD{x}{u}} = -\PDSq{x}{p} + \mu \left( \PDSq{y}{} \cancel{\PD{x}{u}} +\PDSq{z}{} \cancel{\PD{x}{u}} \right)
\end{equation}

so 

\begin{equation}\label{eqn:continuumL16:150}
\frac{d^2 p}{dx^2} = 0
\end{equation}

or

\begin{equation}\label{eqn:continuumL16:170}
p(x, t) = p_0(t) - G x
\end{equation}

\begin{equation}\label{eqn:continuumL16:190}
G = -\frac{dp}{dx}
\end{equation}

in general is a function of $t$, but constant in space.  Navier-Stokes then takes the form

\begin{equation}\label{eqn:continuumL16:210}
\rho \PD{t}{u} = G(t) + \mu \left( \PDSq{y}{u} +\PDSq{z}{u} \right)
\end{equation}

\subsection{Impulsively started flow.}

Reading: \S 2 from \cite{acheson1990elementary}

Let's consider a flow driven by a moving boundary.  We have two ways that we can look at such a flow, the first of which is with the fluid fixed and the boundary moving and the second is with the fluid moving and the boundary fixed.  This is depicted in

FIXME: F3

These two possible viewpoints can be called the Eulerian and the Lagrangian views where

\begin{itemize}
\item Lagrangian: the observer is moving with the fluid.
\item Eulerian: the observer is fixed in space, watching the fluid.
\end{itemize}

With a flow of the form

\begin{equation}\label{eqn:continuumL16:230}
\Bu = u(y, t) \xcap
\end{equation}

NS equation

\begin{equation}\label{eqn:continuumL16:250}
\boxed{
\PD{t}{u} = \nu \PDSq{y}{u}.
}
\end{equation}

Our boundary value constraints are

\begin{equation}\label{eqn:continuumL16:270}
u(0, t) = 
\left\{
\begin{array}{l l}
0 & \quad \mbox{for $t < 0$} \\
U & \quad \mbox{for $t \ge 0$}
\end{array}
\right.
\end{equation}

and $u \rightarrow 0$ as $y \rightarrow \infty$.

If we make a transformation to dimensionless arguments

\begin{equation}\label{eqn:continuumL16:290}
u \rightarrow U
\end{equation}

so that

\begin{equation}\label{eqn:continuumL16:310}
\frac{u}{U} \rightarrow \text{dimensionless}
\end{equation}

Then we require of the parameters

\begin{equation}\label{eqn:continuumL16:330}
y, \nu, t \rightarrow \frac{y}{\sqrt{\nu t}}
\end{equation}

so that we have a characteristic length scale of the form

\begin{equation}\label{eqn:continuumL16:350}
\delta \rightarrow \sqrt{\nu t}
\end{equation}

We can find an approximate solution

\begin{equation}\label{eqn:continuumL16:370}
\frac{U}{t} \approx \frac{\nu U}{\delta^2}.
\end{equation}

We can introduce a similarity variable (often hard to find), of the form

\begin{equation}\label{eqn:continuumL16:390}
\eta = \frac{y}{2 \sqrt{\nu t}}.
\end{equation}

FIXME: try attacking this more systematically using Fourier or Laplace transforms (probably a Laplace transform, because of our initial conditions).

Let's use our similarity variable and see what happens.  With

\begin{equation}\label{eqn:continuumL16:410}
u = U f(\eta),
\end{equation}

we find

\begin{align*}
\PD{y}{u} 
&= \PD{\eta}{u} \PD{y}{\eta} \\
&= U f' \inv{2 \sqrt{\nu t}}
\end{align*}

where 

\begin{equation}\label{eqn:continuumL16:430}
f' = \PD{\eta}{f}
\end{equation}

We then find

\begin{equation}\label{eqn:continuumL16:450}
\PDSq{y}{u} = U f'' \inv{4 \nu t}
\end{equation}

and

\begin{align*}
\PD{t}{u} 
&= \PD{\eta}{u} \PD{t}{\eta} \\
&= - u f' \frac{\eta}{2 t}
\end{align*}

FIXME: commented out and probably too hard to follow.
%\begin{equation}\label{eqn:continuumL16:470}
%- \cancel{U} f' \frac{\eta}{\cancel{2 t}} = \cancel{\eta} \cancel{U} f'' \frac{ 2 \cancel{4} \cancel{\nu} \cancel{t}}
%\end{equation}

or

\begin{equation}\label{eqn:continuumL16:490}
f'' + 2 \eta f' = 0
\end{equation}

Our solution is

\begin{equation}\label{eqn:continuumL16:510}
f(\eta) = A \erf(\eta) + B,
\end{equation}

where our error function is

\begin{equation}\label{eqn:continuumL16:530}
\erf(\eta) = \sqrt{2}{\pi} \int_0^\eta e^{-s^2} ds.
\end{equation}

Recall that 

\begin{enumerate}
\item $\erf(0) = 0$
\item $\erf(\infty) = 1$
\item $\erf(-\eta) = -\erf(\eta)$
\end{enumerate}

FIXME: show.

With 

\begin{align*}
u(0, t) = U &\rightarrow f(0) = 1 \\
u(\infty, t) = 0 &\rightarrow f(\infty) = 0
\end{align*}

we find

\begin{equation}\label{eqn:continuumL16:550}
A = 0
\end{equation}
\begin{equation}\label{eqn:continuumL16:570}
B = 1
\end{equation}

so that 

\begin{equation}\label{eqn:continuumL16:590}
\boxed{
u(y, t) = U(1 - \erf(\eta))
}
\end{equation}

where (again)

\begin{equation}\label{eqn:continuumL16:610}
\eta = \frac{y}{2 \sqrt{\nu t}}.
\end{equation}

Explictly, this is

\begin{equation}\label{eqn:continuumL16:590}
u(y, t) = U_0(1 - \erf\left(\frac{y}{2 \sqrt{\nu t}}\right))
\end{equation}

It turns out that for any fluids, regardless of the viscosities, the thickness of the boundary layers generally vary as a linear function of $\sqrt{\nu t}$ as in

FIXME: F4.

\EndArticle
