\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY454H1S Continuum Mechanics.  Lecture 13: Hydrostatics.  Surface normals and tangent vectors.  Taught by Prof. K. Das.}
\label{chap:continuumL13}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumL13.pdf}
\date{Mar 2, 2012}
\gitRevisionInfo{continuumL13}

\beginArtWithToc
%\beginArtNoToc

%\section{Reading.}
%FIXME: Reading: \S XX from \cite{acheson1990elementary}?

\section{Disclaimer.}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{Midterm discussion.}

We had been asked to solve a 1D flow problem with one fixed surface, and one moving surface, as illustrated in

FIXME: F1

we end up showing that our solutions are of the forms

FIXME: F2.

Note that the last sort of solution, that of return flow, was discussed in the context of surface tension.

\section{Hydrostatics}

Consider a sample volume of water, not moving with respect to the rest of the surrounding water.  If it is not moving the forces must be in balance.  What are the forces acting on this bit of fluid, considering a cylinder of the fluid above it as in 

FIXME: F3.

We have

\begin{equation}\label{eqn:continuumL13:20}
h A_w \rho g + p_A A_w = p_w A_w
\end{equation}

so
\begin{equation}\label{eqn:continuumL13:40}
p_w = h \rho g + p_A
\end{equation}

If we were to replace this blob of water with something of equal density, it should not change the dynamics (or statics) of the situations and that would not move.

We call this the

\begin{definition}
\emph{(Buoyancy force)}
\label{dfn:continuumL13:60}
Buoyancy force = 
weight of the equivalent volume of water - weight of the foreign body.
\end{definition}

If the densities are not equal, then we'd have motion of the new bit of mass as depicted in

FIXME: F4.

Consider a volume of ice floating on the surface of water, one with solid ice and one with partially frozen ice (with water or air or dirt or an anchor or anything else in it)

FIXME: F5

No matter the situation, the water level will not change if the ice melts, because the total weight of the displaced water must have been matched by the weight of the unmelted ice plus additives.

Now what happens when we have fluid flows?  Consider:

FIXME: F6

Conservation of mass is going to mean that the masses of fluid flowing through any pair of cross sections will have to be equal

\begin{equation}\label{eqn:continuumL13:80}
\rho_1 A_1 v_1 = \rho_2 A_2 v_2,
\end{equation}

With incompressible fluids ($\rho = \rho_1 = \rho_2$) we have

\begin{equation}\label{eqn:continuumL13:100}
A_1 v_1 = A_2 v_2,
\end{equation}

so that if 

\begin{equation}\label{eqn:continuumL13:120}
A_1 > A_2,
\end{equation}

we must have
\begin{equation}\label{eqn:continuumL13:140}
v_1 < v_2,
\end{equation}

to balance this.

In class this was illustrated with a pair of computer animations, one showing the deformation of patches of the fluid, and another showing how the velocities vary through the channel.  This is crudely depicted in 

FIXME: F7

We see the same behavior for channels that return to the original diameter after widening as in

FIXME: F8

If we consider half of such a channel as in

FIXME: F9a

considering the flow around a small triangular section we must have a pressure gradient, which induces a vorticity flow.  We'd see something similar in a rectangular channel where there is a block in the channel.

Now let's consider the hydrostatics case again, with an arbitrarily weird channel as in

FIXME: F10

This was also illustrated with a glass blown container in class as in 

FIXME: F11

In this real apparatus, we didn't have exactly the same height (because of bubbles and capillary effects (surface tension induced meniscus curves), but we see first hand what we are talking about.

To account for this, we need to consider the situation in pieces as in

FIXME: F12

Breaking down the total pressure effects into individual bits, any column of fluid contributes to the pressure below it, even if that column of fluid is not directly on top of a continuous column of fluid all the way to the ``bottom''.

\section{Surface tension: normals and tangents.}

Consider a surface with some variation

FIXME: F13

We can construct an equation for the surface

\begin{equation}\label{eqn:continuumL13:160}
z = h(x, t),
\end{equation}

or equivalently

\begin{equation}\label{eqn:continuumL13:180}
\phi = z - h(x, t) = 0.
\end{equation}

If $d$ is the average height, with the $\eta(x,t)$ the variation of the height from this average, we can also write

% FIXME: had z = d + \eta(x, t) on board?
\begin{equation}\label{eqn:continuumL13:200}
h = d + \eta(x, t)
\end{equation}

and for the surface

% FIXME: had: z - \eta(x, t) = \text{constant} on the board?
\begin{equation}\label{eqn:continuumL13:220}
\phi = d - \eta(x, t) = 0
\end{equation}

We can generalize this and define a surface function as one that satisfies

\begin{equation}\label{eqn:continuumL13:220b}
\phi = d - \eta(x, t) = \text{constant}.
\end{equation}

Consider a small section of a 2D surface as in

FIXME: F14

With $\phi = \text{constant}$ on the surface, we have for $\phi = \phi(x, y, z)$

\begin{equation}\label{eqn:continuumL13:240}
d\phi = 0 
\end{equation}

or, in coordinates

\begin{align*}
d\phi &= 
\PD{x}{\phi} dx
+\PD{y}{\phi} dy
+\PD{z}{\phi} dz \\
&= \spacegrad \phi \cdot d\Br
\end{align*}

Pictorially we see that $d\Br$ is tangential to the surface, but since we also have

\begin{equation}\label{eqn:continuumL13:360}
\spacegrad \phi \cdot d\Br = 0,
\end{equation}

the implication is that the gradient is normal to the surface

\begin{equation}\label{eqn:continuumL13:380}
d\Br \perp \spacegrad \phi.
\end{equation}

We can therefore construct the unit normal by scaling the gradient

\begin{equation}\label{eqn:continuumL13:260}
\ncap = \frac{\spacegrad \phi}{\Abs{\spacegrad \phi}},
\end{equation}

since the direction of $\ncap$ is $\spacegrad \phi$.

For example, in our case where $\phi = y - h(x, t)$, we have

\begin{equation}\label{eqn:continuumL13:280}
\spacegrad \phi = \xcap \left( -\PD{x}{h} \right) + \ycap
\end{equation}

This has norm

\begin{equation}\label{eqn:continuumL13:300}
\Abs{\spacegrad \phi} = \sqrt{ 1 + \left( \PD{x}{h} \right)^2 }
\end{equation}

and our unit normal is

\begin{equation}\label{eqn:continuumL13:320}
\ncap = 
\frac{\xcap \left( -\PD{x}{h} \right) + \ycap}{
\sqrt{ 1 + \left( \PD{x}{h} \right)^2 }
}
\end{equation}

By inspection we can also express the unit tangent, and we have for both

\begin{align}\label{eqn:continuumL13:340}
\ncap &= 
\inv{
\sqrt{ 1 + \left( \PD{x}{h} \right)^2 }
}
\left( -\PD{x}{h}, 1 \right)  \\
\taucap &= 
\inv{
\sqrt{ 1 + \left( \PD{x}{h} \right)^2 }
}
\left( 1, \PD{x}{h} \right)
\end{align}

%\EndArticle
\EndNoBibArticle
