\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY454H1S Continuum Mechanics.  Lecture 21: Stability.  Taught by Prof. K. Das.}
\label{chap:continuumL21}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumL21.pdf}
\date{Mar 29, 2012}
\gitRevisionInfo{continuumL21}
\keywords{Navier-Stokes, PHY454H1S} 

\beginArtWithToc
%\beginArtNoToc

\section{Disclaimer.}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{Stability.  Some graphical illustrations.}

What do we mean by stability?  A configuration is stable if after a small disturbance it returns to it's original position.  A couple systems to consider are shown in

FIXME: F1

We can examine how a displacement $\delta x$ changes with time after making it.  In a stable configuration without friction we will induce an oscillation.  With friction we'll have a damping effect.  For the parabolic well, these are illustrated in

FIXME: F2

For the inverted parabola our displacement takes the form

FIXME: F3

For the ball on the table, assuming some friction that stops the ball, fairly quickly, we'll have a displacement as illustrated in

FIXME: F1c''

\section{Characterizing stability}

Let's suppose that our displacement can be described in exponential form

\begin{equation}\label{eqn:continuumL21:10}
\delta x \sim e^{\sigma t}
\end{equation}

where $\sigma$ is the \textit{growth rate of pertubation}, and is in general a complex number of the form

\begin{equation}\label{eqn:continuumL21:30}
\sigma = \sigma_\text{R} + i \sigma_\text{I}
\end{equation}

\subsection{Case I.  Oscillatory unstability}

A system of the form

\begin{align*}
\sigma_{\text{R}} &= 0 \\
\sigma_{\text{I}} &> 0
\end{align*}

\textit{oscillatory unstable}.  An example of this is the undamped parabolic system illustrated above.

\subsection{Case II.  Marginal unstability.}

FIXME:
%\begin{align*}
%\delta x 
%&\sim e^{\sigma_{\text{R}} t} e^{i \simgaI t} \\
%&\sim e^{\sigma_{\text{R}} t} \left( \cos \sigma_{\text{I}} t + i \sin \simgaI t \right)
%\end{align*}

We'll call systems of the form

\begin{align*}
\sigma_{\text{I}} &= 0 \\
\sigma_{\text{R}} &> 0
\end{align*}

\textit{marginally unstable}.  We can have unstable systems with $\sigma_{\text{I}} \ne 0$ but still $\sigma_{\text{R}} > 0$, but these are less common.

\subsection{Case III.  Neutral stability.}

\begin{align*}
\sigma &= 0 \\
\sigma_{\text{R}} &= \sigma_{\text{I}} &= 0
\end{align*}

An example of this was the billard table example where the ball moved to a new location on the table after being bumped slightly.

\section{A mathematical description.}

For fluids we have to bring in the heat equation (which isn't in the scope of this course to derive).  Our equations are the Navier-Stokes

\begin{equation}\label{eqn:continuumL21:50}
\rho \PD{t}{\Bu} + \rho (\Bu \cdot \spacegrad) \Bu = - \spacegrad p + \mu \spacegrad^2 \Bu - \rho \zcap g
\end{equation}

For steady state with $\Bu = 0$ initially (our base state), we'll call the following the equation of the base state

\begin{equation}\label{eqn:continuumL21:70}
\boxed{
\spacegrad p_s = -\rho \zcap g
}
\end{equation}

FIXME: should this be:

\begin{equation}\label{eqn:continuumL21:70bb}
\boxed{
\spacegrad p_s = -\rho_s \zcap g
}
\end{equation}

We'll allow pertubations of each of our variables

\begin{align*}
\Bu &= \Bu_\text{base} + \delta \Bu = 0 + \delta \Bu \\
p &= p_s + \delta p \\
\rho &= \rho_s + \delta \rho
\end{align*}

After pertubation NS takes the form

\begin{equation}\label{eqn:continuumL21:90}
(\rho_s + \delta \rho )\PD{t}{(0 + \delta \Bu)} + (\rho_s + \delta \rho) ((0 + \delta \Bu) \cdot \spacegrad) (0 + \delta \Bu) = - \spacegrad (p_s + \delta p) + \mu \spacegrad^2 (0 + \delta \Bu) - (\rho_s + \delta \rho) \zcap g
\end{equation}

Retaining only terms that are of first order of smallness.

\begin{equation}\label{eqn:continuumL21:110}
\rho_s \PD{t}{\delta \Bu} = - \spacegrad p_s - \spacegrad \delta p + \mu \spacegrad^2 \delta \Bu - \rho_s \zcap g - \delta \rho \zcap g
\end{equation}

applying our equation of base state \ref{eqn:continuumL21:70}, we have
%\spacegrad p_s = -\rho_s \zcap g
\begin{equation}\label{eqn:continuumL21:110b}
\rho_s \PD{t}{\delta \Bu} = \cancel{\rho_s \zcap g} - \spacegrad \delta p + \mu \spacegrad^2 \delta \Bu - \cancel{\rho_s \zcap g} - \delta \rho \zcap g,
\end{equation}

or

\begin{equation}\label{eqn:continuumL21:110c}
\rho_s \PD{t}{\delta \Bu} = - \spacegrad \delta p + \mu \spacegrad^2 \delta \Bu - \delta \rho \zcap g.
\end{equation}

we can write

\begin{equation}\label{eqn:continuumL21:130}
\left( \PD{t}{} - \nu \spacegrad^2 \right) \delta \Bu = -\inv{\rho_s} \spacegrad \delta p - \frac{\delta \rho}{\rho_s} \zcap g
\end{equation}

Applying the divergence operation on both sides, and using $\spacegrad \Bu = 0$ so that $\spacegrad \delta \Bu = 0$ we have

\begin{equation}\label{eqn:continuumL21:150}
\left( \PD{t}{} - \nu \spacegrad^2 \right) \cancel{\spacegrad \cdot \delta \Bu} = -\inv{\rho_s} \spacegrad^2 \delta p - (\zcap \cdot \spacegrad ) \frac{\delta \rho}{\rho_s} g,
\end{equation}

or

\begin{equation}\label{eqn:continuumL21:170}
\inv{\rho_s} \spacegrad^2 \delta p = - (\zcap \cdot \spacegrad ) \frac{\delta \rho}{\rho_s} g.
\end{equation}

Assuming that $\rho_s$ is constant (actually that's already been done above), we can cancel it, leaving

\begin{equation}\label{eqn:continuumL21:190}
\spacegrad^2 \delta p = - (\zcap \cdot \spacegrad ) g \delta \rho = -g \PD{z}{} \delta \rho.
\end{equation}

operating once more with $\PDi{z}{}$ we have

\begin{equation}\label{eqn:continuumL21:210}
\spacegrad^2 \PD{z}{\delta p} = -g \PDSq{z}{\delta \rho}.
\end{equation}

NEW SECTION?  Missed where this came from.  Taking curls?

\begin{equation}\label{eqn:continuumL21:230}
\left( \PD{t}{} - \nu \spacegrad^2 \right) \delta \omega = -\inv{\rho_s} \PD{z}{\delta p} - \frac{\delta \rho}{\rho_s} g
\end{equation}

\begin{align*}
\left( \PD{t}{} - \nu \spacegrad^2 \right) \spacegrad^2 \delta \omega 
&= -\inv{\rho_s} \PD{z}{ \spacegrad^2 \delta p} - \frac{g}{\rho_s} \spacegrad^2 \delta \rho \\
&= -\frac{g}{\rho_s} \PDSq{z}{\delta \rho} - \frac{g}{\rho_s} \spacegrad^2 \delta \rho \\
&= 
-\frac{g}{\rho_s} \left( 
\PDSq{x}{}
+\PDSq{y}{}
\right)
\delta \rho \\
&=
g \alpha \left( 
\PDSq{x}{}
+\PDSq{y}{}
\right)
\delta T
\end{align*}

in the last step we use the following assumed relation for temperature

\begin{equation}\label{eqn:continuumL21:250}
\delta \rho = - \rho_s \alpha \delta T
\end{equation}

(where does this come from?)

We have finally

\begin{equation}\label{eqn:continuumL21:290}
\left( \PD{t}{} - \nu \spacegrad^2 \right) \spacegrad^2 \delta \omega 
= 
g \alpha \left( 
\PDSq{x}{}
+\PDSq{y}{}
\right)
\delta T
\end{equation}

Solving this is the Rayleigh-Benard instabilit problem.

\begin{equation}\label{eqn:continuumL21:270}
\left( \PD{t}{} - \kappa \spacegrad^2 \right) \delta T = \Delta T \frac{\delta \omega}{d}
\end{equation}

%\section{Stuff.}

%FIXME: Reading: \S XX from \cite{acheson1990elementary}

%\EndArticle
\EndNoBibArticle
