\documentclass[]{eliblog}

\usepackage{color}
%\usepackage{txfonts} % for xi

\input{../peeters_macros.tex}

\author{Peeter Joot}
\email{peeter.joot@utoronto.ca, 920798560}
%\input{../peeter_prologue_print.tex}

\chapter{PHY454H1S Continuum mechanics.  Problem Set 3.  Velocity scaling, non-dimensionalisation, boundary layers.}
%\label{chap:continuumProblemSet3}
%\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumProblemSet3.pdf}
\date{Mar 23, 2012}
%\gitRevisionInfo{continuumProblemSet3}

\beginArtNoToc
\section{Disclaimer.}

This problem set is as yet ungraded.

\section{Problem Q1.}
\subsection{Background.}

In fluid convection problems one can make several choices for characteristic velocity scales.  Some choices are given below for example:

\begin{enumerate}
\item $U_1 = \frac{g \alpha d^2 \nabla T}{\nu}$
\item $U_2 = \frac{\nu}{d}$
\item $U_3 = \sqrt{ g \alpha d \nabla T }$
\item $U_4 = \kappa/ d$
\end{enumerate}

where $g$ is the acceleration due to gravity, $\alpha = (\PDi{T}{V})/V$ is the coefficient of volume expansion, $d$ length scale associated with the problem, $\nabla T$ is the applied temperature difference, $\nu$ is the kinematic viscosity and $\kappa$ is the thermal diffusivity.  

\begin{enumerate}
\item For $U_1$:

Observing that 

\begin{equation}\label{eqn:continuumProblemSet3:n}
\left[ \PD{t}{\Bu} \right] = [ \nu \spacegrad^2 \Bu ]
\end{equation}

we must have

\begin{equation}\label{eqn:continuumProblemSet3:n}
[\nu] = \inv{[t] [\spacegrad^2]} = \inv{T} L^2
\end{equation}

We also find

\begin{equation}\label{eqn:continuumProblemSet3:n}
[\alpha] = \inv{[V]} \left[ \PD{T}{V} \right] = \inv{[K]},
\end{equation}

so that

\begin{equation}\label{eqn:continuumProblemSet3:n}
[U_1] = \frac{L}{T \cancel{T}} \cancel{\inv{K}} L^2 \cancel{K} \frac{\cancel{T}}{L^2} = \frac{L}{T}
\end{equation}

\item For $U_2$:

\begin{equation}\label{eqn:continuumProblemSet3:n}
[U_2] = \frac{L^2}{T} \inv{L} = \frac{L}{T}.
\end{equation}

\item For $U_3$

\begin{equation}\label{eqn:continuumProblemSet3:n}
[U_3] = \sqrt{ \frac{L}{T^2} \inv{K} L K } = \frac{L}{T}
\end{equation}

\item For $U_4$

According to \href{http://scienceworld.wolfram.com/physics/ThermalDiffusivity.html}{http://scienceworld.wolfram.com/physics/ThermalDiffusivity.html}, the thermal diffusivity is defined by

\begin{equation}\label{eqn:continuumProblemSet3:n}
\PD{t}{T} = \kappa \spacegrad^2 T
\end{equation}

so that 

\begin{equation}\label{eqn:continuumProblemSet3:n}
[\kappa] = \inv{[t][\spacegrad^2]} = \frac{L^2}{T}
\end{equation}

That gives us

\begin{equation}\label{eqn:continuumProblemSet3:n}
[U_4] = \frac{L^2}{T} \inv{L} = \frac{L}{T}.
\end{equation}
\end{enumerate}

We've verified that all of these have dimensions of velocity.

\subsection{Part 1.  Statement.  Check whether the dimensions match in each case above.}
\subsection{Solution.}

\subsection{Part 2.  Statement. Pure liquid.}

For pure liquid, say pure water at room temperature, one has the following estimates in cgs units:

\begin{align*}
\alpha &\sim 10^{-4} \\
\kappa &\sim 10^{-3} \\
\nu &\sim 10^{-2}
\end{align*}

For a $d \sim 1 \text{cm}$ layer depth and a ten degree temperature drop convective velocities have been experimentally measured of about $10^{-2}$.
% \text{cm}/\text{s}  
With $g \sim 10^{-3}$, calculate the values of $U_1$, $U_2$, $U_3$, and $U_4$.  Which ones of the characteristic velocities $(U_1$, $U_2$, $U_3, U_4)$ do you think are suitable for nondimensionalising the velocity in Navier-Stokes/Energy equation describing the water convection problem?

We have

\begin{itemize}
\item 

\begin{equation}\label{eqn:continuumProblemSet3:n}
U_1 \sim 10^{-3} 10^{-4} (1)^2 10^1 \inv{10^{-2}} = 10^{-4}
\end{equation}

\item

\begin{equation}\label{eqn:continuumProblemSet3:n}
U_2 \sim 10^{-2}/1 = 10^{-2}
\end{equation}

\item

\begin{equation}\label{eqn:continuumProblemSet3:n}
U_3 \sim \sqrt{ 10^{-3} 10^{-4} (1) 10^1 } = 10^{-3}
\end{equation}

\item

\begin{equation}\label{eqn:continuumProblemSet3:n}
U_4 \sim 10^{-3}/1 = 10^{-3}
\end{equation}
\end{itemize}

Use of $U_2 = \nu/d$ gives the closest match to the measured characteristic velocity of $10^{-2}$.

\subsection{Solution.}

\subsection{Part 3.  Statement.  Mantle convection.}

For mantle convection, we have

\begin{align*}
\alpha &\sim 10^{-5} \\
\nu &\sim 10^{21} \\
\kappa &\sim 10^{-2} \\
d &\sim 10^8 \\
\nabla T &\sim 10^3,
\end{align*}

and the actual convective mantle velocity is $10^{-8}$.  Which of the characteristic velocities should we use to nondimensionalise Navier-Stokes/Energy equations describing mantle convection?
\subsection{Solution}

Let's compute the characteristic velocities again with the mantle numbers

\begin{align}\label{eqn:continuumProblemSet3:n}
U_1 &\sim \frac{10^{-3} 10^{-5} 10^{16} 10^3 }{10^{21}} = 10^{-10} \\
U_2 &\sim \frac{10^{21}}{10^8} = 10^{13} \\
U_3 &\sim \sqrt{ 10^{-3} 10^{-5} 10^8 10^3 } \sim 10^1 \\
U_4 &\sim \frac{10^{-2}}{10^8} = 10^{-10}
\end{align}

Both $U_1$ and $U_4$ come close to the actual convective mantle velocity of $10^{-8}$.  Use of $U_1$ to nondimensionalise is probably best, since it has more degrees of freedom, and includes the gravity term that is probably important for such large masses.

\section{Problem Q2.}
\subsection{Statement}

Nondimensionalise N-S equation

\begin{equation}\label{eqn:continuumL19:n}
\rho \PD{t}{\Bu} + \rho (\Bu \cdot \spacegrad) \Bu = - \spacegrad p + \mu \spacegrad^2 \Bu + \rho g \zcap
\end{equation}

where $\zcap$ is the unit vector in the $z$ direction.  You may scale:

\begin{itemize}
\item velocity with the characteristic velocity $U$,
\item time with $R/U$, where $R$ is the characteristic length scale,
\item pressure with $\rho U^2$,
\end{itemize}

Reynolds number $\text{Re} = R U \rho/ \mu$ and Froude number $\text{Fr} = g R/U$.

\subsection{Solution}

\section{Problem Q3.}
\subsection{Statement}

In case of Stokes' boundary layer problem (see class note) calculate shear stress on the plate $y = 0$.   What is the phase difference between the velocity of the plate $U(t) = U_0 \cos\omega t$ and the shear stress on the plate?

\subsection{Solution}

We found in class that the velocity of the fluid was given by

\begin{equation}\label{eqn:continuumL19:n}
u(y, t) = U_0 e^{-\lambda y} \cos(\lambda y - \omega t)
\end{equation}

where

\begin{equation}\label{eqn:continuumL19:n}
\lambda = \sqrt{\frac{\omega}{2 \nu}}
\end{equation}

Calculating our shear stress we find

\begin{align*}
\mu \PD{y}{u} 
&= U_0 \lambda \mu e^{-\lambda y}
\left(
-1
-
 \sin(\lambda y - \omega t)
\right)
\end{align*}

and on the plate ($y = 0$) this is just

\begin{equation}\label{eqn:continuumL19:n}
\evalbar{\mu \PD{y}{u}}{y = 0} = U_0 \lambda \mu (-1 + \sin(\omega t)).
\end{equation}

We've got a constant term, plus one that is sinusoidal.  Observing that 

\begin{align}\label{eqn:continuumProblemSet3:n}
\cos x &= \Real( e^{ix} )  \\
\sin x &= \Real( -i e^{ix} ) = \Real( e^{i (x - \pi/2)} ),
\end{align}

The phase difference between the non-constant portions of the shear stress at the plate, and the plate velocity $U(t) = U_0 \cos\omega t$ is just $-\pi/2$.  The shear stress at that points lags by 90 degrees.

%\EndArticle
\EndNoBibArticle
