\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY454H1S\\Continuum Mechanics.  Lecture 8: Phasor description of elastic waves.  Fluid dynamics.  Taught by Prof. K. Das.}
\label{chap:continuumL8}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumL8.pdf}
\date{Feb 3, 2012}
\revisionInfo{continuumL8.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Disclaimer.}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{Review.  Elastic wave equation}

Starting with

\begin{equation}\label{eqn:continuumL8:10}
\rho \PDSq{t}{\Be} = (\lambda + \mu) \spacegrad (\spacegrad \cdot \Be) + \mu \spacegrad^2 \Be
\end{equation}

and applying a divergence operation we find

\begin{align}\label{eqn:continuumL8:30}
\rho \PDSq{t}{\theta} &= C_L^2 \spacegrad^2 \theta \\
\theta &= \spacegrad \cdot \Be \\
C_L^2 &= \frac{\lambda + 2\mu}{\rho}.
\end{align}

This is the P-wave equation.  Applying a curl operation we find

\begin{align}\label{eqn:continuumL8:50}
\rho \PDSq{t}{\Bomega} &= C_T^2 \spacegrad^2 \Bomega \\
\Bomega &= \spacegrad \cross \Be \\
C_T^2 &= \frac{\lambda + 2\mu}{\rho}.
\end{align}

This is the S-wave equation.  We also found that

\begin{equation}\label{eqn:continuumL8:70}
\frac{C_L}{C_T} > 1,
\end{equation}

and concluded that P waves are faster than S waves.  What we haven't shown is that the P waves are longitudinal, and that the S waves are transverse.

Assuming a gradient and curl description of our displacement

\begin{equation}\label{eqn:continuumL8:90}
\Be = \spacegrad \phi + \spacegrad \cross \BH = \BP + \BS,
\end{equation}

we found

\begin{align}\label{eqn:continuumL8:110}
(\lambda + 2 \mu) \spacegrad^2 \phi - \rho \PDSq{t}{\phi} &= 0 \\
\mu \spacegrad^2 \BH - \rho \PDSq{t}{\BH} &= 0,
\end{align}

allowing us to separately solve for the P and the S wave solutions respectively.  Now, let's introduce a phasor representation (again following \S 22 of the text \cite{landau1960theory})

\begin{align}\label{eqn:continuumL8:130}
\phi &= A \exp\left( i ( \Bk \cdot \Bx - \omega t) \right) \\
\BH &= \BB \exp\left( i ( \Bk \cdot \Bx - \omega t) \right)
\end{align}

Operating with the gradient we find

\begin{align*}
\BP
&= \spacegrad \phi \\
&= \Be_k \partial_k A \exp\left( i ( \Bk \cdot \Bx - \omega t) \right) \\
&= \Be_k \partial_k A \exp\left( i ( k_m x_m - \omega t) \right) \\
&= \Be_k i k_k A \exp\left( i ( k_m x_m - \omega t) \right) \\
&= i \Bk A \exp\left( i ( \Bk \cdot \Bx - \omega t) \right) \\
&= i \Bk \phi
\end{align*}

We can also write

\begin{equation}\label{eqn:continuumL8:150}
\BP = \Bk \phi'
\end{equation}

where $\phi'$ is the derivative of $\phi$ ``with respect to its argument''.   Here argument must mean the entire phase $\Bk \cdot \Bx - \omega t$

\begin{equation}\label{eqn:continuumL8:170}
\phi' = \frac{ d\phi( \Bk \cdot \Bx - \omega t )}{ d(\Bk \cdot \Bx - \omega t) } = i \phi
\end{equation}

For the curl term we find

\begin{align*}
\BS
&= \spacegrad \cross \BH \\
&= \Be_a \partial_b H_c \epsilon_{a b c} \\
&= \Be_a \partial_b \epsilon_{a b c} B_c \exp\left( i ( \Bk \cdot \Bx - \omega t) \right) \\
&= \Be_a \partial_b \epsilon_{a b c} B_c \exp\left( i ( k_m x_m - \omega t) \right) \\
&= \Be_a i k_b \epsilon_{a b c} B_c \exp\left( i ( \Bk \cdot \Bx - \omega t) \right) \\
&= i \Bk \cross \BH
\end{align*}

Again writing
\begin{equation}\label{eqn:continuumL8:190}
\BH' = \frac{ d\BH( \Bk \cdot \Bx - \omega t )}{ d(\Bk \cdot \Bx - \omega t) } = i \BH
\end{equation}

we can write the S wave as

\begin{equation}\label{eqn:continuumL8:210}
\BS = \Bk \cross \BH'
\end{equation}

\subsection{Some waves illustrated.}

The following wave types were noted, but not defined:

\begin{itemize}
\item Love wave
\item Rayleigh wave
\end{itemize}

Some illustrations from the class notes were also shown.

\section{Fluid dynamics.}

%FIXME: Reading: \S XX from \cite{landau1960theory}.

In fluid dynamics we look at displacements with respect to time as illustrated in

FIXME: F1.

\begin{equation}\label{eqn:continuumL8:230}
d\Bx' = d\Bx + d\Bu \delta t
\end{equation}

In index notation

\begin{align*}
dx_i'
&= dx_i + du_i \delta t \\
&= dx_i + \PD{x_j}{u_i} dx_j \delta t
\end{align*}

We define

\begin{equation}\label{eqn:continuumL8:250}
e_{ij} = \inv{2} \left(
\PD{x_j}{u_i} +
\PD{x_i}{u_j} \right)
\end{equation}

a symmetric tensor.  We also define

\begin{equation}\label{eqn:continuumL8:270}
\omega_{ij} = \inv{2} \left(
\PD{x_j}{u_i}
-\PD{x_i}{u_j} \right)
\end{equation}

Effect of $e_{ij}$ when diagonalized

\begin{equation}\label{eqn:continuumL8:290}
e_{ij} =
=
\begin{bmatrix}
e_{11} & 0 & 0 \\
0 & e_{22} & 0 \\
0 & 0 & e_{33}
\end{bmatrix}
\end{equation}

FIXME: check:
\begin{align}\label{eqn:continuumL8:310}
dx_1' &= ( 1 + u_1 \delta t) dx_1 \\
dx_2' &= ( 1 + u_2 \delta t) dx_2 \\
dx_3' &= ( 1 + u_3 \delta t) dx_3
\end{align}

Let's find the matrix form of the antisymmetric tensor.  We find

\begin{equation}\label{eqn:continuumL8:330}
\omega_{11} = \omega_{22} = \omega_{33} = 0
\end{equation}

Introducing a vorticity vector

\begin{equation}\label{eqn:continuumL8:350}
\Bomega = \spacegrad \cross \Bu
\end{equation}

we find

\begin{align}\label{eqn:continuumL8:370}
\omega_{12} &= \inv{2}\left( \PD{x_2}{u_1} -\PD{x_1}{u_2} \right) = - \inv{2} (\spacegrad \cross \Bu)_3 \\
\omega_{23} &= \inv{2}\left( \PD{x_3}{u_2} -\PD{x_2}{u_3} \right) = - \inv{2} (\spacegrad \cross \Bu)_1 \\
\omega_{31} &= \inv{2}\left( \PD{x_1}{u_3} -\PD{x_3}{u_1} \right) = - \inv{2} (\spacegrad \cross \Bu)_2
\end{align}

Writing

\begin{equation}\label{eqn:continuumL8:390}
\Omega_i = \inv{2} \omega_i
\end{equation}

we find the matrix form of this antisymmetric tensor

\begin{equation}\label{eqn:continuumL8:410}
\omega_{ij}
=
\begin{bmatrix}
0 & -\Omega_3 & \Omega_2 \\
\Omega_3 & 0 & -\Omega_1 \\
-\Omega_2 & \Omega_1 & 0 \\
\end{bmatrix}
\end{equation}

\begin{align*}
dx_1'
&= dx_1 + \left( \cancel{\omega_{11}} dx_1 + \omega_{12} dx_2 + \omega_{13} dx_3 \right) \delta t \\
&= dx_1 + \left( \omega_{12} dx_2 + \omega_{13} dx_3 \right) \delta t \\
&= dx_1 + \left( \Omega_2 dx_3 - \Omega_3 dx_2 \right) \delta t
\end{align*}

Doing this for all components we find

\begin{equation}\label{eqn:continuumL8:430}
d\Bx' = d\Bx + (\BOmega \cross d\Bx) \delta t.
\end{equation}

The tensor $\omega_{ij}$ implies rotation of a control volume with an angular velocity $\BOmega = \Bomega/2$ (half the vorticity vector).

In general we have

\begin{equation}\label{eqn:continuumL8:450}
dx_i' = dx_i + e_{ij} dx_j \delta t + \omega_{ij} dx_j \delta t
\end{equation}

\subsection{Making sense of things.}

After this first fluid dynamics lecture I was left troubled.  We'd just been barraged with a set of equations pulled out of a magic hat, with no notion of where they came from.  Unlike the contiuum strain tensor, which was derived by considering differences in squared displacements, we have an antisymmetric term now.  Why did we have no such term considering solids?

After a bit of thought I think I see where things are coming from.  We have essentially looked at a first order decomposition of the displacement (per unit time) of a point in terms of symmetric and antisymmetric terms.  This is really just a gradient evaluation, split into coordinates

\begin{align*}
x_i' 
&= x_i + (\spacegrad u_i) \cdot d\Bx \delta t \\
&= x_i + \PD{x_j}{u_i} dx_j \delta t \\
&= x_i + 
\inv{2}
\left(\PD{x_j}{u_i} 
+\PD{x_j}{u_i} 
\right)
dx_j \delta t 
+
\inv{2}
\left(\PD{x_j}{u_i} 
-\PD{x_j}{u_i} 
\right)
dx_j \delta t  \\
&=
x_i + e_{ij} dx_j \delta t + \omega_{ij} dx_j \delta t
\end{align*}

Here, as in the solids case, we have

\begin{equation}\label{eqn:continuumL8:n}
\Bu = \Bx' - \Bx
\end{equation}

Employing vector notation we can write our first order displacement as

\begin{align*}
\Bx' 
&= \Bx + \Be_i (\spacegrad u_i) \cdot d\Bx \delta t \\
&= \Bx + \Be_i \inv{2}
\left(
\spacegrad u_i d\Bx
+d\Bx
\spacegrad u_i
\right)
\delta t
\end{align*}

\EndArticle
