\documentclass[]{eliblog}

\usepackage{color}
%\usepackage{txfonts} % for xi

\input{../peeters_macros.tex}

\author{Peeter Joot}
\email{peeter.joot@utoronto.ca, 920798560}
%\input{../peeter_prologue_print.tex}

\chapter{PHY456H1S Problem Set 1.}
%\label{chap:continuumProblemSet1}
%\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumProblemSet1.pdf}
%\date{Feb 3, 2012}
%\revisionInfo{continuumProblemSet1.tex}

\beginArtNoToc
%\section{Disclaimer.}
%
%This problem set is as yet ungraded.

\section{Problem Q1.}
\subsection{Statement}

For the stress tensor

\begin{equation}\label{eqn:continuumProblemSet1:10}
\sigma = 
\begin{bmatrix}
6 & 0 & 2 \\
0 & 1 & 1 \\
2 & 1 & 3
\end{bmatrix}
\text{M Pa}
\end{equation}

Find the corresponding strain tensor, assuming an isotropic solid with Young's modulus $E = 200 \times 10^9 \text{N}/\text{m}^2$ and Poisson's ration $\nu = 0.35$.

\subsection{Solution}

We need to express the relation between stress and strain in terms of Young's modulus and Poisson's ratio.  In terms of Lam\'e parameters our model for the relations between stress and strain for an isotropic solid was given as

\begin{equation}\label{eqn:continuumProblemSet1:110}
\sigma_{ij} = \lambda e_{kk} \delta_{ij} + 2 \mu e_{ij}.
\end{equation}

Computing the trace

\begin{equation}\label{eqn:continuumProblemSet1:130}
\sigma_{kk} = (3 \lambda + 2 \mu) e_{kk},
\end{equation}

allows us to invert the relationship

\begin{equation}\label{eqn:continuumProblemSet1:150}
2 \mu e_{ij} = \sigma_{ij} - \lambda \frac{\sigma_{kk}}{3 \lambda + 2 \mu} \delta_{ij}.
\end{equation}

In terms of Poisson's ratio $\nu$ and Young's modulus $E$, our Lam\'e parameters were found to be

\begin{align}\label{eqn:continuumProblemSet1:170}
\lambda &= \frac{ E \nu }{(1 - 2 \nu)(1 + \nu)} \\
\mu &= \frac{E}{2(1 + \nu)},
\end{align}

and

\begin{align*}
3 \lambda + 2 \mu
&= \frac{ 3 E \nu }{(1 - 2 \nu)(1 + \nu)} + \frac{E}{1 + \nu} \\
&= \frac{E}{1 + \nu} \left( \frac{3 \nu}{1 - 2 \nu} + 1\right) \\
&= \frac{E}{1 + \nu} \frac{1 + \nu}{1 - 2 \nu} \\
&= \frac{E}{1 - 2 \nu}.
\end{align*}

Our stress strain model for the relationship for an isotropic solid becomes
we find

\begin{align*}
\frac{E}{1 + \nu} e_{ij} 
&= 
\sigma_{ij} 
-
\frac{ E \nu }{(1 - 2 \nu)(1 + \nu)} \frac{1 - 2 \nu}{E}
\sigma_{kk} \delta_{ij} \\
&=
\sigma_{ij} 
-
\frac{ \nu }{1 + \nu} 
\sigma_{kk} \delta_{ij} \\
\end{align*}

or

\begin{equation}\label{eqn:continuumProblemSet1:190}
e_{ij} 
= 
\inv{E}
\left(
(1 + \nu)
\sigma_{ij} 
-
\nu
\sigma_{kk} \delta_{ij} 
\right).
\end{equation}

As a sanity check note that this matches (5.12) of \cite{landau1960theory}, although they use a notation of $\sigma$ instead of $\nu$ for Poisson's ratio.  We are now ready to tackle the problem.  First we need the trace of the stress tensor

\begin{equation}\label{eqn:continuumProblemSet1:210}
\sigma_{kk} = (6 + 1 + 3) \text{M Pa} = 10 \text{M Pa},
\end{equation}

\begin{align*}
e_{ij} 
= 
\inv{E}
\left(
(1 + \nu)
\begin{bmatrix}
6 & 0 & 2 \\
0 & 1 & 1 \\
2 & 1 & 3
\end{bmatrix}
-
10 \nu
\begin{bmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1 \\
\end{bmatrix}
\right)
\text{M Pa} \\
&=
\inv{E}
\left(
\begin{bmatrix}
6 & 0 & 2 \\
0 & 1 & 1 \\
2 & 1 & 3
\end{bmatrix}
+ 0.35
\begin{bmatrix}
-4 & 0 & 2 \\
0 & -9 & 1 \\
2 & 1 & -7
\end{bmatrix}
\right)
\text{M Pa} \\
\end{align*}

Expanding out the last bits of arithmetic the strain tensor is found to have the form

\begin{equation}\label{eqn:continuumProblemSet1:230}
e_{ij} 
= 
\begin{bmatrix}
 23 & 0 & 13.5 \\
 0 & -10.75 & 6.75 \\
 13.5 & 6.75 & 2.75 
\end{bmatrix}
 10^{-6}.
\end{equation}

Note that this is dimensionless, unlike the stress.

\section{Problem Q2.}
\subsection{Statement}

Small displacement field in a material is given by

\begin{align}\label{eqn:continuumProblemSet1:30}
e_1 &= 2 x_1 x_2 \\
e_2 &= x_3^2 \\
e_3 &= x_1^2 - x_3 
\end{align}

Find

\begin{enumerate}
\item the infinitesimal strain tensor $e_{ij}$,
\item the principal strains and the corresponding principal axes at $(x_1, x_2, x_3) = (1, 2, 4)$,
\item Is the body under compression or expansion?
\end{enumerate}

\subsection{Solution}

\section{Problem Q3.}
\subsection{Statement}

The stress tensor at a point has components given by 

\begin{equation}\label{eqn:continuumProblemSet1:50}
\sigma = 
\begin{bmatrix}
1 & -2 & 2 \\
-2 & 3 & 1 \\
2 & 1 & -1
\end{bmatrix}.
\end{equation}

Find the traction vector across an area normal to the unit vector

\begin{equation}\label{eqn:continuumProblemSet1:70}
\ncap = ( \sqrt{2} \Be_1 - \Be_2 + \Be_3)/2
\end{equation}

Can you construct a tangent vector $\Btau$ on this plane by inspection?  What are the components of the force per unit area along the normal $\ncap$ and tangent $\Btau$ on that surface?  (hint: projection of the traction vector.)

\subsection{Solution}

\section{Problem Q4.}
\subsection{Statement}

The stress tensor of a body is given by

\begin{equation}\label{eqn:continuumProblemSet1:90}
\sigma = 
\begin{bmatrix}
A \cos x & y^2 & C x \\
y^2 & B \sin y & z \\
C x & z & z^3
\end{bmatrix}
\end{equation}

Determine the constant $A$, $B$, and $C$ if the body is in equilbrium.

\EndArticle
