\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY454H1S\\Continuum Mechanics.  Lecture 2.  Intro.  Taught by Prof. K. Das.}
\label{chap:continuumL2}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumL2.pdf}
\date{Jan 13, 2012}
\revisionInfo{continuumL2.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Introduction.}

Mechanics could be defined as the study of effects of forces and displacements on a physical body.

FIXME: F1.

In continuum mechanics we have a physical body and we are interested in the internal motions in the object.  

FIXME: F2.

For the first time considering mechanics we have to introduce the concepts of fields to make progress tackling these problems.

We will have use of the following types of fields

\begin{itemize}
\item Scalar fields.  $3^0$ components.  Examples: density, Temperature, ...
\item Vector fields.  $3^1$ components.  Examples: Force, velocity.
\item Tensor fields.  $3^2$ components.  Examples: stress, strain.
\end{itemize}

We have to consider objects (a control volume) that is small enough that we can consider that we have a point in space limit for the quantities of density and velocity.  At the same time we cannot take this limiting process to the extreme, since if we use a control volume that is sufficiently small, quantum and inter-atomic effects would have to be considered.

FIXME: F3.

\section{Stress and Strain.}

\begin{definition}
\emph{(Stress)}
\label{dfn:continuumL2:10}
Measure of the Internal force on the surfaces.
\end{definition}

\begin{definition}
\emph{(Strain)}
\label{dfn:continuumL2:30}
Measure of the deformation of the body.
\end{definition}

Utilizing summation convention consider a set of small internal displacements $u_1, u_2, u_3$ to the $x, y, z$ coordinates so that the transformation $x_i \rightarrow x_i'$ is related by

FIXME: F4

\begin{align}\label{eqn:continuumL2:50}
u_i &= x_i' - x_i \\
x_i' &= g(x_i) \\
u_i &= f(x_i)
\end{align}

(ie: $x_i'$ is a function of all the initial coordinates, as are the displacements $u_i$).

\begin{equation}\label{eqn:continuumL2:70}
dx_i' = dx_i + du_i
\end{equation}

\begin{align}\label{eqn:continuumL2:90}
dl &= \sqrt{dx_k dx_k} \\
dl' &= \sqrt{d{x'}_k d{x'}_k}
\end{align}

or

\begin{equation}\label{eqn:continuumL2:110}
{dl'}^2 = 
(dx_k + du_k)
(dx_k + du_k)
= 
dl^2 + 2 dx_k du_k + du_k du_k
\end{equation}

with 

\begin{equation}\label{eqn:continuumL2:130}
du_i = \PD{x_k}{u_i} dx_k
\end{equation}

we have

\begin{equation}\label{eqn:continuumL2:150}
du_i^2 = 
\PD{x_k}{u_i} dx_k
\PD{x_l}{u_i} dx_l
\end{equation}

\begin{align*}
{dl'}^2 
&= 
dl^2 
+ 2 \PD{x_k}{u_i} dx_k dx_i 
+ \PD{x_i}{u_l} 
\PD{x_k}{u_l} 
dx_i dx_k \\
&= 
dl^2 
+ 
\left(
\PD{x_k}{u_i} 
+
\PD{x_i}{u_k} 
\right)
dx_k dx_i 
+ \PD{x_i}{u_l} 
\PD{x_k}{u_l} 
dx_i dx_k \\
&=
dl^2 
+ 
2 e_{ik} dx_i dx_k
\end{align*}

We write

\begin{equation}\label{eqn:continuumL2:170}
{dl'}^2 - dl^2 = 2 e_{ik} dx_i dx_k
\end{equation}

where we define the \emph{strain tensor} as

\begin{equation}\label{eqn:continuumL2:190}
e_{ik} = \inv{2} \left(
\left(
\PD{x_k}{u_i} 
+
\PD{x_i}{u_k} 
\right)
+ \PD{x_i}{u_l} 
\PD{x_k}{u_l} 
\right)
\end{equation}

Here $e_{ik}$ is a $3 \times 3$ matrix in Cartesian coordinates

\begin{equation}\label{eqn:continuumL2:210}
\begin{bmatrix}
e_{11} & e_{12} & e_{13} \\
e_{21} & e_{22} & e_{23} \\
e_{31} & e_{32} & e_{33} \\
\end{bmatrix}
\end{equation}

We see from \ref{eqn:continuumL2:190} that $e_{ik}$ is symmetric, so we have

\begin{align}\label{eqn:continuumL2:n}
e_{21} &= e_{12} \\
e_{31} &= e_{13} \\
e_{32} &= e_{23}
\end{align}

Because any real symmetric matrix can be diagonalized we can write in some coordinate system

\begin{equation}\label{eqn:continuumL2:n}
\bar{e}_{ik} = 
\begin{bmatrix}
\bar{e}_{11} & 0 & 0 \\
0 & \bar{e}_{22} & 0 \\
0 & 0 & \bar{e}_{33} \\
\end{bmatrix}
\end{equation}

\begin{align}\label{eqn:continuumL2:n}
{dx_1'}^2 &= (1 + 2 \bar{e}_{11}) dx_1^2 \\
{dx_2'}^2 &= (1 + 2 \bar{e}_{22}) dx_2^2 \\
{dx_3'}^2 &= (1 + 2 \bar{e}_{33}) dx_3^2
\end{align}

If our changes are small enough we can also write approximately

\begin{align}\label{eqn:continuumL2:n}
{dx_1'}^2 &\approx (1 + 2 e_{11}) dx_1^2 \\
{dx_2'}^2 &\approx (1 + 2 e_{22}) dx_2^2 \\
{dx_3'}^2 &\approx (1 + 2 e_{33}) dx_3^2
\end{align}

FIXME: some hand waving here to think through.

We are also free to define a volume element

\begin{equation}\label{eqn:continuumL2:n}
dV' = 
dx_1'
dx_2'
dx_3'
\approx
(1 + e_{11})
(1 + e_{22})
(1 + e_{33})
dx_1 dx_2 dx_3
\end{equation}

\begin{equation}\label{eqn:continuumL2:n}
dV' = (1 + e_{11} +e_{22} +e_{33} ) dV
\end{equation}

So the change of volume is given by the trace

\begin{equation}\label{eqn:continuumL2:n}
dV' = ( 1 + e_{ii} )^2 dV
\end{equation}

\EndArticle
%\EndNoBibArticle
