\documentclass[]{eliblog}

\usepackage{color}
%\usepackage{txfonts} % for xi

\input{../peeters_macros.tex}

\author{Peeter Joot}
\email{peeter.joot@utoronto.ca, 920798560}
%\input{../peeter_prologue_print.tex}

\chapter{PHY454H1S Continuum mechanics.  Problem Set 2.  XXX}
%\label{chap:continuumProblemSet2}
%\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumProblemSet2.pdf}
%\date{Mar 2, 2012}
%\gitRevisionInfo{continuumProblemSet2}

\beginArtNoToc
%\section{Disclaimer.}
%
%This problem set is as yet ungraded.

\section{Problem Q1.}
\subsection{Statement}

Imagine a steady rectilinear blood flow of the form $\Bu = u(y) \ycap$ through a two dimensional atery.  It is driven by a constant pressure gradient $G = -dp/dx$ maintained by an external `heart'.  The top and bottom walls of the arery are $2h$ distance apart and the fluid satisfies no-slip boundary conditions at the walls.  Assuming that the fluid is Newtonian,

\begin{enumerate}
\item Show that the Navier-Stokes equation reduces to

\begin{equation}\label{eqn:continuumProblemSet2:n}
\frac{d^2 u}{dy^2} = -\frac{G}{\mu}
\end{equation}

where $\mu$ is the viscosity of the blood.

\item Show that the velocity profile of the fluid inside the artery is a parabolic profile.
\item What is the maximum speed of the fluid?  Draw the velocity profile to show where the maximum speed occurs inside the artery.
\item If due to smoking etc., the vicosity of the blood gets doubled, then what should be the new pressure gradient to be maintained by the `heart' to keep the liquid flux through the arery at the same level as the non-smoking one?
\end{enumerate}

\subsection{Solution.  Part 1.  Navier-Stokes equation for the system.}

In component form 

\subsection{Solution.  Part 2.  Velocity profile.}
\subsection{Solution.  Part 3.  Maximum speed of the flow.}
\subsection{Solution.  Part 4.  Effects of viscosity doubling.}

\section{Problem Q2.}
\subsection{Statement}

Consider steady simple shearing flow with no imposed pressure gradient $(G = 0)$ of a two layer fluid with viscosity

\begin{equation}\label{eqn:continuumProblemSet2:n}
\mu =
\left\{
\begin{array}{l l}
\mu_1 & \quad \mbox{$-h < y < 0,$} \\
\mu_2 & \quad \mbox{$0 < y < h.$}
\end{array}
\right.
\end{equation}

The boundary conditions are no-slip at the lower plate $(y = -h)$ and at $y = 0$.  The top plate is moving with a volocity $-U$ at $y = h$ and fluid is sticking to it.  using the continuity of tangential (shear) stress at the interface ($y = 0$)

\begin{itemize}
\item Derive the velocity profile of the two fluids.
\item Calculate the maximum speed.
\item Calculate the mean speed.
\item Calculate the flux (the volume flow rate.)
\item Calculate the tangential force (per unit width) $F_x$ on the strip $0 \le x \le L$ of the wall $y = -h$.
\item Calculate the tangential force (per unit width) $F_x^0$ on the strip $0 \le x \le L$ at the interfce $y = 0$ by the top fluid on the lower fluid.
\end{itemize}

\subsection{Part 1.  Derive the velocity profile of the two fluids.}
\subsection{Part 2.  Calculate the maximum speed.}
\subsection{Part 3.  Calculate the mean speed.}
\subsection{Part 4.  Calculate the flux}
\subsection{Part 5.  Tangential force on the wall.}
\subsection{Part 6.  Tangential force on the lower fluid.}

\section{Problem Q3.}
\subsection{Statement}

If on top of the problem described above a constant pressure gradient $G = -dp/dx$ is applied between the boundaries $y = \pm h$, describe qualitatively what type of flow profile you would expect in the steady state.  Draw the velocity profiles for two cases (i) $\mu_1 > \mu_2$ (ii) $\mu_1 < \mu_2$.  Explain your result.

\subsection{Solution}

%\EndArticle
\EndNoBibArticle
