%
% Copyright © 2016 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%{
\input{../blogpost.tex}
\renewcommand{\basename}{curlAndDivergenceVectorUniqueness}
\renewcommand{\dirname}{notes/phy1520/}
%\newcommand{\dateintitle}{}
%\newcommand{\keywords}{}

\input{../peeter_prologue_print2.tex}

\usepackage{peeters_layout_exercise}
\usepackage{peeters_braket}
\usepackage{peeters_figures}
\usepackage{siunitx}

\beginArtNoToc

\generatetitle{Does the divergence and curl uniquely determine the vector}
%\chapter{Does the divergence and curl uniquely determine the vector}
%\label{chap:curlAndDivergenceVectorUniqueness}
% \citep{sakurai2014modern} pr X.Y
% \citep{pozar2009microwave}
% \citep{qftLectureNotes}
% \citep{griffiths1999introduction}

Here's a homework question from ece1228 that screams for solution using Geometric Algebra techniques.  If I submitted such a solution, then my Prof probably wouldn't know what I was doing, so I'll probably also have to try to solve it another way too.

\makeproblem{Helmholtz theorem}{emt:problemSet1:5}{ 
Prove the first Helmholtz's theorem, i.e. if vector \(\BM_1\) is defined by its divergence

\begin{dmath}\label{eqn:emtProblemSet1Problem5:20}
\spacegrad \cdot \BM_1 = s
\end{dmath}

and its curl
\begin{dmath}\label{eqn:emtProblemSet1Problem5:40}
\spacegrad \cross \BM_1 = \BC 
\end{dmath}

within a region and its normal component \( \BM_{1\txtn} \) over the boundary, then \( \BM_1 \) is 
uniquely specified.

Note: Assume there is a vector \( \BM_2 \) with its divergence and curl equal to \( s \) and \( \BC \)
respectively, then show that \( \BM_1 = \BM_2 \) .
} % makeproblem

\makeanswer{emt:problemSet1:5}{ 

It will be important to note that \( \BC \) cannot be arbitrary.  A divergence constraint must be imposted on \( \BC \).  In particular, because \( \BC = \spacegrad \cross \BM_1 \), and \( \spacegrad \cdot (\spacegrad \cross \BM_1) = 0 \), the vector \( \BC \) is required to have zero divergence.

With that said, dropping the suffix from \( \BM_1 \), the gradient of the vector can be written as a single even grade multivector

\begin{dmath}\label{eqn:curlAndDivergenceVectorUniqueness:60}
\spacegrad \BM
= \spacegrad \cdot \BM + I \spacegrad \cross \BM
= s + I \BC.
\end{dmath}

Unlike the divergence or curl, the gradient is invertible, with the \R{3} Green's function

\begin{dmath}\label{eqn:curlAndDivergenceVectorUniqueness:80}
\begin{aligned}
G(\Bx ; \By) &= \inv{4 \pi} \frac{ \Bx - \By }{\Abs{\Bx - \By}^3} \\
\spacegrad \BG &= \spacegrad \cdot \BG = \delta(\Bx - \By).
\end{aligned}
\end{dmath}

This Green's function result is taken from \citep{doran2003gap}, where it is used to generalize the Cauchy integral equations to higher dimensions.

The inversion equation is an application of the Fundamental Theorem of (Geometric) Calculus

\begin{dmath}\label{eqn:curlAndDivergenceVectorUniqueness:100}
\oint_{\partial V} G d^2 \Bx \BM
=
\int_V G d^3 \Bx \lrspacegrad \BM 
=
\int_V d^3 \Bx (G \spacegrad) \BM 
+
\int_V d^3 \Bx G (\spacegrad \BM)
=
\int_V d^3 \Bx \delta(\Bx - \By) \BM 
+
\int_V d^3 \Bx G \lr{ s + I \BC }.
\end{dmath}

In the first volume integral the gradient operates bidirectionally on both \( G \) and \( \BM \).  Because \( d^3 \Bx \propto I \) in \R{3} we can commute it with any grades.  With \( d^3 \Bx = I dV \), and \( d^2 \Bx \ncap = I dA \), we have

\begin{dmath}\label{eqn:curlAndDivergenceVectorUniqueness:120}
\BM(\By)
=
\inv{4\pi} \oint_{\partial V} dA \frac{ \Bx - \By }{\Abs{\Bx - \By}^3} \ncap \BM
-
\inv{4\pi} \int_V dV \frac{ \Bx - \By }{\Abs{\Bx - \By}^3} \lr{ s + I \BC }.
\end{dmath}

The trivectors grades on the RHS must sum to zero.  Omitting those for now, 
%Since \( \BM \) is a vector the vector and trivector grades on the RHS must sum to zero.  
with \( \Br = \Bx - \By \), the vector grade selection is

\begin{dmath}\label{eqn:curlAndDivergenceVectorUniqueness:121}
\BM(\By)
=
\inv{4\pi} \oint_{\partial V} dA \inv{ r^2 } \gpgradeone{ \rcap \ncap \BM }
-
\inv{4\pi} \int_V dV \frac{ s \rcap }{ r^2 } 
-
\inv{4\pi} \int_V dV \frac{ \rcap \cdot (I\BC) }{ r^2 }
=
\inv{4\pi} \oint_{\partial V} dA \inv{ r^2 } \gpgradeone{ \rcap \ncap \BM }
+
\inv{4\pi} \int_V dV \frac{ \rcap \cross \BC - s \rcap }{ r^2 },
\end{dmath}

where the cross product comes from the expansion

\begin{dmath}\label{eqn:curlAndDivergenceVectorUniqueness:140}
\rcap \cdot (I\BC)
=
\gpgradeone{ \rcap I\BC }
=
I (\rcap \wedge \BC)
=
-\rcap \cross \BC.
\end{dmath}

The vector grade selection can be expanded as 

\begin{dmath}\label{eqn:curlAndDivergenceVectorUniqueness:220}
\gpgradeone{ \rcap \ncap \BM }
=
\lr{ \rcap (\ncap \cdot \BM) + \rcap \cdot (\ncap \wedge \BM) }
=
\rcap (\ncap \cdot \BM) + (\rcap \cdot \ncap) \BM - \ncap ( \rcap \cdot \BM ).
\end{dmath}

This looks like it is probably largest when all of \( \rcap, \ncap, \BM \) are colinear, so will likely have magnitude \( \gpgradeone{ \rcap \ncap \BM } < \Abs{\BM} \).  What we can say for sure is 

\begin{dmath}\label{eqn:curlAndDivergenceVectorUniqueness:240}
\Abs{\lr{ \inv{4\pi} \oint_{\partial V} dA \inv{ r^2 } \gpgradeone{ \rcap \ncap \BM } }}
\ge
\inv{4\pi} \oint_{\partial V} dA \inv{ r^2 } \Abs{ \gpgradeone{ \rcap \ncap \BM } }
\ge
\inv{4\pi} \oint_{\partial V} dA \inv{ r^2 } 3 \Abs{ \BM },
\end{dmath}

so if \( \lim_{\Bx' \rightarrow \infty} \ifrac{\Abs{\BM(\Bx')}}{\Abs{\Bx - \Bx'}^2 } \rightarrow 0 \), this whole surface integral will be killed,
leaving \( \BM \) uniquely determined by a weighted integral of the curl and divergence values \( \BC \) and \( s \) respectively

%\begin{boxed}\label{eqn:curlAndDivergenceVectorUniqueness:200}
\boxedEquation{eqn:curlAndDivergenceVectorUniqueness:200}{
\BM(\Bx)
=
\inv{4\pi} \int_V dV' \frac{ \BC(\Bx') \cross (\Bx - \Bx') + (\Bx - \Bx') s(\Bx')}{ \Abs{\Bx - \Bx'}^3 }.
}
%\end{boxed}

The trivector grade component of \cref{eqn:curlAndDivergenceVectorUniqueness:120} was neglected.  For that to be zero means that we must have

\begin{dmath}\label{eqn:curlAndDivergenceVectorUniqueness:160}
\int_V dV \frac{ \rcap \cdot \BC }{ r^2 }
=
\oint_{\partial V} dA \inv{ r^2 } I ( \rcap \wedge \ncap \wedge \BM )
=
\oint_{\partial V} dA \inv{ r^2 } (\ncap \cross \rcap) \cdot \BM.
\end{dmath}

Again requiring that \( \lim_{\Bx' \rightarrow \infty} \ifrac{\Abs{\BM(\Bx')}}{\Abs{\Bx - \Bx'}^2 } \rightarrow 0 \)
, the area integral is killed in the limit.  Somehow the LHS integral must be zero.  It isn't clear that \( \BC \cdot \rcap/r^2 \) should be zero, however, this expression is vaguely reminicent of an integrand related to \( \spacegrad \cdot \BC \), a quantity required to be zero, after a divergence theorem transformation.
To see how this is related to the remaining integral in the trivector grade selection observe that

\begin{equation}\label{eqn:curlAndDivergenceVectorUniqueness:260}
\spacegrad \inv{\Abs{\Br - \Br'}} 
=
-\spacegrad' \inv{\Abs{\Br - \Br'}} 
=
-\frac{\Bx - \Bx'}{\Abs{\Bx - \Bx'}^3}.
\end{equation}

This result, the zero divergence requirement for \( \BC \), and the chain rule can be used to rewrite the LHS of \cref{eqn:curlAndDivergenceVectorUniqueness:160}

\begin{dmath}\label{eqn:curlAndDivergenceVectorUniqueness:280}
\int_V dV \frac{ \rcap \cdot \BC(\Bx') }{ r^2 }
=
\int_V dV \spacegrad' \inv{\Abs{\Bx - \Bx'} } \cdot \BC(\Bx')
=
\int_V dV \spacegrad' \cdot \frac{\BC(\Bx')}{\Abs{\Bx - \Bx'} }
-\int_V dV \frac{\spacegrad' \cdot \BC(\Bx')}{ \Abs{\Bx - \Bx'} }
=
\int_{\partial V} d\BA \cdot \frac{\BC(\Bx')}{\Abs{\Bx - \Bx'} }
-\int_V dV \frac{\spacegrad' \cdot \BC(\Bx')}{ \Abs{\Bx - \Bx'} }.
\end{dmath}

Requiring that \( \lim_{\Bx' \rightarrow \infty} \ifrac{\BC(\Bx')}{\Abs{\Bx - \Bx'} } \rightarrow 0 \) and imposing the zero divergence requirement on \( \BC \), we see that all the trivector grade contributions in \cref{eqn:curlAndDivergenceVectorUniqueness:120} are zero.
}

%}
\EndArticle
