\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Notes and problems for Desai Chapter V.}
\label{chap:desaiCh5}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2010/desaiCh5.pdf}
\date{Oct 18, 2010}
\revisionInfo{desaiCh5.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

Chapter V notes for \cite{desai2009quantum}.

\section{Notes}
\section{Problems}

\subsection{Problem 1.}

\subsubsection{Statement.}
Obtain $S_x, S_y, S_z$ for spin 1 in the representation in which $S_z$ and $S^2$ are diagonal.

\subsubsection{Solution.}

For spin 1, we have

\begin{align}\label{eqn:desaiCh5:100}
S^2 = 1 (1+1) \hbar^2 \BOne
\end{align}

and are interested in the states $\ket{1,-1}, \ket{1, 0}, and \ket{1,1}$.  If, like angular momentum, we assume that we have for $m_s = -1,0,1$

\begin{align}\label{eqn:desaiCh5:101}
S_z \ket{1,m_s} = m_s \hbar \ket{1, m_s}
\end{align}

and introduce a column matrix representations for the kets as follows

\begin{align}\label{eqn:desaiCh5:102}
\ket{1,1} &=
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix} \\
\ket{1,0} &=
\begin{bmatrix}
0 \\
1 \\
0
\end{bmatrix} \\
\ket{1,-1} &=
\begin{bmatrix}
0 \\
0 \\
-1
\end{bmatrix},
\end{align}

then we have, by inspection
\begin{align}\label{eqn:desaiCh5:103}
S_z &= \hbar
\begin{bmatrix}
1 & 0 & 0 \\
0 & 0 & 0 \\
0 & 0 & -1
\end{bmatrix}.
\end{align}

Note that, like the Pauli matrices, and unlike angular momentum, the spin states $\ket{-1, m_s}, \ket{0, m_s}$ have not been considered.  Do those have any physical interpretation?

That question aside, we can procede as in the text, utilizing the ladder operator commutators

\begin{align}\label{eqn:desaiCh5:104}
S_{\pm} &= S_x \pm i S_y,
\end{align}

to determine the values of $S_x$ and $S_y$ indirectly.  We find

\begin{align}\label{eqn:desaiCh5:105}
\antisymmetric{S_{+}}{S_{-}} &= 2 \hbar S_z \\
\antisymmetric{S_{+}}{S_{z}} &= -\hbar S_{+} \\
\antisymmetric{S_{-}}{S_{z}} &= \hbar S_{-}.
\end{align}

Let
\begin{align}\label{eqn:desaiCh5:106}
S_{+} &=
\begin{bmatrix}
a & b & c \\
d & e & f \\
g & h & i
\end{bmatrix}.
\end{align}

Looking for equality between $\antisymmetric{S_{z}}{S_{+}}/\hbar = S_{+}$, we find
\begin{align}\label{eqn:desaiCh5:107}
\begin{bmatrix}
0 & b & 2 c \\
-d & 0 & f \\
-2g & -h & 0
\end{bmatrix}
&=
\begin{bmatrix}
a & b & c \\
d & e & f \\
g & h & i
\end{bmatrix},
\end{align}

so we must have
\begin{align}\label{eqn:desaiCh5:108}
S_{+} &=
\begin{bmatrix}
0 & b & 0 \\
0 & 0 & f \\
0 & 0 & 0
\end{bmatrix}.
\end{align}

Furthermore, from $\antisymmetric{S_{+}}{S_{-}} = 2 \hbar S_z$, we find
\begin{align}\label{eqn:desaiCh5:109}
\begin{bmatrix}
\Abs{b}^2 & 0 & 0 \\
0 & \Abs{f}^2 - \Abs{b}^2 & 0 \\
0 & 0 & -\Abs{f}^2
\end{bmatrix}
&=
2 \hbar^2
\begin{bmatrix}
1 & 0 & 0 \\
0 & 0 & 0 \\
0 & 0 & -1
\end{bmatrix}.
\end{align}

We must have $\Abs{b}^2 = \Abs{f}^2 = 2 \hbar^2$.  We could probably pick any
$b = \sqrt{2} \hbar e^{i\phi}$, and $f = \sqrt{2} \hbar e^{i\theta}$, but assuming we have no reason for a non-zero phase we try

\begin{align}\label{eqn:desaiCh5:110}
S_{+}
&=
\sqrt{2} \hbar
\begin{bmatrix}
0 & 1 & 0 \\
0 & 0 & 1 \\
0 & 0 & 0
\end{bmatrix}.
\end{align}

Putting all the pieces back together, with $S_x = (S_{+} + S_{-})/2$, and $S_y = (S_{+} - S_{-})/2i$, we finally have
\begin{align}\label{eqn:desaiCh5:111}
S_x &=
\frac{\hbar}{\sqrt{2}}
\begin{bmatrix}
0 & 1 & 0 \\
1 & 0 & 1 \\
0 & 1 & 0
\end{bmatrix} \\
S_y &=
\frac{\hbar}{\sqrt{2} i}
\begin{bmatrix}
0 & 1 & 0 \\
-1 & 0 & 1 \\
0 & -1 & 0
\end{bmatrix} \\
S_z &=
\hbar
\begin{bmatrix}
1 & 0 & 0 \\
0 & 0 & 0 \\
0 & 0 & -1
\end{bmatrix}.
\end{align}

A quick calculation verifies that we have $S_x^2 + S_y^2 + S_z^2 = 2 \hbar \BOne$, as expected.

\subsection{Problem 2.}
\subsubsection{Statement.}

Obtain eigensolution for operator $A = a \sigma_y + b \sigma_z$.  Call the eigenstates $\ket{1}$ and $\ket{2}$, and determine the probabilities that they will correspond to $\sigma_x = +1$.

\subsubsection{Solution.}

The first part is straight forward, and we have
\begin{align*}
A &= a \PauliY + b \PauliZ \\
&=
\begin{bmatrix}
b & -i a \\
ia & -b
\end{bmatrix}.
\end{align*}

Taking $\Abs{A - \lambda I} = 0$ we get

\begin{align}\label{eqn:desaiCh5:202}
\lambda &= \pm \sqrt{a^2 + b^2},
\end{align}

with eigenvectors proportional to
\begin{align}\label{eqn:desaiCh5:203}
\ket{\pm} &=
\begin{bmatrix}
i a \\
b \mp \sqrt{a^2 + b^2}
\end{bmatrix}
\end{align}

The normalization constant is $1/\sqrt{2 (a^2 + b^2) \mp 2 b \sqrt{a^2 + b^2}}$.  Now we can call these $\ket{1}$, and $\ket{2}$ but what does the last part of the question mean?  What's meant by $\sigma_x = +1$?

Asking the prof about this, he says:

``I think it means that the result of a measurement of the x component of spin is $+1$. This corresponds to the eigenvalue of $\sigma_x$ being $+1$. The spin operator $S_x$ has eigenvalue $+\hbar/2$''.

Aside: Question to consider later.  Is is significant that $\bra{1} \sigma_x \ket{1} = \bra{2} \sigma_x \ket{2} = 0$?

So, how do we translate this into a mathematical statement?

First let's recall a couple of details.  Recall that the x spin operator has the matrix representation

\begin{align}\label{eqn:desaiCh5:204}
\sigma_x = \PauliX.
\end{align}

This has eigenvalues $\pm 1$, with eigenstates $(1,\pm 1)/\sqrt{2}$.  At the point when the x component spin is observed to be $+1$, the state of the system was then

\begin{align}\label{eqn:desaiCh5:205}
\ket{x+} =
\inv{\sqrt{2}}
\begin{bmatrix}
1 \\
1
\end{bmatrix}
\end{align}

Let's look at the ways that this state can be formed as linear combinations of our states $\ket{1}$, and $\ket{2}$.  That is

\begin{align}\label{eqn:desaiCh5:206}
\inv{\sqrt{2}}
\begin{bmatrix}
1 \\
1
\end{bmatrix}
&=
\alpha \ket{1}
+ \beta \ket{2},
\end{align}

or

\begin{align}\label{eqn:desaiCh5:207}
\begin{bmatrix}
1 \\
1
\end{bmatrix}
&=
\frac{\alpha}{\sqrt{(a^2 + b^2) - b \sqrt{a^2 + b^2}}}
\begin{bmatrix}
i a \\
b - \sqrt{a^2 + b^2}
\end{bmatrix}
+\frac{\beta}{\sqrt{(a^2 + b^2) + b \sqrt{a^2 + b^2}}}
\begin{bmatrix}
i a \\
b + \sqrt{a^2 + b^2}
\end{bmatrix}
\end{align}

Letting $c = \sqrt{a^2 + b^2}$, this is

\begin{align}\label{eqn:desaiCh5:208}
\begin{bmatrix}
1 \\
1
\end{bmatrix}
&=
\frac{\alpha}{\sqrt{c^2 - b c}}
\begin{bmatrix}
i a \\
b - c
\end{bmatrix}
+\frac{\beta}{\sqrt{c^2 + b c}}
\begin{bmatrix}
i a \\
b + c
\end{bmatrix}.
\end{align}

We can solve the $\alpha$ and $\beta$ with Cramer's rule, yielding
\begin{align*}
\begin{vmatrix}
1 & i a \\
1 & b - c
\end{vmatrix}
&=
\frac{\beta}{\sqrt{c^2 + b c}}
\begin{vmatrix}
i a  & i a \\
b + c & b - c
\end{vmatrix} \\
\begin{vmatrix}
1 & i a \\
1 & b + c
\end{vmatrix}
&=
\frac{\alpha}{\sqrt{c^2 - b c}}
\begin{vmatrix}
i a  & i a \\
b - c & b + c
\end{vmatrix},
\end{align*}

or
\begin{align}\label{eqn:desaiCh5:209}
\alpha &= \frac{(b + c - ia)\sqrt{c^2 - b c}}{2 i a c} \\ %= \frac{(a -i(b + c))\sqrt{1 - b/c}}{2 a} \\
\beta &= \frac{(b - c - ia)\sqrt{c^2 + b c}}{-2 i a c} %= \frac{(a + i(b - c))\sqrt{1 + b/c}}{2 a}.
\end{align}

It is $\Abs{\alpha}^2$ and $\Abs{\beta}^2$ that are probabilities, and after a bit of algebra we find that those are

\begin{align}\label{eqn:desaiCh5:210}
\Abs{\alpha}^2 = \Abs{\beta}^2 = \inv{2},
\end{align}

so if the x spin of the system is measured as $+1$, we have a $50\%$ chance that the measured eigenvalue for the operator $A$ would be $\sqrt{a^2 + b^2}$ (ie: with state $\ket{1}$.

Is that what the question was asking?  I think that I've actually got it backwards.  I think that the question was asking for the probability of finding state $\ket{x+}$ (measuring a spin 1 value for $\sigma_x$) given the state $\ket{1}$ or $\ket{2}$.

So, suppose that we have

\begin{align}\label{eqn:desaiCh5:211}
\mu_{+} \ket{x+} + \nu_{+} \ket{x-} &= \ket{1} \\
\mu_{-} \ket{x+} + \nu_{-} \ket{x-} &= \ket{2},
\end{align}

or (considering both cases simulaneously), 
\begin{align*}
\mu_{\pm}
\begin{bmatrix}
1 \\
1
\end{bmatrix}
+ \nu_{\pm}
\begin{bmatrix}
1 \\
-1
\end{bmatrix}
&= 
\inv{\sqrt{ c^2 \mp b c }} 
\begin{bmatrix}
i a \\
b \mp c
\end{bmatrix} \\
\implies \\
\mu_{\pm}
\begin{vmatrix}
1 & 1 \\
1 & -1
\end{vmatrix}
&= 
\inv{\sqrt{ c^2 \mp b c }} 
\begin{vmatrix}
i a & 1 \\
b \mp c & -1
\end{vmatrix},
\end{align*}

or
\begin{align}\label{eqn:desaiCh5:212}
\mu_{\pm} &= 
\frac{ia + b \mp c}{2 \sqrt{c^2 \mp bc}} .
\end{align}

Unsuprisingly, this mirrors the previous scenerio and we find that we have a probability $\Abs{\mu}^2 = 1/2$ of measuring a spin 1 value for $\sigma_x$ when the state of the operator $A$ has been measured as $\pm \sqrt{a^2 + b^2}$ (ie: in the states $\ket{1}$, or $\ket{2}$ respectively).

No measurement of the operator $A = a \sigma_y + b\sigma_z$ gives a biased prediction of the state of the state $\sigma_x$.  Loosely, this seems to justify calling these operators orthogonal.  This is consistent with the geometrical antisymetric nature of the spin components where we have $\sigma_y \sigma_x = -\sigma_x \sigma_y$, just like two orthogonal vectors under the Clifford product.

\subsection{Problem 3.}
\subsubsection{Statement.}

Obtain the expectation values of $S_x, S_y, S_z$ for the case of a spin $1/2$ particle with the spin pointed in the direction of a vector with azimuthal angle $\beta$ and polar angle $\alpha$.

\subsubsection{Solution.}

TODO.

\subsection{Problem 4.}
\subsubsection{Statement.}
\subsubsection{Solution.}

TODO.

\subsection{Problem 5.}
\subsubsection{Statement.}
\subsubsection{Solution.}

TODO.

\subsection{Problem 6.}
\subsubsection{Statement.}

If a Hamiltonian is given by $\Bsigma \cdot \Bn$ where $\Bn = (\sin\alpha\cos\beta, \sin\alpha\sin\beta, \cos\alpha)$, determine the time evolution operator as a 2 x 2 matrix.  If a state at $t = 0$ is given by 

\begin{align}\label{eqn:desaiCh5:600}
\ket{\phi(0)} = 
\begin{bmatrix}
a \\
b
\end{bmatrix},
\end{align}

then obtain $\ket{\phi(t)}$.

\subsubsection{Solution.}

Before diving into the meat of the problem, observe that a tidy factorization of the Hamiltonian is possible as a composition of rotations.  That is

\begin{align*}
H 
&= \Bsigma \cdot \Bn \\
&= \sin\alpha \sigma_1 ( \cos\beta + \sigma_1 \sigma_2 \sin\beta ) + \cos\alpha \sigma_3 \\
&= \sigma_3 \left(
\cos\alpha 
+ \sin\alpha \sigma_3 \sigma_1 e^{ i \sigma_3 \beta }
\right) \\
&= 
\sigma_3 \exp\left( \alpha i \sigma_2 
\exp\left( \beta i \sigma_3 
\right)
\right)
\end{align*}

So we have for the time evolution operator

\begin{align}\label{eqn:desaiCh5:610}
U(\Delta t) 
&=
\exp( -i \Delta t H /\hbar )
= 
\exp \left(
- \frac{\Delta t}{\hbar} i \sigma_3 \exp\Bigl( \alpha i \sigma_2 
\exp\left( \beta i \sigma_3 
\right)
\Bigr)
\right).
\end{align}

Does this really help?  I guess not, but it is nice and tidy.

Returning to the specifics of the problem, we note that squaring the Hamiltonian produces the identity matrix

\begin{align}\label{eqn:desaiCh5:615}
(\Bsigma \cdot \Bn)^2 &= I \Bn^2 = I.
\end{align}

This allows us to exponentiate $H$ by inspection utilizing

\begin{align}\label{eqn:desaiCh5:620}
e^{i \mu (\Bsigma \cdot \Bn) } = I \cos\mu + i (\Bsigma \cdot \Bn) \sin\mu
\end{align}

Writing $\sin\mu = S_\mu$, and $\cos\mu = C_\mu$, we have
\begin{align}\label{eqn:desaiCh5:625}
\Bsigma \cdot \Bn &=
\begin{bmatrix}
C_\alpha & S_\alpha e^{-i\beta} \\
S_\alpha e^{i\beta} & -C_\alpha
\end{bmatrix},
\end{align}

and thus
\begin{align}\label{eqn:desaiCh5:630}
U(\Delta t) = \exp( -i \Delta t H /\hbar )
=
\begin{bmatrix}
C_{\Delta t/\hbar} -i S_{\Delta t/\hbar} C_\alpha & -i S_{\Delta t/\hbar} S_\alpha e^{-i\beta} \\
-i S_{\Delta t/\hbar} S_\alpha e^{i\beta} & C_{\Delta t/\hbar} + i S_{\Delta t/\hbar} C_\alpha
\end{bmatrix}.
\end{align}

Note that as a sanity check we can calculate that $ U(\Delta t) U(\Delta t)^\dagger = 1$ as expected.

Now for $\Delta t = t$, we have 
\begin{align}\label{eqn:desaiCh5:640}
U(t,0) 
\begin{bmatrix}
a \\
b
\end{bmatrix}
&=
\begin{bmatrix}
a C_{t/\hbar} -a i S_{t/\hbar} C_\alpha  - b i S_{t/\hbar} S_\alpha e^{-i\beta} \\
-a i S_{t/\hbar} S_\alpha e^{i\beta} + b C_{t/\hbar} + b i S_{t/\hbar} C_\alpha
\end{bmatrix}.
\end{align}

It doesn't seem terribly illuminating to multiply this all out, but we can factor the results slightly to tidy it up.  That gives us

\begin{align}\label{eqn:desaiCh5:650}
U(t,0) 
\begin{bmatrix}
a \\
b
\end{bmatrix}
&=
\cos(t/\hbar)
\begin{bmatrix}
a \\
b
\end{bmatrix}
+ 
\sin(t/\hbar) \cos\alpha
\begin{bmatrix}
-a \\
b
\end{bmatrix}
+ i
\sin(t/\hbar) \sin\alpha
\begin{bmatrix}
b e^{-i\beta} \\
-a e^{i \beta}
\end{bmatrix}
\end{align}

\subsection{Problem 7.}
\subsubsection{Statement.}
\subsubsection{Solution.}

TODO.

\subsection{Problem 8.}
\subsubsection{Statement.}
\subsubsection{Solution.}

TODO.

\subsection{Problem 9.}
\subsubsection{Statement.}
\subsubsection{Solution.}

TODO.

\EndArticle
