\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Desai Chapter 9 notes and problems.}
\label{chap:desaiCh9}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2010/desaiCh9.pdf}
\date{Nov 19, 2010}
\revisionInfo{desaiCh9.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

Chapter 9 notes for \cite{desai2009quantum}.

\section{Notes}
\section{Problems}

\subsection{Problem 4.}
\subsubsection{Statement.}

Consider the following two-dimensional harmonic oscilator problem:

\begin{align}\label{eqn:desaiCh9:400}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial x^2}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial y^2}
+ \inv{2} K_1 x^2 u
+ \inv{2} K_2 y^2 u
= E u
\end{align}

where $(x,y)$ are the coordinates of the particle.  Use the separation of variables technique to obtain the energy eigenvalues.  Discuss the degeneracy in the eigenvalues if $K_1 = K_2$.

\subsubsection{Solution.}

Write $u = A(x) B(y)$.  Substitute and dividing throughout by $u$ we have

\begin{align}\label{eqn:desaiCh9:401}
\left( -\frac{\hbar^2}{2m} \frac{A''}{A} + \inv{2} K_1 x^2 \right)
+\left( -\frac{\hbar^2}{2m} \frac{B''}{B} + \inv{2} K_2 y^2 \right)
= E
\end{align}

Introduction of a pair of constants $E_1, E_2$ for each of the independent terms we have

\begin{align}\label{eqn:desaiCh9:403}
H_1 A &= -\frac{\hbar^2}{2m} A'' + \inv{2} K_1 x^2 A = E_1 A \\
H_2 B &= -\frac{\hbar^2}{2m} B'' + \inv{2} K_1 y^2 B = E_2 B \\
H &= H_1 + H_2 \\
E  &= E_1 + E_2
\end{align}

For each of these equations we have a set of quantized eigenvalues and can write

\begin{align}\label{eqn:desaiCh9:404}
E_{1m} &= \left(m + \inv{2}\right) \hbar \sqrt{\frac{K_1}{m}} \\
E_{2n} &= \left(n + \inv{2}\right) \hbar \sqrt{\frac{K_2}{m}} \\
H_1 A_m(x) &= E_{1m} A_m(x) \\
H_2 A_n(y) &= E_{2n} B_n(y)
\end{align}

The complete eigenstates are then

\begin{align}\label{eqn:desaiCh9:405}
u_{mn}(x,y) &= A_m(x) B_n(y)
\end{align}

with total energy satisfying
\begin{align}\label{eqn:desaiCh9:406}
H u_{mn}(x,y) &=
\frac{\hbar}{\sqrt{m}} \left( \left(m + \inv{2}\right) \sqrt{K_1} + \left(n + \inv{2}\right) \sqrt{K_2} \right) u_{mn}(x,y)
\end{align}

A general state requires a double sum over the possible combinations of states $\Psi = \sum_{mn} c_{mn} u_{mn}$, however if $K_1 = K_2 = K$, we cannot distinguish between $u_{mn}$ and $u_{nm}$ based on the energy eigenvalues

\begin{align}\label{eqn:desaiCh9:407}
H u_{mn}(x,y) &= \hbar\sqrt{\frac{K}{m}} \left( m + n + 1 \right) u_{mn}(x,y) = H u_{nm}(x,y)
\end{align}

In this case, we can write the wave function corresponding to a general state for the system as just $\Psi = \sum_{m+ n = \text{constant}} c_{mn} u_{mn}$.  This reduction in the cardinality of this set of basis eigenstates is the degeneracy to be discussed.

\subsection{Problem 5,6.}
\subsubsection{Statement.}

Consider now a variation on Problem 4 in which we have a coupled oscillator with the potential given by

\begin{align}\label{eqn:desaiCh9:500}
V(x,y) = \inv{2} K \Bigl( x^2 + y^2 + 2 \lambda x y \Bigr)
\end{align}

Obtain the energy eigenvalues by changing variables $(x,y)$ to $(x', y')$ such that the new potential is quadratic in $(x', y')$, without the coupling term.

\subsubsection{Solution.}

This has the look of a diagonalization problem so we write the potential in matrix form

\begin{align}\label{eqn:desaiCh9:501}
V(x,y)
= \inv{2} K
\begin{bmatrix}
x & y
\end{bmatrix}
\begin{bmatrix}
1 & \lambda \\
\lambda & 1
\end{bmatrix}
\begin{bmatrix}
x \\ y
\end{bmatrix} = \inv{2} K \tilde{X} M X
\end{align}

The similarity transformation required is
\begin{align}\label{eqn:desaiCh9:502}
M = \inv{\sqrt{2}}
\begin{bmatrix}
1 & 1 \\
1 & -1
\end{bmatrix}
\begin{bmatrix}
1+ \lambda & 0 \\
0 & 1 - \lambda
\end{bmatrix}
\inv{\sqrt{2}}
\begin{bmatrix}
1 & 1 \\
1 & -1
\end{bmatrix}
\end{align}

Our change of variables is therefore
\begin{align}\label{eqn:desaiCh9:503}
X' =
\inv{\sqrt{2}}
\begin{bmatrix}
1 & 1 \\
1 & -1
\end{bmatrix}
X
=
\inv{\sqrt{2}}
\begin{bmatrix}
x + y \\
x - y
\end{bmatrix}
\end{align}

Our Laplacian should also remain diagonal under this orthonormal transformation, but we can verify this by expanding out the partials explicitly
\begin{align}\label{eqn:desaiCh9:504}
\PD{x}{} &=
\PD{x}{x'}\PD{x'}{}
+\PD{x}{y'}\PD{y'}{} = \inv{\sqrt{2}} \left( \PD{x'}{} + \PD{y'}{} \right) \\
\PD{y}{} &=
\PD{y}{x'}\PD{x'}{} +\PD{y}{y'}\PD{y'}{}
= \inv{\sqrt{2}}
\left( \PD{x'}{} - \PD{y'}{} \right)
\end{align}

Squaring and summing we have
\begin{align}\label{eqn:desaiCh9:505}
\frac{\partial^2}{\partial x^2} +
\frac{\partial^2}{\partial y^2}
&=
\inv{2} \left( \PD{x'}{} + \PD{y'}{} \right)^2
+\inv{2} \left( \PD{x'}{} - \PD{y'}{} \right)^2
=
\frac{\partial^2}{\partial {x'}^2} +
\frac{\partial^2}{\partial {y'}^2}
\end{align}

Our transformed Hamiltonian operator is thus

\begin{align}\label{eqn:desaiCh9:506}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {x'}^2}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {y'}^2}
+ \inv{2} K(1+\lambda) {x'}^2 u
+ \inv{2} K(1-\lambda) {y'}^2 u
= E u
\end{align}

So, provided $\Abs{\lambda} < 1$, the energy eigenvalue equation is given by \ref{eqn:desaiCh9:406} with $K_1 = K(1+ \lambda)$, and $K_2 = K(1 -\lambda)$.

\subsection{Problem 7.}
\subsubsection{Statement.}

Consider two coupled harmonic oscillators in one dimension of natural length $a$ and spring constant $K$ connecting three particles located at $x_1, x_2$, and $x_3$.  The corresponding Schr\"{o}dinger equation is given as

\begin{align}\label{eqn:desaiCh9:700}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {x_1}^2}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {x_2}^2}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {x_3}^2}
+ \frac{K}{2}
\left(
(x_2 - x_1 - a)^2
+(x_3 - x_2 - a)^2
\right) u
= E u
\end{align}

Obtain the energy eigenvalues using the matrix method.

\subsubsection{Solution.}

Let's start with an initial simplifying substutition to get rid of the factors of $a$.  Write

\begin{align}\label{eqn:desaiCh9:701}
r_1 &= x_1 + a \\
r_2 &= x_2 \\
r_3 &= x_3 - a
\end{align}

These were picked so that the differences in our quadratic terms involve only factors of $r_k$

\begin{align}\label{eqn:desaiCh9:702}
x_2 - x_1 - a &= r_2 - r_1 \\
x_3 - x_2 - a &= r_3 - r_2
\end{align}

Schr\"{o}dinger's equation is now

\begin{align}\label{eqn:desaiCh9:700a}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {r_1}^2}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {r_2}^2}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {r_3}^2}
+ \frac{K}{2}
\left(
(r_2 - r_1)^2
+(r_3 - r_2)^2
\right) u
= E u
\end{align}

Putting our potential into matrix form, we have

\begin{align}\label{eqn:desaiCh9:703}
V(r_1, r_2, r_3) &=
\frac{K}{2}
\left(
(r_2 - r_1)^2
+(r_3 - r_2)^2
\right)
=
\frac{K}{2}
\begin{bmatrix}
r_1 & r_2 & r_3
\end{bmatrix}
\begin{bmatrix}
1 & -1 & 0 \\
-1 & 2 & -1 \\
0 & -1 & 1
\end{bmatrix}
\begin{bmatrix}
r_1 \\ r_2 \\ r_3
\end{bmatrix}
\end{align}

This symmetric matrix, let's call it M
\begin{align}\label{eqn:desaiCh9:704}
M=
\begin{bmatrix}
1 & -1 & 0 \\
-1 & 2 & -1 \\
0 & -1 & 1
\end{bmatrix}
\end{align}

has eigenvalues $0,1,3$, with orthonormal eigenvectors
\begin{align}\label{eqn:desaiCh9:705}
e_0 &=
\inv{\sqrt{3}}
\begin{bmatrix}
1 \\
1 \\
1
\end{bmatrix} \\
e_1 &=
\inv{\sqrt{2}}
\begin{bmatrix}
1 \\
0 \\
-1
\end{bmatrix} \\
e_3 &=
\inv{\sqrt{6}}
\begin{bmatrix}
1 \\
-2 \\
1
\end{bmatrix}
\end{align}

Writing

\begin{align}\label{eqn:desaiCh9:706}
U = [e_0 e_1 e_3]
=
\begin{bmatrix}
\inv{\sqrt{3}} & \inv{\sqrt{2}}  & \inv{\sqrt{6}}  \\
\inv{\sqrt{3}} & 0  & -\frac{2}{\sqrt{6}}  \\
\inv{\sqrt{3}} & -\inv{\sqrt{2}}  & \inv{\sqrt{6}}
\end{bmatrix}
\end{align}

\begin{align}\label{eqn:desaiCh9:707}
M = U
\begin{bmatrix}
0 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 3
\end{bmatrix}
\tilde{U}
=
U D \tilde{U}
\end{align}

Writing $R' = \tilde{U} R$, and $\spacegrad' = \tilde{U} \spacegrad$, we see that the Laplacian has no mixed partial terms after transformation

\begin{align*}
\spacegrad' \cdot \spacegrad' 
&= 
(\tilde{U} \spacegrad)^{\tilde{}} \tilde{U} \spacegrad \\
&= 
\tilde{\spacegrad } \spacegrad \\
&= 
\spacegrad \cdot \spacegrad
\end{align*}

Schr\"{o}dinger's equation is then just
\begin{align}\label{eqn:desaiCh9:708}
\left( -\frac{\hbar^2}{2m} {\spacegrad'}^2 + \frac{K}{2} \tilde{R'} D R' \right) u = E u
\end{align}

Or
\begin{align}\label{eqn:desaiCh9:708b}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {r_1'}^2}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {r_2'}^2}
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {r_3'}^2}
+ \frac{K}{2}
\left(
{r_2'}^2
+3 {r_3'}^2
\right) u
= E u
\end{align}

Separation of variables provides us with one free particle wave equation, and two harmonic oscillator equations

\begin{align}\label{eqn:desaiCh9:708c}
-\frac{\hbar^2}{2m} \frac{\partial^2 u_1}{\partial {r_1'}^2} &= E_1 u_1 \\
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {r_2'}^2} + \frac{K}{2} {r_2'}^2 u_2 &= E_2 u_2 \\
-\frac{\hbar^2}{2m} \frac{\partial^2 u}{\partial {r_3'}^2} + \frac{3 K}{2} {r_3'}^2 u_3 &= E_3 u_3
\end{align}

We can borrow the Harmonic oscillator energy eigenvalues from problem 4 again with $K_1 = K$, and $K_2 = 3 K$.

\subsection{Problem 8.}
\subsubsection{Statement.}

As a variation of Problem 7 assume that the middle particle at $x_2$ has a different mass $M$.  Reduce this problem to the form of Problem 7 by a scale change in $x_2$ and then use the matrix method to obtain the energy eigenvalues.

\subsubsection{Solution.}

We write $\sqrt{M} x_2 = \sqrt{m} x_2', x_1 + a = x_1', x_3 - a = x_3'$, and then  Schr\"{o}dinger's equation takes the form

\begin{align}\label{eqn:desaiCh9:800}
\left( -\frac{\hbar^2}{2m} {\spacegrad'}^2 + V(X') \right) u &= E u 
\end{align}
\begin{align}\label{eqn:desaiCh9:801}
V(X') = \frac{K}{2} 
\left( 
\left( \sqrt{\frac{m}{M}} x_2' - x_1' 
\right)^2
+\left( -\sqrt{\frac{m}{M}} x_2' + x_3' 
\right)^2
\right)
\end{align}

With $\mu = \sqrt{m/M}$, we have
\begin{align}\label{eqn:desaiCh9:802}
V(X') = \frac{K}{2} 
\tilde{X'}
\begin{bmatrix}
1 & -\mu & 0 \\
-\mu & 2 \mu^2 & -\mu \\
0 & -\mu & 1
\end{bmatrix}
X'
\end{align}

We find that this symmetric matrix has eigenvalues $0, 1, 1 + 2 \mu^2$, and eigenvectors

\begin{align}\label{eqn:desaiCh9:803}
e_0 &=
\inv{\sqrt{1 + 2 \mu^2}}
\begin{bmatrix}
\mu \\ 1 \\ \mu
\end{bmatrix} \\
e_1 &=
\inv{\sqrt{2}}
\begin{bmatrix}
1 \\ 0 \\ -1
\end{bmatrix} \\
e_{1+ 2 \mu^2} &=
\inv{\sqrt{2 + 4 \mu^2}}
\begin{bmatrix}
1 \\ -2 \mu \\ 1
\end{bmatrix} 
\end{align}

The rest of the problem is now no different than the tail end of Problem 7, and we end up with $K_1 = K$, $K_2 = (1 + 2 \mu^2) K$.

\EndArticle
