\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Notes and problems for Desai Chapter 26.}
\label{chap:desaiCh26}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2010/desaiCh26.pdf}
\date{Dec 9, 2010}
\revisionInfo{desaiCh26.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

Chapter 26 notes for \cite{desai2009quantum}.

\section{Guts}

\subsection{Trig relations.}

To verify equations 26.3-5 in the text it's worth noting that 

\begin{align*}
\cos(a + b) 
&= \Re( e^{ia} e^{ib} ) \\
&= \Re( (\cos a + i \sin a)( \cos b + i \sin b) ) \\
&= \cos a \cos b - \sin a \sin b
\end{align*}

and
\begin{align*}
\sin(a + b) 
&= \Im( e^{ia} e^{ib} ) \\
&= \Im( (\cos a + i \sin a)( \cos b + i \sin b) ) \\
&= \cos a \sin b + \sin a \cos b
\end{align*}

So, for 
\begin{align}\label{eqn:desaiCh26:10}
x &= \rho \cos\alpha \\
y &= \rho \sin\alpha 
\end{align}

the transformed coordinates are
\begin{align*}
x' 
&= \rho \cos(\alpha + \phi) \\
&= \rho (\cos \alpha \cos \phi - \sin \alpha \sin \phi) \\
&= x \cos \phi - y \sin \phi
\end{align*}

and
\begin{align*}
y' 
&= \rho \sin(\alpha + \phi) \\
&= \rho (\cos \alpha \sin \phi + \sin \alpha \cos \phi) \\
&= x \sin \phi + y \cos \phi \\
\end{align*}

This allows us to read off the rotation matrix.  Without all the messy trig, we can also derive this matrix with geometric algebra.

\begin{align*}
\Bv' 
&= e^{- \Be_1 \Be_2 \phi/2 } \Bv e^{ \Be_1 \Be_2 \phi/2 } \\
&= v_3 \Be_3 + (v_1 \Be_1 + v_2 \Be_2) e^{ \Be_1 \Be_2 \phi } \\
&= v_3 \Be_3 + (v_1 \Be_1 + v_2 \Be_2) (\cos \phi + \Be_1 \Be_2 \sin\phi) \\
&= v_3 \Be_3 
+ \Be_1 (v_1 \cos\phi - v_2 \sin\phi)
+ \Be_2 (v_2 \cos\phi + v_1 \sin\phi)
\end{align*}

Here we use the Pauli-matrix like identities

\begin{align}\label{eqn:desaiCh26:20}
\Be_k^2 &= 1 \\
\Be_i \Be_j &= -\Be_j \Be_i,\quad i\ne j
\end{align}

and also note that $\Be_3$ commutes with the bivector for the $x,y$ plane $\Be_1 \Be_2$.  We can also read off the rotation matrix from this.

\subsection{Infinitesimal transformations.}

Recall that in the problems of Chapter 5, one representation of spin one matrices were calculated \chapcite{desaiCh5}.  Since the choice of the basis vectors was arbitrary in that exersize, we ended up with a different representation.  For $S_x, S_y, S_z$ as found in (26.20) and (26.23) we can also verify easily that we have eigenvalues $0, \pm \hbar$.  We can also show that our spin kets in this non-diagonal representation have the following column matrix representations:

\begin{align}\label{eqn:desaiCh26:30}
\ket{1,\pm 1}_x 
&=
\inv{\sqrt{2}} \begin{bmatrix}
0 \\
1 \\
\pm i
\end{bmatrix} \\
\ket{1,0}_x 
&=
\begin{bmatrix}
1 \\
0 \\
0 
\end{bmatrix} \\
\ket{1,\pm 1}_y 
&=
\inv{\sqrt{2}} \begin{bmatrix}
\pm i \\
0 \\
1 
\end{bmatrix} \\
\ket{1,0}_y 
&=
\begin{bmatrix}
0 \\
1 \\
0 
\end{bmatrix} \\
\ket{1,\pm 1}_z 
&=
\inv{\sqrt{2}} \begin{bmatrix}
1 \\
\pm i \\
0
\end{bmatrix} \\
\ket{1,0}_z 
&=
\begin{bmatrix}
0 \\
0 \\
1
\end{bmatrix} 
\end{align}

\subsection{Verifying the commutator relations.}

Given the (summation convention) matrix representation for the spin one operators

\begin{equation}\label{eqn:desaiCh26:40}
(S_i)_{jk} = - i \hbar \epsilon_{ijk},
\end{equation}

let's demonstrate the commutator relation of (26.25).

\begin{align*}
{\antisymmetric{S_i}{S_j}}_{rs} 
&=
(S_i S_j - S_j S_i)_{rs} \\
&=
\sum_t (S_i)_{rt} (S_j)_{ts} - (S_j)_{rt} (S_i)_{ts} \\
&=
(-i\hbar)^2 \sum_t \epsilon_{irt} \epsilon_{jts} - \epsilon_{jrt} \epsilon_{its} \\
&=
-(-i\hbar)^2 \sum_t \epsilon_{tir} \epsilon_{tjs} - \epsilon_{tjr} \epsilon_{tis} \\
\end{align*}

Now we can employ the summation rule for sums products of antisymmetic tensors over one free index (4.179) 

\begin{equation}\label{eqn:desaiCh26:50}
\sum_i 
\epsilon_{ijk} \epsilon_{iab}
= 
\delta_{ja}
\delta_{kb}
-\delta_{jb}
\delta_{ka}.
\end{equation}

Continuing we get
\begin{align*}
{\antisymmetric{S_i}{S_j}}_{rs} 
&=
-(-i\hbar)^2 \left(
\delta_{ij}
\delta_{rs}
-\delta_{is}
\delta_{rj}
-\delta_{ji}
\delta_{rs}
+\delta_{js}
\delta_{ri} \right) \\
&=
(-i\hbar)^2 \left( 
\delta_{is}
\delta_{jr}
-
\delta_{ir} 
\delta_{js}
\right)
\\
&=
(-i\hbar)^2 \sum_t \epsilon_{tij} \epsilon_{tsr}
\\
&=
i\hbar \sum_t \epsilon_{tij} (S_t)_{rs}
\qquad\square
\end{align*}

\subsection{General infinitesimal rotation.}

Equation (26.26) has for an infinitesimal rotation counterclockwise around the unit axis of rotation vector $\Bn$

\begin{equation}\label{eqn:desaiCh26:60}
\BV' = \BV + \epsilon \Bn \cross \BV.
\end{equation}

Let's derive this using the geometric algebra rotation expression for the same

\begin{align*}
\BV' 
&=
e^{-I\Bn \alpha/2}
\BV 
e^{I\Bn \alpha/2} \\
&=
e^{-I\Bn \alpha/2}
\left(
(\BV \cdot \Bn)\Bn
+(\BV \wedge \Bn)\Bn
\right)
e^{I\Bn \alpha/2} \\
&=
(\BV \cdot \Bn)\Bn
+(\BV \wedge \Bn)\Bn
e^{I\Bn \alpha}
\end{align*}

We note that $I\Bn$ and thus the exponential commutes with $\Bn$, and the projection component in the normal direction.  Similarily $I\Bn$ anticommutes with $(\BV \wedge \Bn) \Bn$.  This leaves us with

\begin{align*}
\BV' 
&=
(\BV \cdot \Bn)\Bn
\left(
+(\BV \wedge \Bn)\Bn
\right)
( \cos \alpha + I \Bn \sin\alpha)
\end{align*}

For $\alpha = \epsilon \rightarrow 0$, this is

\begin{align*}
\BV' 
&=
(\BV \cdot \Bn)\Bn
+(\BV \wedge \Bn)\Bn
( 1 + I \Bn \epsilon) \\
&=
(\BV \cdot \Bn)\Bn 
+(\BV \wedge \Bn)\Bn
+\epsilon I^2(\BV \cross \Bn)\Bn^2 \\
&=
\BV
+ \epsilon (\Bn \cross \BV) \qquad\square
\end{align*}

\EndArticle
