\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Notes and problems for Desai chapter III.}
\label{chap:desaiCh3}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2010/desaiCh3.pdf}
\date{Oct 1, 2010}
\revisionInfo{desaiCh3.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Notes.}

Chapter III notes and problems for \cite{desai2009quantum}.

FIXME:
Some puzzling stuff in the interaction section and superposition of time-dependent states sections.  Work through those here.

\section{Problems}

\subsection{Problem 1. Virial Theorem.}
\subsubsection{Statement.}

With the assumption that $\expectation{\Br \cdot \Bp}$ is independent of time, and 

\begin{align}\label{eqn:desaiCh3:100}
H = \frac{\Bp^2}{2m} + V(\Br) = T + V
\end{align}

show that 

\begin{align}\label{eqn:desaiCh3:102}
2 \expectation{T} = \expectation{ \Br \cdot \spacegrad V}.
\end{align}

\subsubsection{Solution.}

I floundered with this a bit, but found the required hint in \href{http://www.physicsforums.com/showthread.php?t=164682}{physicsforums}.  We can start with the Hamiltonian time derivative relation

\begin{align}\label{eqn:desaiCh3:103}
i\hbar \frac{d A_H}{dt} = \antisymmetric{A_H}{H}
\end{align}

So, with the assumption that $\expectation{\Br \cdot \Bp}$ is independent of time, and the use of a stationary state $\ket{\psi}$ for the expectation calculation we have

\begin{align*}
0 &=
\frac{d}{dt} \expectation{\Br \cdot \Bp}  \\
&=
\frac{d}{dt} \bra{\psi} \Br \cdot \Bp \ket{\psi} \\
&=
\bra{\psi} 
\frac{d}{dt} ( \Br \cdot \Bp ) \ket{\psi} \\
&= 
\inv{i\hbar} \expectation{ \antisymmetric{ \Br \cdot \Bp }{H} } \\
&= 
-\expectation{ \antisymmetric{ \Br \cdot \spacegrad }{\frac{\Bp^2}{2m}} } 
-\expectation{ \antisymmetric{ \Br \cdot \spacegrad }{V(\Br)} }.
\end{align*}

The exercise now becomes one of evaluating the remaining commutators.  For the Laplacian commutator we have

\begin{align*}
\antisymmetric{ \Br \cdot \spacegrad }{\spacegrad^2} \psi
&=
x_m \partial_m \partial_n \partial_n \psi 
- \partial_n \partial_n x_m \partial_m \psi \\
&=
x_m \partial_m \partial_n \partial_n \psi 
- \partial_n \partial_n \psi 
- \partial_n x_m \partial_n \partial_m \psi \\
&=
x_m \partial_m \partial_n \partial_n \psi 
- \partial_n \partial_n \psi 
- \partial_n \partial_n \psi 
- x_m \partial_n \partial_n \partial_m \psi \\
&=
- 2 \spacegrad^2 \psi
\end{align*}

For the potential commutator we have

\begin{align*}
\antisymmetric{ \Br \cdot \spacegrad }{V(\Br)} \psi
&=
x_m \partial_m V \psi 
-V x_m \partial_m \psi  \\
&=
x_m (\partial_m V) \psi 
x_m V \partial_m \psi 
-V x_m \partial_m \psi  \\
&=
\Bigl( \Br \cdot (\spacegrad V) \Bigr) \psi
\end{align*}

Putting all the $\hbar$ factors back in, we get

\begin{align}\label{eqn:desaiCh3:104}
2 \expectation{ \frac{\Bp^2}{2m} } = \expectation{ \Br \cdot (\spacegrad V) },
\end{align}

which is the desired result.

Followup: why assume $\expectation{\Br \cdot \Bp}$ is independent of time?

%\href{http://www.caelestis.de/dateien/UEA05_2.pdf}{http://www.caelestis.de/dateien/UEA05_2.pdf} (ie: we need the Hamiltonian commutator).


\subsection{Problem 2. Application of virial theorem.}

Calculate $\expectation{T}$ with $V = \lambda \ln(r/a)$.

\begin{align*}
\Br \cdot \spacegrad V 
&= r \rcap \cdot \rcap \lambda \PD{r}{\ln(r/a)} \\
&= \lambda r \inv{a} \frac{a}{r} \\
&= \lambda  \\
\implies \\
\expectation{T} &= \lambda/2
\end{align*}

\subsection{Problem 3. Heisenberg Position operator representation.}

\subsubsection{Part I.}
Express $x$ as an operator $x_H$ for $H = \Bp^2/2m$.

With 

\begin{align*}
\bra{\psi} x \ket{\psi} = \bra{\psi_0} U^\dagger x U \ket{\psi_0}
\end{align*}

We want to expand 
\begin{align*}
x_H 
&= U^\dagger x U \\
&= e^{i H t/\hbar} x e^{-iH t/\hbar} \\
&= \sum_{k,l = 0}^\infty \inv{k!} \inv{l!} 
\left(\frac{i H t}{\hbar}\right)^k 
x 
\left(\frac{-i H t}{\hbar}\right)^l .
\end{align*}

We to evaluate $H^k x H^l$ to proceed.  Using $p^n x = -i \hbar n p^{n-1} + x p^n$, we have

\begin{align*}
H^k x 
&= \inv{(2m)^k} p^2k x \\
&= \inv{(2m)^k} \left( -i \hbar (2k) p^{2k -1} + x p^2k \right) \\
&= x H^k + \inv{2m} (-i \hbar) (2k) p p^{2(k-1)}/(2m)^{k-1} \\
&= x H^k - \frac{i \hbar k}{m} p H^{k-1}.
\end{align*}

This gives us
\begin{align*}
x_H 
&= x - \frac{i \hbar p }{m} \sum_{k,l=0}^\infty \frac{k}{k!} \inv{l!}
\left(\frac{i t}{\hbar}\right)^k H^{k-1 + l}
\left(\frac{-i t}{\hbar}\right)^l  \\
&= x - \frac{i \hbar p i t }{m \hbar} 
\end{align*}

Or
\begin{align}\label{eqn:desaiCh3:303}
x_H 
&= x + \frac{p t }{m} 
\end{align}


\subsubsection{Part II.}
Express $x$ as an operator $x_H$ for $H = \Bp^2/2m + V$ with $V = \lambda x^m$.

In retrospect, for the first part of this problem, it would have been better to use the series expansion for this exponential sandwich

Or, in explicit form
\begin{align}\label{eqn:desaiCh3:304}
e^A B e^{-A}
&=
B 
+ \inv{1!} \antisymmetric{A}{B}
+ \inv{2!} 
\antisymmetric{A}{\antisymmetric{A}{B}}
+ \cdots
\end{align}

Doing so, we'd find for the first commutator

\begin{align}\label{eqn:desaiCh3:305}
\frac{i t}{2m \hbar} \antisymmetric{\Bp^2}{x} = \frac{t p}{m},
\end{align}

so that the series has only the first two terms, and we'd obtain the same result.  That seems like a logical approach to try here too.  For the first commutator, we get the same $tp/m$ result since $\antisymmetric{V,x} = 0$.

Employing 
\begin{align}\label{eqn:desaiCh3:306}
x^n p = i \hbar n x^{n-1} + p x^n,
\end{align}

I find 

\begin{align*}
\left( \frac{i t}{\hbar} \right)^2 \antisymmetric{H}{\antisymmetric{H}{x}} 
&= \frac{i \lambda t^2}{\hbar m } \antisymmetric{x^n}{p}  \\
&= - \frac{n t^2 \lambda}{m} x^{n-1} \\
&= - \frac{n t^2 V}{m x} \\
\end{align*}

The triple commutator gets no prettier, and I get

\begin{align*}
\left( \frac{i t}{\hbar} \right)^3 \antisymmetric{H}{\antisymmetric{H}{\antisymmetric{H}{x}}}
&= 
\frac{it}{\hbar} \antisymmetric{ \frac{\Bp^2}{2m} + \lambda x^n}{ - \frac{n t^2 V}{m x} } \\
&= 
-\frac{it}{\hbar} \frac{n t^2 }{m } \frac{\lambda}{2m} \antisymmetric{\Bp^2}{ x^{n-1}} \\
&= \cdots \\
&= \frac{n(n-1)t^3 V}{ 2 m^2 x^3 } (i \hbar n + 2 p x).
\end{align*}

Putting all the pieces together this gives

\begin{align}\label{eqn:desaiCh3:307}
x_H =
e^{iH t/\hbar} x e^{-iH t/\hbar}  &= 
x + \frac{tp}{m} - \frac{n t^2 V}{ 2 m x} 
+ \frac{n(n-1)t^3 V}{ 12 m^2 x^3 } (i \hbar n + 2 p x) + \cdots
\end{align}

If there is a closed form for this it isn't obvious to me.  Would a fixed lower degree potential function shed any more light on this.  How about the Harmonic oscillator Hamiltonian

\begin{align}\label{eqn:desaiCh3:308}
H = \frac{p^2}{2m} + \frac{m \omega^2 }{2} x^2
\end{align}

... this one works out nicely since there's an even-odd alternation.

Get

\begin{align}\label{eqn:desaiCh3:309}
x_H = x \cos (\omega^2 t^2 /2) + 
\frac{ p t }{m} \frac{\sin( \omega^2 t^2/2)}{ \omega^2 t^2/2 }
\end{align}

I'd not expect such a tidy result for an arbitrary $V(x) = \lambda x^n$ potential.

\subsection{Problem 4. Feynman-Hellman relation.}

For continuously parametrized eigenstate, eigenvalue and Hamiltonian $\ket{\psi(\lambda)}$, $E(\lambda)$ and $H(\lambda)$ respectively, we can relate the derivatives

\begin{align*}
\PD{\lambda}{} ( H \ket{\psi} ) &= \PD{\lambda}{} ( E \ket{\psi} ) \\
\implies \\
\PD{\lambda}{H} \ket{\psi} +H \PD{\lambda}{\ket{\psi}} &= \PD{\lambda}{E} \ket{\psi} + E \PD{\lambda}{\ket{\psi} } 
\end{align*}

Left multiplication by $\bra{\psi}$ gives

\begin{align*}
\bra{\psi}\PD{\lambda}{H} \ket{\psi} +\bra{\psi}H \PD{\lambda}{\ket{\psi}} &= \bra{\psi}\PD{\lambda}{E} \ket{\psi} +  E \bra{\psi}\PD{\lambda}{\ket{\psi} } \\
\implies \\
\bra{\psi}\PD{\lambda}{H} \ket{\psi} +(\bra{\psi}E) \PD{\lambda}{\ket{\psi}} &= \bra{\psi}\PD{\lambda}{E} \ket{\psi} +  E \bra{\psi}\PD{\lambda}{\ket{\psi} } \\
\implies \\
\bra{\psi}\PD{\lambda}{H} \ket{\psi} &= \PD{\lambda}{E} \braket{\psi}{\psi},
\end{align*}

which provides the desired identity
\begin{align}\label{eqn:desaiCh3:4}
\PD{\lambda}{E} 
= \bra{\psi(\lambda)}\PD{\lambda}{H} \ket{\psi(\lambda)}
%= \expectation{ \psi(\lambda) \PD{\lambda}{H} \psi(\lambda) }
\end{align}

\subsection{Problem 5. }
\subsection{Problem 6. }
\subsection{Problem 7. }
\subsection{Problem 8. }
\subsection{Problem 9. }
\subsection{Problem 10. }
\subsection{Problem 11. commutator of angular momentum with Hamiltonian.}

Show that $\antisymmetric{\BL}{H} = 0$, where $H = \Bp^2/2m + V(r)$.

This follows by considering $\antisymmetric{\BL}{\Bp^2}$, and $\antisymmetric{\BL}{V(r)}$.  Let

\begin{align}\label{eqn:desaiCh3:1100}
L_{jk} = x_j p_k - x_k p_j,
\end{align}

so that 

\begin{align}\label{eqn:desaiCh3:1101}
\BL = \Be_i \epsilon_{ijk} L_{jk}.
\end{align}

We now need to consider the commutators of the operators $L_{jk}$ with $\Bp^2$ and $V(r)$.

Let's start with $p^2$.  In particular

\begin{align*}
\Bp^2 x_m p_n
&=
p_k p_k x_m p_n \\
&=
p_k (p_k x_m) p_n \\
&=
p_k (-i\hbar \delta_{km} + x_m p_k) p_n \\
&=
-i\hbar p_m p_n + (p_k x_m) p_k p_n \\
&=
-i\hbar p_m p_n + (-i \hbar \delta_{km} + x_m p_k ) p_k p_n \\
&=
-2 i\hbar p_m p_n + x_m p_n \Bp^2.
\end{align*}

So our commutator with $\Bp^2$ is

\begin{align*}
\antisymmetric{L_{jk}}{\Bp^2}
&=
(x_j p_k - x_j p_k) \Bp^2 
-( -2 i\hbar p_j p_k + x_j p_k \Bp^2 +2 i\hbar p_k p_j - x_k p_j \Bp^2 ).
\end{align*}

Since $p_j p_k = p_k p_j$, all terms cancel out, and the problem is reduced to showing that 

\begin{align*}
\antisymmetric{\BL}{H} &= \antisymmetric{\BL}{V(r)} = 0.
\end{align*}

Now assume that $V(r)$ has a series representation

\begin{align*}
V(r) &= \sum_j a_j r^j = \sum_j a_j (x_k x_k)^{j/2}
\end{align*}

We'd like to consider the action of $x_m p_n$ on this function

\begin{align*}
x_m p_n V(r) \Psi
&= -i \hbar x_m \sum_j a_j \partial_n (x_k x_k)^{j/2} \Psi \\
&= -i \hbar x_m \sum_j a_j (j x_n (x_k x_k)^{j/2-1} + r^j \partial_n \Psi) \\
&= -\frac{i \hbar x_m x_n}{r^2} \sum_j a_j j r^j + 
x_m V(r) p_n \Psi
\end{align*}

\begin{align*}
L_{mn} V(r) 
&=
(x_m p_n - x_n p_m) V(r) \\
&= 
-\frac{i \hbar x_m x_n}{r^2} \sum_j a_j j r^j
+\frac{i \hbar x_n x_m}{r^2} \sum_j a_j j r^j 
+ 
V(r) (x_m p_n - x_n p_m )
\\
&= 
V(r) L_{mn}
\end{align*}

Thus $\antisymmetric{L_{mn}}{V(r)} = 0$ as expected, implying $\antisymmetric{\BL}{H} = 0$.

\EndArticle
