\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Errata for Feynman's Quantum Electrodynamics (Addison-Wesley)?}
\label{chap:feynmanQEDerrata}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2010/feynmanQEDerrata.pdf}
\date{May 28, 2010}
\revisionInfo{feynmanQEDerrata.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

I got a nice present today which included one of \href{http://www.amazon.com/Quantum-Electrodynamics-Advanced-Book-Classics/dp/0201360756/ref=sr_1_1?ie=UTF8&s=books&qid=1275092228&sr=8-1}{Feynman's QED books}.  I noticed some early mistakes, and since I can't find an errata page anywhere, I'll collect them here.

\section{Third Lecture}
\subsection{Page 6 typos.}

The electric field is given in terms of only the scalar potential
\begin{align*}
\BE = -\spacegrad \phi + \partial \phi/ \partial t,
\end{align*}

and should be
\begin{align*}
\BE = -\spacegrad \phi - \inv{c} \partial \BA/ \partial t.
\end{align*}

The invariant gauge transformation for the vector and scalar potentials are given as

\begin{align*}
\BA' &= \BA = \spacegrad \chi \\
\phi' &= \phi + \partial \chi / \partial t
\end{align*}

But these should be
\begin{align*}
\BA' &= \BA + \spacegrad \chi \\
\phi' &= \phi - \inv{c} \partial \chi / \partial t
\end{align*}

The sign was crossed on the scalar potential transformation.  Feynman is also probably used to using $c=1$, but he doesn't do that explicitly at a different point on the page, so including it here is proper.

\subsection{Page 7 notes.}

The units in the transformation for the wave function don't look right.  We want to transform the Pauli equation

\begin{align*}
i \hbar \PD{t}{\Psi} = \inv{2 m} \left( \Bp - \frac{e}{c} \BA \right)^2 \Psi + e \phi \Psi,
\end{align*}

with a transformation of the form
\begin{align*}
\BA' &= \BA + \spacegrad \chi \\
\phi' &= \phi - \inv{c} \PD{t}{\chi} \\
\Psi' &= e^{-i \mu} \Psi,
\end{align*}

Where $\mu \propto \chi$ is presumed, and we want to find the proportionality constant required for invariance.  With $\Bp = - i \hbar \spacegrad$ we have

\begin{align*}
\Bp \Psi' 
&=
-i \hbar \spacegrad e^{-i \mu} \Psi \\
&=
-i \hbar \left( 
-i (\spacegrad \mu) e^{-i \mu} \Psi 
+ e^{-i \mu} \spacegrad \Psi  
\right) \\
&=
+ e^{-i \mu} \left( \Bp + \hbar \spacegrad \mu \right) \Psi,
\end{align*}

so
\begin{align*}
(\Bp -\frac{e}{c} \BA' )\Psi' 
&=
e^{-i \mu} \left( \Bp - \frac{e}{c} \BA - \spacegrad (\hbar \mu + \frac{e}{c} \chi) \right) \Psi.
\end{align*}

For the time partial we have

\begin{align*}
\PD{t}{\Psi'} &= e^{-i \mu} \PD{t}{\Psi} -i \PD{t}{\mu} e^{-i \mu} \Psi,
\end{align*}

and the scalar potential term transforms as
\begin{align*}
e \phi' \Psi'
&=
e \left( \phi - \inv{c} \PD{t}{\chi} \right) e^{-i \mu } \Psi
\end{align*}

Putting the pieces together we have

\begin{align*}
i \hbar e^{-i \mu}
\left( \PD{t}{} -i \PD{t}{\mu} \right) \Psi 
&=
\inv{2m}
\left(\Bp -\frac{e}{c} \BA -\frac{e}{c} \spacegrad \chi \right)
e^{-i \mu} \left( \Bp - \frac{e}{c} \BA - \spacegrad (\hbar \mu + \frac{e}{c} \chi) \right) \Psi 
+ e \left( \phi - \inv{c} \PD{t}{\chi} \right) e^{-i \mu } \Psi
\end{align*}

We need one more intermediate result, that of

\begin{align*}
\Bp e^{-i \mu } \BD
&= 
- i \hbar e^{-i \mu} \left( -i (\spacegrad \mu) + \spacegrad \right) \BD \\
&= 
e^{-i\mu} (\Bp - \hbar \spacegrad \mu) \BD.
\end{align*}

So we have
\begin{align*}
i \hbar \PD{t}{\Psi}
+\hbar \PD{t}{\mu} \Psi 
&=
\inv{2m}
\left(\Bp - \hbar \spacegrad \mu -\frac{e}{c} \BA -\frac{e}{c} \spacegrad \chi \right)
\left( \Bp - \frac{e}{c} \BA - \spacegrad (\hbar \mu + \frac{e}{c} \chi) \right) \Psi 
+ e \left( \phi - \inv{c} \PD{t}{\chi} \right) \Psi.
\end{align*}

To get rid of the $\mu$, and $\chi$ time partials we need

\begin{align*}
\hbar \PD{t}{\mu} = - \frac{e}{c} \PD{t}{\chi}
\end{align*}

Or 
\begin{align*}
\mu = - \frac{e}{c\hbar} \chi
\end{align*}

This also kills off all the additional undesirable terms in the transformed $\BP^2$ operator (with $\BP = \Bp - e \BA/c$), leaving the invariant transformation completely specified

\begin{align*}
\BA' &= \BA + \spacegrad \chi \\
\phi' &= \phi - \inv{c} \PD{t}{ \chi } \\
\Psi' &= \exp\left( i \frac{e}{ \hbar c } \chi \right) \Psi,
\end{align*}

This is a fair bit different than Feynman's result, but since he starts with the wrong electrodynamic gauge transformation, that's not too unexpected.

\section{Second Lecture}

This isn't errata, but I found the following required slight exploration.  He gives (implicitly)

\begin{align*}
\overline{\sin^2(\omega t - \BK \cdot \Bx)} = \inv{2}
\end{align*}

Is this an average over space and time?  How would one do that?  What do we get just integrating this over the volume?  That dot product is $\BK \cdot \Bx = 2 \pi \left(\frac{m}{\lambda_1} x + \frac{n}{\lambda_2} y + \frac{o}{\lambda_3} z \right)$.  Our average over the volume, for $m \ne 0$, using \href{http://www.wolframalpha.com/input/?i=\int+sin^2(a+x+%2B+b)+dx}{wolfram alpha to do the dirty work}, is then

\begin{align*}
&\inv{\lambda_1 \lambda_2 \lambda_3} 
\int_{z=0}^{\lambda_3} dz
\int_{y=0}^{\lambda_2} dy
\int_{x=0}^{\lambda_1}
dx \sin^2 \left( 
-\frac{2 \pi m x}{\lambda_1} 
-\frac{2 \pi n y}{\lambda_2} 
-\frac{2 \pi o z}{\lambda_3} 
+ \omega t \right) \\
&=
\inv{\lambda_1 \lambda_2 \lambda_3} 
\int_{z=0}^{\lambda_3} dz
\int_{y=0}^{\lambda_2} dy
{\left.
\frac{-\lambda_1}{4 \pi m} \left( 
-\frac{2 \pi m }{\lambda_1} x 
-\frac{2 \pi n y}{\lambda_2} 
-\frac{2 \pi o z}{\lambda_3} 
+ \omega t \right)
\right\vert}_{x=0}^{\lambda_1} \\
&-
\inv{\lambda_1 \lambda_2 \lambda_3} 
\int_{z=0}^{\lambda_3} dz
\int_{y=0}^{\lambda_2} dy
{\left.
\frac{-\lambda_1}{8 \pi m} 
\sin \left( 2 \left(
-\frac{2 \pi m }{\lambda_1} x 
-\frac{2 \pi n y}{\lambda_2} 
-\frac{2 \pi o z}{\lambda_3} 
+ \omega t \right) \right)
\right\vert}_{x=0}^{\lambda_1}
\end{align*}

Since the sine integral vanishes, we have just $1/2$ as expected regardless of the angular frequency $\omega$.  Okay, that makes sense now.  Looks like $\omega$ is only relevant for the single $\BK = 0$ Fourier component, but that likely doesn't matter since I seem to recall that the $\BK = 0$ Fourier component of this oscillators in a box problem was entirely constant (and perhaps zero?).

%\EndArticle
\EndNoBibArticle
