%
%
%
% Copyright © 2012 Peeter Joot
% All Rights Reserved
%
% This file may be reproduced and distributed in whole or in part, without fee, subject to the following conditions:
%
% o The copyright notice above and this permission notice must be preserved complete on all complete or partial copies.
%
% o Any translation or derived work must be approved by the author in writing before distribution.
%
% o If you distribute this work in part, instructions for obtaining the complete version of this file must be included, and a means for obtaining a complete version provided.
%
%
% Exceptions to these rules may be granted for academic purposes: Write to the author and ask.
%
%
%
\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

%\usepackage[english]{babel}
%\usepackage{media9}
\chapter{A Fourier series refresher.}

\label{chap:fourierSeries}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/fourierSeries.pdf}
\date{Apr 3, 2012}
\gitRevisionInfo{fourierSeries}
\keywords{Fourier series,Fourier coefficient}

\beginArtWithToc
%\beginArtNoToc
%\wordpresscategory{}

\section{Motivation.}

I'd used the wrong scaling in a Fourier series over a $[0, 1]$ interval.  Here's a reminder to self what the right way to do this is.

\section{Guts}

Suppose we have a function that is defined in terms of a trigonometric Fourier sum

\begin{equation}\label{eqn:fourierSeries:10}
\phi(x) = \sum c_k e^{i \omega k x},
\end{equation}

where the domain of interest is $x \in [a, b]$.  Stating the problem this way avoids any issue of existence.  We know $c_k$ exists, but just want to find what they are given some other representation of the function.

Multiplying and integrating over our domain we have

\begin{equation}\label{eqn:fourierSeries:30}
\begin{aligned}
\int_a^b \phi(x) e^{-i \omega m x} dx 
&= \sum c_k \int_a^b e^{i \omega (k -m) x} dx \\
&= c_m (b - a) + \sum_{k \ne m} \frac{e^{i \omega(k-m) b} - e^{i \omega(k-m)a}}{i \omega (k -m)} .
\end{aligned}
\end{equation}

We want all the terms in the sum to be be zero, requiring equality of the exponentials, or

\begin{equation}\label{eqn:fourierSeries:50}
e^{i \omega (k -m) (b -a )} = 1,
\end{equation}

or

\begin{equation}\label{eqn:fourierSeries:70}
\omega = \frac{2 \pi}{b - a}.
\end{equation}

This fixes our Fourier coefficients

\begin{equation}\label{eqn:fourierSeries:90}
c_m = \inv{b - a} \int_a^b \phi(x) e^{- 2 \pi i m x/(b - a)} dx.
\end{equation}

Given this, the correct (but unnormalized) Fourier basis for a $[0, 1]$ interval would be the functions $e^{2 \pi i x}$, or the sine and cosine equivalents.

%\EndArticle
\EndNoBibArticle
