\input{../peeter_prologue.tex}

\chapter{Energy and momentum for Fourier transform solutions to homogeneous Maxwell equation.}
\label{chap:ftMaxVacuum}
\blogpage{http://sites.google.com/site/peeterjoot/math2009/ftMaxVacuum.pdf}
\date{Dec 21, 2009}
\revisionInfo{ftMaxVacuum.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Motivation and notation.}

In \chapcite{fourierMaxVac}, building on \cite{complexFieldEnergy} a derivation for the energy and momentum density was derived for an assumed Fourier series solution to the homogeneous Maxwell's equation.  Here we move to the continous case examining Fourier transform solutions and the associated energy and momentum density.

A complex (phasor) representation is implied, so taking real parts when all is said and done is required of the fields.  For the energy momentum tensor the Geometric Algebra form, modified for complex fields, is used

\begin{align}\label{eqn:ftMaxVacuum:1}
T(a) = -\frac{\epsilon_0}{2} \Real \Bigl( \conjugateStar{F} a F \Bigr).
\end{align}

The assumed four vector potential will be written

\begin{align}\label{eqn:ftMaxVacuum:2}
A(\Bx, t) = A^\mu(\Bx, t) \gamma_\mu = \inv{(\sqrt{2 \pi})^3} \int A(\Bk, t) e^{i \Bk \cdot \Bx } d^3 \Bk.
\end{align}

Subject to the requirement that $A$ is a solution of Maxwell's equation

\begin{align}\label{eqn:ftMaxVacuum:3}
\grad (\grad \wedge A) = 0.
\end{align}

To avoid latex hell, no special notation will be used for the Fourier coefficients $A(\Bk, t)$.  When convient and unambiguous, the $\Bk$ and $t$ dependence will be implied.

Having picked a time and space respresentation for the field, it will be natural to express both the four potential and the gradient as scalar plus spatial vector, instead of using the Dirac basis.  For the gradient this is

\begin{align}\label{eqn:ftMaxVacuum:4}
\grad &= \gamma^\mu \partial_\mu = (\partial_0 - \spacegrad) \gamma_0 = \gamma_0 (\partial_0 + \spacegrad),
\end{align}

and for the four potential (or the Fourier transform functions), this is

\begin{align}\label{eqn:ftMaxVacuum:5}
A &= \gamma_\mu A^\mu = (\phi + \BA) \gamma_0 = \gamma_0 (\phi - \BA).
\end{align}

\section{Setup}

The field bivector $F = \grad \wedge A$ is required for the energy momentum tensor.  This is

\begin{align*}
\grad \wedge A
&= \inv{2}\left( \rgrad A - A \lgrad \right) \\
&= \inv{2}\left( (\rpartial_0 - \rspacegrad) \gamma_0 \gamma_0 (\phi - \BA)
- 
(\phi + \BA) \gamma_0 \gamma_0 (\lpartial_0 + \lspacegrad)
\right) \\
&= -\spacegrad \phi -\partial_0 \BA + \inv{2}(\rspacegrad \BA - \BA \lspacegrad) 
\end{align*}

This last term is a spatial curl and the field is then

\begin{align}\label{eqn:ftMaxVacuum:6}
F = -\spacegrad \phi -\partial_0 \BA + \spacegrad \wedge \BA 
\end{align}

Applied to the Fourier representation this is

\begin{align}\label{eqn:ftMaxVacuum:7}
F = 
\inv{(\sqrt{2 \pi})^3} \int 
\left( 
- \inv{c} \dot{\BA}
- i \Bk \phi
+ i \Bk \wedge \BA
\right)
e^{i \Bk \cdot \Bx } d^3 \Bk.
\end{align}

The energy momentum tensor is then

\begin{align}\label{eqn:ftMaxVacuum:8}
T(a) &= -\frac{\epsilon_0}{2 (2 \pi)^3} \Real \iint 
\left( 
- \inv{c} \conjugateStar{\dot{\BA}}(\Bk',t)
- i \Bk' \conjugateStar{\phi}(\Bk', t)
+ i \Bk' \wedge \conjugateStar{\BA}(\Bk', t)
\right)
a
\left( 
- \inv{c} \dot{\BA}(\Bk, t)
- i \Bk \phi(\Bk, t)
+ i \Bk \wedge \BA(\Bk, t)
\right)
e^{i (\Bk -\Bk') \cdot \Bx } d^3 \Bk d^3 \Bk'.
\end{align}

Integrating this over all space and identification of the delta function

\begin{align}\label{eqn:ftMaxVacuum:9}
\delta(\Bk) \equiv \inv{(2 \pi)^3} \int e^{i \Bk \cdot \Bx} d^3 \Bx,
\end{align}

reduces the tensor to a single integral in the continuous angular wave number space of $\Bk$.

\begin{align}\label{eqn:ftMaxVacuum:10}
\int T(a) d^3 \Bx &= -\frac{\epsilon_0}{2} \Real \int 
\left( 
- \inv{c} \conjugateStar{\dot{\BA}}
- i \Bk \conjugateStar{\phi}
+ i \Bk \wedge \conjugateStar{\BA}
\right)
a
\left( 
- \inv{c} \dot{\BA}
- i \Bk \phi
+ i \Bk \wedge \BA
\right)
d^3 \Bk.
\end{align}

Writing 

\begin{align}\label{eqn:ftMaxVacuum:11}
\int T(\gamma^0) d^3 \Bx = (H + \BP) \gamma_0,
\end{align}

and observing that $\gamma_0$ commutes with spatial bivectors and anticommutes with spatial vectors one has

\begin{align}\label{eqn:ftMaxVacuum:12}
H + \BP =
-\frac{\epsilon_0}{2} \Real \int 
\gpgrade{
\left( 
- \inv{c} \conjugateStar{\dot{\BA}}
- i \Bk \conjugateStar{\phi}
+ i \Bk \wedge \conjugateStar{\BA}
\right)
\left( 
\inv{c} \dot{\BA}
+ i \Bk \phi
+ i \Bk \wedge \BA
\right)
}{0,1}
d^3 \Bk.
\end{align}

The scalar and spatial vector grade selection operator has been added for convienience and does not change the result since those are neccessarily the only grades anyhow.

\EndArticle
