%\input{../peeter_prologue_print.tex}
\input{../peeter_prologue_widescreen.tex}

\chapter{Notes on Goldstein's Route's procedure.}
\label{chap:goldsteinRouth}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2010/goldsteinRouth.pdf}
\date{Mar 3, 2010}
\revisionInfo{goldsteinRouth.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Motivation.}

Attempting study of \cite{goldstein1951cm} section 7-2 on Routh's procedure has been giving me some trouble.  It wasn't ``sinking in'', indicating a fundamental misunderstanding, or at least a requirement to work some examples.  Here I pick a system, the spherical pendulum, which has the required ignorable coordinate, to illustrate the ideas for myself with something less abstract.

\section{Spherical pendulum example.}

The Lagrangian for the pendulum is

\begin{align}\label{eqn:goldsteinRouth:1}
\LL = \inv{2} m r^2 \left( \dot{\theta}^2 + \dot{\phi}^2 \sin^2 \theta \right) - m g r ( 1 + \cos\theta ),
\end{align}

and our conjugate momenta are therefore

\begin{align}\label{eqn:goldsteinRouth:2}
p_\theta &= \PD{\dot{\theta}}{\LL} = m r^2 \dot{\theta} \\
p_\phi &= \PD{\dot{\phi}}{\LL} = m r^2 \sin^2\theta \dot{\phi}.
\end{align}

That's enough to now formulate the Hamiltonian $H = \dot{\theta} p_\theta + \dot{\phi} p_\phi - \LL$, which is

\begin{align}\label{eqn:goldsteinRouth:3}
H = H(\theta, p_\theta, p_\phi) = \inv{2 m r^2 } (p_\theta)^2 + \inv{2 m r^2 \sin^2\theta} (p_\phi)^2 + m g r ( 1 + \cos\theta ).
\end{align}

We've got the ignorable coordinate $\phi$ here, since the Hamiltonian has no explicit dependence on it.  In the Hamiltonian formalism the constant of motion associated with this comes as a consequence of evaluating the Hamiltonian equations.  For this system, those are

\begin{align}\label{eqn:goldsteinRouth:4}
\PD{\theta}{H} &= - \dot{p}_\theta \\
\PD{\phi}{H} &= - \dot{p}_\phi \\
\PD{p_\theta}{H} &= \dot{\theta} \\
\PD{p_\phi}{H} &= \dot{\phi},
\end{align}

Or, explicitly,
\begin{align}\label{eqn:goldsteinRouth:5}
- \dot{p}_\theta &= -m g r \sin\theta - \frac{\cos\theta}{2 m r^2 \sin^3 \theta} (p_\phi)^2  \\
- \dot{p}_\phi &= 0 \\
\dot{\theta} &= \inv{m r^2 } p_\theta \\
\dot{\phi} &= \inv{m r^2 \sin^2 \theta} p_\phi.
\end{align}

The second of these provides the integration constant, allowing us to write, $p_\phi = \alpha$.  Once this is done, our Hamiltonian example is reduced to one complete set of conjugate coordinates,

\begin{align}\label{eqn:goldsteinRouth:6}
H(\theta, p_\theta, \alpha) = \inv{2 m r^2 } (p_\theta)^2 + \inv{2 m r^2 \sin^2\theta} \alpha^2 + m g r ( 1 + \cos\theta ).
\end{align}

Goldstein notes that the behavior of the cyclic coordinate follows by integrating

\begin{align}\label{eqn:goldsteinRouth:7}
\dot{q}_n = \PD{\alpha}{H}.
\end{align}

In this example $\alpha = p_\theta$, so this is really just one of our Hamiltonian equations

\begin{align}\label{eqn:goldsteinRouth:8}
\dot{\phi} = \PD{p_\phi}{H}.
\end{align}

Okay, good.  First part of the mission is accomplished.  The setup for Routh's procedure no longer has anything mysterious to it.

Now, Goldstein defines the Routhian as

\begin{align}\label{eqn:goldsteinRouth:9}
R = p_i \dot{q}_i - \LL,
\end{align}

where the index $i$ is summed only over the cyclic (ignorable) coordinates.  For this spherical pendulum example, this is $q_i = \phi$, and $p_i = m r^2 \sin^2 \theta \dot{\phi}$, for

\begin{align}\label{eqn:goldsteinRouth:10}
R = \inv{2} m r^2 \left( -\dot{\theta}^2 + \dot{\phi}^2 \sin^2 \theta \right) + m g r ( 1 + \cos\theta ).
\end{align}

Now, we should also have for the non-cyclic coordinates, just like the Euler-Lagrange equations 

\begin{align}\label{eqn:goldsteinRouth:11}
\PD{\theta}{R} = \frac{d}{dt} \PD{\dot{\theta}}{R}.
\end{align}

Evaluating this we have

\begin{align}\label{eqn:goldsteinRouth:12}
m r^2 \sin\theta \cos\theta \dot{\phi}^2 - m g r \sin\theta = \frac{d}{dt} \left( -m r^2 \dot{\theta} \right).
\end{align}

It would be reasonable now to compare this the $\theta$ Euler-Lagrange equations, but evaluating those we get

\begin{align}\label{eqn:goldsteinRouth:13}
m r^2 \sin\theta \cos\theta \dot{\phi}^2 + m g r \sin\theta = \frac{d}{dt} \left( m r^2 \dot{\theta} \right).
\end{align}

Bugger.  We've got a sign difference on the $\dot{\phi}^2$ term.

\section{Simpler planar example.}

Having found an inconsistency with Routhian formalism and the concrete example of the spherical pendulum which has a cyclic coordinate as desired, let's step back slightly, and try a simpler example, artificially constructed

\begin{align}\label{eqn:goldsteinRouth:20}
\LL = \inv{2} m (\dot{x}^2 + \dot{y}^2 ) - V(x).
\end{align}

Our Hamiltonian and Routhian functions are

\begin{align}\label{eqn:goldsteinRouth:21}
H &= \inv{2} m (\dot{x}^2 + \dot{y}^2 ) + V(x) \\
R &= \inv{2} m (-\dot{x}^2 + \dot{y}^2 ) + V(x) 
\end{align}

For the non-cyclic coordinate we should have

\begin{align}\label{eqn:goldsteinRouth:22}
\PD{x}{R} = \frac{d}{dt} \PD{\dot{x}}{R},
\end{align}

which is

\begin{align}\label{eqn:goldsteinRouth:23}
V'(x) = \frac{d}{dt}\left( - m \dot{x} \right)
\end{align}

Okay, good, that's what is expected, and exactly what we get from the Euler-Lagrange equations.  This looks good, so where did things go wrong in the spherical pendulum evaluation.

\EndArticle
