%
% Copyright © 2016 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%{
\input{../blogpost.tex}
\renewcommand{\basename}{helmholtzSolution}
%\renewcommand{\dirname}{notes/phy1520/}
\renewcommand{\dirname}{notes/ece1228-electromagnetic-theory/}
%\newcommand{\dateintitle}{}
%\newcommand{\keywords}{}

\input{../peeter_prologue_print2.tex}

\usepackage{peeters_layout_exercise}
\usepackage{peeters_braket}
\usepackage{peeters_figures}
\usepackage{siunitx}
%\usepackage{mhchem} % \ce{}
%\usepackage{macros_bm} % \bcM
%\usepackage{txfonts} % \ointclockwise

\beginArtNoToc

\generatetitle{Helmholtz solution}
%\chapter{Helmholtz solution}
%\label{chap:helmholtzSolution}

We know that the Laplace equation

\begin{dmath}\label{eqn:helmholtzSolution:40}
\spacegrad^2 \phi = \rho/\epsilon,
\end{dmath}

has solution

\begin{dmath}\label{eqn:helmholtzSolution:60}
\phi(\Bx) = \inv{4 \pi \epsilon} \int \frac{\rho(\Bx')}{\Abs{\Br - \Br'}} dV'.
\end{dmath}

We also know that the wave equation
\begin{dmath}\label{eqn:helmholtzSolution:20}
\spacegrad^2 \phi = \inv{v^2} \PDSq{t}{\phi}
\end{dmath}

has solutions of the form

%f(x,t) = f(x - t/v)
\begin{dmath}\label{eqn:helmholtzSolution:240}
\phi(\Bx, t) = f( t \pm \frac{\Bbeta}{\omega} \cdot \Bx ),
\end{dmath}

or
\begin{dmath}\label{eqn:helmholtzSolution:260}
\phi(\Bx, t) = f( t \pm \frac{\Bbeta}{\omega} \cdot \Bx ),
\end{dmath}

\begin{dmath}\label{eqn:helmholtzSolution:300}
\beta^2 \phi = \rho/\epsilon
\end{dmath}

In class it was argued that the solution of the non-homogeneous Helmholtz equation

\begin{dmath}\label{eqn:helmholtzSolution:20}
\spacegrad^2 \phi + \beta^2 \phi = \rho/\epsilon
\end{dmath}

should have the same form, but retarded in time

\begin{dmath}\label{eqn:helmholtzSolution:80}
\phi(\Bx, t)
= \inv{4 \pi \epsilon} \int \frac{\rho(\Bx', t - R/v)}{\Abs{\Br - \Br'}} dV'
= \Real \lr{
\inv{4 \pi \epsilon} \int \frac{\rho(\Bx') e^{j \omega( t - R/v)}}{\Abs{\Br - \Br'}} dV'
}.
\end{dmath}

so if the time domain solution is \( \phi(\Bx, t) = \Real \lr{ \phi(\Bx) e^{j \omega t} } \), the time harmonic potential is

\begin{dmath}\label{eqn:helmholtzSolution:100}
\phi(\Bx)
=
\inv{4 \pi \epsilon} \int \frac{\rho(\Bx') e^{-j \omega R/v}}{\Abs{\Br - \Br'}} dV'.
\end{dmath}

This was written as

\begin{dmath}\label{eqn:helmholtzSolution:120}
\phi(\Bx)
=
\inv{4 \pi \epsilon} \int \frac{\rho(\Bx') e^{-j \beta R}}{\Abs{\Br - \Br'}} dV'.
\end{dmath}

where we must have
\begin{dmath}\label{eqn:helmholtzSolution:140}
\beta = \omega/v.
\end{dmath}

The justification for the time retardation was physical, arguing that there is a delay required for any change to manifest at a distance.

\paragraph{TEM, TM, AND TE modes}

The TEM mode is that for which

\begin{equation}\label{eqn:helmholtzSolution:180}
\BH \cdot \zcap = \BE \cdot \zcap = 0,
\end{equation}

whereas the TM mode is one for which

\begin{equation}\label{eqn:helmholtzSolution:200}
\BH \cdot \zcap = 0,
\end{equation}

and the TE mode is one for which
\begin{equation}\label{eqn:helmholtzSolution:220}
\BE \cdot \zcap = 0.
\end{equation}

Strictly speaking these are the \( \textrm{TEM}^z, \textrm{TM}^z \) and \( \textrm{TE}^z \) modes respectively.

See \citep{cheng1989field} for a justification that \( \BJ \) must be axial (like a coaxial cable) for a TEM mode satisfying

\begin{dmath}\label{eqn:helmholtzSolution:160}
\spacegrad \cross \BH = \BJ + j \omega \epsilon \BE.
\end{dmath}

From the expansion of the ``Master solution equation'', one way of constructing a TEM mode was pointed out in the notes.  A similar TEM solution, can be constructed in a similar swapping \( \BA \) and \( \BF \).  HW: Do this as an exersize as well as walking through the details.

\paragraph{TE}

The class notes give the  \( \textrm{TE}^z \) solution as
\begin{dmath}\label{eqn:helmholtzSolution:280}
\begin{aligned}
E_x &= -j \inv{\omega \mu \epsilon} \frac{\partial^2 A_z}{\partial_x \partial_y} ...
\end{aligned}
\end{dmath}

HW: In class it was claimed that \( \BE \cdot \BH = 0 \), but that wasn't obvious by inspection.

Note: be mindful that some books will express this with one of the field components equal to zero.  This is due to the fact that a rotation of coordinates can kill one of the field components.  HW: show this.

\paragraph{more on tunnelling.}
%superluminal tunnelling through wave guides: big -> small -> big.  similar to the tunnelling in prisms coupled by dielelectric even past critical angle.

%}
\EndArticle
%\EndNoBibArticle
