%
% Copyright © 2016 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%{
\input{../blogpost.tex}
\renewcommand{\basename}{hw1228}
%\renewcommand{\dirname}{notes/phy1520/}
\renewcommand{\dirname}{notes/ece1228-electromagnetic-theory/}
%\newcommand{\dateintitle}{}
%\newcommand{\keywords}{}

\input{../peeter_prologue_print2.tex}

\usepackage{peeters_layout_exercise}
\usepackage{peeters_braket}
\usepackage{peeters_figures}
\usepackage{siunitx}
\usepackage{mhchem} % \ce{}
%\usepackage{txfonts} % \ointclockwise

\beginArtNoToc

\generatetitle{XXX}
%\chapter{XXX}
%\label{chap:hw1228}
% \citep{sakurai2014modern} pr X.Y
% \citep{pozar2009microwave}
% \citep{qftLectureNotes}
% \citep{doran2003gap}
% \citep{jackson1975cew}
% \citep{griffiths1999introduction}

\paragraph{Displacement current.}

It will be left to homework to show that without the displacement current \( \PDi{t}{\BD} \), these equations will not satisfy conservation relations.

\paragraph{magnetic field lines}

For homework: show that magnetic field lines must close on themselves when there are no magnetic sources (zero divergence).  This is opposed to electric fields that spread out from the charge.

\paragraph{Helmholtz:}

Extra homework question (bonus) : can knowledge of the tangential components of the fields also be used to uniquely determine \( \BE \)?

\paragraph{Homework question:}

Starting from Maxwell's equations, in particular \( \oint \BE \cdot d\Bs = Q/\epsilon_0 \) that \( \BE = E_r \rcap \), and explicitly demonstrate why there is no \( \theta \) or \( \phi \) dependencies in this field.  Also calculate the potential \( V \propto 1/r \) associated with an electric field, and show that \( \BE = -\spacegrad V \), and show that this implies that \( -\int_b^a \BE \cdot d\Bl = V_a - V_b \).


\paragraph{Infinite self energy, Born and Infeld vacuum response}

HW: Consider a hydrogen atom unexcited and in thermal equilibrium. Calculate the magnitude of the electric field due to its nucleus at the site of its electron.

%\paragraph{HW: Show that potential at point A for an electric dipole is given by ...}
%\paragraph{HW: From above it can be shown (Poynting)}
%\paragraph{Poynting time harmonic: conjugate relations.}
%\paragraph{HW: k cross E, and k cross H relations}

\paragraph{surface currents}
%{eqn:emtLecture4:160}
(showing this may be given as a homework problem ... if not do it).

\paragraph{HW: reflection coefficients}

\paragraph{HW: druid}

Assuming that each gold atom contributes 1 conduction electron (\(Z=1\)) calculate the number of conduction electrons \(n\) that exist in a cubic centimeter of the gold. 
The mass density of the gold, is 19.3 [\si{g/cm^3}] and its molar mass (atomic mass) is \(Mm = 196.967 [\si{g/mol}]\). The number density of conduction electrons, \(n\) , for most metals is in the range of \( 10^22 to 25 \times 10^22 \) electrons per cubic centimeter. Answer: For \ce{Au} this number is \( \sim 5.9 10 [\si{1/cm^3}] \).

\paragraph{HW: druid}

Using the equipartition of energy theorem from classical physics calculate the average electron velocity at room temperature.

%}
\EndArticle
%\EndNoBibArticle
