%
% Copyright © 2013 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
% pick one:
%\input{../assignment.tex}
\input{../blogpost.tex}
\renewcommand{\basename}{hyperSphereVolume}
\renewcommand{\dirname}{notes/phy452/}
%\newcommand{\dateintitle}{}
\newcommand{\keywords}{hypersphere, volume, Jacobian, Euclidean space, Minkowski space, Geometric Algebra, Gamma function}

\input{../peeter_prologue_print2.tex}

\beginArtNoToc

\generatetitle{hypersphere volume calculation the easy way}
\chapter{hypersphere volume calculation the easy way}
%\label{chap:hyperSphereVolume}
\section{Motivation}

In problem set three I solved the hypersphere volume problem the hard way.  I did it the way that I thought was obvious, starting with the spherical coordinate hypervolume element for an Euclidean space.  That build on the volume element calculation I'd previously done for four dimensional Euclidean and Minkowski spaces in \citep{gabook:nvolume}.  This time I avoided any use of Geometric Algebra and wrote out the volume element directly using a Jacobian transformation instead of a wedge product (that leads to the same Jacobian).  I then proceeded to integrate the volume element, come up with a recurrence relation for the volume, solve that for even and odd powers, and find a common expression using Gamma functions that was correct for even and odd powers.  It was a labourious process, but satisfying since I'd previously tried this calculation and not succeeded.

As and after I did this calculation, which I'll post later, I figured there had to have been an easier way.  In the midterm prep reading of \S 5.5 of \citep{ma1985statistical} I found just that method.  It's done there in about six lines, using a trick that seemed really sneaky, but after a bit of reflection, very sensible.  I'll outline that here for 2 and 3 dimensions to show the technique, and the reader can generalize if desired.

\section{Guts}

% this is to produce the sites.google url and version info and so forth (for blog posts)
%\vcsinfo
\EndArticle
%\EndNoBibArticle
