\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Some details for the radial solution of the hydrogen atom.}
\label{chap:hyrdogenLaguerre}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2010/hyrdogenLaguerre.pdf}
\date{Nov 24, 2010}
\revisionInfo{hyrdogenLaguerre.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

For the hydrogen atom, after some variable substitutions the radial part of the Schr\"{o}dinger equation takes the form

\begin{equation}\label{eqn:hyrdogenLaguerre:10}
\frac{d^2 R_l}{d\rho^2} + \frac{2}{\rho} \frac{d R_l}{d\rho} + \left( \frac{\lambda}{\rho} - \frac{l(l+1)}{\rho^2} - \inv{4} \right) R_l = 0
\end{equation}

In \cite{desai2009quantum} it is argued that the functions $R_l$ are of the form

\begin{equation}\label{eqn:hyrdogenLaguerre:20}
R_l = \rho^s L(\rho) e^{-\rho/2}
\end{equation}

where $L$ is a polynomial in $\rho$, specifically Laguerre polynomials.  Let's look at some of those details a bit more closely.

\section{Guts}

The first part of the argument comes from considering the $\rho \rightarrow \infty$ case, where Schr\"{o}dinger's equation is approximately

\begin{equation}\label{eqn:hyrdogenLaguerre:10a}
\frac{d^2 R_l}{d\rho^2} - \inv{4} R_l \approx 0.
\end{equation}

This large $\rho$ approximation has solutions $e^{\pm \rho/2}$, and we take the negative sign case as physically meaningful in order for the wave function to be normalizable.

Next it is argued that polynomial multiples of this will also be approximate solutions.  Utilizing monomial multiple of the decreasing exponential as a trial solution, let's compute how this fits into the radial Schr\"{o}dinger's equation \ref{eqn:hyrdogenLaguerre:10} above.  Write

\begin{equation}\label{eqn:hyrdogenLaguerre:15}
R_l = \rho^s e^{-\rho/2}
\end{equation}

The derivatives are

\begin{align*}
R_l' &= \rho^{s-1} \left( s -\frac{\rho}{2}\right) e^{-\rho/2} \\
R_l'' &= 
\rho^{s-2} 
\left( s (s-1) -s \rho +\frac{1}{4} \rho^2 
\right)
e^{-\rho/2}
\end{align*}

and substitution yields
%\rho^{s-2} 
%e^{-\rho/2}
%\left( 
%s (s-1) 
%-s \rho 
%+\frac{1}{4} \rho^2 
%+2s 
%-\rho 
%+\lambda \rho 
%- l(l+1) 
%- \frac{\rho^2}{4} 
%\right)
%&=
\begin{equation}\label{eqn:hyrdogenLaguerre:16}
\rho^{s-2} 
e^{-\rho/2}
\left( 
(s - \rho) (s+1) 
+\lambda \rho 
- l(l+1) 
\right)
\end{equation}

There are two things that this can show.  The first is that for $\rho \rightarrow \infty$ this produces a polynomial with degree $s-2$ and $s-1$ terms multiplied by the exponential, and we have approximately

\begin{equation}\label{eqn:hyrdogenLaguerre:17}
\rho^{s-1} 
e^{-\rho/2}
(\lambda - s - 1)
\end{equation}

The $s-1$ terms will dominate the polynomial, but the exponential dominate all, approaching zero for $\rho \rightarrow \infty$, just as the non-polynomial multiplied $e^{-\rho/2}$ approximate solution will.  This confirms that in the limit this polynomial multiplied exponential still has the desired behavior in the large $\rho$ limit.  Also observe that in the limit of small $\rho$ we have approximately

\begin{equation}\label{eqn:hyrdogenLaguerre:18}
\rho^{s-2} 
e^{-\rho/2}
\left( 
s (s+1) - l(l+1) 
\right)
\end{equation}

Since $\rho^{s-2} \rightarrow \infty$ as $\rho \rightarrow 0$, we require either a different trial solution, or $s=l$ to have a normalizable wavefunction.

Before settling on $s=l$ let's compute the derivatives for a more general trial function, of the form \ref{eqn:hyrdogenLaguerre:20}, and substitute those.  After a bit of computation we find

\begin{equation}\label{eqn:hyrdogenLaguerre:19}
R_l' = \rho^{s-1} e^{-\rho/2} \left( \left( s - \frac{\rho}{2} \right) L + \rho L'
\right)
\end{equation}
\begin{equation}\label{eqn:hyrdogenLaguerre:19b}
R_l'' = \rho^{s-2} e^{-\rho/2} \left( 
\left( s(s-1) - s \rho + \frac{\rho^2}{4} \right) L 
+\left( 2 s \rho -\rho^2 \right) L'
+ \rho^2 L''
\right)
\end{equation}

Putting these together and substitution back into \ref{eqn:hyrdogenLaguerre:10} yields

\begin{equation}\label{eqn:hyrdogenLaguerre:19c}
0 = \rho^{s-2} e^{-\rho/2} \left( 
L \left( (s-\rho)(s+1) + \rho \lambda -l (l+1)
\right) 
+\rho L' \left( 2 (s+1) -\rho \right) 
+ \rho^2 L''
\right)
\end{equation}

In the $\rho \rightarrow 0$ limit where the $\rho^{s-2}$ terms dominate \ref{eqn:hyrdogenLaguerre:50} becomes
\begin{equation}\label{eqn:hyrdogenLaguerre:50}
0 \approx
\rho^{s-2} L \left(
s(s+1) - l(l+1)
\right)
\end{equation}

Again, this provides the $s=l$ or $s = -(l+1)$ possibilities from the text, and we discard $s=-(l+1)$ due to non-normalizability.  A side question.  How does one solve integer equations like this?

\subsection{What remains?}

With $s=l$ killing off the $\rho^{s-2}$ terms, what is our differential equation for $L$?

\begin{equation}\label{eqn:hyrdogenLaguerre:100}
0 = 
\rho L''
+L' \left( 2 (l+1) -\rho \right) 
+L \left( \lambda - (l+1) \right) 
\end{equation}

Comparing this to \cite{wiki:laguerre} we have something pretty close to the stated differential equation for the Laguerre polynomial.  Ours is of the form

\begin{equation}\label{eqn:hyrdogenLaguerre:100}
0 = 
\rho L''
+L' \left( m -\rho \right) 
+L n,
\end{equation}

whereas the wikipedia article has $m=1$ explicitly.  No change of variables involving a scalar multiplicative factor for $\rho$ appears to be able to get it into that form, and I am guessing this is the differential equation for the associated Laguerre polynomial (something not stated in the wikipedia article).

Let's derive the recurrence relations for the coefficients, and work out the first few such polynomials to compare.

\EndArticle
%\EndNoBibArticle
