\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Maxwell's equation in STA.}
\label{chap:maxwellsFromGA}
%\useCCL
%\blogpage{http://sites.google.com/site/peeterjoot/math2010/maxwellsFromGA.pdf}
\date{Nov 8, 2010}
\revisionInfo{maxwellsFromGA.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Reciprocal frame.}

Associated with the basis $\{\gamma_\mu\}$ is the reciprocal basis $\{\gamma^\mu = \inv{\gamma^\mu}\}$, satisfying the relation

\begin{align*}
\gamma_\mu \cdot \gamma^\nu &= {\delta_\mu}^\nu 
\end{align*}

These reciprocal frame vectors differ only by a sign, with $\gamma^0 = \gamma_0$, and $\gamma^k = -\gamma_k$.

A vector may be represented in either upper or lower index coordinates $a = a^\mu \gamma_\mu = a_\mu \gamma^\mu$, where the coordinates may be extracted by taking dot products with the basis vectors or their reciprocals

\begin{align*}
a \cdot \gamma^\nu &= a^\nu \\
a \cdot \gamma_\nu &= a_\nu
\end{align*}

\section{Space time gradient}

The spacetime gradient, like the gradient in a Euclidean space, is defined such that the directional derivative relationship is satisfied

\begin{align*}
a \cdot \grad F(x)= \lim_{\tau \rightarrow 0} \frac{F(x + a\tau) - F(x)}{\tau}
\end{align*}

One can show that this requires the definition of the gradient to be
\begin{align*}
\grad &= \inv{\gamma_\mu} \PD{x^\mu}{} \equiv \gamma^\mu \partial_\mu \\
      &= \inv{\gamma^\mu} \PD{x_\mu}{} \equiv \gamma_\mu \partial^\mu.
\end{align*}

Written out explicitly with $x = ct \gamma_0 + x^k \gamma_k$, these partials are
\begin{align*}
\partial_0 &= \partial^0 = \inv{c} \PD{t}{} \\
\partial_k &= \PD{x^k}{} \\
\partial^k &= \PD{x_k}{} = -\partial_k \\
\end{align*}

\section{Space time split}

Pre or post multiplication by the time like basis vector $\gamma_0$ serves to split a four vector into a scalar and spatial vector components.  With $x = x^\mu \gamma_\mu$ we have

\begin{align*}
x \gamma_0 &= x^0 + x^k \gamma_k \gamma_0 \\
\gamma_0 x &= x^0 - x^k \gamma_k \gamma_0 
\end{align*}

As these bivectors $\gamma_k \gamma_0$ square to unity, they serve as a spatial basis.  Utilizing the Pauli matrix notation, these are written $\sigma_k = \gamma_k \gamma_0$.  Spatial vectors in STA are denoted in boldface, so for the spacetime split

\begin{align*}
\gamma_0 x = x^0 + x^k \sigma_k,
\end{align*}

one writes $\Bx = x^k \sigma_k$.  

\section{blah}
\begin{align*}
F &= \grad \wedge A = \BE + I c \BB \\
A &= \gamma_\mu A^\mu \\
J &= \gamma_\mu J^\mu \\
I &= \gamma_0 \gamma_1 \gamma_2 \gamma_3 \\
\end{align*}

\section{Maxwell's equation in Geometric Algebra.}

Like differential forms and tensor notations, Geometric Algebra allows Maxwell's equations to be united into a compact form

\begin{align*}
\grad F = J/\epsilon_0 c.
\end{align*}

\section{Basic notation.}
Where 
\begin{align*}
\grad &= \gamma^\mu \partial_\mu \\
F &= \grad \wedge A = \BE + I c \BB \\
A &= \gamma_\mu A^\mu \\
J &= \gamma_\mu J^\mu \\
I &= \gamma_0 \gamma_1 \gamma_2 \gamma_3 \\
\partial_0 &= \inv{c} \PD{t}{} \\
\partial_k &= \PD{x^k}{} \\
\gamma_\mu \cdot \gamma^\nu &= {\delta_\mu}^\nu 
\end{align*}

With a metric choice of $\gamma_0^2 = -\gamma_k^2 = 1$, the Geometric Algebra with basis $\{\gamma_\mu\}$ and its associated reciprocal basis $\{\gamma^\mu\}$ is known as the Space Time Algebra (STA).  Spatial vectors, designated with bold above, are bivectors in STA, as in the following example for the electric field

\begin{align*}
\BE &= E^k \sigma_k \\
\sigma^k = \sigma_k &= \gamma_k \gamma_0.
\end{align*}

Note that because of the mixed signature these bivectors $\sigma_k$ all square to unity, like a unit vector in a Euclidian vector space.

\section{Recovering the traditional vector form of Maxwell's equations.}

Recovery of the vector form of Maxwell's equations requires expansion of

\begin{align*}
\gamma_0 \left( \grad F - J/\epsilon_0 c \right) = 0 
\end{align*}

To do so first note that 
\begin{align*}
\gamma_0 \grad &= \partial_t + \spacegrad 
\end{align*}

where a bold font is used for the spatial gradient.

\begin{align*}
\spacegrad &= \sigma^k \partial_k
\end{align*}

Similarily, multiplication of the four vector current density also has scalar and spatial vector components.  With $\BJ = J^k \sigma_k$, and $\rho = J^0/c$ this is

\begin{align*}
\gamma_0 J &= c \rho - \BJ.
\end{align*}

One obtains
\begin{align*}
\left( \inv{c} \partial_t + \spacegrad \right) (\BE + I c \BB) - \frac{\rho}{\epsilon_0} + \frac{\BJ}{\epsilon_0 c} = 0.
\end{align*}

Noting that the pseudoscalar $I$ commutes with all spatial vectors, that $I^2 = -1$, $\epsilon_0 \mu_0 c^2 = 1$, and $\spacegrad \BX = \spacegrad \cdot \BX + I \spacegrad \cross \BX$ for spatial vectors $\BX$, one can expand and regroup this yielding

\begin{align*}
\left( \spacegrad \cdot \BE - \frac{\rho}{\epsilon_0} \right)
- 
c \left( \spacegrad \cross \BB - \mu_0 \epsilon_0 \PD{t}{\BE} - \mu_0 \BJ \right)
+ \left( \spacegrad \cross \BE + \PD{t}{\BB} \right)
+ I c \left( \spacegrad \cdot \BB \right)
= 0 
\end{align*}

In the spatial basis $\{\sigma_k\}$ we have scalar, vector, bivector, and trivector grades.  Equating each to zero recovers all of Maxwell's equations in their traditional vector form.

%\EndArticle
\EndNoBibArticle
