%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\input{../blogpost.tex}
\renewcommand{\basename}{modernOpticsLecture8}
\renewcommand{\dirname}{notes/phy485/}
\newcommand{\keywords}{Optics, PHY485H1F}
\input{../peeter_prologue_print2.tex}
\beginArtNoToc
\generatetitle{PHY485H1F Modern Optics.  Lecture 8: Coherence (cont.).  Taught by Prof.\ Joseph Thywissen}
%\chapter{Coherence (cont.)}
\label{chap:modernOpticsLecture8}

\section{Disclaimer}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{Zoology of interferometers}

\makedefinition{Coherence}{dfn:modernOpticsLecture8:1}{(Operational definition)  Something measured by an interferometer}

\paragraph{Types of dual path interferometers}

FIXME: F1-F5

\paragraph{Types of multi-path interferometers}

FIXME: F6-F8

\section{Lloyd's interferometer}

Using a virtual ray we can think of the Lloyd's interferometer setup as equivalent to a Young's double slit setup as illustrated in

FIXME: F9-F10

Consider two sources as in

FIXME: F11

Looking at this mathematically we have

\begin{dmath}\label{eqn:modernOpticsLecture8:n}
I 
= \expectation{ \Abs{\Psi}^2 }
= \expectation{ \Abs{\Psi(\Br_1, t) + \Psi(\Br_2, t) }^2 }
= 
I(\Br_1) + 
I(\Br_2) + 
2 \Real \expectation{ \Psi(\Br_1, t) \Psi^\conj(\Br_2, t) }
\end{dmath}

All the action is in the cross term.  The portion of this that is hard to calculate, we call the \underline{Mutual coherence}

\begin{dmath}\label{eqn:modernOpticsLecture8:n}
\Gamma_{12} \equiv
\expectation{ \Psi(\Br_1, t) \Psi^\conj(\Br_2, t) }
\end{dmath}

\section{Types of coherence}

\subsection{Longitudinal coherence}

Consider the measurement of the relative interference at two points as in 

FIXME: F12

where we have a device that measures the relative interference at these points as in 

FIXME: F13,F14

where we suppose that there's something that has introduced a small amount of delay or path length.  The extra pathlength like a time delay

\begin{dmath}\label{eqn:modernOpticsLecture8:n}
\tau = \frac{s_2 - s_1}{c}
\end{dmath}

With a \underline{coherence time} defined as

\begin{dmath}\label{eqn:modernOpticsLecture8:n}
\tau_{\text{coh}} = \inv{\Delta w}
\end{dmath}

where $\Delta w$ is the spectral width of the source.

We will show that if

\begin{dmath}\label{eqn:modernOpticsLecture8:n}
s_2 - s_1 \ll c \tau_{\text{coh}}
\end{dmath}

we have good visibility.

We want to think about what happens when the source gets broad as in 

FIXME: F16

\subsection{Transverse coherence}

As illustrated in

FIXME: F15

with 

\begin{dmath}\label{eqn:modernOpticsLecture8:n}
\text{length} = \frac{\lambda}{\Delta \theta_s}
\end{dmath}

we will show that we get a good fringe if

\begin{dmath}\label{eqn:modernOpticsLecture8:n}
x \ll 
\frac{\lambda}{\Delta \theta_s}
\end{dmath}

A \underline{point source} is one for which $\Delta \theta_s \rightarrow 0$, so that $\lambda/\Delta \theta_s \rightarrow \infty$.

We want to think about what happens when the source gets big.

\section{Questions}

\question{Doesn't the intensity loss in the $P_1,P_2$ linear interference setup matter?}

FIXME: F19

If $R$ is small, then the resulting intensities are similar.

\subsection{Mathematical treatment of the mutual coherence}

TO BE CONTINUED.

%\vcsinfo
%\EndArticle
\EndNoBibArticle
