\makeproblem{Fresnel Lens}{modernOptics:problemSet2:2}{ 

In this problem you will design a fresnel lens that focuses a plane wave onto a line (ie, a cylindrical lens). Design it for wavelength $\lambda$ and a focal length $f$. The on-axis Fresnel diffraction integral we found in class was 
$$ \Psi = \frac{\Psi_\infty }{2 i} \int\int du \, dv \, \exp{[i \frac{\pi}{2} (u^2 + v^2)]}, $$
where $u = x' \sqrt{2 (r_S^{-1} + r^{-1})/\lambda} $, and a similar expression applied for $v$ as a function of $y'$. The goal of the design is to maximize the intensity at the focus of the lens, located at $\Br = \langle 0, 0, f \rangle$. We'll call this amplitude $\Psi_f$ and this intensity $I_f$.

\makesubproblem{ 
Integrate out the $y'$ direction, so that we are only dealing with an integral of $u$. Give an expression for the contribution to $\Psi_f$ from an open segment from $u=a$ to $u=b$, in terms of the Fresnel Integrals $\mathcal{S}(u)$ and $\mathcal{C}(u)$.
}{modernOptics:problemSet2:2a}
\makesubproblem{ 
As a first step in the design, consider a Fresnel lens with only one opening, from $-u_0$ to $u_0$. Plot the resultant intensity $I_f/I_\infty$ versus $u_0$. {\em You will need numerical evaluation of the Fresnel integrals for parts \ref{modernOptics:problemSet2:2b} and \ref{modernOptics:problemSet2:2e} of this problem.}
}{modernOptics:problemSet2:2b}
\makesubproblem{ 
The next step is to find subsequent open regions  (I'll call these zones) of the mask that most increase $I_f$. These zones need to chosen to have a phase that matches the amplitude passing through through the central zone. {\bf Illustrate this principle with a drawing of the Cornu spiral}, shown what segments should be blocked.
}{modernOptics:problemSet2:2c}
\makesubproblem{ 
From (c), show that this criterion results in zone edges at $\pm u_m$, where $u_m = \sqrt{3/2 + 2 m}$, and $m=\{ 0, 1, 2, \ldots \}$. {\em Hint: recall that the angle of the spiral is $\beta = \frac{\pi}{2} u^2$. }
}{modernOptics:problemSet2:2d}
\makesubproblem{ 
Calculate the increase in intensity with three open zones and with five open zones.
}{modernOptics:problemSet2:2e}
\makesubproblem{ 
Give a pathlength interpretation of the expression for $u_m$.
}{modernOptics:problemSet2:2f}
\makesubproblem{ 
If this is an x-ray Fresnel lens, where $\lambda=10$\,nm and $f=10$\,cm, how big is the central opening? Take the plane wave limit $r_s \rightarrow \infty$. If we can only fabricate zones as small as 1$\mu$m, how many zones could we make?
}{modernOptics:problemSet2:2g}
}

\makeanswer{modernOptics:problemSet2:2}{ 
\makesubanswer{TODO.}{modernOptics:problemSet2:2a}
\makesubanswer{TODO.}{modernOptics:problemSet2:2b}
\makesubanswer{TODO.}{modernOptics:problemSet2:2c}
\makesubanswer{TODO.}{modernOptics:problemSet2:2d}
\makesubanswer{TODO.}{modernOptics:problemSet2:2e}
\makesubanswer{TODO.}{modernOptics:problemSet2:2f}
\makesubanswer{TODO.}{modernOptics:problemSet2:2g}
}
