\makeproblem{Inside the Fabry Perot}{modernOptics:problemSet3:2}{
What does the energy density $u(x) \equiv \langle | \Psi(x,t) |^2 \rangle$ look like {\em inside} the Fabry Perot etalon? Assume a monochromatic travelling wave $\Psi(x,t) = \sqrt{u_0} \exp{(i k x - i \omega t)}$ is normally incident on a cavity. The mirrors are two surfaces with with transmission $t = \sqrt{T} \exp{(i \delta_t)}$, reflecivity $r = \sqrt{R} \exp{(i \delta_r)}$, and $T=1-R$. The length of the cavity is $L$. As in class, the round-trip phase shift will be called $\Delta$.
\makesubproblem{
{\bf Show that} $u(x)$ in the cavity can be expressed in the form 

\begin{dmath}\label{eqn:modernOptics:problemSet3:2:20}
u = u_0 A \, \left(1 + R + 2 \sqrt{R} \cos [\phi(x)] \right), 
\end{dmath}

and {\bf give} $A$ and $\phi(x)$ in terms of $R$, $\Delta$, and $k$. This should explain why we say the $m$th order resonance is when there are $m$ standing-wave anti-nodes in the cavity.
}{modernOptics:problemSet3:2a}

\makesubproblem{Peak energy density}{modernOptics:problemSet3:2b}

Find an expression the {\bf peak energy density} $u_{\mathrm{max}}$ in the cavity, and {\bf plot} $u_{\mathrm{max}}/ u_0$ versus $\Delta$, for $R=0.8$. What is the {\bf resonance condition}? {\bf By how much} can $u_{\mathrm{max}}$ exceed $u_0$? Where is this extra energy density coming from if energy is conserved?

\makesubproblem{Standing waves?}{modernOptics:problemSet3:2c}

Is there {\bf always a standing wave} in the cavity, even off resonance? Write an expression for the {\bf visibility}, (max-min)/(max+min), of the energy density $u(x)$. One way to measure this visibility would be to look at the optical forces on an atom in the cavity.

\makesubproblem{Compare to class}{modernOptics:problemSet3:2d}

Finally, make sure your expression also makes sense when compared to the $I_T$ we found in class. Since it's only the forward-going component that is transmitted through the last mirror, the intensity outside the cavity is $u_0 A T$. (You don't need to prove this.) {\bf Does this reproduce $I_T$?}
} % makeproblem

\makeanswer{modernOptics:problemSet3:2}{
\makeSubAnswer{XXX}{modernOptics:problemSet3:2a}
\makeSubAnswer{XXX}{modernOptics:problemSet3:2b}
\makeSubAnswer{XXX}{modernOptics:problemSet3:2c}
\makeSubAnswer{XXX}{modernOptics:problemSet3:2d}
%\paragraph{Part \ref{modernOptics:problemSet3:2a}.  XXX}
} % makeanswer

