\makeproblem{\bf Spatial coherence and grating spectrometers}{modernOptics:problemSet3:3}{
A look inside a grating spectrometer reveals that incident light is passed through a series of slits to increase the transverse spatial coherence. In this problem, we'll try to understand why. For all of the parts below, consider a grating of $N$ slits, periodicity $a$, and width $b$. 

For parts \ref{modernOptics:problemSet3:3c} and \ref{modernOptics:problemSet3:3d}, neglect the envelope due to a finite slit width $b$, and consider only the sharp diffraction peaks.

\makesubproblem{Wavelength resolution}{modernOptics:problemSet3:3a}

For a collimated ($k_{s,y}=0$), monochromatic source illuminating $N$ slits, diffraction peaks would have an angular width of $\Delta \theta = \lambda / (N a \cos{\theta})$, for the first order of diffraction. {\bf re-derive this result} for yourself. Show that this gives a wavelength resolution is $\Delta \lambda = \lambda/N m$

\makesubproblem{Intensity}{modernOptics:problemSet3:3b}

Next, consider how the output intensity of the grating shifts if the input comes in at an angle $\theta_s$. {\bf Write an expression for $I(\theta_s, \theta)$.} You can also use the variables $k_{s,y}=k \sin \theta_s$ and $k_{y}=k \sin \theta$, as we did in class.

\makesubproblem{Resolution of spectrometer}{modernOptics:problemSet3:3c}
If the incident beam has an angular spread $\Delta \theta_s$ around normal incidence, what is the {\bf resolution $\Delta \lambda$ of the spectrometer?} Calculate this in the lmit of large $N$, or $N \gg \lambda / a \Delta \theta_s$, where the angular width of the diffracted light is comletely determined by the angular width of the incident light. 

\makesubproblem{Decreased coherence length at the grating}{modernOptics:problemSet3:3d}

An alternate view of part \ref{modernOptics:problemSet3:3c} is that by broadening the angular distribution of the source, we also decrease the transverse coherence length at the grating. The number of slits leading to coherent diffraction is reduced to some $N_{\mathrm{eff}}$, which is the number of slits within one coherence lenth $\ell_{\mathrm{tc}} = \lambda / \Delta \theta_s$. {\bf Sketch a diagram explaing this.} The frequency resolution of the spectrometer is then reduced from $\lambda/N$ to $\lambda/N_{\mathrm{eff}}$. (for order $m=1$) {\bf Compare this to the result you found in \ref{modernOptics:problemSet3:3c}.}

} % makeproblem

\makeanswer{modernOptics:problemSet3:3}{
\makeSubAnswer{XXX}{modernOptics:problemSet3:3a}
\makeSubAnswer{XXX}{modernOptics:problemSet3:3b}
\makeSubAnswer{XXX}{modernOptics:problemSet3:3c}
\makeSubAnswer{XXX}{modernOptics:problemSet3:3d}
%\paragraph{Part \ref{modernOptics:problemSet3:3a}.  XXX}
} % makeanswer

