%
% Copyright © 2014 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\input{../blogpost.tex}
\renewcommand{\basename}{multiphysicsL5}
\renewcommand{\dirname}{notes/ece1254/}
\newcommand{\keywords}{Condensed matter physics, ECE1254H}
\input{../peeter_prologue_print2.tex}

\usepackage{kbordermatrix}

\beginArtNoToc

\generatetitle{ECE1254H Modeling of Multiphysics Systems.  Lecture 5: Numerical error and conditioning.  Taught by Prof.\ Piero Triverio}
%\chapter{Numerical error and conditioning}
\label{chap:multiphysicsL5}

\section{Disclaimer}

Peeter's lecture notes from class.  These may be incoherent and rough.

\section{Numerical error and conditioning}

Related to a theorem on one of the slides:

\makedefinition{Strictly diagonally dominant}{dfn:multiphysicsL5:20}{

A matrix \( [ M_{ij} ] \) is strictly diagonally dominant if

\begin{dmath}\label{eqn:multiphysicsL5:40}
	\Abs{M_{ii}} > \sum_{j \ne i} \Abs{ M_{ij} } \forall i
\end{dmath}
}

For example, the stamp matrix

\begin{equation}\label{eqn:multiphysicsL5:60}
\kbordermatrix{
    & i      & j \\
i & \inv{R}  & -\inv{R} \\
j & -\inv{R} & \inv{R}
}
\end{equation}

is not strictly diagonally dominant.  For row $i$ this strict dominance can be achieved by adding a reference resistor

\begin{equation}\label{eqn:multiphysicsL5:80}
\kbordermatrix{
    & i      & j \\
i & \inv{R_0} + \inv{R}  & -\inv{R} \\
j & -\inv{R} & \inv{R}
}
\end{equation}

However, even with strict dominance, we can have trouble with ill posed (perturbative) systems.

Round off error examples with double precision 

\begin{dmath}\label{eqn:multiphysicsL5:100}
\lr{ 1 - 1 } + \pi 10^{-17} = \pi 10^{-17},
\end{dmath}

vs.

\begin{dmath}\label{eqn:multiphysicsL5:120}
\lr{ 1 + \pi 10^{-17} } -1 = 0.
\end{dmath}

This is demonstrated by

\begin{verbatim}
#include <stdio.h>
#include <math.h>

// produces:
// 0 3.14159e-17
int main()
{
   double d1 = (1 + M_PI * 1e-17) - 1 ;
   double d2 = M_PI * 1e-17 ;

   printf( "%g %g\n", d1, d2 ) ;

   return 0 ;
}
\end{verbatim}

Note that a union and bitfield \citep{peeterjRoundingCpp} can be useful for exploring double precision representation.

\section{Exploring uniqueness and existence}

For a matrix system \( \overbar{M} x = \overbar{b} \) in column format, with

\begin{dmath}\label{eqn:multiphysicsL5:140}
\begin{bmatrix}
\overbar{M}_1 & \overbar{M}_2 & \cdots & \overbar{M}_N 
\end{bmatrix}
\begin{bmatrix}
	x_1 \\
	x_2 \\
	\vdots \\
	x_N
\end{bmatrix}
= \overbar{b}.
\end{dmath}

This can be written as

\begin{dmath}\label{eqn:multiphysicsL5:160}
\mathLabelBox
{
	x_1 
}
{weight}
	\overbar{M}_1 + x_2 \overbar{M}_2 + \cdots x_N \overbar{M}_N = \overbar{b}.
\end{dmath}

Linear dependence means

\begin{dmath}\label{eqn:multiphysicsL5:180}
y_1 \overbar{M}_1 + y_2 \overbar{M}_2 + \cdots y_N \overbar{M}_N = 0,
\end{dmath}

or \( M \overbar{y} = 0 \).

With a linear dependency an additional solution, given solution \( \overbar{x} \) is \( \overbar{x}^1 = \overbar{x} + \alpha y \).  This becomes relevant for numerical processing since for a system 

\begin{dmath}\label{eqn:multiphysicsL5:220}
M \overbar{x}^1 = \overbar{b}
\end{dmath}

we can often find 

\begin{dmath}\label{eqn:multiphysicsL5:200}
M \overbar{x} + \alpha M \overbar{y} = \overbar{b},
\end{dmath}

where \( \alpha M \overbar{y} \) is of order \( 10^{-20} \).

\section{Perturbation and norms}

Consider a perturbation to the system \( M \overbar{x} = \overbar{b} \)

\begin{dmath}\label{eqn:multiphysicsL5:240}
	\lr{ M + \delta M } \lr{ \overbar{x} + \delta \overbar{x} } = \overbar{b}.
\end{dmath}

Some vector norms

\begin{itemize}
\item \(L_1\) norm

\begin{dmath}\label{eqn:multiphysicsL5:260}
\Norm{ \overbar{x} }_1 = \sum_i \Abs{ x_i }
\end{dmath}

\item \(L_2\) norm

\begin{dmath}\label{eqn:multiphysicsL5:280}
\Norm{ \overbar{x} }_2 = \sqrt{ \sum_i \Abs{ x_i }^2 }
\end{dmath}

\item \(L_\infty\) norm

\begin{dmath}\label{eqn:multiphysicsL5:300}
\Norm{ \overbar{x} }_\infty = \max_i \Abs{ x_i }.
\end{dmath}
\end{itemize}

These are illustrated  for \( \overbar{x} = (x_1, x_2) \) in

F2

\section{matrix norm}

For

\begin{dmath}\label{eqn:multiphysicsL5:320}
\overbar{y} = M \overbar{x}
\end{dmath}

F3

The 1-norm is

\begin{dmath}\label{eqn:multiphysicsL5:340}
\Norm{ M } = \max_{ \Norm{\overbar{x}} = 1 } \Norm{ M \overbar{x} },
\end{dmath}

and the 2-norm is

\begin{dmath}\label{eqn:multiphysicsL5:360}
\Norm{ M }_2 = \max_{ \Norm{\overbar{x}}_2 = 1 } \Norm{ M \overbar{x} }_2.
\end{dmath}

\EndArticle
%\EndNoBibArticle
