\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Laplacian of Pressure and Vorticity.}
\label{chap:nsVorticity}
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/nsVorticity.pdf}
\date{Mar 16, 2012}
\gitRevisionInfo{nsVorticity}

%\beginArtWithToc
\beginArtNoToc

\section{Motivation.}

Supposing that we are neglecting the non-linear term of the NS equation.  For incompressible steady state flow, we have

\begin{subequations}
\begin{equation}\label{eqn:qmTwoExamReflection:5}
0 = -\spacegrad p + \mu \spacegrad^2 \Bu 
\end{equation}
\begin{equation}\label{eqn:qmTwoExamReflection:n}
0 = \spacegrad \cdot \Bu
\end{equation}
\end{subequations}

How do we actually solve this beastie?

\subsection{Separation of variables?}

Considering this in 2D, assuming no z-dependence, with $\Bu = \Bu(x, y) = (u, v)$ we have

\begin{align}\label{eqn:nsVorticity:30}
0 &= -\partial_x p + \mu (\partial_{xx} + \partial_{yy} )u \\
0 &= -\partial_y p + \mu (\partial_{xx} + \partial_{yy} )v \\
0 &= \partial_x u + \partial_y v.
\end{align}

Attempting separation of variables seems like something reasonable to try.  With

\begin{align}\label{eqn:nsVorticity:50}
u &= X(x) Y(y) \\
v &= R(x) S(y) \\
p &= P(x) Q(y)
\end{align}

we get

\begin{align}\label{eqn:nsVorticity:70}
0 &= -P' Q + \mu (X'' Y + X Y'') \\
0 &= -P Q' + \mu (R'' S + R S'') \\
0 &= X' Y + R S'
\end{align}

I couldn't find a way to substitute any of these into the other that would allow me to separate them, but perhaps I wasn't clever enough.  

\subsection{In terms of vorticity?}

The idea of substuting the zero divergence equation $\spacegrad \cdot \Bu$ will clearly lead to something a bit simpler.  Treating the Laplacian as a geometric (clifford) product of two gradients we have

\begin{align*}
0 
&= -\spacegrad p + \mu \spacegrad^2 \Bu \\
&= -\spacegrad p + \mu \spacegrad ( \spacegrad \Bu ) \\
&= -\spacegrad p + \mu \spacegrad ( \cancel{\spacegrad \cdot \Bu} + \spacegrad \wedge \Bu ) \\
&= -\spacegrad p + \mu \spacegrad ( \spacegrad \wedge \Bu ) \\
&= \spacegrad \left( -p + \mu \spacegrad \wedge \Bu \right)
\end{align*}

Writing out the vorticity (bivector) in component form, and writing $i = \Be_1 \wedge \Be_2 = \Be_1 \Be_2$ for the 2D pseudoscalar, we have

\begin{align*}
\spacegrad \wedge \Bu 
&= (\Be_1 \partial_x + \Be_2 \partial_y) \wedge (\Be_1 u + \Be_2 v) \\
&= \Be_1 \Be_2 (\partial_x v - \partial_y u ) \\
&= i (\partial_x v - \partial_y u )
\end{align*}

It seems natural to write

\begin{equation}\label{eqn:qmTwoExamReflection:90}
\Theta = \partial_x v - \partial_y u,
\end{equation}

so that Navier-Stokes takes the form

\begin{equation}\label{eqn:qmTwoExamReflection:110}
0 = \spacegrad (-p + i \mu \Theta ).
\end{equation}

Operating on this from the left with another gradient we find that this combination of pressure and vorticity must satisfy the following multivector Laplacian equation

\begin{equation}\label{eqn:qmTwoExamReflection:130}
0 = \spacegrad^2 (-p + i \mu \Theta ).
\end{equation}

However, note that $\spacegrad^2$ is a scalar operator, and this zero identity has both scalar and pure bivector components.  Both must separately equal zero

\begin{subequations}
\label{eqn:qmTwoExamReflection:150}
\begin{equation}\label{eqn:qmTwoExamReflection:170}
0 = \spacegrad^2 p 
\end{equation}
\begin{equation}\label{eqn:qmTwoExamReflection:190}
0 = \spacegrad^2 \Theta.
\end{equation}
\end{subequations}

So, it seems like a reasonable way to tackle a problem of this form may be to take a general solution to the Laplacian equation (i.e. linear combinations of harmonmic functions), and then fit those to our specific boundary value constraints.

Note that we can obtain \ref{eqn:qmTwoExamReflection:150} much more directly, if we know that's what we want to do.  Just operate on \ref{eqn:qmTwoExamReflection:5} with the gradient from the left right off the bat.  We find

\begin{align*}
0 
&= -\spacegrad^2 p + \mu \spacegrad^3 \Bu \\
&= -\spacegrad^2 p + \mu \spacegrad^2 (\spacegrad \Bu) \\
&= -\spacegrad^2 p + \mu \spacegrad^2 (\spacegrad \wedge \Bu) \\
\end{align*}

Again, we have a multivector equation scalar and bivector parts, that must separately equal zero.  With the magnitude of the vorticity $\Theta$  given by \ref{eqn:qmTwoExamReflection:90}, we once again obtain \ref{eqn:qmTwoExamReflection:150}.  This can be done in plain old vector algebra as well by operating on the left not by the gradient, but separately with a divergence and curl operator.

\section{Now what?}



\EndArticle
%\EndNoBibArticle
