\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Degeneracy and diagonalization}
\label{chap:pertubationDegeneracy}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/pertubationDegeneracy.pdf}
\date{Nov 7, 2011}
\revisionInfo{pertubationDegeneracy.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

In class it was mentioned that to deal with perturbation around a degenerate energy eigenvalue, we needed to diagonalize the perturbing Hamiltonian.  I didn't follow those arguments completely, and I'd like to revisit those here.

\section{A four state Hamiltonian.}

Problem set 3, problem 1, was to calculate the energy eigenvalues for the following Hamiltonian

\begin{align}\label{eqn:pertubationDegeneracy:10}
H &= H_0 + \lambda H' \\
H_0 &=
\begin{bmatrix}
a & 0 & 0 & 0 \\
0 & b & 0 & 0 \\
0 & 0 & c & 0 \\
0 & 0 & 0 & c 
\end{bmatrix} \\
H' &=
\begin{bmatrix}
\alpha 		& 0 		& \nu 		& \eta \\
0 		& \beta 	& 0 		& \mu \\
\nu^\conj 	& 0 		& \gamma 	& 0 \\
\eta^\conj 	& \mu^\conj 	& 0 		& \delta
\end{bmatrix}
\end{align}

This is more complicated that the two state problem that are solved exactly in \S 13.1.1 in the text \cite{desai2009quantum}, but differs from the (possibly) infinite dimensional problem that was covered in class.  Unfortunately, the solution provided to this problem didn't provide the illumination I expected, so let's do it again, calculating the perturbed energy eigenvalues for the degenerate levels, from scratch.

Can we follow the approach used in the text for the two (only) state problem.  For the two state problem, it was assumed that the perturbed solution could be expressed as a superposition of the two states that formed the basis for the unperturbed Hilbert space.  That is

\begin{equation}\label{eqn:pertubationDegeneracy:30}
\ket{\psi} = m \ket{1} + n \ket{2}
\end{equation}

For the two state problem, assuming that the perturbed energy eigenvalue is $E$, and the unperturbed energy eigenvalue is $E^0$ we find

\begin{align*}
0 
&= (H - E) \ket{\psi} \\
&= (H_0 + \lambda H') \ket{\psi} - E \ket{\psi} \\
&= 
(H_0 + \lambda H') 
(m \ket{1} + n \ket{2})
-E 
(m \ket{1} + n \ket{2})
\\
&=
\lambda H' (m \ket{1} + n \ket{2}) 
-E^0 
(m \ket{1} + n \ket{2}) \\
&=
(-E^0 + \lambda H')
\begin{bmatrix}
\ket{1} & \ket{2} 
\end{bmatrix}
\begin{bmatrix}
m \\
n
\end{bmatrix}
\end{align*}

Left multiplying by the brakets we find

\begin{align*}
0 &= 
\begin{bmatrix}
\bra{1} \\
\bra{2} 
\end{bmatrix}
(H - E)\ket{\psi} \\
&=
\left(
(E^0 - E)I + \lambda
\begin{bmatrix}
\bra{1} H' \ket{1} & \bra{1} H' \ket{2} \\
\bra{2} H' \ket{1} & \bra{2} H' \ket{2} 
\end{bmatrix}
\right)
\begin{bmatrix}
m \\
n
\end{bmatrix}
\end{align*}

Or

\begin{equation}\label{eqn:pertubationDegeneracy:50}
\left( (E^0 - E)I + \lambda \begin{bmatrix} H_{ij}' \end{bmatrix} \right)
\begin{bmatrix}
m \\
n
\end{bmatrix}
= 0.
\end{equation}

Observe that there was no assumption about the dimensionality of $H_0$ and $H'$ here, just that the two degenerate energy levels had eigenvalues $E^0$ and a pair of eigenkets $\ket{1}$ and $\ket{2}$ such that $H_0 \ket{i} = E^0 \ket{i}, i \in [1,2]$.  It's clear that we can use a similar argument for any degeneracy degree.  It's also clear how to proceed, since we have what almost amounts to a characteristic equation for the degenerate subspace of Hilbert space for the problem.

Because $H'$ is Hermitian, a diagonalization

\begin{equation}\label{eqn:pertubationDegeneracy:70}
H' = U^\conj D U
\end{equation}

can be found.  To solve for $E$ we can take the determinant of the matrix factor of \ref{eqn:pertubationDegeneracy:50}, and because $I = U^\conj U$ we have

\begin{align*}
0 &=
\begin{vmatrix}
(E^0 -E) U^\conj I U + \lambda U^\conj D U 
\end{vmatrix} \\
&=
\begin{vmatrix}
U^\conj
\end{vmatrix}
\begin{vmatrix}
(E^0 -E)I + \lambda D 
\end{vmatrix}
\begin{vmatrix}
U
\end{vmatrix} \\
&=
\begin{vmatrix}
E^0 -E + \lambda d_1 & 0 \\
0 & E^0 -E + \lambda d_2
\end{vmatrix} \\
&=
(E^0 -E + \lambda d_1)(E^0 -E + \lambda d_2)
\end{align*}

So our energy eigenvalues associated with the perturbed state are (exactly) 

\begin{equation}\label{eqn:pertubationDegeneracy:n}
E = E^0 + \lambda d_1, E^0 + \lambda d_2.
\end{equation}

It's a bit curious seeming that only the energy eigenvalues associated with the degeneracy play any part in this result, but there is some intuitive comfort in this idea.  Without the perturbation, we can't do an energy measurement that would distinguish one or the other of the eigenkets for the degenerate energy level, so it doesn't seem unreasonable that a perturbed energy level close to the original can be formed by superposition of these two states, and thus the perturbed energy eigenvalue for the new system would then be related to only those degenerate levels.

Observe that in the problem set three problem we had a diagonal initial Hamiltonian $H_0$, that doesn't have an impact on the argument above, since that portion of the Hamiltonian only has a diagonal contribution to the result found in \ref{eqn:pertubationDegeneracy:50}, since the identity $H_0 \ket{i} = c \ket{i}, i \in [3,4]$ removes any requirement to know the specifics of that portion of the matrix element of $H_0$.

\section{Repeating the perturbation argument with explicit diagonalization.}

\EndArticle
