\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Phase Space and Trajectories}
\label{chap:phaseSpaceAndTrajectories}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/phaseSpaceAndTrajectories.pdf}
\date{Feb 29, 2012}
\gitRevisionInfo{phaseSpaceAndTrajectories}

\beginArtWithToc
%\beginArtNoToc

\section{Phase space and phase trajectories.}

The phase space and phase trajectories are the space of $p$'s and $q$'s of a mechanical system (always even dimensional, with as many $p$'s as $q$'s for N particles in 3d: 6N dimensional space).

The state of a mechanical system $\equiv$ the point in phase space.
Time evolution $\equiv$ a curve in phase space.

Example: 1 dim system, say a harmmonic oscillator.

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
H = \frac{p^2}{2m} + \frac{1}{2} m \omega^2 q^2
\end{equation}

Our phase space can be illustrated as an ellipse as in

FIXME: F1

where the phase space trajectories of the SHO.  The equation describing the ellipse is

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
E = \frac{p^2}{2m} + \frac{1}{2} m \omega^2 q^2,
\end{equation}

which we can put into standard elliptical form as

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
1 = \left( \frac{p}{\sqrt{2 m E}}\right)^2 + \left(\sqrt{\frac{m}{2 E}} \omega\right) q^2
\end{equation}

\subsection{Applications of $H$.}

\begin{itemize}
\item Classical stat mech.
\item transition into QM -> via Poisson brackets.
\item mathematical theorems about phase space ``flow''.
\item pertubation theory.
\end{itemize}

\subsection{Poisson brackets.}

Poisson brackets arisse very naturally if one asks about the time evolution of a function $f(p, q, t)$ on phase space.

\begin{align*}
\ddt{} f(p_i, q_i, t)
&=
\sum_i
 \PD{p_i}{f} \PD{t}{p_i}
+ \PD{q_i}{f} \PD{t}{q_i}
+ \PD{t}{f}
&=
\sum_i
- \PD{p_i}{f} \PD{q_i}{H}
+ \PD{q_i}{f} \PD{p_i}{H}
+ \PD{t}{f}
\end{align*}

Define the commutator of $H$ and $f$ as

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\antisymmetric{H}{f} =
\sum_i
\PD{p_i}{H}
\PD{q_i}{f}
-
\PD{q_i}{H}
\PD{p_i}{f}
\end{equation}

This is the Poisson bracket of $H(p,q,t)$ with $f(p,q,t)$, defined for arbitrary functions on phase space.

Note that other conventions for sign exist (apparently in Landau and Lifshitz uses the opposite).

So we have

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\ddt{} f(p_i, q_i, t) = \antisymmetric{H}{f} + \PD{t}{f}.
\end{equation}

Corrolaries:

If $f$ has no explicit time dependence $\PDi{t}{f} = 0$ and if $\antisymmetric{H}{f} = 0$, then $f$ is an integral of motion.

In QM conserved quantities are the ones that commute with the Hamiltonian operator.

To see the analogy better, recall def of PB

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\antisymmetric{f}{g} =
\sum_i
\PD{p_i}{f}
\PD{q_i}{g}
-
\PD{q_i}{f}
\PD{p_i}{g}
\end{equation}

Properties of BP

\begin{itemize}
\item antisymmetric

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\antisymmetric{f}{g} = -\antisymmetric{g}{f}.
\end{equation}

\item linear

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\antisymmetric{a f + b h}{g} =
a \antisymmetric{f}{g}
+ b\antisymmetric{h}{g}.
\end{equation}
\end{itemize}

\subsubsection{Example.  Compute $p$, $q$ commutators.}

\begin{align*}
\antisymmetric{p_i}{p_j} 
&= 
\sum_k
\PD{p_k}{p_i}
\cancel{
\PD{q_k}{p_j}
}
-
\cancel{
\PD{q_k}{p_i}
}
\PD{p_k}{p_j} \\
&= 0
\end{align*}

So 

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\antisymmetric{p_i}{p_j} = 0
\end{equation}

Similarily $\antisymmetric{q_i}{q_j} = 0$.

How about

\begin{align*}
\antisymmetric{q_i}{p_j} 
&= 
\sum_k
\cancel{
\PD{p_k}{q_i}
}
\cancel{
\PD{q_k}{p_j}
}
-
\PD{q_k}{q_i}
\PD{p_k}{p_j} \\
&= 
-
\sum_k
\delta_{ik}
\delta_{jk}
&=
-\delta_{ij}
\end{align*}

So 

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\antisymmetric{q_i}{p_j} = -\delta_{ij}.
\end{equation}

This provides a systematic (axiomatic) way to ``quantize'' a classical mechanics system, where we make replacements

\begin{align}\label{eqn:phaseSpaceAndTrajectories:n}
q_i &\rightarrow \hat{q}_i \\
p_i &\rightarrow \hat{p}_i
\end{align}

and

\begin{align}\label{eqn:phaseSpaceAndTrajectories:n}
\antisymmetric{q_i}{p_j} &= -\delta_{ij} \rightarrow i \hbar \delta_{ij} \\
H(p, q, t) &\rightarrow \hat{H}(\hat{p}, \hat{q}, t)
\end{align}

So 

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\frac{\antisymmetric{\hat{q}_i}{\hat{p}_j}}{-i \hbar } = - \delta_{ij}
\end{equation}

Our quantization of time evolution is therefore

\begin{align}\label{eqn:phaseSpaceAndTrajectories:n}
\ddt{} \hat{q}_i &= \inv{-i\hbar} \antisymmetric{\hat{H}}{\hat{q}_i} \\
\ddt{} \hat{p}_i &= \inv{-i\hbar} \antisymmetric{\hat{H}}{\hat{p}_i}.
\end{align}

These are the Heisenberg equations of motion in QM.

\subsubsection{Conserved quantities.}

For conserved quantities $f$, functions of $p$'s $q$'s, we have

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\antisymmetric{f}{H} = 0
\end{equation}

Considering the components $M_i$, where 

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\BM = \Br \cross \Bp,
\end{equation}

We can show (FIXME: do it) that our poisson brackets obey

\begin{align}\label{eqn:phaseSpaceAndTrajectories:n}
\antisymmetric{M_x}{M_y} &= -M_z \\
\antisymmetric{M_y}{M_z} &= -M_x \\
\antisymmetric{M_z}{M_x} &= -M_y
\end{align}

FIXME: check this, because the sign could be off.

These are the analogue of the momentum commutator relationships from QM right here in classical mechanics.

Considering the symmetries that lead to this conservation relationship, it is actually possible to show that rotations in 4D space lead to these symmetries and the conservation of the Runge-Lenz vector.

\section{}

In 

FIXME: F2

where we have

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
T = \frac{2 \pi}{\omega(t)} = \sqrt{\frac{l(t)}{g}}.
\end{equation}

Imagine that we change the length $l(t)$ very \underline{slowly} so that

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
T \frac{\dot{l}}{l} \ll 1
\end{equation}

where $T$ is the period of oscillation.  This is what's called an adiabatic change, where the change of $\omega$ is small over a period.

It turns out that if this rate of change is slow, then there is actually an invariant, and

\begin{equation}\label{eqn:phaseSpaceAndTrajectories:n}
\frac{E}{\omega},
\end{equation}

is the so-called ``adiabatic invariant''.  There's an important application to this (and some relations to QM).  Imagine that we have a particle bounded by two walls, where the walls are moved very slowly as in

FIXME: F3.

This can be used to derive the adiabatic equation for an ideal gas (also using the equipartition theorem).

%\EndArticle
\EndNoBibArticle
