\input{../peeter_prologue.tex}

\chapter{Polar form for the gradient and Laplacian.}
\label{chap:polarGradAndLaplacian}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2009/polarGradAndLaplacian.pdf}
\date{Dec 1, 2009}
\revisionInfo{$RCSfile: polarGradAndLaplacian.tex,v $ Last $Revision: 1.1 $ $Date: 2009/12/02 05:32:53 $}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

The text \cite{byron1992mca} has an excellent general derivation of a number of forms of the gradient, divergence, curl and Laplacian.

This is actually done, not starting with the usual cartesian forms, but more general definitions.

\begin{align}\label{eqn:divGradCurLap:1}
(\text{grad}\  \phi)_i &= \lim_{ds_i \rightarrow 0} \frac{\phi(q_i + dq_i) - \phi(q_i)}{ds_i} \\
\text{div}\  \BV &= \lim_{\Delta \tau \rightarrow 0} \inv{\Delta \tau} \int_\sigma \BV \cdot d\Bsigma \\
(\text{curl}\  \BV) \cdot \Bn &= \lim_{\Delta \sigma \rightarrow 0} \inv{\Delta \sigma} \oint_\lambda \BV \cdot d\Blambda \\
\text{Laplacian}\  \phi &= \text{div} (\text{grad} \phi).
\end{align}

These are then shown to imply the usual Cartesian definitions, plus provide the means to calculate the general relationships in whatever coordinate system you like.  All in all one can't beat this approach, and I'm not going to try to replicate it, because I can't improve it in any way by doing so.

Given that, what do I have to say on this topic?  Way way back in first year electricity and magnetism, my dictator of a prof, the intimidating but diminutive Dmitrevsky, yelled at us repeatedly that one cannot just dot the gradient to form the Laplacian.  As far as he was concerned one can only say

\begin{align}\label{eqn:divGradCurLap:2}
\text{Laplacian}\  \phi &= \text{div} (\text{grad} \phi),
\end{align}

and never never never, 

\begin{align}\label{eqn:divGradCurLap:3}
\text{Laplacian}\  \phi &= (\spacegrad \cdot \spacegrad) \phi.
\end{align}

Because ``this only works in cartesian coordinates''.  This is something that has bugged me ever since, because my assumption was that, provided one was careful, why would a change to an alternate coordinate system matter.  The gradient is still the gradient, so it seems to me that this ought to be a general way to calculate things.

Here we explore the validity of the dictatorial comments of Prof Dmitrevsky.  The key to reconsiling intuition and his statement turns out to lie with the fact that one has to let the gradient operate on the unit vectors in the non cartesian representation as well as the partials, something that wasn't clear as a first year student.  Provided that this is done, the plain old dot product procedure yields the expected results.

This exploration will utilize a two dimensional space, transforming from cartesian to polar form representation.  I'll also utilize a geometric algebra representation of the polar unit vectors since it is so convienent.

\section{The gradient in polar form.}

\EndArticle
%\EndNoBibArticle
