\documentclass[]{eliblog}

\usepackage{color}

\input{../peeters_macros.tex}


\author{Peeter Joot}
\email{peeter.joot@utoronto.ca, 920798560}
%\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY356 Problem Set II.}
\label{chap:qmIproblemSet1}
%\useCCL
%\blogpage{http://sites.google.com/site/peeterjoot/math2010/qmIproblemSet1.pdf}
%\date{Oct X, 2010}
%\revisionInfo{qmIproblemSet1.tex}

\beginArtNoToc
\section{Problem 1.}

A particle of mass $m$ is free to move along the x-direction such that $V(X)=0$. Express the time evolution operator $U(t,t_0)$ defined by Eq. (2.166) using the momentum eigenstates $\ket{p}$ with delta-function normalization. Find $\bra{x} U(t,t0) \ket{x'}$,  where $\ket{x}$ and $\ket{x'}$ are position eigenstates.  What is the physical meaning of this expression?

\section{Solution.}

We can expand the time evolution operator in series

\begin{align*}
U(t,t_0) 
&= e^{-i H(t-t_0)/\hbar} \\
&= e^{ -i P^2 (t-t_0)/ 2m \hbar } \\
&= 1 + \sum_{k=1}^\infty \inv{k!} \left( -i \frac{P^2 (t-t_0)}{2m \hbar} \right)^k.
\end{align*}

We can now evaluate the momentum matrix element $\bra{p} U(t,t_0) \ket{p'}$, which will essentially require the value of $\bra{p} P^{2k} \ket{p'}$.  That is

\begin{align*}
\bra{p} P^{2k} \ket{p'}
&= \bra{p} P^{2k-1} P \ket{p'} \\
&= \bra{p} P^{2k-1} \ket{p'} p' \\
&= \cdots \\
&= \braket{p}{p'} (p')^{2k}.
\end{align*}

The momentum matrix element is therefore reduced to

\begin{align}\label{eqn:qmIproblemSet2:1}
\bra{p} U(t,t_0) \ket{p'}
&=
\braket{p}{p'} \exp\left( -i \frac{p^2 (t-t_0)}{2m \hbar} \right)
= \delta(p-p') \exp\left( -i \frac{p^2 (t-t_0)}{2m \hbar} \right)
\end{align}

For the position matrix element we have a similar sum
\begin{align*}
\bra{x} U(t,t_0) \ket{x'} 
&= 
\braket{x}{x'} 
+ \sum_{k=1}^\infty \inv{k!} \bra{x} \left( -i \frac{P^2 (t-t_0)}{2m \hbar} \right)^k \ket{x'},
\end{align*}

and require $\bra{x} P^{2k} \ket{x'}$ to continue.  That is

\begin{align*}
\bra{x} P^{2k} \ket{x'}
&=
\int dx''\bra{x} P^{2k-1} \ket{x''}\bra{x''} P \ket{x'} \\
&=
\int dx''\bra{x} P^{2k-1} \ket{x''} \delta(x''-x') (-i\hbar) \frac{d}{dx'} \\
&=
\bra{x} P^{2k-1} \ket{x'} (-i\hbar) \frac{d}{dx'} \\
&= \cdots \\
&= \braket{x}{x'} \left( (-i\hbar) \frac{d}{dx'} \right)^{2k}
\end{align*}

Our position matrix element is therefore the differential operator

\begin{align}\label{eqn:qmIproblemSet2:10}
\bra{x} U(t,t_0) \ket{x'} 
&=
\braket{x}{x'} \left( \frac{i (t-t_0)\hbar}{2m} \frac{d^2}{d{x'}^2} \right)
=\delta(x-x') \left( \frac{i (t-t_0)\hbar}{2m} \frac{d^2}{d{x'}^2} \right)
\end{align}

Finally, we need to determine the physical meaning of such a matrix element operator.  

With the delta function that this matrix element operator includes it really only takes on a meaning with a convolution integral.  One such integral would be

\begin{align*}
\int dx' \bra{x} U \ket{x'} \braket{x'}{\phi_0} 
&=
\bra{x} U \ket{\phi_0} \\
&=
\braket{x}{\phi(t)} \\
&=
\phi(x,t)
\end{align*}

Consider two additional contexts where this matrix element arises.  We can find this matrix element by expanding a normalized inner product for the time evolved state
\begin{align*}
1 
&= \braket{\phi(t)}{\phi(t)} \\
&= \bra{\phi(t)} U \ket{\phi_0} \\
&= \iint dx dx' \bra{\phi(t)} \ket{x} \bra{x} U \ket{x'}\bra{x'} \ket{\phi_0} \\
&= \iint dx dx' \phi^\conj(x,t) \bra{x} U \ket{x'} \phi(x,0).
\end{align*}

Here we see the matrix element as an operator weighting factor in the probability density relating the initial and final time states, but also relating two positions.  

This matrix element also arises in the expectation value of the time evolution operator with respect to the initial time state.  That is

\begin{align*}
\expectation{U} 
&= \bra{\phi_0} U \ket{\phi_0} \\
&= \braket{\phi_0}{\phi} = \int dx \phi^\conj(x,0) \phi(x,t) \\
&= \iint dx dx' \bra{\phi_0} \ket{x} \bra{x} U \ket{x'}\bra{x'} \ket{\phi_0} \\
&= \iint dx dx' \phi^\conj(x,0) \bra{x} U \ket{x'} \phi(x,0)
\end{align*}

Again we have something akin to a probability density or weighting factor relating the initial and final time states.

Based on this, I would be inclined to state that the position matrix element of the time evolution operator $\bra{x} U \ket{x'}$ represents something akin to an operator form of probability amplitude for a particle to travel between two points.

%\EndArticle
\EndNoBibArticle
