\documentclass[]{eliblog}

\usepackage{color}
%\usepackage{txfonts} % for xi

\input{../peeters_macros.tex}

\author{Peeter Joot}
\email{peeter.joot@utoronto.ca, 920798560}
%\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY356 Problem Set 5.}
\label{chap:qmIproblemSet5}
%\useCCL
%\blogpage{http://sites.google.com/site/peeterjoot/math2010/qmIproblemSet5.pdf}
\date{Nov 25, 2010}
\revisionInfo{qmIproblemSet5.tex}

\beginArtNoToc
\section{Problem.}
\subsection{Statement}

A particle of mass m moves along the x-direction such that $V(X)=\inv{2}KX^2$. Is the state $u(\xi) = B \xi e^{+\xi^2/2}$, where $\xi$ is given by Eq. (9.60), $B$ is a constant, and time $t=0$, an energy eigenstate of the system?  What is probability per unit length for measuring the particle at position $x=0$ at $t=t_0>0$?  Explain the physical meaning of the above results.

\subsection{Solution}
\subsubsection{Is this state an energy eigenstate?}

Recall that $\xi = \alpha x$, $\alpha = \sqrt{m\omega/\hbar}$, and $K = m \omega^2$.  With this variable substitution Schr\"{o}dinger's equation for this harmonic oscillator potential takes the form

\begin{equation}\label{eqn:qmIproblemSet5:10}
\frac{d^2 u}{d\xi^2} - \xi^2 u = \frac{2 E }{\hbar\omega} u
\end{equation}

While we can blindly substitute a function of the form $\xi e^{\xi^2/2}$ into this to get

\begin{align*}
\frac{d^2 u}{d\xi^2} - \xi^2 u 
&= 
\frac{d u}{d\xi} \left( 1 + \xi^2 \right) e^{\xi^2/2} - \xi^3 e^{\xi^2/2} \\
&= 
\left( 2 \xi + \xi + \xi^3 \right) e^{\xi^2/2} - \xi^3 e^{\xi^2/2} \\
&= 
3 \xi e^{\xi^2/2} 
\end{align*}

and formally make the identification $E = 3 \omega \hbar/2 = (1 + 1/2) \omega \hbar$, this isn't a normalizable wavefunction, and has no physical relevance, unless we set $B = 0$.

By changing the problem, this state could be physically relavent.  We'd require a potential of the form

\begin{equation}\label{eqn:qmIproblemSet5:11}
V(x) =
\left\{
\begin{array}{l l}
V_l & \quad \mbox{if $x < a$} \\
\inv{2} K x^2 & \quad \mbox{if $a < x < b$} \\
V_r & \quad \mbox{if $x > b$} \\
\end{array}
\right.
\end{equation}

where $V_l$ and $V_r$ are constants.  For such a potential, within the harmonic well, the general solution would be a superposition of states

\begin{equation}\label{eqn:qmIproblemSet5:19}
u(x,t) = \sum_n H_n(\xi) \Bigl(A_n e^{-\xi^2/2} + B_n e^{\xi^2/2} \Bigr) e^{-i E_n t/\hbar}.
\end{equation}

Normalization would not prohibit non-zero $B_n$ values in that situation, and the problem becomes one of matching the boundary value constraints.

\subsubsection{Probability per unit length at $x=0$.}

Suppose we form the wavefunction for a superposition of all the normalizable states

\begin{equation}\label{eqn:qmIproblemSet5:20}
u(x,t) = \sum_n A_n H_n(\xi) e^{-\xi^2/2} e^{-i E_n t/\hbar}
\end{equation}

Here it is assumed that the $A_n$ coefficients yield unit probability

\begin{equation}\label{eqn:qmIproblemSet5:30}
\int \Abs{u(x,0)}^2 dx = \sum_n \Abs{A_n}^2 = 1
\end{equation}

The question of probability per unit length is much easier asked of a pure state, for which we have just a single contributor

\begin{equation}\label{eqn:qmIproblemSet5:20b}
u(x,t) = N_n H_n(\xi) e^{-\xi^2/2} e^{-i E_n t/\hbar}.
\end{equation}

We cannot answer the question for the probability that the particle is found at the specific $x=0$ position at $t=t_0$, but we can answer the question for the probability that a particle is found in an interval surrounding a specific point at this time.  We could for example ask for the probabilities that the particle could be found in any of the regions $[-1,0]$, $[-1/2,1/2]$, or $[0,1]$.  It seems reasonable to use $[-1/2,1/2]$ to answer this question, for which we have

\begin{align*}
\text{Probability that particle is found between -0.5 and 0.5}
&=
N_n^2 \int_{x=-1/2}^{1/2} H_n^2(\xi) e^{-\xi^2} dx \\
&=
\frac{N_n^2}{\alpha} \int_{\xi=-\alpha/2}^{\alpha/2} H_n^2(\xi) e^{-\xi^2} d\xi \\
\end{align*}

Inserting $N_n$ we have
\begin{equation}\label{eqn:qmIproblemSet5:20e}
P =
\inv{ \sqrt{\pi} 2^n n!} \int_{\xi=-\alpha/2}^{\alpha/2} H_n^2(\xi) e^{-\xi^2} d\xi 
\end{equation}

Can this be evaluated any further without knowledge of $n$ specifically?

For the impure state of \ref{eqn:qmIproblemSet5:20} we have for the probability density

\begin{align*}
\Abs{u}^2 
&= 
\sum_{m,n} 
A_n A_m^\conj H_n(\xi) H_m(\xi) e^{-\xi^2} e^{-i (E_n - E_m)t_0/\hbar} \\
&=
\sum_n 
\Abs{A_n}^2 (H_n(\xi))^2 e^{-\xi^2} 
+\sum_{m \ne n} 
A_n A_m^\conj H_n(\xi) H_m(\xi) e^{-\xi^2} e^{-i (E_n - E_m)t_0/\hbar} \\
&=
\sum_n 
\Abs{A_n}^2 (H_n(\xi))^2 e^{-\xi^2} 
+\sum_{m \ne n} 
A_n A_m^\conj H_n(\xi) H_m(\xi) e^{-\xi^2} e^{-i (E_n - E_m)t_0/\hbar} \\
&=
\sum_n 
\Abs{A_n}^2 (H_n(\xi))^2 e^{-\xi^2} 
+\sum_{m < n} 
H_n(\xi) H_m(\xi)
\left(
A_n A_m^\conj 
e^{-\xi^2} e^{-i (E_n - E_m)t_0/\hbar}
+A_m A_n^\conj 
e^{-\xi^2} e^{-i (E_m - E_n)t_0/\hbar}
\right) \\
&=
\sum_n 
\Abs{A_n}^2 (H_n(\xi))^2 e^{-\xi^2} 
+2 \sum_{m < n} 
H_n(\xi) H_m(\xi)
e^{-\xi^2} 
\Re \left(
A_n A_m^\conj 
e^{-i (E_n - E_m)t_0/\hbar}
\right) \\
&=
\sum_n 
\Abs{A_n}^2 (H_n(\xi))^2 e^{-\xi^2}  \\
&\quad+2 \sum_{m < n} 
H_n(\xi) H_m(\xi)
e^{-\xi^2}
\left(
\Re ( A_n A_m^\conj ) \cos( (n - m)\omega t_0)
+\Im ( A_n A_m^\conj ) \sin( (n - m)\omega t_0)
\right) \\
\end{align*}

For this mixed state wavefunction the question of the probability to find the particle within $1/2$ unit of $x=0$ in either direction is now

\begin{align*}
P 
&= \int_{-1/2}^{1/2} \Abs{u(x)}^2 dx \\
&=
\sum_n 
\frac{\Abs{A_n}^2 }{\alpha}
\int_{-\alpha/2}^{\alpha/2} d\xi
(H_n(\xi))^2 e^{-\xi^2} \\
&+\frac{2}{\alpha} \sum_{m < n} 
\left(
\Re ( A_n A_m^\conj ) \cos( (n - m)\omega t_0)
+\Im ( A_n A_m^\conj ) \sin( (n - m)\omega t_0)
\right) 
\int_{-\alpha/2}^{\alpha/2} d\xi
H_n(\xi) H_m(\xi)
e^{-\xi^2}
\end{align*}

It is interesting that the time dependence of the probability per unit length is time dependent only for a mixed state.

\subsubsection{The probability for the first few pure states.}

We can try to evaluate \ref{eqn:qmIproblemSet5:20e} explicitly for a couple of values of $n$, however, we can do no better than utilization of the error function to express this.  We have for the first couple Hankle functions

\begin{align}\label{eqn:qmIproblemSet5:200}
H_0(\xi) &= 1 \\
H_1(\xi) &= 2 \xi \\
H_2(\xi) &= 4 \xi^2 - 2
\end{align}

So for the first few pure states, we get for each of the respective probabilities in this interval (at $t = t_0$ or any other time)
\begin{align*}
\inv{ \sqrt{\pi} }
\int_{-{\alpha/2}}^{\alpha/2} H_0^2(\xi) e^{-\xi^2} d\xi 
&= 
\inv{ \sqrt{\pi} }
\int_{-{\alpha/2}}^{\alpha/2} e^{-\xi^2} d\xi 
= 
% integrate e^(-x^2)/sqrt(pi)
\erf({\alpha/2}) \\
%%%
\inv{ 2 \sqrt{\pi} }
\int_{-{\alpha/2}}^{\alpha/2} H_1^2(\xi) e^{-\xi^2} d\xi 
&= 
\frac{2}{ \sqrt{\pi} }
\int_{-{\alpha/2}}^{\alpha/2} \xi^2 e^{-\xi^2} d\xi 
= 
% integrate 2 x^2 e^(-x^2)/sqrt(pi)
\erf({\alpha/2}) - \frac{\alpha}{\sqrt{\pi}} e^{-\alpha^2/4} 
\\
\inv{ 8 \sqrt{\pi} }
\int_{-{\alpha/2}}^{\alpha/2} H_2^2(\xi) e^{-\xi^2} d\xi 
&= 
\inv{ 2 \sqrt{\pi} }
\int_{-{\alpha/2}}^{\alpha/2} (2 \xi^2 -1)^2 e^{-\xi^2} d\xi 
= 
% integrate (2 x^2 - 1)^2 e^(-x^2)/2/sqrt(pi)
\erf({\alpha/2}) - \frac{1}{2 \sqrt{\pi}} \alpha e^{-\alpha^2/4} (1 + \alpha^2/2)
\end{align*}

To get a feel for this observe that we have $m_e/\hbar = 8 637.99381 s / m^2$, so for an electron, so when $\omega$ is not small our parameter $\alpha = \sqrt{m \omega/\hbar}$ is fairly large.  This means that we can make the approximations $\erf(\alpha/2) \approx 1$, and $e^{-\alpha^2/4} \approx 0$.

On the other hand, for small $\alpha$, we have $\erf(\alpha/2) \approx \alpha/\sqrt{\pi}$.

FIXME: negative probability in second integral for small $\alpha$?

%\begin{align*}
%\text{for pure state of electron with $E = E_0$} P(x \in [-1/2,1/2]) 
%\approx \sqrt{2/\pi} &= 0.8 \\
%\text{for pure state of electron with $E = E_1$} P(x \in [-1/2,1/2]) 
%\approx &= 1 \\
%\text{for pure state of electron with $E = E_1$} P(x \in [-1/2,1/2]) 
%\approx &= 3/8 = 0.4 \\
%\end{align*}

For large $\alpha$ we 

We see that physically the probability that a particle will be found very close to the origin is highly dependent energy level associated with the state that the particle is found in and its associated wave function.  A low energy state does not necessarily mean that the particle will be found at a small displacement from the origin, as in the case in a classical system.  Also different from the classical system is that when the state the particle is expressed as a superposition of basis states, then the probability for the particle to be found in the interval is also time dependent.  That probability oscillates with sine and cosine dependence around a steady state probability fixed by the sum of the pure state probabilities.

\subsubsection{A final note.  Probability for non-normalizable state earlier in the problem statement.}

In case this question was asking about the probability for the particle to be found in the region for the non-normalizable state $u(x) = B \xi e^{\xi^2/2}$, we can answer this too.  Since this requires $B=0$, that probability is also simply zero.

%\EndArticle
\EndNoBibArticle
