%
% Copyright © 2015 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\input{../blogpost.tex}
\renewcommand{\basename}{gmLecture1}
\renewcommand{\dirname}{notes/phy1520/}
%\newcommand{\dateintitle}{}
%\newcommand{\keywords}{}

\input{../peeter_prologue_print2.tex}

\usepackage{peeters_layout_exercise}
\usepackage{peeters_braket}
\usepackage{peeters_figures}

\beginArtNoToc

\generatetitle{Introduction}
%\chapter{Introduction}
%\label{chap:gmLecture1}

Text \citep{sakurai2014modern} (revised edition).

\paragraph{Classical mechanics}

We'll be talking about one body physics for most of this course.  In classical mechanics we can figure out the particle trajectories using both of \( (\Br, \Bp \), where

\begin{dmath}\label{eqn:qmLecture1:20}
\begin{aligned}
\ddt{\Br} &= \inv{m} \Bp \\
\ddt{\Bp} &= \spacegrad V
\end{aligned}
\end{dmath}

A two dimensional phase space shows the trajectory of a point particle subject to some equations of motion

\paragraph{Quantum mechanics}

%\begin{dmath}\label{eqn:qmLecture1:n}
\boxedEquation{eqn:qmLecture1:n}{
\Hbar = 1.
}
%\end{dmath}

In QM we are no longer allowed to think of position and momentum, but have to start asking about state vectors \( \ket{\Psi} \).

We'll consider the state vector with respect to some basis, for example, in a position basis, we write

\begin{dmath}\label{eqn:qmLecture1:40}
\braket{ x }{\Psi } = \Psi(x),
\end{dmath}

a complex numbered ``wave function'', the probability amplitude for a particle in \( \ket{\Psi} \) to be in the vincinity of \( x \).

We could also consider the state in a momentum basis

\begin{dmath}\label{eqn:qmLecture1:60}
\braket{ p }{\Psi } = \Psi(p),
\end{dmath}

a probability amplitude with respect to momentum \( p \).

More precisely, 

\begin{dmath}\label{eqn:qmLecture1:80}
\Abs{\Psi(x)}^2 dx \ge 0
\end{dmath}

is the probability of finding the particle in the range \( (x, x + dx ) \).  To have meaning as a probability, we require

\begin{dmath}\label{eqn:qmLecture1:100}
\int_{-\infty}^\infty \Abs{\Psi(x)}^2 dx = 1.
\end{dmath}

The average position can be calculated using this PDF.  For example

\begin{dmath}\label{eqn:qmLecture1:120}
\expectation{x} = \int_{-\infty}^\infty \Abs{\Psi(x)}^2 x dx,
\end{dmath}

or 
\begin{dmath}\label{eqn:qmLecture1:140}
\expectation{f(x)} = \int_{-\infty}^\infty \Abs{\Psi(x)}^2 f(x) dx.
\end{dmath}

Similarily, calculation of an average of a function of momentum can be expressed as

\begin{dmath}\label{eqn:qmLecture1:160}
\expectation{f(p)} = \int_{-\infty}^\infty \Abs{\Psi(p)}^2 f(p) dp.
\end{dmath}

\paragraph{Transformation from a position to momentum basis}

We have a problem, if we which to compute an average in momentum space such as \( \expectation{p} \), when given a wavefunction \( \Psi(x) \).

How do we convert 

\begin{dmath}\label{eqn:qmLecture1:180}
\Psi(p) 
\Psi(x),
\stackrel{?}{\leftrightarrow}
\end{dmath}

or equivalently
\begin{dmath}\label{eqn:qmLecture1:200}
\braket{p}{\Psi} = 
\leftrightarrow 
\stackrel{?}{\leftrightarrow}
\braket{x}{\Psi}.
\end{dmath}

Such a conversion can be performed by virtue of an the assumption that we have a complete orthonormal basis, for which we can introduce identity operations such as

\begin{dmath}\label{eqn:qmLecture1:220}
\int_{-\infty}^\infty \ket{p}\bra{p} = 1,
\end{dmath}

or
\begin{dmath}\label{eqn:qmLecture1:240}
\int_{-\infty}^\infty \ket{x}\bra{x} = 1
\end{dmath}

FIXME: justification?:

\begin{dmath}\label{eqn:qmLecture1:260}
\begin{aligned}
\int_{-\infty}^\infty \Psi{p}\ket{p} &= \ket{\Psi} \\
\int_{-\infty}^\infty \Psi{x}\ket{x} &= \ket{\Psi}
\end{aligned}
\end{dmath}

Some interpretations:

\begin{enumerate}
\item \( \ket{x_0} \leftrightarrow \text{sits at} x = x_0 \)
\item \( \braket{x}{x'} \leftrightarrow \delta(x - x') \)
\item \( \braket{p}{p'} \leftrightarrow \delta(p - p') \)
\item \( \braket{x}{p'} = \frac{e^{i p x}}{\sqrt{V}} \), where \( V \) is the volume of the box containing the particle.  We'll treat the case where the box bolume is infinite later.
\end{enumerate}

The conversion from a position basis to momentum space is now possible 

\begin{dmath}\label{eqn:qmLecture1:280}
\braket{p}{\Psi} 
=
\Psi(p)
= \int_{-\infty}^\infty \braket{p}{x} \braket{x}{\Psi} dx 
= \int_{-\infty}^\infty \frac{e^{-ip x}}{\sqrt{V}} \Psi(x) dx
\end{dmath}

where \( \braket{p}{x} = \braket{x}{p}^\conj \).  The momentum space to position space conversion can be written as

\begin{dmath}\label{eqn:qmLecture1:300}
\Psi(x) 
= \int_{-\infty}^\infty \frac{e^{ip x}}{\sqrt{V}} \Psi(p) dp.
\end{dmath}

Now we can go back and figure out the an expectation

\begin{dmath}\label{eqn:qmLecture1:320}
\expectation{p}
=
\int \Psi^\conj(p) \Psi(p) p d p
=
\int dp 
\lr{
\int_{-\infty}^\infty \frac{e^{ip x}}{\sqrt{V}} \Psi^\conj(x) dx
}
\lr{
\int_{-\infty}^\infty \frac{e^{-ip x'}}{\sqrt{V}} \Psi(x') dx'
}
p d p.
\end{dmath}

FIXME : Performing this integral we find

\begin{dmath}\label{eqn:qmLecture1:340}
p e^{i p x} \leftrightarrow -i \PD{x}{} e^{i p x},
\end{dmath}

so

\begin{dmath}\label{eqn:qmLecture1:360}
\expectation{p}
=
\int dp dx dx'
\lr{ \frac{e^{ip x}}{\sqrt{V}} \Psi^\conj(x)  }
\lr{ i \frac{\Psi(x')}{\sqrt{V}} \PD{x'}{} e^{-ip x'} }
= \int \frac{dx dx'}{V} \Psi^\conj(x) \Psi(x) i \PD{x'}{} \lr{ \int dp e^{i p x - i p x'} }
= \int \frac{dx dx'}{V} \Psi^\conj(x) \Psi(x) i \PD{x'} V \delta(x - x')
= \int dx dx' \Psi^\conj(x) \Psi(x) i \PD{x'}{} \delta(x - x')
= \int dx dx' \Psi^\conj(x) \lr{ -i \PD{x'}{\Psi(x')}} \delta(x - x')
= \int dx \Psi^\conj(x) \lr{ -i \PD{x}{} } \Psi(x).
\end{dmath}

Allowing identifications like

\begin{dmath}\label{eqn:qmLecture1:380}
p \leftrightarrow -i \PD{x}{}
\end{dmath}
\begin{dmath}\label{eqn:qmLecture1:400}
p^2 \leftrightarrow - \PDSq{x}{}
\end{dmath}

\paragraph{Matrix interpreation}

\begin{enumerate}
\item Ket's \( \ket{\Psi} \leftrightarrow \text{column vector} \)
\item Bra's \( \bra{\Psi} \leftrightarrow {\text{row vector}}^\conj \)
\item Operators \( \leftrightarrow \) matrices that act on vectors.
\end{enumerate}

\begin{dmath}\label{eqn:qmLecture1:420}
\hat{p} \ket{\Psi} \rightarrow \ket{\Psi'}
\end{dmath}

\paragraph{Time evolution}

For a state subject to the equations of motion given by the Hamiltonian operator \( \hat{H} \)

\begin{dmath}\label{eqn:qmLecture1:440}
i \PD{t}{} \ket{\Psi} = \hat{H} \ket{\Psi},
\end{dmath}

the time evolution is given by
\begin{dmath}\label{eqn:qmLecture1:460}
\ket{\Psi(t)} = e^{-i \hat{H} t} \ket{\Psi(0)}.
\end{dmath}

\paragraph{Incomplete information}

We'll need to introduce the concept of Density matrices.  This will bring us to concepts like entanglement.

\EndArticle
