\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY456H1F: Quantum Mechanics II.  Lecture 12 (Taught by Mr. Federico Duque Gomez).  WKB Method}
\label{chap:qmTwoL12}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/qmTwoL12.pdf}
\date{Oct 19, 2011}
\revisionInfo{qmTwoL12.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Disclaimer.}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{WKB (Wentzel-Kramers-Brillouin) Method.}

This is covered in \S 24 in the text \cite{desai2009quantum}.  Also \S 8 of \cite{griffiths2005introduction}.

\begin{equation}\label{eqn:qmTwoL12:10}
-\frac{\hbar^2}{2m} \frac{d^2 U}{dx^2} + V(x) U(x) = E U(x)
\end{equation}

or

\begin{equation}\label{eqn:qmTwoL12:30}
\frac{d^2 U}{dx^2} + \frac{2m}{\hbar^2} (E - V(x)) U(x) = 0
\end{equation}

FIXME: fig 1

\begin{equation}\label{eqn:qmTwoL12:50}
U \propto e^{\pm i k x}
\end{equation}

in the well, and 

\begin{equation}\label{eqn:qmTwoL12:70}
U \propto e^{\pm \kappa x}
\end{equation}

outside of the well, where

\begin{align}\label{eqn:qmTwoL12:90}
k &= \frac{2m (E - V)}{\hbar},\qquad E > V \\
\kappa &= \frac{2m (V - E)}{\hbar}, \qquad V > E
\end{align}

In general we can hope for something similar, looking for

\begin{align}\label{eqn:qmTwoL12:110}
k^2(x) &= \frac{2m (E - V(x))}{\hbar},\qquad E > V \\
\kappa^2(x) &= \frac{2m (V(x) - E)}{\hbar}, \qquad V > E
\end{align}

Schr\"{o}dinger becomes

\begin{equation}\label{eqn:qmTwoL12:130}
\frac{d^2 U(x)}{dx^2} + k^2(x) U(x) = 0
\end{equation}

Try

\begin{equation}\label{eqn:qmTwoL12:150}
U(x) = A e^{i \phi(x)},
\end{equation}

with 

\begin{equation}\label{eqn:qmTwoL12:170}
\phi(x) = \phi_R(x) + i \phi_I(x) 
\end{equation}

With 

\begin{align*}
(e^{i \phi})'' 
&=
(i \phi' e^{i \phi})'  \\
&=
(i \phi')^2 e^{i \phi} + i \phi'' e^{i \phi}
\end{align*}

Schr\"{o}dinger becomes

\begin{equation}\label{eqn:qmTwoL12:190}
- (\phi'(x))^2 + i \phi''(x) + k^2(x) = 0
\end{equation}

For the first round of approximation we assume

\begin{equation}\label{eqn:qmTwoL12:210}
\phi''(x) \approx 0,
\end{equation}

and obtain

\begin{equation}\label{eqn:qmTwoL12:230}
(\phi'(x))^2 = k^2(x),
\end{equation}

or
\begin{equation}\label{eqn:qmTwoL12:250}
\phi'(x) = \pm k(x).
\end{equation}

A second round of approximation we use \ref{eqn:qmTwoL12:250} and obtain

\begin{equation}\label{eqn:qmTwoL12:250b}
\phi''(x) = \pm k'(x)
\end{equation}

Plugging back into \ref{eqn:qmTwoL12:190} we have

\begin{equation}\label{eqn:qmTwoL12:270}
-(\phi'(x))^2 \pm i k'(x) + k^2(x) = 0,
\end{equation}

or

\begin{equation}\label{eqn:qmTwoL12:290}
\begin{aligned}
\phi'(x) 
&= \pm \sqrt{ \pm i k'(x) + k^2(x) } \\
&= \pm k(x) \sqrt{ 1 + \pm i \frac{k'(x)}{k^2(x)} } 
\end{aligned}
\end{equation}

and if
\begin{equation}\label{eqn:qmTwoL12:310}
\frac{k'(x)}{k^2(x)} \ll 1, 
\end{equation}

we have

\begin{equation}\label{eqn:qmTwoL12:330}
\phi'(x) 
= \pm k(x) \pm i \frac{k'(x)}{2 k(x)}  
\end{equation}

Integrating

\begin{align*}
\phi(x) 
&= \pm \int dx k(x) \pm i \int dx \frac{k'(x)}{2 k(x)}  + \text{const} \\
&= \pm \int dx k(x) \pm i \ln k(x) + \text{const} 
\end{align*}

Going back to our wavefunction, if $E > V(x)$ we have

\begin{align*}
U(x) 
&\sim A e^{i \phi(x)} \\
&= \exp \left(i\left( 
\pm \int dx k(x) \pm i \ln k(x) + \text{const} 
\right)\right) \\
&\sim \exp \left(i\left( 
\pm \int dx k(x) \pm i \ln k(x) 
\right)\right) \\
&= e^{\pm i \int dx k(x)} e^{\mp \inv{2} \ln k(x)} \\
\end{align*}

or
\begin{equation}\label{eqn:qmTwoL12:350}
U(x) \propto \inv{\sqrt{k(x)}} e^{\pm i \int dx k(x)} 
\end{equation}

Similarly for $E < V(x)$ we can find

\begin{equation}\label{eqn:qmTwoL12:370}
U(x) \propto \inv{\sqrt{\kappa(x)}} e^{\pm i \int dx \kappa(x)}.
\end{equation}

\paragraph{Validity}
\begin{enumerate}
\item V(x) changes very slowly $\implies k'(x)$ small, and $k(x) = \sqrt{2 m (E - V(x))}/\hbar$.
\item $E$ very far away from the potential $\Abs{(E - V(x))/V(x)} \gg 1$.
\end{enumerate}

\section{Examples}

FIXME: fig2

FIXME: fig3

WKB won't work at the turning points in this figure since our main assumption was that

\begin{equation}\label{eqn:qmTwoL12:390}
\Abs{\frac{k'(x)}{k^2(x)}} \ll 1
\end{equation}

so we get into trouble where $k(x) \sim 0$.  There are some methods for dealing with this.  Our text as well as Griffiths give some examples, but they require Bessel functions and more complex mathematics.

The idea is that one finds the WKB solution in the regions of validity, and then looks for a polynomial solution in the patching region where we are closer to the turning point, probably requiring lookup of various special functions.

This power series method is also outlined in \cite{wiki:wkb}, where solutions to connect the regions are expressed in terms of Airy functions.

\EndArticle
