\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY456H1F: Quantum Mechanics II.  Lecture 22 (Taught by Prof J.E. Sipe).  Scattering (cont.)}
\label{chap:qmTwoL22}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2011/qmTwoL22.pdf}
\date{Nov XX, 2011}
\revisionInfo{qmTwoL22.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Disclaimer.}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{Scattering.  Recap}

READING: \S 19, \S 20 of the text \cite{desai2009quantum}.

We used a positive potential of the form

FIXME: F1

\begin{equation}\label{eqn:qmTwoL22:10}
-\frac{\hbar^2}{2 \mu} \PDSq{x}{\psi_k(x)} + V(x) \psi_k(x) = \frac{\hbar^2 k^2}{2 \mu}
\end{equation}

for $x \ge x_3$

\begin{equation}\label{eqn:qmTwoL22:30}
\psi_k(x) = C e^{i k x}
\end{equation}

\begin{equation}\label{eqn:qmTwoL22:50}
\phi_k(x) = \ddx{\psi_k(x)}
\end{equation}

for $x \ge x_3$ 

\begin{equation}\label{eqn:qmTwoL22:70}
\phi_k(x) = i k C e^{i k x}
\end{equation}

\begin{align}\label{eqn:qmTwoL22:90}
\ddx{\phi_k(x)} &= \phi_k(x) \\
-\frac{\hbar^2}{2 \mu} \ddx{\phi_k(x)} &= - V(x) \psi_k(x) + \frac{\hbar^2 k^2}{2 \mu}
\end{align}

integrate these equations back to $x_1$.

For $x \le x_1$

\begin{equation}\label{eqn:qmTwoL22:110}
\phi_k(x) = A e^{i k x} + B e^{-i k x},
\end{equation}

where both $A$ and $B$ are proportional to $C$, dependent on $k$.

There are cases where we can solve this analytically (one of these is on our problem set).

Alternatively, write as (so long as $A \ne 0$)

\begin{equation}\label{eqn:qmTwoL22:130}
\begin{array}{l l l}
\phi_k(x) 
&\rightarrow e^{i k x} + \beta_k e^{-i k x} & \qquad \mbox{for $x < x_1$} \\
&\rightarrow \gamma_k e^{i k x} & \qquad \mbox{for $x > x_2$}
\end{array}
\end{equation}

Now want to consider the problem of no potential in the interval of interest, and our window bounded potential as in 

FIXME: F3.

where we model our particle as a wave packet as we found can have the fourier transform description, for $t_{\text{initial}} < 0$, of

\begin{equation}\label{eqn:qmTwoL22:150}
\psi(x, \ti) = \int \frac{dk}{\sqrt{2 \pi} \alpha(k, \ti) e^{i k x}
\end{equation}

Returning to the 

same coefficients, the solution of the Sch eqn for problem with the potential

% FIXME: BB referral to \ref{eqn:qmTwoL22:130}
%\begin{equation}\label{eqn:qmTwoL22:170}
%\psi(x, \ti) = \int \frac{dk}{\sqrt{2 \pi} \alpha(k, \ti) \psi_k(x)
%\end{equation}
%
For $x \le x_1$, 

\begin{equation}\label{eqn:qmTwoL22:190}
\psi(x, t) = \psi_i(x, t) + \psi_r(x, t)
\end{equation}

where 

\begin{align}\label{eqn:qmTwoL22:210}
\psi_i(x, t) &= \int \frac{dk}{\sqrt{2 \pi} \alpha(k, \ti) e^{i k x} \\
\psi_r(x, t) &= \int \frac{dk}{\sqrt{2 \pi} \alpha(k, \ti) \beta_k e^{-i k x}
\end{align}

For $x > x_2$ 

\psi(x, t) = \psi_t(x, t) 

and

\begin{equation}\label{eqn:qmTwoL22:230}
\psi_t(x, t) = \int \frac{dk}{\sqrt{2 \pi} \alpha(k, \ti) \gamma_k e^{i k x}
\end{equation}

Look at 

\begin{equation}\label{eqn:qmTwoL22:n}
\psi_r(x, t) = \chi(-x, t)
\end{equation}

where

\begin{equation}\label{eqn:qmTwoL22:n}
\begin{aligned}
\chi(x, t) 
&= \int \frac{dk}{\sqrt{2 \pi} \alpha(k, \ti) \beta_k e^{i k x} \\
&\approx 
&= \beta_{k_0} \int \frac{dk}{\sqrt{2 \pi} \alpha(k, \ti) e^{i k x} 
\end{aligned}
\end{equation}

for $t = \ti$, this is nonzero for $x < x_1$.

so for $x < x_1$

\begin{equation}\label{eqn:qmTwoL22:n}
\psi_r(x, \ti) = 0
\end{equation}

In the same way, for $x > x_2$ 

\begin{equation}\label{eqn:qmTwoL22:n}
\psi_t(x, \ti) = 0.
\end{equation}

What hasn't been proved is that the wavefunction is also zero in the $[x_1, x_2]$ interval.

\EndArticle
