\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY456H1F: Quantum Mechanics II.  Lecture 15 (Taught by Prof J.E. Sipe).  Rotation operator in spin space}
\label{chap:qmTwoL15}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/qmTwoL15.pdf}
\date{Oct 31, 2011}
\revisionInfo{qmTwoL15.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Disclaimer.}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{Rotation operator in spin space.}

We can formally expand our rotation operator in Taylor series

\begin{equation}\label{eqn:qmTwoL15:10}
e^{-i \theta \ncap \cdot \BS/\hbar}
= 
I 
+
\left(-i \theta \ncap \cdot \BS/\hbar\right)
+
\inv{2!}
\left(-i \theta \ncap \cdot \BS/\hbar\right)^2
+
\inv{3!}
\left(-i \theta \ncap \cdot \BS/\hbar\right)^3
+ \cdots
\end{equation}

or
\begin{align*}
e^{-i \theta \ncap \cdot \Bsigma/2}
&= 
I 
+
\left(-i \theta \ncap \cdot \Bsigma/2\right)
+
\inv{2!}
\left(-i \theta \ncap \cdot \Bsigma/2\right)^2
+
\inv{3!}
\left(-i \theta \ncap \cdot \Bsigma/2\right)^3
+ \cdots \\
&=
\sigma_0 
+
\left(\frac{-i \theta}{2}\right) (\ncap \cdot \Bsigma)
+
\inv{2!} \left(\frac{-i \theta}{2}\right) (\ncap \cdot \Bsigma)^2
+
\inv{3!} \left(\frac{-i \theta}{2}\right) (\ncap \cdot \Bsigma)^3
+ \cdots \\
&=
\sigma_0 
+
\left(\frac{-i \theta}{2}\right) (\ncap \cdot \Bsigma)
+
\inv{2!} \left(\frac{-i \theta}{2}\right) \sigma_0
+
\inv{3!} \left(\frac{-i \theta}{2}\right) (\ncap \cdot \Bsigma) 
+ \cdots \\
&=
\sigma_0 \left( 1 - \inv{2!}\left(\frac{\theta}{2}\right)^2 + \cdots \right) 
+
(\ncap \cdot \Bsigma) \left( \frac{\theta}{2} - \inv{3!}\left(\frac{\theta}{2}\right)^3 + \cdots \right) \\
&=
\cos(\theta/2) \sigma_0 + \sin(\theta/2) (\ncap \cdot \Bsigma)
\end{align*}

where we've used the fact that $(\ncap \cdot \Bsigma)^2 = \sigma_0$.

So our representation of the spin operator is

\begin{equation}\label{eqn:qmTwoL15:30}
\begin{aligned}
e^{-i \theta \ncap \cdot \BS/\hbar} 
&\rightarrow
\cos(\theta/2) \sigma_0 + \sin(\theta/2) (\ncap \cdot \Bsigma) \\
&=
\cos(\theta/2) \sigma_0 + \sin(\theta/2) 
\left(n_x \PauliX + n_y \PauliY + n_z \PauliZ \right) \\
&=
\begin{bmatrix}
\cos(\theta/2) -i n_z \sin(\theta/2) & -i (n_x -i n_y) \sin(\theta/2) \\
-i (n_x + i n_y) \sin(\theta/2) & \cos(\theta/2) +i n_z \sin(\theta/2) 
\end{bmatrix}
\end{aligned}
\end{equation}

Note that, in particular, 

\begin{equation}\label{eqn:qmTwoL15:50}
e^{-2 \pi i \ncap \cdot \BS/\hbar} \rightarrow \cos\pi \sigma_0 = -\sigma_0
\end{equation}

This ``rotates'' the ket, but introduces a phase factor.

Can do this in general for other degrees of spin, for $s = 1/2, 3/2, 5/2, \cdots$.

\paragraph{Unfortunate interjection by me:} Mentioned the half angle rotation operator that requires a half angle operator sandwich.  Sipe thought I might be talking about a Heisenburg picture representation, where we have something like this in expectation values

\begin{equation}\label{eqn:qmTwoL15:70}
\ket{\psi'} = e^{-i \theta \ncap \cdot \BJ/\hbar} \ket{\psi}
\end{equation}

so that 
\begin{equation}\label{eqn:qmTwoL15:90}
\bra{\psi'}
\mathcal{O}
\ket{\psi'} = \bra{\psi} 
e^{i \theta \ncap \cdot \BJ/\hbar} 
\mathcal{O}
e^{-i \theta \ncap \cdot \BJ/\hbar} 
\ket{\psi}
\end{equation}

\section{Spin dynamics}

At least classically, the angular momentum of charged objects is associated with a magnetic moment

FIXME: F1.

\begin{equation}\label{eqn:qmTwoL15:110}
\Bmu = I A \Be_\perp
\end{equation}

In our scheme, following the (cgs?) text conventions of \cite{desai2009quantum}, where the $\BE$ and $\BB$ have the same units, we write

\begin{equation}\label{eqn:qmTwoL15:130}
\Bmu = \frac{I A}{c} \Be_\perp
\end{equation}

For a charge moving in a circle

FIXME: F2

\begin{equation}\label{eqn:qmTwoL15:150}
\begin{aligned}
I 
&= \frac{\text{charge}}{\text{time}} \\
&= 
\frac{\text{distance}}{\text{time}} \frac{\text{charge}}{\text{distance}} \\
&= 
\frac{q v}{ 2 \pi r}
\end{aligned}
\end{equation}

so the magnetic moment is 

\begin{equation}\label{eqn:qmTwoL15:170}
\begin{aligned}
\mu 
&= \frac{q v}{ 2 \pi r} \frac{\pi r^2}{c}  \\
&= \frac{q }{ 2 m c } (m v r) \\
&= \gamma L
\end{aligned}
\end{equation}

Here $\gamma$ is the gyromagnetic ratio

Recall that we have a torque 

FIXME: F3

\begin{equation}\label{eqn:qmTwoL15:190}
\BT = \Bmu \cross \BB
\end{equation}

tending to line up $\Bmu$ with $\BB$.  The energy is then

\begin{equation}\label{eqn:qmTwoL15:210}
-\Bmu \cdot \BB
\end{equation}

Also recall that this torque leads to precession 

\begin{equation}\label{eqn:qmTwoL15:230}
\ddt{\BL} = \BT = \gamma \BL \cross \BB
\end{equation}

FIXME: F4

with precession frequency 

\begin{equation}\label{eqn:qmTwoL15:250}
\Bomega = - \gamma \BB.
\end{equation}

For a current due to a moving electron

\begin{equation}\label{eqn:qmTwoL15:270}
\gamma = -\frac{e}{2 m c} < 0
\end{equation}

where we are, here, writing for charge on the electron $-e$.  

\paragraph{Question: steady state currents only?}.  Yes, this is only true for steady state currents.

For the translational motion of an electron, even if it is not moving in a steady way, regardless of it's dynamics

\begin{equation}\label{eqn:qmTwoL15:290}
\Bmu_0 = - \frac{e}{2 m c} \BL
\end{equation}

Now, back to quantum mechanics, we turn $\Bmu_0$ into a dipole moment operator and $\BL$ is ``promoted'' to an angular momentum operator.

\begin{equation}\label{eqn:qmTwoL15:310}
H_{\text{int}} = - \Bmu_0 \cdot \BB
\end{equation}

What about the ``spin''?

Perhaps 

\begin{equation}\label{eqn:qmTwoL15:330}
\Bmu_s = \gamma_s \BS
\end{equation}

we write this as 
\begin{equation}\label{eqn:qmTwoL15:350}
\Bmu_s = g \left( -\frac{e}{ 2 m c} \right)\BS
\end{equation}

so that

\begin{equation}\label{eqn:qmTwoL15:370}
\gamma_s = - \frac{g e}{ 2 m c} 
\end{equation}

Experimentally, one finds to very good approximation

\begin{equation}\label{eqn:qmTwoL15:390}
g = 2
\end{equation}

There was a lot of trouble with this in early quantum mechanics where people got things wrong, and cancelled the wrong factors of $2$.

In fact, Dirac's relativistic theory for the electron predicts $g=2$.

When this is measured experimentally, one does not get exactly $g=2$, and a theory that also incorporates photon creation and destruction and the interaction with the electron with such (virtual) photons.  We get

\begin{equation}\label{eqn:qmTwoL15:410}
\begin{aligned}
g_{\text{theory}} 
&= 2 \left(
1.001159652140 (\pm 28)
\right) \\
g_{\text{experimental}} 
&= 2 \left(
1.0011596521884 (\pm 43)
\right)
\end{aligned}
\end{equation}

Feynman compared this to measuring the distance between New York and Las Angeles to within the accuracy of a human hair.

\EndArticle
