\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY456H1F: Quantum Mechanics II.  Lecture L5 (Taught by Prof J.E. Sipe).  Pertubation theory and degeneracy}
\label{chap:qmTwoL5}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/qmTwoL5.pdf}
\date{Sept 23, 2011}
\revisionInfo{qmTwoL5.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Disclaimer.}

Peeter's lecture notes from class.  May not be entirely coherent.

\section{Issues concerning degeneracy.}

\subsection{When the perturbed state is non-degenerate.}

Suppose the state of interest is non-degenerate but others are

FIXME: diagram.  states designated by dashes labelled $n1$, $n2$, $n3$ degeneracy $\alpha = 3$ for energy $E_n^{(0)}$.

This is no problem except for notation, and if the analysis is repeated we find

\begin{align}\label{eqn:qmTwoL5:10}
E_s &= E_s^{(0)} + \lambda {H_{ss}}' + \lambda^2 
\sum_{m \ne s, \alpha} 
\frac{\Abs{{H_{m \alpha ; s}}'}^2 }
{ E_s^{(0)} - E_m^{(0)} } 
+ \cdots
\\
\ket{\bar{\psi}_s} &= \ket{{\psi_s}^{(0)}} + \lambda
\sum_{m \ne s, \alpha} 
\frac{{H_{m \alpha ; s}}'}
{ E_s^{(0)} - E_m^{(0)} } \ket{{\psi_{m \alpha}}^{(0)}}
+ \cdots,
\end{align}

where

\begin{equation}\label{eqn:qmTwoL4:30}
{H_{m \alpha ; s}}' =
\bra{{\psi_{m \alpha}}^{(0)}} H' \ket{{\psi_{s \alpha}}^{(0)}}
\end{equation}

\subsection{When the perturbed state is also degenerate.}

FIXME: diagram.  states designated by dashes labelled $n1$, $n2$, $n3$ degeneracy $\alpha = 3$ for energy $E_n^{(0)}$, and states designated by dashes labelled $s1$, $s2$, $s3$ degeneracy $\alpha = 3$ for energy $E_s^{(0)}$.

If we just blindly repeat the derivation for the non-degenerate case we would obtain

\begin{align}\label{eqn:qmTwoL5:50}
E_s &= E_s^{(0)} + \lambda {H_{s1 ; s1}}' 
+ \lambda^2 
\sum_{m \ne s, \alpha} 
\frac{\Abs{{H_{m \alpha ; s1}}'}^2 }
{ E_s^{(0)} - E_m^{(0)} } 
+ \lambda^2 
\sum_{\alpha \ne 1} 
\frac{\Abs{{H_{s \alpha ; s1}}'}^2 }
{ E_s^{(0)} - \color{red}{E_s^{(0)}} } 
+ \cdots
\\
\ket{\bar{\psi}_s} &= \ket{{\psi_s}^{(0)}} 
+ \lambda
\sum_{m \ne s, \alpha} 
\frac{{H_{m \alpha ; s}}'}
{ E_s^{(0)} - E_m^{(0)} } \ket{{\psi_{m \alpha}}^{(0)}}
+ \lambda
\sum_{\alpha \ne s1} 
\frac{{H_{s \alpha ; s1}}'}
{ E_s^{(0)} - \color{red}{E_s^{(0)}} } \ket{{\psi_{s \alpha}}^{(0)}}
+ \cdots,
\end{align}

where

\begin{equation}\label{eqn:qmTwoL4:70}
{H_{m \alpha ; s1}}' =
\bra{{\psi_{m \alpha}}^{(0)}} H' \ket{{\psi_{s1}}^{(0)}}
\end{equation}

Note that the $E_s^{(0)} -E_s^{(0)}$ is NOT a typo, and why we run into trouble.  There is one case where a pertubation approach is still possible.  That case is if we happen to have

\begin{equation}\label{eqn:qmTwoL5:90}
\bra{{\psi_{m \alpha}}^{(0)}} H' \ket{{\psi_{s1}}^{(0)}} = 0.
\end{equation}

That may not be obvious, but if one returns to the original derivation, the right terms cancel so that one will not end up with the $0/0$ problem.

\subsection{Diagonalizing the pertubation Hamiltonian.}

Suppose that we do not have this special zero condition that allows the pertubation treatment to remain valid.  What can we do.  It turns out that we can make use of the fact that the pertubation Hamiltonian is Hermitian, and diagonalize the matrix

\begin{equation}\label{eqn:qmTwoL5:110}
\bra{{\psi_{s \alpha}}^{(0)}} H' \ket{{\psi_{s \beta}}^{(0)}} 
\end{equation}

In the example of a two fold degeneracy, this amounts to us choosing not to work with the states

\begin{equation}\label{eqn:qmTwoL5:130}
\ket{\psi_{s1}^{(0)}}, \ket{\psi_{s2}^{(0)}},
\end{equation}

both some linear combinatations of the two
\begin{align}\label{eqn:qmTwoL5:150}
\ket{\psi_{sI}^{(0)}} &= a_1 \ket{\psi_{s1}^{(0)}} + b_1 \ket{\psi_{s2}^{(0)}} \\
\ket{\psi_{sII}^{(0)}} &= a_2 \ket{\psi_{s1}^{(0)}} + b_2 \ket{\psi_{s2}^{(0)}} 
\end{align}

In this new basis, once found, we have

\begin{equation}\label{eqn:qmTwoL5:170}
\bra{{\psi_{s \alpha}}^{(0)}} H' \ket{{\psi_{s \beta}}^{(0)}} = \mathcal{H}_\alpha \delta_{\alpha \beta}
\end{equation}

Utilizing this to fix the previous, one would get if the analysis was repeated correctly

\begin{align}\label{eqn:qmTwoL5:190}
E_{s\alpha} &= E_s^{(0)} + \lambda {H_{s\alpha ; s\alpha}}' 
+ \lambda^2 
\sum_{m \ne s, \alpha} 
\frac{\Abs{{H_{m \beta ; s \alpha}}'}^2 }
{ E_s^{(0)} - E_m^{(0)} } 
+ \cdots
\\
\ket{\bar{\psi}_{s \alpha}} &= \ket{{\psi_{s \alpha}}^{(0)}} 
+ \lambda
\sum_{m \ne s, \beta} 
\frac{{H_{m \beta ; s \alpha}}'}
{ E_s^{(0)} - E_m^{(0)} } \ket{{\psi_{m \beta}}^{(0)}}
+ \cdots.
\end{align}

We see that a degenerate state can be split by applying pertubation

FIXME: diagram.  $E_s^{(0)}$ as one energy level without pertubation, and as two distinct levels with pertubation.

\paragraph{guess} I'll bet that this is the origin of the spectral line splitting, especially given that an atom like hydrogen has degenerate states.

\section{Review of dynamics.}

This is covered in \S 3 of the text \cite{desai2009quantum}.

\EndArticle
