\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY450H1S.  Relativistic Electrodynamics Lecture 18 (Taught by Prof. Erich Poppitz).  Green's function solution to Maxwell's equation.}
\label{chap:relativisticElectrodynamicsL18}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsL18.pdf}
\date{Mar 9, 2011}
\revisionInfo{relativisticElectrodynamicsL18.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Reading.}

Covering chapter 8 material from the text \cite{landau1980classical}.

Covering \href{http://www.physics.utoronto.ca/~poppitz/epoppitz/PHY450_files/RelEMpp136-146.pdf}{lecture notes pp. 136-146}: continued reminder of electrostatic Green’s function (136); the retarded Green’s function of the d’Alembert operator: derivation and properties (137-140); the solution of the d’Alembert equation with a source: retarded potentials (141-142); retarded time: meaning and the Lienard-Wiechert potentials (143-146) [Wednesday, Mar. 9...]

\section{Solving the forced wave equation.}

See the notes for a complex variables and Fourier transform method of deriving the Green's function.  In class, we'll just pull it out of a magic hat.  We wish to solve

\begin{equation}\label{eqn:relativisticElectrodynamicsL18:n}
\square A^k = \partial_i \partial^i A^k = \frac{4 \pi}{c} j^k
\end{equation}

(with a $\partial_i A^i = 0$ gauge choice).

Our Green's method utilizes

\begin{equation}\label{eqn:relativisticElectrodynamicsL18:n}
\square_{(\Bx, t)} G(\Bx - \Bx', t - t') = \delta^3( \Bx - \Bx') \delta( t - t')
\end{equation}

If we know such a function, our solution is simple to obtain

\begin{equation}\label{eqn:relativisticElectrodynamicsL18:n}
A^k(\Bx, t) 
= \int d^3 \Bx' dt' \frac{4 pi}{c} j^k(\Bx', t') G(\Bx - \Bx', t - t') 
\end{equation}

Proof:
\begin{align*}
\square_{(\Bx, t)} A^k(\Bx, t) 
&= 
\int d^3 \Bx' dt' \frac{4 pi}{c} j^k(\Bx', t') 
\square_{(\Bx, t)} 
G(\Bx - \Bx', t - t') \\
&=
\int d^3 \Bx' dt' \frac{4 pi}{c} j^k(\Bx', t') 
\delta^3( \Bx - \Bx') \delta( t - t') \\
&=
\frac{4 pi}{c} j^k(\Bx, t) 
\end{align*}

Claim:

\begin{equation}\label{eqn:relativisticElectrodynamicsL18:n}
G(\Bx, t) = \frac{\delta(t - \Abs{\Bx}/c)}{4 pi \Abs{\Bx} }
\end{equation}

This is the retarded Green's function of the operator $\square$, where

\begin{equation}\label{eqn:relativisticElectrodynamicsL18:n}
\square G(\Bx, t) = \delta^3(\Bx) \delta(t)
\end{equation}

Proof:

\begin{equation}\label{eqn:relativisticElectrodynamicsL18:n}
\inv{c^2} \PDSq{t}{} \frac{\delta(t - \Abs{\Bx}/c)}{4 pi \Abs{\Bx} }
=
\inv{c^2} \frac{\delta''(t - \Abs{\Bx}/c)}{4 pi \Abs{\Bx} }
\end{equation}

\begin{align*}
\spacegrad \cdot \spacegrad 
\frac{\delta(t - \Abs{\Bx}/c)}{4 pi \Abs{\Bx} }
&=
\spacegrad^2 \left( \frac{1}{4 pi \Abs{\Bx} } \right) \delta(t - \Abs{\Bx}/c)
+ 
\frac{\delta(t - \Abs{\Bx}/c)}{4 pi \Abs{\Bx} }
\end{align*}

FIXME: on paper.

\section{}

Consider a charged particle moving on a worldline 

\begin{equation}\label{eqn:relativisticElectrodynamicsL18:n}
(c t, \Bx_c(t))
\end{equation}

($c$ for classical)

For this particle 

\begin{align}\label{eqn:relativisticElectrodynamicsL18:n}
\rho(\Bx, t) &= e \delta^3(\Bx - \Bx_c(t)) \\
\Bj(\Bx, t) &= e \dot{\Bx}_c(t) \delta^3(\Bx - \Bx_c(t))
\end{align}

\begin{align*}
\begin{bmatrix}
A^0(\Bx, t) 
\BA(\Bx, t) 
\end{bmatrix}
&= 
\inv{c} 
\int d^3 \Bx' dt' 
\frac{ \delta( t - t' - \Abs{\Bx - \Bx'}/c }{\Abs{\Bx - \Bx'}}
\begin{bmatrix}
c e \\
e \dot{\Bx}_c(t)
\end{bmatrix}
\delta^3(\Bx - \Bx_c(t)) \\
&=
\int_{-\infty}^\infty 
\frac{ \delta( t - t' - \Abs{\Bx - \Bx_c(t')}/c }{\Abs{\Bx_c(t') - \Bx}}
\begin{bmatrix}
e \\
e \frac{\dot{\Bx}_c(t)}{c}
\end{bmatrix}
\end{align*}

PICTURE: light cones, and curved worldline.  Pick an arbitrary point $(\Bx_0, t_0)$, and draw the past light cone, looking at where this intersects with the tradjectory

For the arbitrary point $(\Bx_0, t_0)$ we see that this point and the retarded time $(\Bx_c(t_r), t_r)$ obey the relation

\begin{equation}\label{eqn:relativisticElectrodynamicsL18:n}
c (t_0 - t_r) = \Abs{\Bx_0 - \Bx_c(t_r)}
\end{equation}

This retarded time is unique.  There is only one such intersection.

Our job is to calculate

\begin{equation}\label{eqn:relativisticElectrodynamicsL18:n}
\int_{-\infty}^\infty \delta(f(x)) g(x) = \frac{g(x_\conj)}{f'(x_\conj)}
\end{equation}

where $f(x_\conj) = 0$.

\begin{equation}\label{eqn:relativisticElectrodynamicsL18:n}
f(t') = t - t' - \Abs{\Bx - \Bx_c(t')}/c
\end{equation}

\begin{align*}
\PD{t'}{f} 
&= -1 - \inv{c} \PD{t'}{} \sqrt{ (\Bx - \Bx_c(t')) \cdot (\Bx - \Bx_c(t')) }
&= -1 + \inv{c} \PD{t'}{} \frac{(\Bx - \Bx_c(t')) \cdot \Bv_c(t_r)}{\Abs{\Bx - \Bx_c(t_r)}}
\end{align*}

\EndArticle
