\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY450H1S.  Relativistic Electrodynamics Lecture 14 (Taught by Simon Freedman).  Wave equation in Coulomb and Lorentz gauges.}
\label{chap:relativisticElectrodynamicsL14}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsL14.pdf}
\date{Feb 16, 2011}
\revisionInfo{relativisticElectrodynamicsL14.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Reading.}

Covering chapter 4 material from the text \cite{landau1980classical}.

Covering \href{http://www.physics.utoronto.ca/~poppitz/e-poppitz/PHY450_files/RelEMpp103-114.pdf}{lecture notes pp.103-114}: the wave equation in the relativistic Lorentz gauge (114-114) [Tuesday, Feb. 15; Wednesday, Feb.16]...

Covering \href{http://www.physics.utoronto.ca/~poppitz/e-poppitz/PHY450_files/RelEMpp114-127.pdf}{lecture notes pp. 114-127}: reminder on wave equations (114); reminder on Fourier series and integral (115-117); Fourier expansion of the EM potential in Coulomb gauge and equation of motion for the spatial Fourier components (118-119); the general solution of Maxwell's equations in vacuum (120-121) [Tuesday, Mar. 1]; properties of monochromatic plane EM waves (122-124); energy and energy flux of the EM field and energy conservation from the equations of motion (125-127)  [Wednesday, Mar. 2]

\section{Trying to understand ``c''}

\begin{align}\label{eqn:relativisticElectrodynamicsL14:10}
\spacegrad \cdot \BE &= 0 \\
\spacegrad \cross \BB &= \inv{c} \PD{t}{\BE}
\end{align}

Maxwell's equations in a vacuum were

\begin{align}\label{eqn:relativisticElectrodynamicsL14:30}
\spacegrad (\spacegrad \cdot \BA) &= \spacegrad^2 \BA  -\inv{c} \PD{t}{} \spacegrad \phi - \inv{c^2} \frac{\partial^2 \BA}{\partial t^2} \\
\spacegrad \cdot \BE &= - \spacegrad^2 \phi - \inv{c} \PD{t}{\spacegrad \cdot \BA} 
\end{align}

There's a redundancy here since we can change $\phi$ and $\BA$ without changing the EOM

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:50}
(\phi, \BA) \rightarrow (\phi', \BA')
\end{equation}

with

\begin{align}\label{eqn:relativisticElectrodynamicsL14:70}
\phi &= \phi' + \inv{c} \PD{t}{\chi} \\
\BA &= \BA' - \spacegrad \chi
\end{align}

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:90}
\chi(\Bx, t) = c \int dt \phi(\Bx, t)
\end{equation}

which gives 

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:110}
\phi' = 0
\end{equation}

\begin{align}\label{eqn:relativisticElectrodynamicsL14:130}
(\phi, \BA) \sim (\phi = 0, \BA')
\end{align}

Maxwell's equations are now

\begin{align*}
\spacegrad (\spacegrad \cdot \BA') &= \spacegrad^2 \BA'  - \inv{c^2} \frac{\partial^2 \BA'}{\partial t^2} \\
\PD{t}{\spacegrad \cdot \BA'}  &= 0
\end{align*}

Can we make $\spacegrad \cdot \BA'' = 0$, while $\phi'' = 0$.

\begin{align}\label{eqn:relativisticElectrodynamicsL14:150}
\underbrace{\phi}_{=0} &= \underbrace{\phi'}_{=0} + \inv{c} \PD{t}{\chi'} \\
\end{align}

We need 
\begin{equation}\label{eqn:relativisticElectrodynamicsL14:170}
\PD{t}{\chi'} = 0
\end{equation}

How about $\BA'$

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:190}
\BA' = \BA'' - \spacegrad \chi'
\end{equation}

We want the divergence of $\BA'$ to be zero, which means

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:210}
\spacegrad \cdot \BA' = \underbrace{\spacegrad \cdot \BA''}_{=0} - \spacegrad^2 \chi'
\end{equation}

So we want

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:230}
\spacegrad^2 \chi' = \spacegrad \cdot \BA'
\end{equation}

Can we solve this?

Recall that in electrostatics we have

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:250}
\spacegrad \cdot \BE = 4 \pi \rho
\end{equation}

and 

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:270}
\BE = -\spacegrad \phi
\end{equation}

which meant that we had 

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:290}
\spacegrad^2 \phi = 4 \pi \rho
\end{equation}

This has the identical form (with $\phi \sim \chi$, and $4 \pi \rho \sim \spacegrad \cdot \BA'$).

While we aren't trying to actually solve this (just show that it can be solved).  One way to look at this problem is that it is just a Laplace equation, and we could utilize a Green's function solution if desired.

FIXME: What's the Green's function?

What are the Maxwell's vacuum equations now?

With the second gauge substitution we have

\begin{align*}
\spacegrad (\spacegrad \cdot \BA'') &= \spacegrad^2 \BA''  - \inv{c^2} \frac{\partial^2 \BA''}{\partial t^2} \\
\PD{t}{\spacegrad \cdot \BA''}  &= 0
\end{align*}

but we can utilize

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:310}
\spacegrad \cross (\spacegrad \cross \BA) = \spacegrad (\spacegrad \cdot \BA) - \spacegrad^2 \BA,
\end{equation}

to reduce Maxwell's equations (after dropping primes) to just

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:330}
\inv{c^2} \frac{\partial^2 \BA''}{\partial t^2} - \Delta \BA = 0
\end{equation}

where 
\begin{equation}\label{eqn:relativisticElectrodynamicsL14:350}
\Delta = \spacegrad^2 = \spacegrad \cdot \spacegrad = 
\frac{\partial^2}{\partial x^2}
+\frac{\partial^2}{\partial y^2}
+\frac{\partial^2}{\partial y^2}
\end{equation}

Note that for this to be correct we have to also explicitly include the gauge condition used.  This particular gauge is called the \underline{Coulomb gauge}.

\begin{align}\label{eqn:relativisticElectrodynamicsL14:370}
\phi &= 0 \\
\spacegrad \cdot \BA'' &= 0 
\end{align}

\section{Claim: EM waves propagate with speed $c$ and are transverse.}

\paragraph{Note:} Is the Coulomb gauge Lorentz invariant?
\paragraph{No.} We can boost which will introduce a non-zero $\phi$.

The gauge that is Lorentz Invariant is the ``Lorentz gauge''.  This one uses

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:390}
\partial_i A^i = 0
\end{equation}

Recall that Maxwell's equations are

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:410}
\partial_i F^{ij} = j^j = 0
\end{equation}

where 

\begin{align}\label{eqn:relativisticElectrodynamicsL14:430}
\partial_i &= \PD{x^i}{} \\
\partial^i &= \PD{x_i}{}
\end{align}

Writing out the equations in terms of potentials we have
\begin{align*}
0 &= \partial_i (\partial^i A^j - \partial^j A^i)  \\
&= \partial_i \partial^i A^j - \partial_i \partial^j A^i \\
&= \partial_i \partial^i A^j - \partial^j \partial_i A^i \\
\end{align*}

So, if we pick the gauge condition $\partial_i A^i = 0$, we are left with just 

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:450}
0 = \partial_i \partial^i A^j
\end{equation}

Can we choose ${A'}^i$ such that $\partial_i A^i = 0$?

Our gauge condition is 

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:470}
A^i = {A'}^i + \partial^i \chi
\end{equation}

Hit it with a derivative for

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:490}
\partial_i A^i = \partial_i {A'}^i + \partial_i \partial^i \chi
\end{equation}

If we want $\partial_i A^i = 0$, then we have

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:510}
-\partial_i {A'}^i = \partial_i \partial^i \chi = \left( \inv{c^2} \frac{\partial^2}{\partial t^2} - \Delta \right) \chi
\end{equation}

This is the physicist proof.  Yes, it can be solved.  To really solve this, we'd want to use Green's functions.

Returning to Maxwell's equations we have

\begin{align}\label{eqn:relativisticElectrodynamicsL14:530}
0 &= \partial_i \partial^i A^j \\
0 &= \partial_i A^i ,
\end{align}

where the first is Maxwell's equation, and the second is our gauge condition.

Observe that the gauge condition is now a Lorentz scalar.

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:550}
\partial^i A_i \rightarrow \partial^j {O_j}^i {O_i}^k A_k
\end{equation}

But the Lorentz transform matrices multiply out to identity, in the same way that they do for the transformation of a plain old four vector dot product $x^i y_i$.

\section{What happens with a Massive vector field?}

FIXME: go through this section and fill in the details.

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:570}
S = \int d^4 x \left( \inv{4} F^{ij} F_{ij} + \frac{m^2}{2} A^i A_i \right)
\end{equation}

We call this a mass term because the units require it.

\begin{align*}
\delta S 
&= 0  \\
&= \int d^4 x \left( \partial^i \partial_i A_j - \partial_j \partial_i A^i \delta A^i + m^2 A_i \delta A^i \right)
\end{align*}

We get 

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:590}
\left( (\Delta - m^2) g_{ij} - \partial_i \partial_j \right) A^i = 0
\end{equation}

FIXME: The EOM (DIY)

\begin{equation}\label{eqn:relativisticElectrodynamicsL14:610}
m^2 \partial_i A^i = 0
\end{equation}

With the mass term, the Lorentz gauge is already chosen for us?

\EndArticle
