\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}
%\usepackage{cancel}

\chapter{PHY450H1S.  Relativistic Electrodynamics Lecture 26 (Taught by Prof. Erich Poppitz).  FIXME.}
\label{chap:relativisticElectrodynamicsL26}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsL26.pdf}
\date{April 5, 2011}
\revisionInfo{relativisticElectrodynamicsL26.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Reading.}

Covering chapter 8 \S 65 material from the text \cite{landau1980classical}.

Covering \href{http://www.physics.utoronto.ca/~poppitz/epoppitz/PHY450_files/RelEMpp181-195.pdf}{pp. 181-195}: (182-189) [Tuesday, Mar. 29]; the EM potentials to order $(v/c)^2$ (190-193); the ``Darwin Lagrangian.  and Hamiltonian for a system of nonrelativistic charged particles to order $(v/c)^2$ and its many uses in physics (194-195) [Wednesday, Mar. 30]

Next week (last topic): attempt to go to the next order $(v/c)^3$ - radiation damping, the limitations of classical electrodynamics, and the relevant time/length/energy scales.

\section{Recap.}

A system of N charged particles $m_a, q_a ; a \in [1, N]$ closed system and nonrelativistic, $v_a/c \ll 1$.  In this case we can incorporate EM effects in a Largrangian \underline{ONLY} involving particles (EM field not a dynamical DOF).  In \underline{general case}, this works to $O((v/c)^2)$, because at $O((v/c))$ system radiation effects occur.

In a specific case, when
 
\begin{equation}\label{eqn:relativisticElectrodynamicsL26:n}
\frac{m_1}{q_1} = \frac{m_2}{q_2} = \frac{m_3}{q_3} = \cdots
\end{equation}

we can do that (meaning use a Lagrangian with particles only) to O((v/c)^4) brecause 



The Lagrangian for our particle after the gauge transformation is

\begin{equation}\label{eqn:relativisticElectrodynamicsL25:330b}
\LL_a = \inv{2} m_a \Bv_a^2 - \frac{m_a}{8} \frac{\Bv_a^4}{c^2} 
-\sum_{b \ne a} \frac{q_a q_b}{\Abs{\Bx_a(t) - \Bx_b(t)}}
+\sum_b q_a q_b \frac{\Bv_a \cdot \Bv_b + (\Bn \cdot \Bv_a) (\Bn \cdot \Bv_b)}{2 c^2 \Abs{\Bx - \Bx_b} }.
\end{equation}

Next time we'll probably get to the Lagrangian for the entire system.  It was hinted that this is called the Darwin Lagrangian (after Charles Darwin's grandson).

We find for whole system

\begin{equation}\label{eqn:relativisticElectrodynamicsPS6P1:n}
\LL =  \sum_a \LL_a + \inv{2} \sum_a \LL_a (interaction)
\end{equation}

\begin{equation}\label{eqn:relativisticElectrodynamicsPS6P1:n}
\LL = \inv{2} \sum_a m_a \Bv_a^2 - \sum_a \frac{m_a}{8} \frac{\Bv_a^4}{c^2} 
-\sum_{ a < b} \frac{q_a q_b}{\Abs{\Bx_a(t) - \Bx_b(t)}}
+\sum_b q_a q_b \frac{\Bv_a \cdot \Bv_b + (\Bn \cdot \Bv_a) (\Bn \cdot \Bv_b)}{2 c^2 \Abs{\Bx - \Bx_b} }.
\end{equation}

This is the Darwin Lagrangian (also Charles).  The Darwin Hamiltonian is then

\begin{equation}\label{eqn:relativisticElectrodynamicsPS6P1:n}
H = \sum_a \frac{p_a}{2 m_a} \Bv_a^2 - \sum_a \frac{p_a^4}{8 m_a^3 c^2} 
+\sum_{ a < b} \frac{q_a q_b}{\Abs{\Bx_a(t) - \Bx_b(t)}}
- \sum_{ a < b } \frac{q_a q_b}{ 2 c^2 m_a m_b } \frac{\Bp_a \cdot \Bp_b + (\Bn_{a b} \cdot \Bp_a) (\Bn_{ a b } \cdot \Bp_b)}{\Abs{\Bx - \Bx_b} }.
\end{equation}

\section{Incorporating radiation effects as a friction term.}

To $O((v/c)^3)$ obvious problem due to radiation (system not closed).  We'll incorporate radiation via a function term in the EOM

Again consider the dipole system

m \ddot{z} = -k z
\omega^2 = \frac{k}{m}

m \ddot{z} = -\omega^2 m z

gives

\ddt{}\left( \frac{m}{2} \zdot^2 + \frac{ m \omega^2 }{2} z^2 \right) = 0 

(because there's no radiation).

The energy radiated per unit time averaged per period is

P = \frac{ 2 e^2 }{ 3 c^3} \Abs{\ddot{z}}^2

We'll modify the EOM

m \ddot{z} = -\omega^2 m z + f_{\text{radiation}}

m \ddot{z} \dot{z} = -\omega^2 m z \dot{z} + f_{\text{radiation}} \dot{z}

\ddt{} 
\left( m \dot{z}^2 + \omega^2 m z^2 \right) 
= 
f_{\text{radiation}} \dot{z} \equiv \ddt{\mathcal{E}}

Average over period both sides

\expectation{ \ddt{\mathcal{E}} } = 
\expectation{ f_{\text{radiation}} \dot{z} }
=
- \frac{2 e^2 }{3 c^3} \Abs{\ddot{z}^2}

(where we demand this last equality).

\paragraph{Claim:} 

f_{\text{radiation}} = \frac{2 e^2 }{3 c^3} \dddot{z}

\paragraph{Proof:} 

We need to show 

\expectation{ f_{\text{radiation}} }
= - \frac{2 e^2 }{3 c^3} \Abs{\ddot{z}^2}


CONTINUED ON PAPER.

\EndArticle
