\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY450H1S.  Relativistic Electrodynamics Lecture 23 (Taught by Prof. Erich Poppitz).  Energy Momentum Tensor.}
\label{chap:relativisticElectrodynamicsL23}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsL23.pdf}
\date{Mar 23, 2011}
\revisionInfo{relativisticElectrodynamicsL23.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Reading.}

FIXME:
Covering chapter 8 material from the text \cite{landau1980classical}.

Covering \href{http://www.physics.utoronto.ca/~poppitz/epoppitz/PHY450_files/RelEMpp166-180.pdf}{lecture notes pp. 173-178:} the force on a surface element of a body (177-178); a plane wave example (179-180).

%http://www.physics.utoronto.ca/~poppitz/epoppitz/PHY450_files/RelEMpp181-195.pdf
pp. 181-195: [read on your own: on diagonalizability of energy-momentum tensor (181)]; the Lagrangian for a system of nonrelativistic charged particles to zeroth order in $(v/c)$: electrostatic energy of a system of charges and .mass renormalization. (182-189) [Tuesday, Mar. 29]; the EM potentials to order $(v/c)^2$ (190-193); the ``Darwin Lagrangian.  and Hamiltonian for a system of nonrelativistic charged particles to order $(v/c)^2$ and its many uses in physics (194-195) [Wednesday, Mar. 30]

Next week (last topic): attempt to go to the next order $(v/c)^3$ - radiation damping, the limitations of classical electrodynamics, and the relevant time/length/energy scales.

\section{Recap.}

Last time we found that spacetime translation invariance led to the four conservation relations

\partial_k T^{k m} = 0

where

T^{k m} = \inv{4 \pi} \left( F^{k j} F^{m l} g_{j l} + \inv{4} g^{k m} F_{i j} F^{i j}

last time we found for $m = 0$

\inv{c} \PDt T^{0 0} + \PD{x^\alpha}{} T^{\alpha 0} = 0

Here

T^{0 0} &= \inv{8 \pi} (\BE^2 + \BB^2) & \mbox{energy density} \\
c T^{\alpha 0} &= \BS^\alpha &= \mbox{energy flux}

\section{Spatial components of $T^{k m}$}

Now for $m = 1,2,3$ we write

\partial_k T^{k \alpha} = 0

so we write

\PDt \BS^\alpha/c^2 + \partial_\beta T^{\beta \alpha} = 0

\EndArticle
