\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY450H1S.  Relativistic Electrodynamics Lecture 10 (Taught by Prof. Erich Poppitz).  FIXME.}
\label{chap:relativisticElectrodynamicsL10}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsL10.pdf}
\date{Feb 8, 2011}
\revisionInfo{relativisticElectrodynamicsL10.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Reading.}

Covering chapter N material from the text \cite{landau1980classical}.

Covering \href{}{}

http://www.physics.utoronto.ca/~poppitz/e-poppitz/PHY450_files/RelEMpp74-83.pdf

pp. 74-83: gauge transformations in 3-vector language (74); energy of a relativistic particle in EM field (75); variational
principle and equation of motion in 4-vector form (76-77); the field strength tensor (78-80); the fourth equation of motion (81);  Lorentz transformation of the strength tensor (82) [Tuesday, Feb. 8] [extra reading for the mathematically minded: gauge field, strength tensor, and gauge transformations in differential form language, not to be covered in class (83)] 

http://www.physics.utoronto.ca/~poppitz/e-poppitz/PHY450_files/RelEMpp84-102.pdf

pp. 84-102: Lorentz invariants of the electromagnetic field (84-86); Bianchi identity and the first half of Maxwell’s equations (87-90); relativity, gauge invariance, and superposition principles and the action for the electromagnetic field coupled to charged particles (91-95); the 4-current and its physical interpretation (96-102), including a needed mathematical interlude on delta-functions of functions (98-100) [Wednesday, Feb. 8; Thursday, Feb. 10]

http://www.physics.utoronto.ca/~poppitz/e-poppitz/PHY450_files/RelEMpp103-113.pdf

pp.103-113: variational principle for the electromagnetic field and the relevant boundary conditions (103-105); the second set of Maxwell’s equations from the variational principle (106-108); Maxwell’s equations in vacuum and the wave equation in the nonrelativistic Coulomb gauge (109-111); the wave equation in the relativistic Lorentz gauge (112-113) [Tuesday, Feb. 15; Wednesday, Feb.16]...

\section{.}

\section{What is the significance to this claim?}

Having argued that under a gauge transformation

A_i \rightarrow A_i + \partial_i \chi

the action for a particle changes by a boundary term 

\begin{equation}\label{eqn:lorentzForceHamiltonian:n}
- \frac{e}{c} ( \chi(x_b) - \chi(x_a) )
\end{equation}

Because $S$ changes by a boundary term only, variation problem is not affected.  The extremal trajectories are then the same, hence the EOM are the same.

\paragraph{A less high brow demonstration}

A_i = (\phi, -\BA)
A^i = (\phi, \BA)

A_0 -> A_0 + \PD{x^0}{\chi}
-\BA -> -\BA + \PD{\Bx}{\chi}

\phi -> \phi + \inv{c}\PD{t}{\chi}
\BA -> \BA - \spacegrad \chi

E = - \grad \phi - \inv{c} \pdt \BA -> 
- \grad \phi - \inv{c} \grad \pdt \chi 
- \grad \phi - \inv{c} \pdt \BA 
...
= E

B = \spacegrad \cross BA 
-> 
\spacegrad \cross BA  - \spacegrad (\cross \spacegrad \chi)

EM fields (as opposed to potentials) do not change under grauge transformations

{A_i} description is hugely redundant

Despite that, local \LL and H can only be written into A_i

\LL = - mc^2 \InvGamma + \fracec \BA \cdot \Bv - e \phi

we define the energy as 

\E = \Bv \cdot \pdV \LL - \LL

This is not neccessarily a conserved quantity, but we define it as the energy anyways (we don't really have a Hamiltonian when the fields are time dependent).  In general we have

\ddt \E = -pdt \LL

\E = \underbrace{\frac{m c^2}{\InvGamma}}_{\conj} + \Bv \cdot \pdV \left( \frac{e}{c} \BA \cdot \Bv - e \phi
\right)

The contribution of $\conj$ to $\E$ of $\LL = -m c^2 \InvGamma$

\E = \frac{m c^2}{\InvGamma} + \underbrac{e \phi}_{\text{"potential"}}

\frac{d}{dt} \E_\{\text{kinetic}} = e \BE \cdot \Bv

One way that this follows is from 

\ddt \Bp = e \left( \BE + ... ) lorentz force

\section{}

Using $ds = \sqrt{ dx^i dx_i } $ our action can be rewritten

S 
&= \int ( -m c ds - \fracec u^i A_i ds ) \\
&= \int ( -m c ds - \fracec dx^i A_i ) \\
&= \int ( -m c \sqrt{ dx^i dx_i} - \fracec dx^i A_i ) \\

x^i(\tau) is a worldline $x^i(0) = a^i$, $x^i(1) = b^i$, 

We want $\delta S = S[ x + \delta x ] - S[ x ] = 0$ (to linear order in $\delta x$)

\delta \sqrt{ dx^i dx_i } 
&= \inv{ 2 \sqrt{ dx^i dx_i }} \delta (dx^j dx_j)

\delta (dx^j dx_j) 
&= \delta ( dx^j g_{jk} dx^k ) \\
&= \delta ( dx^j ) g_{jk} dx^k ) + dx^j g_{jk} \delta ( dx^k ) \\
&= \delta ( dx^j ) g_{jk} dx^k ) + dx^k g_{kj} \delta ( dx^j ) \\
&= 2 \delta ( dx^j ) g_{jk} dx^k ) \\
&= 2 \delta ( dx^j ) dx_j 

\paragraph{TIP:} If this goes to quick, or there is any disbelief, write these all out explicitly as $dx^j dx_j = dx^0 dx_0 + dx^1 dx_1 + dx^2 dx_2 + dx^3 dx_3$ and compute it that way.

\delta A_i = A_i(x + \delta x) - A_i = \pd{x^j}{A_i} \delta x^j

(chain rule)

Completing the above we have
\delta \sqrt{ dx^i dx_i } 
= \inv{ \sqrt{ dx^i dx_i }} \delta (dx^j) dx_j

and our action variance is now

\delta S 
&= \int ( -m c \delta \sqrt{ dx^i dx_i} - \fracec d \delta x^i A_i -fracec dx^i \delta A_i ) \\
&= \int ( -m c \frac{ dx^j g_{jk} d \delta x^k }{\sqrt{ dx^i dx_i} - \fracec d \delta x^i A_i - \fracec dx^i \pd{x^j}{A_i} \delta x^j ) \\
&= \int ( -m c u^j g_{jk} d \delta x^k - \fracec d \delta x^i A_i - \fracec dx^i \pd{x^j}{A_i} \delta x^j ) \\
&= \int ( -m c u^j g_{jk} \delta x^k + d u^j g_{jk} \delta x^k m c + d ( -\fracec \delta x^i A_i + \fracec \delta x^i d A_i ) - \fracec dx^i \pd{x^j}{A_i} \delta x^j ) \\
&= \int ( -m c u^j g_{jk} d \delta x^k + d u^j g_{jk} + d ( -\fracec \delta x^i A_i + \fracec \delta x^i d A_i ) - \fracec dx^i \pd{x^j}{A_i} \delta x^j ) \\

\delta S = 
\int d ( - m c u^j g_{jk} \delta x^k - \fracec A_i \delta x^i
+ \int m c d u^i g_{ij} \delta x^j + \fracec \delta x^l d A_l - \fracec \delta x^i \pd x^i A_j d x^j

The first term is

 - mc u^i \delta x_i - \fracec A_i \delta x^i \vert^b_a = 0

since \delta x^i |vert_a = \delta x^i \vert_b = 0 (the variation at the endpoints is zero)

We are left with

\int ds m c \dds u^j g_{jk} \delta x^k + \fracec \delta x^l \pd x^k A_l \dds x^k ds - \fracec \pd x^l A_i \dds x^i ds
\int ds \delta x^l  ...

\delta S = \int ds \delta x^l ( m c \dds u_l + \fracec u^k ( \partial_k A_l - \partial_l A_k )

Since this is true for all variations $\delta x^l$, which is arbitrary, the interior part is zero everywhere in the tragectory

m c \dds u_l 
= - \fracec ( \pd x^l A_k - \pd x_k A_l ) u^k
= \fracec ( \pd x^k A_l - \pd x_l A_k ) u^k

m c \dds u_l = \fracec ( \underbrace{ \pd x^k A_l - \pd x_l A_k }_{F_{lk} }) u^k

The quantity $F_{lk}$ is called the electromagnetic field strength tensor ( 4-tensor, rank-2 antisymmetric ).

We write

\dds m c u_l = \fracec F_{lk} u^k 

\Abs{ F_{lk} } = 
\begin{bmatrix}
0 & E_x & E_y & E_z \\
-E_x & 0 & -B_z & B_y \\
-E_y & B_z & 0 & -B_x \\
-E_z & -B_y & B_x & 0
\end{bmatrix}

\EndArticle
