\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY450H1S.  Relativistic Electrodynamics Tutorial 9 (TA: Simon Freedman).  Some worked problems.  EM reflection.  Stress energy tensor for simple configurations.}
\label{chap:relativisticElectrodynamicsT9}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsT9.pdf}
\date{Mar 30, 2011}
\revisionInfo{relativisticElectrodynamicsT9.tex}

\beginArtWithToc
%\beginArtNoToc

\section{HW6. Question 3. (Non subtle hints about how important this is (i.e. for the exam)}

This is problem 1 from \S 47 of the text \cite{landau1980classical}.

Determine the force exerted on a wall from which an incident plane EM wave is reflected (w/ reflection coefficient $R$) and incident angle $\theta$.

Solution from the book

\begin{equation}\label{eqn:relativisticElectrodynamicsT9:10}
f_\alpha = - \sigma_{\alpha \beta} n_\beta - {\sigma'}_{\alpha \beta} n_\beta
\end{equation}

Here $\sigma_{\alpha \beta}$ is the Maxwell stress tensor for the incident wave, and ${\sigma'}_{\alpha \beta}$ is the Maxwell stress tensor for the reflected wave, and $n_\beta$ is normal to the wall.

\begin{align}\label{eqn:relativisticElectrodynamicsT9:30}
f_\perp &= - \sigma_{\perp \perp} - {\sigma'}_{\perp \perp} \\
f_\parallel &= - \sigma_{\parallel \perp} - {\sigma'}_{\parallel \perp} 
\end{align}

But stating this solution is very non-trivial, because $\sigma_{\alpha \beta}$ is non-linear in $\BE$ and $\BB$.

FIXME: PICTURE: 

The Reflection coefficient can be defined in this case as

\begin{equation}\label{eqn:relativisticElectrodynamicsT9:50}
\Abs{\BE}^2 = R \Abs{\BE'}
\end{equation}

(really a ratio of powers, but those are proportional to $\BE^2$ and ${\BE'}^2$.

Incident:

\begin{align}\label{eqn:relativisticElectrodynamicsT9:70}
E_x &= E \sin\theta \\
E_y &= -E \cos\theta \\
B_z &= E 
\end{align}

Reflected:
\begin{align}\label{eqn:relativisticElectrodynamicsT9:90}
E_x' &= E' \sin\theta \\
E_y' &= E' \cos\theta \\
B_z' &= E'
\end{align}

\begin{align}\label{eqn:relativisticElectrodynamicsT9:110}
E &= E_0 \cos(\Bp \cdot \Bx - \omega t) \\
E' &= \sqrt{R} E_0 \cos(\Bp \cdot \Bx - \omega t)
\end{align}

\begin{equation}\label{eqn:relativisticElectrodynamicsT9:130}
\sigma_{\alpha\beta} = -T^{\alpha\beta} = \inv{4\pi} \left(
\mathcal{E}^\alpha
\mathcal{E}^\beta
+\mathcal{B}^\alpha
\mathcal{B}^\beta
- \inv{2} \delta^{\alpha\beta} ( \vec{\mathcal{E}}^2 + \vec{\mathcal{B}}^2 )
\right)
\end{equation}

Where $\vec{\mathcal{E}}$ and $\vec{\mathcal{B}}$ are the total EM fields.

FIXME: why the total here?

\begin{align}\label{eqn:relativisticElectrodynamicsT9:150}
\vec{\mathcal{E}} &= \BE + \BE' \\
\vec{\mathcal{B}} &= \BB + \BB'
\end{align}

Get 

\begin{align}\label{eqn:relativisticElectrodynamicsT9:170}
\sigma_{3 3} &= \inv{4 \pi} \left( \mathcal{E}^2 - \inv{2} (\mathcal{E}^2 + \mathcal{B}^2) \right) = 0 \\
\sigma_{3 1} &= 0 = \sigma_{3 2} \\
\sigma_{1 1} &= \inv{4 \pi} \left( (\mathcal{E}')^2 - \inv{2} (\mathcal{E}^2 + \mathcal{B}^2) \right) = 0 \\
\end{align}

\begin{align}\label{eqn:relativisticElectrodynamicsT9:190}
\vec{\mathcal{B}}^2 &= (B_z + B_z')^2 = (E + E')^2 \\
\vec{\mathcal{E}}^2 &= (\BE + \BE')^2
\end{align}

so
\begin{align*}
\sigma_{1 1} 
&= \inv{4 \pi} \left( (\mathcal{E}')^2 - \inv{2} ((\mathcal{E}')^2 + \mathcal{E}^2 + (E + E')^2 \right) \\
&= \inv{8 \pi} \left( (\mathcal{E}')^2 - (\mathcal{E})^2 - (E + E')^2 \right) \\
&= \inv{8 \pi} \left( 
(E + E')^2 \sin^2\theta 
-(E - E')^2 \cos^2\theta -(E + E')^2
\right) \\
&= \inv{8 \pi} \left( 
E^2( \sin^2\theta - \cos^2\theta - 1)
(E')^2( \sin^2\theta - \cos^2\theta - 1)
+ 2 E E' (\sin^2\theta + \cos^2\theta -1 )
\right) \\
&= \inv{8 \pi} \left( - 2 E^2 \cos^2\theta - 2 (E')^2 \cos^2\theta \right) \\
&= -\inv{4 \pi} (E^2 + (E')^2) \cos^2\theta \\
&= \sigma_\parallel + {\sigma'}_\parallel
\end{align*}

For $\sigma_{12}$ we have

\begin{align*}
\sigma_{12} 
&= \inv{4 \pi} \mathcal{E}^1 \mathcal{E}^2 \\
&= \inv{4 \pi} (E + E') \sin\theta (-E + E') \cos\theta \\
&= -\inv{4 \pi} (E^2 + (E')^2) \sin\theta \cos\theta 
\end{align*}

So we have

\begin{equation}\label{eqn:relativisticElectrodynamicsT9:210}
\sigma_{\alpha\beta}^{\text{total}} n_\beta = \sigma_{\alpha 1 }^{\text{total}} = \sigma_{\alpha 1} + {\sigma'}_{\alpha 1}
\end{equation}

(where $n_\beta = \delta^{\beta 1}$)

This result, assumed in the text, was non-trivial to derive.  It is also not generally true.  We have 

\begin{align*}
\sigma_{2 2} 
&= \inv{4 \pi} \left( (\mathcal{E}^y)^2 - \inv{2} ( \vec{\mathcal{E}}^2 + \vec{\mathcal{B}}^2 ) \right) \\
&= \inv{8 \pi} \left( (\mathcal{E}^y)^2 - \vec{\mathcal{E}}^2 + \vec{\mathcal{B}}^2 ) \right) \\
&= \inv{8 \pi} \left( 
E^2 ( -1 + \cos^2 \theta - \sin^2\theta )
+{E'}^2 ( -1 + \cos^2 \theta - \sin^2\theta )
+ 2 E E' ( -\cos^2\theta + \sin^2\theta -1 ) \right)
&= -\inv{4 \pi} \left( E^2 \sin^2 \theta + (E')^2 \sin^2 \theta + 2 E E' \right)
\end{align*}

\subsection{Force per unit area?}

\begin{equation}\label{eqn:relativisticElectrodynamicsT9:230}
f_\alpha = n^x \sigma_{x \alpha} 
\end{equation}

Averaged

\begin{align}\label{eqn:relativisticElectrodynamicsT9:250}
\expectation{\sigma_{xx}} &= -\inv{8 \pi} E_0^2 ( 1 + R) \cos^2\theta \\
\expectation{\sigma_{xy}} &= -\inv{8 \pi} E_0^2 ( 1 - R) \sin\theta \cos\theta
\end{align}

\begin{align}\label{eqn:relativisticElectrodynamicsT9:270}
\expectation{\BS} &= -\frac{c}{8 \pi} E_0^2 \ncap \\
\expectation{\BS'} &= -\frac{c}{8 \pi} E_0^2 \ncap'
\end{align}

\begin{equation}\label{eqn:relativisticElectrodynamicsT9:290}
\expectation{\Abs{\BS}} = \text{Work} = W
\end{equation}

\begin{align}\label{eqn:relativisticElectrodynamicsT9:310}
f_x &= n^x \sigma_{x x} = W (1 + R) \cos^2\theta \\
f_y &= n^y \sigma_{x y} = W (1 - R) \sin\theta \cos\theta
f_z &= 0
\end{align}

\section{A problem from Griffiths.}

FIXME: try this.

Two charges $q+$, $q-$ reflected in a plane, separated by distance $a$.  Work out the stress energy tensor from the Coulomb fields of the charges on the plane.

Will get the Coulomb force:

\begin{equation}\label{eqn:relativisticElectrodynamicsT9:330}
\BF = k \frac{q^2}{2 a^2}.
\end{equation}

\section{Infinite parallel plate capacitor}

Write $\sigma_{\alpha\beta}$.

\begin{align}\label{eqn:relativisticElectrodynamicsT9:350}
\BB &= 0  \\
\BE &= - \frac{\sigma}{\epsilon_0} \Be_z
\end{align}

FIXME: derive this.  Observe that we have no distance dependence in the field because it is an infinite plate.

\begin{align}\label{eqn:relativisticElectrodynamicsT9:370}
\sigma_{1 1} &= \left( - \inv{2} \delta^{1 1} \left( \frac{-\sigma}{\epsilon_0} \right)^2 \right) = - \frac{ \sigma^2}{ 2 \epsilon_0^2 } = \sigma_{22} \\
\sigma_{3 3} &= \left( (E^3)^2 - \inv{2} \BE^2 \right)  = - \inv{2} \BE^2 = - \sigma_{2 2}
\end{align}

Force per unit area is then

\begin{align*}
f_\alpha 
&= n_\beta \sigma_{\alpha \beta} \\
&= n_3 \sigma_{\alpha 3}
\end{align*}

So 

\begin{align}\label{eqn:relativisticElectrodynamicsT9:390}
f_1 &= 0 = f_2 \\
f_3 &= \sigma_{3 3} = -\frac{\sigma^2}{2 \epsilon_0^2}
\end{align}

\begin{equation}\label{eqn:relativisticElectrodynamicsT9:410}
\Bf = -\frac{\sigma^2}{2 \epsilon_0^2} \Be_z
\end{equation}

REMEMBER: EXAM WEDNESDAY!
%  Simon Available on Monday.

\EndArticle
