\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY450H1S.  Relativistic Electrodynamics Lecture 19 (Taught by Prof. Erich Poppitz).  Lienard-Wiechert potentials.}
\label{chap:relativisticElectrodynamicsL19}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsL19.pdf}
\date{Mar 15, 2011}
\revisionInfo{relativisticElectrodynamicsL19.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Reading.}

Covering chapter 8 material from the text \cite{landau1980classical}.

FIXME:
Covering \href{http://www.physics.utoronto.ca/~poppitz/epoppitz/PHY450_files/RelEMpp136-146.pdf}{lecture notes pp. 136-146}: the Lienard-Wiechert potentials (143-146) [Wednesday, Mar. 9...]

pp. 147-165: EM fields of a moving source (147-148+HW5); a particle at rest (148); a constant velocity particle (149-152); behavior of EM fields ``at infinity'' for a general-worldline source and radiation (152-153) [Tuesday, Mar. 15]; radiated power (154); fields in the ``wave zone'' and discussions of approximations made (155-159); EM fields due to electric dipole radiation (160-163); Poynting vector, angular distribution, and power of dipole radiation (164-165) [Wednesday, Mar. 16...]

\section{.}

To find $\BE$ and $\BB$ need 

$\PD{t}{t_r}$, and $\spacegrad t_r(\Bx, t)$

where 

\begin{equation}\label{eqn:relativisticElectrodynamicsT4:n}
t - t_r = \Abs{\Bx - \Bx_c(t_r) }
\end{equation}

implicit definition of $t_r(\Bx, t)$

In HW5 you'll show

\begin{equation}\label{eqn:relativisticElectrodynamicsT4:n}
\PD{t}{t_r} = \frac{\Abs{\Bx - \Bx_c(t_r) }}{
\Abs{\Bx - \Bx_c(t_r) } - \frac{\Bv_c }{c} \cdot (\Bx - \Bx_c(t_r))}
\end{equation}

\begin{equation}\label{eqn:relativisticElectrodynamicsT4:n}
\spacegrad t_r = \inv{c} \frac{\Bx - \Bx_c(t_r) }{
\Abs{\Bx - \Bx_c(t_r) } - \frac{\Bv_c }{c} \cdot (\Bx - \Bx_c(t_r))}
\end{equation}

\begin{equation}\label{eqn:relativisticElectrodynamicsT4:n}
\BE(\Bx, t) = \frac{e R}
\end{equation}

\EndArticle
