\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY450H1S.  Relativistic Electrodynamics Tutorial 1 (TA: Simon Freedman).}
\label{chap:relativisticElectrodynamicsT1}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsT1.pdf}
\date{Jan 20, 2011}
\revisionInfo{relativisticElectrodynamicsT1.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Worked question.}

The TA blasted through a problem from Hartle \cite{hartle2003gravity}, section 5.17 (all the while appologizing for going so slow).  I'm going to have to look these notes over carefully to figure out what on Earth he was doing.

\paragraph{Q:}
In a source's rest frame $S$ emits radiation isotripically with a frequency $\omega$ with number flux $f(\text{photons}/\text{cm}^2 s)$.  Moves along x'-axis with speed $V$ in an observer frame ($O$).  What does the energy flux in $O$ look like?

\subsection{A brief intro with four vectors}

A 3-vector: 

\begin{align}\label{eqn:relativisticElectrodynamicsT1:10}
\Ba &= (a_x, a_y, a_z) = (a^1, a^2, a^3) \\
\Bb &= (b_x, b_y, b_z) = (b^1, b^2, b^3)
\end{align}

For this we have the dot product
\begin{equation}\label{eqn:relativisticElectrodynamicsT1:20}
\Ba \cdot \Bb = \sum_{\alpha=1}^3 a^\alpha b^\alpha
\end{equation}

Greek letters in this course (opposite to everybody else in the world, because of Landau and Lifshitz) run from 1 to 3, whereas roman letters are in $0,1,2,3$.

Our interval is 

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:30}
s^2 = (ct)^2 - \Br^2
\end{equation}

We write
\begin{equation}\label{eqn:relativisticElectrodynamicsT1:40}
x^i = (ct, \Br) = (x^0, x^1, x^2, x^3),
\end{equation}

and
\begin{equation}\label{eqn:relativisticElectrodynamicsT1:50}
X \cdot X = (ct)^2 - \Br^2 = (x^0)^2 - \sum_{\alpha=1}^3 (x^\alpha)^2
\end{equation}

Notationally, three vectors in this class will be designated with arrows (although in my notes I'll use bold letters).


\begin{equation}\label{eqn:relativisticElectrodynamicsT1:60}
{S_{X_1, X_2}}^2 = (ct_1 - c t_2)^2 - (\Bx_1 - \Bx_2)^2 = (X_1 - X_2) \cdot (X_1 - X_2)
\end{equation}

So we have

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:70}
X_1 \cdot X_2 = 
x_1^0 x_2^0
- \sum_{\alpha =1}^3 x_1^\alpha x_2^\alpha
\end{equation}

With
\begin{align}\label{eqn:relativisticElectrodynamicsT1:80}
x^0 &= ct \\
x^1 &= x \\
x^2 &= y \\
x^3 &= z 
\end{align}

And these transform as

\begin{align}\label{eqn:relativisticElectrodynamicsT1:90}
{x^0}' &= ct' = \gamma ( ct - \beta x) = \gamma ( x^0 - \beta x^1 ) \\
{x^1}' &= x' = \gamma ( x - \beta ct ) = \gamma ( x^1 - \beta x^0 ) 
\end{align}

We'd write 
\begin{align}\label{eqn:relativisticElectrodynamicsT1:100}
A \cdot B = 
\begin{bmatrix}
a^0 & \Ba^\T 
\end{bmatrix}
\begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & -1 & 0 & 0 \\
0 & 0 & -1 & 0 \\
0 & 0 & 0 & -1 
\end{bmatrix}
\begin{bmatrix}
a^0 \\
\Ba
\end{bmatrix}
= A^\T G B
\end{align}

and we can write the Lorentz boost as a matrix

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:110}
\begin{bmatrix}
\gamma & -\beta \gamma & 0 & 0 \\
-\beta \gamma & \gamma & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 
\end{bmatrix}
\end{equation}

so that 

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:120}
a' \cdot b' = A^\T O^\T G O B
\end{equation}

\subsection{Back to the problem.}

\begin{align}\label{eqn:relativisticElectrodynamicsT1:130}
p^i &= (p^0, \Bp) = \left( \frac{E}{c}, \Bp\right) \\
p^2 &= \frac{E^2}{c^2} -\Bp^2 \\
\Bp &= \hbar \Bk \\
E &= \hbar \omega \\
p^i &= \hbar k^i \\
k^i &= (\omega/c, \Bk)
\end{align}

\begin{align}\label{eqn:relativisticElectrodynamicsT1:140} 
\cos \alpha &= \frac{k^1}{\Bk} = \frac{c k^1}{\omega} \\
\cos \alpha' &= \frac{c {k^1}'}{\omega'} = \frac{\gamma (k^1 + \beta \omega/c)}{\gamma(\omega/c + \beta k^1)}
\end{align}

\paragraph{check 1}

as $\beta \rightarrow 1$, $\cos \alpha' \rightarrow 1$, $\alpha' = 0$

In the original reference frame the radiation was isotropic.  In the new frame how does it change with respect to the angle.

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:150}
f'(\alpha') = ?
\end{equation}

\begin{align}\label{eqn:relativisticElectrodynamicsT1:160}
N &= \int d\Omega dt f(\alpha) = \int d \phi \sin \alpha d\alpha = -2 \pi \int d(\cos\alpha) dt f \\
N' &= -2 \pi \int d(\cos\alpha') dt' f'(\alpha')
\end{align}

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:170}
f'(\alpha') = \left( \frac{d(\cos\alpha)}{d(\cos\alpha'} \right) \left( \frac{dt}{dt'} \right) f
\end{equation}

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:180}
\left( \frac{d(\cos\alpha)}{d(\cos\alpha'} \right) = \left( \frac{d(\cos\alpha')}{d(\cos\alpha} \right)^{-1} = \gamma^2 ( 1 + \beta \cos\alpha)^2
\end{equation}

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:190}
\cos\alpha = \frac{\cos\alpha' - \beta}{a - \beta \cos\alpha'}
\end{equation}

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:200}
f'(\alpha') = \frac{f/\gamma}{(\gamma (1-\beta \cos\alpha'))^2}
\end{equation}

What's our energy density?

\begin{align}\label{eqn:relativisticElectrodynamicsT1:210}
L(\alpha) &= \hbar \omega(\alpha) f(\alpha) = \hbar \omega f \\
L'(\alpha') &= \hbar \omega'(\alpha') f'(\alpha') = \hbar \omega' f
\end{align}

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:220}
\omega'(\alpha') = \gamma( \omega + \beta k^1 )
= \gamma \omega ( 1 + \beta \cos\alpha )
\end{equation}

Aside, $\beta << 1$, 

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:230}
\omega' = \omega ( 1 + \beta \cos\alpha) + O(\beta^2) = \omega + \delta \omega
\end{equation}

\begin{align}\label{eqn:relativisticElectrodynamicsT1:240}
\delta \omega &= \beta, \alpha = 0 		\qquad \text{blue shift} \\
\delta \omega &= -\beta, \alpha = \pi 		\qquad \text{red shift}
\end{align}

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:241}
L'(\alpha') = \frac{L/\gamma}{(\gamma (1 - \beta \cos\alpha'))^3}
\end{equation}

forward, backward ratio

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:250}
L'(0)/L'(\pi) = {\left( \frac{ 1 + \beta }{1-\beta} \right)}^3
\end{equation}

The forward radiation is much bigger than the backwards radiation.

\EndArticle
