\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY450H1S.  Relativistic Electrodynamics Lecture 11 (Taught by Prof. Erich Poppitz).  FIXME.}
\label{chap:relativisticElectrodynamicsL13}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsL13.pdf}
\date{Feb 15, 2011}
\revisionInfo{relativisticElectrodynamicsL13.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Reading.}

Covering chapter 4 material from the text \cite{landau1980classical}.

Covering \href{http://www.physics.utoronto.ca/~poppitz/e-poppitz/PHY450_files/RelEMpp103-113.pdf}{lecture notes pp.103-113}: variational principle for the electromagnetic field and the relevant boundary conditions (103-105); the second set of Maxwell’s equations from the variational principle (106-108); Maxwell’s equations in vacuum and the wave equation in the nonrelativistic Coulomb gauge (109-111)

\section{Review.  Our action.}

\begin{align*}\label{eqn:relativisticElectrodynamicsL13:10}
S 
&= S_{\text{particles}} + S_{\text{interaction}} + S_{\text{EM field}}
&= \sum_A \int_{x_A^i(\tau)} ds ( -m_A c )
- \sum_A 
\frac{e_A}{c}
\int dx_A^i A_i(x_A) 
- \inv{16 \pi c} \int d^4 x F^{ij } F_{ij}
\end{align*}

Our dynamics variables are 

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:30}
\left\{
\begin{array}{l l}
x_A^i(\tau) & \quad \mbox{$A = 1, \cdots, N$} \\
A^i(x) & \quad \mbox{$A = 1, \cdots, N$}
\end{array}
\right.
\end{equation}

Also
\begin{equation}\label{eqn:relativisticElectrodynamicsL13:50}
S_{\text{interaction}} 
= -\inv{c^2} \int d^4x j^i(x) A_i(x)
\end{equation}

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:70}
j^i(x) = \sum_A c e_A \int dx_A^i \delta^4( x - x_A(\tau))
\end{equation}

Variation with respect to $x_A^i(\tau)$ we have

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:90}
m c \dds{u^i_A} = \frac{e}{c} u_A^j F_{ij}
\end{equation}

Note that it's easy to get the sign mixed up here.  With our $(+,-,-,-)$ metric tensor, if the second index is the summation index, we have a positive sign.

Only the $S_{\text{particles}}$ and $S_{\text{interaction}}$ depend on $x_A^i(\tau)$.

\section{The field action variation.}

\paragraph{Today}: We'll find the EOM for $A^i(x)$.  The dynamical degrees of freedom are $A^i(\Bx,t)$

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:110}
S[A^i(\Bx, t)] = -\inv{16 \pi c} \int d^4x F_{ij}F^{ij} - \inv{c^2} \int d^4 x A^i j_i
\end{equation}

Here $j^i$ are treated as ``sources''.

We demand that 

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:130}
\delta S = S[ A^i(\Bx, t) + \delta A^i(\Bx, t)] - S[ A^i(\Bx, t) ] = 0 + O(\delta A)^2
\end{equation}

We need to impose two conditions.
\begin{itemize}
\item At spatial $\infty$, i.e. at $\Abs{\Bx} \rightarrow \infty, \forall t$, we'll impose the condition 

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:150}
\evalbar{A^i(\Bx, t)}{\Abs{\Bx} \rightarrow \infty} \rightarrow 0
\end{equation}

This is sensible, because fields are created by charges, and charges are assumed to be localized in a bounded region.  The field outside charges will $\rightarrow 0$ at $\Abs{\Bx} \rightarrow \infty$.  Later we will treat the integration range as finite, and bounded, then later allow the boundary to go to infitiy.

\item at $t = -T$ and $t = T$ we'll imagine that the values of $A^i(\Bx, \pm T)$ are fixed.

This is analogous to $x(t_i) = x_1$ and $x(t_f) = x_2$ in particle mechanics.

Since $A^i(\Bx, \pm T)$ is given, and equivalent to the initial and final field configurations, our extremes at the boundary is zero

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:170}
\delta A^i(\Bx, \pm T) = 0
\end{equation}

\end{itemize}

PICTURE: a cylinder in spacetime, with an attempt to depict the boundary.

\section{Computing the variation.}

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:190}
\delta S[A^i(\Bx, t)] 
= -\inv{16 \pi c} \int d^4 x \delta (F_{ij}F^{ij}) - \inv{c^2} \int d^4 x \delta(A^i) j_i 
\end{equation}

But 

\begin{align*}
\delta (F_{ij}F^{ij}) 
&= 
\delta(F_{ij}) F^{ij} + F_{ij} \delta(F^{ij}) \\
&= 
2 \delta(F^{ij}) F_{ij} \\
&= 
2 \delta(\partial^i A^j - \partial^j A^i) F_{ij} \\
&= 
2 \delta(\partial^i A^j) F_{ij} - 2 \delta(\partial^j A^i) F_{ij} \\
&= 
2 \delta(\partial^j A^i) F_{ji} - 2 \delta(\partial^j A^i) F_{ij} \\
&= 
4 \delta(\partial^i A^j) F_{ij} \\
&= 
4 F_{ij} \partial^i \delta(A^j) \\
\end{align*}

So
\begin{align*}
\delta S[A^i(\Bx, t)] 
&= -\inv{4 \pi c} \int d^4 x F_{ij} \partial^i \delta(A^j) - \inv{c^2} \int d^4 x j^i \delta(A_i) \\
\end{align*}

\begin{align*}
\int d^4 x F_{ij} \partial^i \delta(A^j) 
=
\int d^4 x F^{ij} \partial_i \delta(A_j) 
=
\int d^4 x F^{ij} \PD{x^i}{} \delta(A_j) 
=
\int d^4 x \PD{x^i}{} \delta A_j F^{ij}
\delta(A^j) 
\end{align*}

DIY.  Integrate by parts.

= ... 0 (zero over the boundary)

Noting that 

\begin{align*}
\PD{x^\alpha}{}\left( \delta A_j F^{\alpha j} \right) \\
\spacegrad \cdot \BC, \BC^\alpha = \delta A_j F^{\alpha j} \\
...
\end{align*}

We are left with

\begin{align*}
\delta S[A^i(\Bx, t)] 
&= -\inv{4 \pi c} \int d^4 x \delta (A_j) \partial_i F^{ij} - \inv{c^2} \int d^4 x j^i \delta(A_i) \\
&= 
\int d^4 x \delta A_j(x)
\left(
-\inv{4 \pi c} \partial_i F^{ij}(x) - \inv{c^2} j^i 
\right)  \\
&= 0
\end{align*}

This gives us

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:210}
\boxed{
\partial_i F^{ij} = \frac{4 \pi}{c} j^j
}
\end{equation}

\section{Unpacking these.}

Recall that 

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:230}
e^{ijkl} \partial_j F_{kl} = 0
\end{equation}

gave us

\begin{align}\label{eqn:relativisticElectrodynamicsL13:250}
\spacegrad \cdot \BB &= 0 \\
\spacegrad \cross \BE &= -\inv{c} \PD{t}{\BB}
\end{align}

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:270}
\partial_\alpha F^{\alpha 0} = \frac{4 \pi}{c} j^0 = 4 \pi \rho
\end{equation}

(since $j^0 = c \rho$).

Because 

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:290}
F^{\alpha 0} = (\BE)^\alpha
\end{equation}

or
\begin{equation}\label{eqn:relativisticElectrodynamicsL13:310}
\spacegrad \cdot \BE = 4 \pi \rho
\end{equation}

The messier one to deal with is

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:330}
\partial_i F^{i\alpha} = \frac{4 \pi}{c} j^\alpha
\end{equation}

...

\begin{align*}
\partial_i F^{i\alpha} 
&= \partial_\beta F^{\beta \alpha} + \partial_0 F^{0 \alpha} \\
&= \partial_\beta F^{\beta \alpha} - \inv{c} \PD{t}{(\BE)^\alpha} \\
\end{align*}

Details: DIY.  We get

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:350}
\frac{4 \pi}{c} j^\alpha
= (\spacegrad \cross \BB)^\alpha - \inv{c} \PD{t}{(\BE)^\alpha} \\
\end{equation}

or

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:370}
\spacegrad \cross \BB - \inv{c} \PD{t}{\BE} = \frac{4 \pi}{c} \Bj
\end{equation}

Summarizing what we know so far, we have

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:390}
\boxed{
\begin{aligned}
\partial_i F^{ij} &= \frac{4 \pi}{c} j^j \\
\epsilon^{ijkl} \partial_j F_{kl} &= 0
\end{aligned}
}
\end{equation}

or in vector form

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:410}
\boxed{
\begin{aligned}
\spacegrad \cdot \BE &= 4 \pi \rho \\
\spacegrad \cross \BB -\inv{c} \PD{t}{\BE} &= \frac{4 \pi}{c} \Bj \\
\spacegrad \cdot \BB &= 0 \\
\spacegrad \cross \BE +\inv{c} \PD{t}{\BB} &= 0
\end{aligned}
}
\end{equation}

\section{Speed of light}

\paragraph{Claim}: ``$c$'' is the speed of EM waves in vacuum.

Study equations in vacuum (no sources, so $j^i = 0$) for $A^i = (\phi, \BA)$.

\begin{align}\label{eqn:relativisticElectrodynamicsL13:430}
\spacegrad \cdot \BE &= 0 \\
\spacegrad \cross \BB &= \inv{c} \PD{t}{\BE} 
\end{align}

where

\begin{align}\label{eqn:relativisticElectrodynamicsL13:450}
\BE &= - \spacegrad \phi - \inv{c} \PD{t}{\BA} \\
\BB &= \spacegrad \cross \BA
\end{align}

In terms of potentials

\begin{align}\label{eqn:relativisticElectrodynamicsL13:470}
\spacegrad \cross (\spacegrad \spacegrad \BA) &= - \inv{c} \spacegrad \PD{t}{\phi} - \inv{c} \frac{\partial^2}{\partial t^2} \BA
-\spacegrad^2 \phi - \inv{c} \spacegrad \cdot \BA &= 0
\end{align}

Can make a gauge transformation

\begin{align}\label{eqn:relativisticElectrodynamicsL13:490}
(\phi, \BA) \rightarrow (\phi', \BA')
\end{align}

with 

\begin{align}\label{eqn:relativisticElectrodynamicsL13:510}
\phi &= \phi' + \inv{c} \PD{t}{\chi} \\
\BA &= \BA' - \spacegrad \chi
\end{align}

Can choose $\chi(\Bx, t)$ to make $\phi' = 0$ ($\forall \phi \exists \chi, \phi' = 0$)

\begin{align}\label{eqn:relativisticElectrodynamicsL13:530}
\inv{c} \PD{t}{} \chi(\Bx, t) = \phi(\Bx, t)
\end{align}

\begin{align}\label{eqn:relativisticElectrodynamicsL13:550}
\chi(\Bx, t) = c \int_{-\infty}^t dt' \phi(\Bx, t')
\end{align}

Can also find a transformation that also allows $\spacegrad \cdot \BA = 0$

This is the Coulomb gauge 

\begin{align}\label{eqn:relativisticElectrodynamicsL13:570}
\phi &= 0 \\
\spacegrad \cdot \BA &= 0
\end{align}

DIY.  Can use this above to kill a bunch of the terms, leaving 

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:590}
\spacegrad \cross (\spacegrad \cross \BA') = -\inv{c^2} \frac{\partial^2 \BA'}{\partial t^2}
\end{equation}

Using 

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:600}
\spacegrad \cross (\spacegrad \cross \BA') = \spacegrad (\spacegrad \cdot \BA') - \spacegrad^2 \BA',
\end{equation}

we have

\begin{equation}\label{eqn:relativisticElectrodynamicsL13:610}
\inv{c^2} \frac{\partial^2 \BA'}{\partial t^2} -\spacegrad^2 \BA' = 0
\end{equation}

which is the wave equation for the propagation of the vector potential $\BA'(\Bx, t)$ through space at velocity $c$.

\EndArticle
