\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{PHY450H1S.  Relativistic Electrodynamics Lecture 20 (Taught by Prof. Erich Poppitz).  Lienard-Wiechert potentials.}
\label{chap:relativisticElectrodynamicsL20}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsL20.pdf}
\date{Mar 15, 2011}
\revisionInfo{relativisticElectrodynamicsL20.tex}

%\beginArtWithToc
\beginArtNoToc

\section{Reading.}

Covering chapter 8 material from the text \cite{landau1980classical}.

FIXME:
Covering \href{http://www.physics.utoronto.ca/~poppitz/epoppitz/PHY450_files/RelEMpp136-146.pdf}{lecture notes pp. 136-146}: the Lienard-Wiechert potentials (143-146) [Wednesday, Mar. 9...]

pp. 147-165: EM fields of a moving source (147-148+HW5); a particle at rest (148); a constant velocity particle (149-152); behavior of EM fields ``at infinity'' for a general-worldline source and radiation (152-153) [Tuesday, Mar. 15]; radiated power (154); fields in the ``wave zone'' and discussions of approximations made (155-159); EM fields due to electric dipole radiation (160-163); Poynting vector, angular distribution, and power of dipole radiation (164-165) [Wednesday, Mar. 16...]

\section{.}

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:10}
A^i(\Bx, t) = \inv{c} \int d^3 \Bx' j^i\left(\Bx', t - \frac{\Abs{\Bx - \Bx'}}{c}\right) \inv{\Abs{\Bx - \Bx'}}
\end{equation}

This integral is over the region of space where the sources $j^i$ are non-vanishing, but this region is limited.  The value $\Abs{\Bx'} \le l$, so we can expand the denominator in multipole expansion

\begin{align*}
\inv{\Abs{\Bx - \Bx'}}
&=
\inv{\sqrt{(\Bx - \Bx')^2}} \\
&=
\inv{\sqrt{\Bx^2 + {\Bx'}^2 - 2 \Bx \cdot \Bx'}} \\
&=
\inv{\Abs{\Bx}} \inv{\sqrt{1 + \frac{{\Bx'}^2}{\Bx^2} - 2 \frac{\xcap}{\Abs{\Bx}} \cdot \Bx'}} \\
&\approx
\inv{\Abs{\Bx}} \inv{\sqrt{1 - 2 \frac{\xcap}{\Abs{\Bx}} \cdot \Bx'}} \\
&\approx
\inv{\Abs{\Bx}} \left(1 + \frac{\xcap}{\Abs{\Bx}} \cdot \Bx' \right).
\end{align*}

so
\begin{equation}\label{eqn:relativisticElectrodynamicsL20:30}
\inv{\Abs{\Bx - \Bx'}}
\approx 
\inv{\Abs{\Bx}} + \frac{\Bx}{\Abs{\Bx}^3} \cdot \Bx' 
\end{equation}

similarily 

\begin{align*}
t - \frac{\Abs{\Bx - \Bx'}}{c} 
&= t - \frac{\Abs{\Bx}}{c} \left( 1 - \frac{\Bx \cdot \Bx'}{\Abs{\Bx}^2} \right) \\
&= t - \frac{\Abs{\Bx}}{c} + \frac{\Bx \cdot \Bx'}{\Abs{\Bx}}
\end{align*}

The retardation effect is the time argument of the $f'$ the source $j^i$

FIXME: what's $f'$?

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:50}
j^i\left(\Bx', t - \frac{\Abs{\Bx}{c}} + \frac{\Bx \cdot \Bx'}{\Abs{\Bx}} + \cdots \right)
\approx
j^i\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right) + j^i \left(\Bx, t - \frac{\Abs{\Bx}}{c}\right) \frac{\Bx \cdot \Bx'}{\Abs{\Bx}}
\end{equation}

to elucidate the physics, imagine that time dependence of the source is periodic with angular frequency $\omega_0$.  Then

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:70}
j^i \approx \omega_0 j^i
\end{equation}

magnitude fo the second term 

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:90}
j^i \frac{\Bx \cdot \Bx'}{\Abs{\Bx}} = \omega_0 \frac{\Bx \cdot \Bx'}{\Abs{\Bx}} = \omega_0 
\end{equation}

while the first time is $\propto j^i$.

Requiring second term much less than the first term means 

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:110}
\Abs{\omega_0 \frac{\Bx \cdot \Bx'}{\Abs{\Bx}}} \ll 1
\end{equation}

But recall 
\begin{equation}\label{eqn:relativisticElectrodynamicsL20:130}
\Abs{\frac{\Bx \cdot \Bx'}{\Abs{\Bx}}} \le l
\end{equation}

therefore

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:150}
\Abs{\omega_0 \frac{\Bx \cdot \Bx'}{\Abs{\Bx}}} \le \frac{\omega_0 l}{c} \ll 1
\end{equation}

This is a physical requirement size of the wavelength of the emitter (if the wavvelength doesn't meet this requirement, this expansion does not work).

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:170}
\frac{\omega_0}{c} = \frac{2 \pi}{c T} \propto \inv{\lambda}
\end{equation}

\paragraph{Moral:} We'll utilize two expandsions (we need two small parameters)

\begin{enumerate}
\item $\Abs{\Bx} \gg l$
\item $\lambda \gg l$
\end{enumerate}

Plugging into our current

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:190}
A^i(\Bx, t) 
\approx \inv{c} \int d^3 \Bx' 
\left( j^i\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right) + j^i \left(\Bx, t - \frac{\Abs{\Bx}}{c}\right) \frac{\Bx \cdot \Bx'}{\Abs{\Bx}} \right)
\left( \inv{\Abs{\Bx}} + \frac{\Bx}{\Abs{\Bx}^3} \cdot \Bx' \right)
\end{equation}

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:210}
A^0(\Bx, t) 
\approx 
\inv{\Abs{\Bx}} \int d^3 \Bx' \rho\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right)
+\inv{\Abs{\Bx}^3} \int d^3 \Bx' \Bx' \rho \left(\Bx', t - \frac{\Abs{\Bx}}{c}\right)
+\inv{\Abs{\Bx}^2} \Bx \cdot \int d^3 \Bx' \Bx' \PD{t}{} \rho\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right)
\end{equation}

Note that this term

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:230}
\int d^3 \Bx' \Bx' \rho\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right) = \Bd,
\end{equation}

is the dipole moment.  In the last term we can pull out the time derivative (because we are integrating over $\Bx'$)

\begin{align*}
\inv{\Abs{\Bx}^2} \Bx \cdot \int d^3 \Bx' \Bx' \PD{t}{} \rho\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right)
&=
\inv{\Abs{\Bx}^2} \Bx \cdot \PD{t}{} \int d^3 \Bx' \Bx' \rho\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right)
\rho\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right)
&=
\inv{\Abs{\Bx}^2} \Bx \cdot \PD{t}{}\Bd \left(t - \frac{\Abs{\Bx}}{c}\right)
\end{align*}

For the spatial components of the current lets just keep the first term

\begin{align*}
A^\alpha(\Bx, t) 
&\approx
\inv{ c \Abs{\Bx}} \int d^3 \Bx' j^\alpha\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right) \\
&=
\inv{ c \Abs{\Bx}} \int d^3 \Bx' (\spacegrad_{\Bx'} x^\alpha \cdot \Bj\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right)  \\
&=
\inv{ c \Abs{\Bx}} \int d^3 \Bx' 
\left(
\spacegrad \left( {x'}^\alpha \Bj \left(\Bx', t - \frac{\Abs{\Bx}}{c}\right) \right)
- {x'}^\alpha \spacegrad_{\Bx'} \cdot \Bj\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right) 
\right) \\
&=
\inv{ c \Abs{\Bx}} \oint_{S^2_\infty} d^2 \Bsigma \cdot {x'}^\alpha \Bj\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right)
+\inv{ c \Abs{\Bx}} \int d^3 \Bx' {x'}^\alpha \PD{t}{}\rho\left(\Bx', t - \frac{\Abs{\Bx}}{c}\right)
\end{align*}

%boxed:
\begin{align}\label{eqn:relativisticElectrodynamicsL20:250}
A^0(\Bx, t) &= \frac{Q\left(t - \frac{\Abs{\Bx}}{c}\right)}{\Abs{\Bx}} + \frac{\Bx \cdot \Bd\left(t - \frac{\Abs{\Bx}}{c}\right)}{Abs{\Bx}^3} + \frac{\Bx \cdot \dot{\Bd}\left(t - \frac{\Abs{\Bx}}{c}\right)}{c \Abs{\Bx}^2} \\
\BA(\Bx, t) &= \inv{c \Abs{\Bx}} \dot{\Bd}\left(t - \frac{\Abs{\Bx}}{c}\right)
\end{align}

\subsection{Example}

Electric dipole radiation.

PICTURE: two closely separated oppositely charges, wiggling along the line connecting them.

$-q$ at rest, while $+q$ oscillates.

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:270}
z_+(t) = z_0 + a \sin\omega t
\end{equation}

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:290}
\Bd(t) = \Be_z q (z_0 + a \sin\omega t)
\end{equation}

Thus

\begin{equation}\label{eqn:relativisticElectrodynamicsL20:310}
A^0(\Bx, t) = 
\end{equation}

\EndArticle
