\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Runge-Lenz vector conservation}
\label{chap:RungeLenz}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/RungeLenz.pdf}
\date{Feb 11, 2012}
\revisionInfo{RungeLenz.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

Notes from Poppitz' phy354 lecture on the Runge-Lenz vector, a less well known conserved quantity for the 3D $1/r$ potentials that can be used to solve the Kepler problem.

\section{Motivation: The Kepler problem.}

We can plug away at the Lagrangian in cylindrical coordinates and find eventually

\begin{equation}\label{eqn:RungeLenz:n}
\int_{\phi_0}^\phi d\phi = \int_{r_0}^r \frac{M}{m r^2} \frac{dr}{\sqrt{\frac{2}{M} ( E - U + \frac{M^2}{2 m r^2}}}
\end{equation}

but this can be messy to solve, where we get eliptic integrals or worse, depending on the potential.

For the special case of the 3D problem where the potential has a $1/r$ form, this is what Poppitz called ``super-integrable''.  With $2N - 1 = 5$ conserved quantites to be found, we've got one more.  Here the form of that last conserved quantity is given, called the Runge-Lenz vector, and we verify that it is conserved.

\section{Runge-Lenz vector}

Given a potential

\begin{equation}\label{eqn:RungeLenz:n}
U = -\frac{\alpha}{r}
\end{equation}

and a Lagrangian

\begin{align}\label{eqn:RungeLenz:n}
\LL &= \frac{m \Bv^2}{2} + \inv{2} \frac{M_z^2}{m r^2} - U \\
M_z &= m r^2 \dot{\phi}^2
\end{align}

and writing the angular momentum as
\begin{equation}\label{eqn:RungeLenz:n}
\BM = m \Br \cross \Bv,
\end{equation}

the Runge-Lenz vector

\begin{equation}\label{eqn:RungeLenz:n}
\BA = \Bv \cross \BM - \alpha \rcap,
\end{equation}

is a conserved quantity.  Let's show this.

%\EndArticle
\EndNoBibArticle
