\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Some potentials.}
\label{chap:somePotentials}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/somePotentials.pdf}
\date{Feb 19, 2012}
\gitRevisionInfo{somePotentials}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

Part of a classical mechanics problem set was to look at what portions of momentum and angular momentum are conserved for various fields.  Since this was also a previous midterm question, I'm expecting that some intuition was expected to be used determine the form of the Lagrangians, with not much effort on finding the precise form of the potentials.

I've asked:

Can the potential of 4.2 (infinite homogeneous cylinder) actually be calculated?  I get infinities if I let the length of the cylinder become unbounded.

Oddly, I can calculate the potential for 4.1 (infinite homogeneous plane) if I start with the force and then calculate the potential, but if I start with the potential the integral also diverges?  That doesn't make any sense, so I'm wondering if I've miscalculated.

For both of these my expectation is that you don't actually care about the exact form of the Lagrangian, and just want us to consider potentials that are dependent on the distance from the plane and the radial distance from the center of the cylinder, but I felt like playing and hit trouble doing the calculations determining the precise form of the Lagrangian.

For the two and three point particle portions of the problems, is your assumption that these particles are fixed in position, or do you really want us to find the components of momentum and angular momentum conservation for the three and four body problems?

\section{Forces and potential for an infinite homogeneous plane.}

\subsection{Calculating the potential from the force.}

For the plane, with $z$ as the distance from the plane, and $(r,\theta)$ coordinates in the plane, as illustrated in

FIXME:F1

the gravitational force from an element of mass on the plane is

\begin{align*}
d\BF 
&= -G \lambda m r dr d\theta \frac{z \zcap - r \rcap}{\Abs{z \zcap - r \rcap}^3 } \\
&= -G \lambda m r dr d\theta \frac{z \zcap - r \Be_1 e^{i \theta} }{(z^2 + r^2)^{3/2} } \\
\end{align*}

Here I've used a Clifford Algebra representation of the unit vector $\rcap$ with $i = \Be_1 \Be_2$.  Should a more conventional physics notation be desired we can consider this the element $\Bsigma \cdot d\BF$, and make the substitutions

\begin{align*}
%1 &\rightarrow \sigma_0 = 
%\begin{bmatrix}
%1 & 0 \\
%0 & 1
%\end{bmatrix}
%\\
\zcap &\rightarrow \Bsigma \cdot \Be_3 = \PauliZ
\\
\rcap &\rightarrow \Bsigma \cdot \Be_1 e^{ (\Bsigma \cdot \Be_1) (\Bsigma \cdot \Be_2) \theta} = \PauliX \cos\theta + \PauliY \sin\theta.
\end{align*}

Integrating for the total force on the test mass, noting that $\int_0^{2 \pi} e^{i\theta} d\theta = 0$, we have

\begin{equation}\label{eqn:classicalMechanicsPs2:n}
\BF = - 2 \pi G \lambda m z \zcap \int_0^\infty r dr \inv{(z^2 + r^2)^{3/2} },
\end{equation}

a substuition $r = z \tan \alpha$ gives us

\begin{align*}
\BF 
&= - 2 \pi G \lambda m z \zcap \int_0^{\pi/2} z \tan \alpha z \sec^2 \alpha d\alpha \inv{z^3 \sec^3 \alpha } \\
&= - 2 \pi G \lambda m \zcap \int_0^{\pi/2} \sin \alpha d\alpha \\
&= 2 \pi G \lambda m \zcap (\cos(\pi/2) - \cos(0)),
\end{align*}

For

\begin{equation}\label{eqn:classicalMechanicsPs2:n}
\BF = -2 \pi G \lambda m \zcap.
\end{equation}

For the Lagrangian problem we want the potential 

\begin{align*}
-\spacegrad \phi &= \BF \\
-\zcap \PD{z}{\phi} &= \\
\end{align*}

or

\begin{equation}\label{eqn:classicalMechanicsPs2:n}
\phi = 2 \pi G \lambda m z.
\end{equation}

This is a reasonable seeming answer.  Our potential is of the form

\begin{equation}\label{eqn:classicalMechanicsPs2:n}
\phi = m g z,
\end{equation}

with
\begin{equation}\label{eqn:classicalMechanicsPs2:n}
g = 2 \pi G \lambda.
\end{equation}

\subsection{Calculating the potential directly.}

If we only want the potential, why start with the force?  We ought to be able to work with potentials directly, and write

\begin{equation}\label{eqn:classicalMechanicsPs2:n}
\phi(\Br) = G m \rho \int \frac{dV'}{\Abs{\Br' - \Br}}.
\end{equation}

It's a quick calculation to verify that this has the signs right, and we find (provided $\Br \ne \Br'$)

\begin{equation}\label{eqn:classicalMechanicsPs2:n}
\BF = - \spacegrad \phi = - G m \rho \int \frac{dV' (\Br - \Br')}{\Abs{\Br - \Br'}}.
\end{equation}

Our position vectors are illustrated in 

FIXME: F2.

Now, suppose we calculate the potential directly

\begin{align*}
\phi(z)
&= G m \rho \int \frac{r dr d\theta}{(z^2 + r^2)^{1/2}} \\
&= 2 \pi G m \rho \int_0^\infty \frac{r dr }{(z^2 + r^2)^{1/2}} \\
&= 2 \pi G m \rho z \int_0^\infty \frac{u du }{(1 + u^2)^{1/2}}.
\end{align*}

Again we can make a tangent substitution 

\begin{equation}\label{eqn:classicalMechanicsPs2:n}
u = \tan\alpha,
\end{equation}

for

\begin{align*}
\phi(z)
&= 2 \pi G m \rho z \int_0^{\pi/2} \frac{\tan \alpha \sec^2 \alpha d \alpha }{\sec\alpha} \\
&= 2 \pi G m \rho z \int_0^{\pi/2} \tan \alpha \sec \alpha d \alpha \\
&= 2 \pi G m \rho z \int_0^{\pi/2} d\alpha \frac{\sin \alpha}{\cos^2 \alpha}
\end{align*}

However, this isn't nicely behaved, since the cosine goes to zero at $\alpha = \pi/2$.  What has gone wrong?

%\EndArticle
\EndNoBibArticle
