% 
% 
% 
% Copyright © 2012 Peeter Joot
% All Rights Reserved
% 
% This file may be reproduced and distributed in whole or in part, without fee, subject to the following conditions:
% 
% o The copyright notice above and this permission notice must be preserved complete on all complete or partial copies.
% 
% o Any translation or derived work must be approved by the author in writing before distribution.
% 
% o If you distribute this work in part, instructions for obtaining the complete version of this file must be included, and a means for obtaining a complete version provided.
% 
% 
% Exceptions to these rules may be granted for academic purposes: Write to the author and ask.
% 
% 
% 
\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Flow between infinite moving inner cylinder and outer cylinder.}
\label{chap:twoCylinders}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/twoCylinders.pdf}
\date{Apr 10, 2012}
\gitRevisionInfo{twoCylinders}
\keywords{Navier-Stokes, PHY454H1S, PHY454H1} 

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

A problem from the 2009 phy1530 final that appears appropriate for phy454 exam prep.

\section{Statement.}

An infinite cylinder of radius $R_1$ is moving with velocity $v$ parallel to its axis.  It is places inside another cylinder of radius $R_2$.  The axes of the two cylinders coincide.  The fluid is incompressible, with viscosity $\mu$ and density $\rho$, the flow is assumed to be stationary, and no external pressure gradient is applied.

\begin{enumerate}
\item Find and sketch the velocity field of the fluid between the cylinders.
\item Find the friction force per unit length acting on each cylinder.
\item Find and sketch the pressure field of the liquid.
\item If an external pressure gradient is present, how do you think your answer will change?  Sketch your expectation for the velocity and pressure in this case.
\end{enumerate}

\section{Solution.}

%\EndArticle
\EndNoBibArticle
