\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

%\newcommand{\br1}[0]{(1)}
%\newcommand{\br2}[0]{(2)}
%\newcommand{\bri}[0]{(i)}
%s/\\br\(.\)/{(\1)}/g

\chapter{Steady state inclined flow down a plane of two layers of incompressible viscous equal density fluids.}
\label{chap:twoLayerInclinedFlow}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/twoLayerInclinedFlow.pdf}
\date{Mar 1, 2012}
\gitRevisionInfo{twoLayerInclinedFlow}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

Here's one of the problems from \S 2 of \cite{acheson1990elementary}, a slight variation on what we did in class.  It seemed to me that it would be a good midterm prep problem to try.

\section{Steady state inclined flow down a plane of two layers of incompressible viscous equal density fluids.}

\subsection{Setup of the equations of motion for this system.}

Our problem is illustrated in 

FIXME: F4

with a plane set at angle $\alpha$, fluid depths of $h^{(1)}$ and $h^{(2)}$ respectively, and viscosities $\mu^{(1)}$ and $\mu^{(2)}$.  We'll write $H = h^{(1)} + h^{(2)}$.  We have a pair of Navier-Stokes equations to solve

\begin{equation}\label{eqn:twoLayerInclinedFlow:10}
\rho \ddt{\Bu^{(i)}} = \rho \PD{t}{\Bu^{(i)}} + \rho (\Bu^{(i)} \cdot \spacegrad) \Bu^{(i)} = - \spacegrad p^{(i)} + \mu^{(i)} \spacegrad^2 \Bu^{(i)} + \mu^{(i)} \spacegrad (\spacegrad \cdot \Bu^{(i)}) + \rho \Bg
\end{equation}

Our steady state and incompressiblity constraints break this into a few independent equations

\begin{align}\label{eqn:twoLayerInclinedFlow:30}
\rho \PD{t}{\Bu^{(i)}} &= 0 \\
\spacegrad \cdot \Bu^{(i)} &= 0 \\
\rho (\Bu^{(i)} \cdot \spacegrad) \Bu^{(i)} &= - \spacegrad p^{(i)} + \mu^{(i)} \spacegrad^2 \Bu^{(i)} + \rho \Bg
\end{align}

Let's require that we initially have no components of the flows in the $y$, or $z$ directions.  As the equivalent of Newton's law for fluid flows, conservation of linear momentum requires that for our steady state problem we have $u_y = u_z = 0$ for the flows in both fluid layers.  

FIXME: should go back and think about momentum conservation in the context of fluids.  I'm now so used to thinking about this as a symmetry issue from a Lagrangian context.  With no Lagrangian here, what would be a robust way treat this in fluids?  What is the Lagrangian for a fluid mechanics system?  I'd imagine that it would be possible to set up a field Lagrangian with velocity fields.

Our problem is now reduced to a problem in four quanties (two velocities and two pressures).  With $\Bu^{(i)} = (u^{(i)}, 0, 0)$ we can restate Navier-Stokes in coordinate form as

\begin{subequations}
\label{eqn:twoLayerInclinedFlow:50}
\begin{equation}\label{eqn:twoLayerInclinedFlow:110}
\partial_x u^{(i)}(x,y,z) = 0
\end{equation}
\begin{equation}\label{eqn:twoLayerInclinedFlow:130}
\rho (u^{(i)} \partial_x u^{(i)} = - \partial_x p^{(i)} + \mu^{(i)} (\partial_{xx} + \partial_{yy} + \partial_{zz}) u^{(i)} + \rho g \sin\alpha 
\end{equation}
\begin{equation}\label{eqn:twoLayerInclinedFlow:150}
0 = - \partial_y p^{(i)} - \rho g \cos\alpha 
\end{equation}
\begin{equation}\label{eqn:twoLayerInclinedFlow:155}
0 = - \partial_z p^{(i)} 
\end{equation}
\end{subequations}

In order to solve this, we have eight simultaneous non-linear PDEs, four unknown functions, plus boundary conditions to consider!

What are the boundary conditions?  One is the ``no-slip'' condition, the experimental observation that velocites match at the interfaces.  So we should have zero velocity for the fluid lying against the plane, and velocity matching between the fluids.  The air above the fluid will also be flowing along at the rate of the uppermost portion of the top layer, but we'll neglect that effect (i.e. considering two layers of equal density and not three, with one having a separate density).  We also have matching of the traction vectors at the interfaces.  

Writing this, it occured to me that I didn't fully understand what motivated that last boundary value condition.  Talking to our Prof about this, the matching of the traction vectors at any point can be thought of as an observational issue, but this is also a force balance issue.  There is an induced velocity in the direction of the traction vector at any given point.  For example, when we have unidirectional flow, we must have no normal component of the traction vector, and only a tangential component, because we have only the tangential flow.

Before continuing to solve our Navier-Stokes equations let's express the condition that the tangential component of the traction vectors match algebraically.

Dropping indexes temporarily, for the normal to the surface $\ncap = (n_1, n_2, n_3) = (0, 1, 0)$ we want to compute

\begin{align*}
T_1 
&= \sigma_{1 k} n_k \\
&= \sigma_{1 k} \delta_{2 k} \\
&= \sigma_{1 2} \\
&= -p \cancel{\delta_{1 2}} + 2 \mu e^{1 2} \\
&= \mu \left( \PD{y}{u_1} + \cancel{\PD{x}{u_2}} \right) 
\end{align*}

So the tangential component of the traction vector is

\begin{equation}\label{eqn:twoLayerInclinedFlow:90}
\BT^{(i)} = \mu^{(i)} \PD{y}{u^{(i)}} \xcap.
\end{equation}

As noted above, this is in fact, the only component of the traction vector, since we don't have any non-horizontal flow.

Our boundary value conditions, what we need in addition to the Navier-Stokes equations of \ref{eqn:twoLayerInclinedFlow:50}, to solve our problem, are the matching at any interface of the following conditions

\begin{subequations}
\label{eqn:twoLayerInclinedFlow:70}
\begin{equation}\label{eqn:twoLayerInclinedFlow:n}
u^{(i)} = u^{(j)} 
\end{equation}
\begin{equation}\label{eqn:twoLayerInclinedFlow:n}
p^{(i)} = p^{(j)} 
\end{equation}
\begin{equation}\label{eqn:twoLayerInclinedFlow:n}
\mu^{(i)} \PD{y}{u^{(i)}} = \mu^{(i)} \PD{y}{u^{(j)}}.
\end{equation}
\end{subequations}

There are actually three interfaces to consider, that of the lower layer liquid with the inclined plane, the interface between the two fluid layers, and the interface between the upper layer fluid and the air above it.

\subsection{Solving our equations of motion.}

Starting with the simplest, the z-coordinate equation, of Navier-Stokes \ref{eqn:twoLayerInclinedFlow:155}, we can conclude that each of the pressures is not a function of z, so that we have

\begin{equation}\label{eqn:twoLayerInclinedFlow:n}
p^{(i)} = p^{(i)}(x, y).
\end{equation}

Using this, we can integrate our y-coordinate Navier-Stokes equation \ref{eqn:twoLayerInclinedFlow:150}, to find

\begin{equation}\label{eqn:twoLayerInclinedFlow:n}
p^{(i)} = - \rho g y \cos\alpha + f^{(i)}(x)
\end{equation}

\EndArticle
