%
% Copyright © 2013 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\input{../blogpost.tex}
\renewcommand{\basename}{twoMassHarmonic}
\renewcommand{\dirname}{notes/classicalmechanics/}
%\newcommand{\dateintitle}{}
\newcommand{\keywords}{Lagrangian, Harmonic oscillator, center of mass, two body problem}

\newcommand{\nought}[0]{\circ}

\input{../peeter_prologue_print2.tex}

\beginArtNoToc

\generatetitle{Two body harmonic oscillator in 3D}
%\chapter{Two body harmonic oscillator in 3D}
%\label{chap:twoMassHarmonic}

\makeproblem{Two body harmonic oscillator in 3D}{pr:twoMassHarmonic:1}{

In lattice problems, we consider normal modes of harmonic coupled systems.  As review, before switching to displacement coordinates, solve the two body problem exactly using absolute coordinates.
} % makeproblem

\makeanswer{pr:twoMassHarmonic:1}{ 

For the system illustrated in our Lagrangian is

FIXME: figure

\begin{dmath}\label{eqn:twoMassHarmonic:20}
\LL = 
\inv{2} m_1 \lr{ \dot{\Br}_1 }^2
+\inv{2} m_2 \lr{ \dot{\Br}_2 }^2
- \frac{K}{2} \lr{ \Br_1 - \Br_2 }^2.
\end{dmath}

We wish to solve the equations of motion

\begin{dmath}\label{eqn:twoMassHarmonic:40}
\ddt{} \spacegrad_{\dot{\Br}_i} \LL = 
\spacegrad_{\Br_i} \LL.
\end{dmath}

Noting that $\spacegrad_\Bx \Ba \cdot \Bx = \Ba$, the coupled system to solve is found to be

\begin{dmath}\label{eqn:twoMassHarmonic:60}
\begin{aligned}
m_1 \ddot{\Br}_1 &= - K \lr{ \Br_1 - \Br_2 } \\
m_2 \ddot{\Br}_2 &= - K \lr{ \Br_2 - \Br_1 }.
\end{aligned}
\end{dmath}

These can be decoupled using differences and sums

\begin{dmath}\label{eqn:twoMassHarmonic:80}
\begin{aligned}
m_2 \lr{ m_1 \ddot{\Br}_1 } - m_1 \lr{ m_2 \ddot{\Br}_2 } &= - (m_1 + m_2) K \lr{ \Br_1 - \Br_2 } \\
m_1 \ddot{\Br}_1 + m_2 \ddot{\Br}_2 &= 0
\end{aligned}
\end{dmath}

The second is the equation for the acceleration of the center of mass $\BR_{\mathrm{CM}}(t)$, and it directly integrable.  With $M = m_1 + m_2$, that is

\begin{dmath}\label{eqn:twoMassHarmonic:100}
M \BR_{\mathrm{CM}}(t) 
= 
m_1 \Br_1 + m_2 \Br_2 = 
 (t - t_\nought) 
M \BV_{\mathrm{CM}}
+ M \BR_{\mathrm{CM}}(t_\nought).
\end{dmath}

The first is the harmonic oscillation about the center of mass position.  Introducing the reduced mass

\begin{dmath}\label{eqn:twoMassHarmonic:120}
\mu = \frac{m_1 m_2}{m_1 + m_2},
\end{dmath}

that oscillation equation is

\begin{dmath}\label{eqn:twoMassHarmonic:140}
\frac{d^2}{dt^2}
\lr{ \Br_1 - \Br_2 }
= -\frac{K}{\mu} 
\lr{ \Br_1 - \Br_2 }.
\end{dmath}

With 

\begin{dmath}\label{eqn:twoMassHarmonic:160}
\omega^2 = \frac{K}{\mu},
\end{dmath}

this difference has the solution

\begin{dmath}\label{eqn:twoMassHarmonic:180}
\Br_1(t) - \Br_2(t) = \lr{ \Br_1(t_\nought) - \Br_2(t_\nought) } \cos\lr{ \omega(t - t_\nought) }.
\end{dmath}

This form of solution allows for arbitrary phase shift, including both sine and cosine solutions.  For example, for cosine solutions, pick $t_\nought$ as the time for which the amplitude difference is maximized.  

Writing $\Delta \Br_\nought = \lr{ \Br_1(t_\nought) - \Br_2(t_\nought) }$, the matrix inversion problem for 
individual $\Br_i$ vectors is

\begin{dmath}\label{eqn:twoMassHarmonic:200}
\begin{bmatrix}
1 & -1 \\
m_1 & m_2 
\end{bmatrix}
\begin{bmatrix}
\Br_1 \\
\Br_2
\end{bmatrix}
=
\begin{bmatrix}
\Delta \Br_\nought
\cos\lr{ \omega(t - t_\nought) } \\
M \BR_{\mathrm{CM}}(t) 
\end{bmatrix},
\end{dmath}

or

\begin{dmath}\label{eqn:twoMassHarmonic:220}
\begin{bmatrix}
\Br_1 \\
\Br_2
\end{bmatrix}
=
\inv{m_2 + m_1}
\begin{bmatrix}
m_2 & 1 \\
-m_1 & 1 
\end{bmatrix}
\begin{bmatrix}
\Delta \Br_\nought
\cos\lr{ \omega(t - t_\nought) } 
\\
M \BR_{\mathrm{CM}}(t) 
\end{bmatrix}
\end{dmath}

The final solution is

\begin{dmath}\label{eqn:twoMassHarmonic:240}
\myBoxed{
\begin{aligned}
\Br_1(t) &=
\frac{\mu}{m_1}
\Delta \Br_\nought
\cos\lr{ \omega(t - t_\nought) } 
+ \BR_{\mathrm{CM}}(t) \\
\Br_2(t) &=
-
\frac{\mu}{m_2}
\Delta \Br_\nought
\cos\lr{ \omega(t - t_\nought) } 
+ \BR_{\mathrm{CM}}(t) 
\end{aligned}
}
\end{dmath}

FIXME: something not right here!  Seems to allow for masses to pass through each other, through the center of mass position.  Think that I need a different general solution to the difference equation, one that allows for both initial position and velocities.
} % makeanswer

%\EndArticle
\EndNoBibArticle
