%
% Copyright © 2016 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\input{../blogpost.tex}
\renewcommand{\basename}{uwaves2}
\renewcommand{\dirname}{notes/ece1236/}
\newcommand{\keywords}{ECE1236H}
\input{../peeter_prologue_print2.tex}

%\usepackage{ece1236}
\usepackage{peeters_braket}
%\usepackage{peeters_layout_exercise}
\usepackage{peeters_figures}
\usepackage{enumerate}
\usepackage{mathtools}
\usepackage{macros_cal}

\beginArtNoToc
\generatetitle{ECE1236H Microwave and Millimeter-Wave Techniques.  Lecture 2: Wave equation.  Taught by Prof.\ G.V. Eleftheriades}
%\chapter{Wave equation}
\label{chap:uwaves2}

\paragraph{Disclaimer}

Peeter's lecture notes from class.  These may be incoherent and rough.

These are notes for the UofT course ECE1236H, Microwave and Millimeter-Wave Techniques, taught by Prof. G.V. Eleftheriades, covering \textchapref{{1}} \citep{pozar2009microwave} content.

\section{Plane waves}

In simple media with no sources

\begin{subequations}
\label{eqn:uwavesLecture2:20}
\begin{equation}\label{eqn:uwavesLecture2:40}
\spacegrad \cross \BE = - \PD{t}{\BB} = - \mu \PD{t}{\BH},
\end{equation}
\begin{equation}\label{eqn:uwavesLecture2:60}
\spacegrad \cdot \BH = 0.
\end{equation}
\begin{equation}\label{eqn:uwavesLecture2:80}
\spacegrad \cross \BH = 
\BJ + \PD{t}{\BD} = 
\lr{ \sigma + \epsilon \PD{t}{} } \BE
\end{equation}
\begin{equation}\label{eqn:uwavesLecture2:100}
\spacegrad \cdot \BE = \frac{\rho}{\epsilon} = 0
\end{equation}
\end{subequations}

Here \( \BJ = \sigma \BE \) is considered to be an induced current, and the \( \rho/\epsilon \) term vanishes since we have no sources.

From Faraday's law \cref{eqn:uwavesLecture2:40}

\begin{dmath}\label{eqn:uwavesLecture2:120}
\spacegrad \cross \lr{ \spacegrad \cross \BE }
= - \mu \PD{t}{} \lr { \spacegrad \cross \BH }
= - \mu \PD{t}{} \lr { \sigma + \epsilon \PD{t}{} } \BE
= - \mu \sigma \PD{t}{\BE} - \epsilon \mu \PDSq{t}{\BE}.
\end{dmath}

We can also use the vector identity

\begin{dmath}\label{eqn:uwavesLecture2:140}
\spacegrad \cross \lr{ \spacegrad \cross \BF } = \spacegrad \lr{ \spacegrad \cdot \BF } - \spacegrad^2 \BF,
\end{dmath}

so 
\begin{dmath}\label{eqn:uwavesLecture2:160}
\spacegrad \cross \lr{ \spacegrad \cross \BE }
=
\spacegrad \lr{ \cancel{\spacegrad \cdot \BE} } - \spacegrad^2 \BE,
\end{dmath}

or
\begin{dmath}\label{eqn:uwavesLecture2:180}
0 = \spacegrad^2 \BE - \mu \sigma \PD{t}{\BE} - \epsilon \mu \PDSq{t}{\BE}.
\end{dmath}

The \( \mu \sigma \PDi{t}{\BE} \) term is a damping contribution.

Similarily for the magnetic field
\begin{dmath}\label{eqn:uwavesLecture2:200}
0 = \spacegrad^2 \BH - \mu \sigma \PD{t}{\BH} - \epsilon \mu \PDSq{t}{\BH}.
\end{dmath}

This is the wave or Helmholtz equation.

Note that for the no loss condition \( \sigma = 0 \), we have the undamped wave equations
\begin{subequations}
\label{eqn:uwavesLecture2:220}
\begin{dmath}\label{eqn:uwavesLecture2:240}
\spacegrad^2 \BE = \epsilon \mu \PDSq{t}{\BE}.
\end{dmath}
\begin{dmath}\label{eqn:uwavesLecture2:260}
\spacegrad^2 \BH = \epsilon \mu \PDSq{t}{\BH}.
\end{dmath}
\end{subequations}

This is one wave equation for each of \( E_x, E_y, E_z, H_x, H_y\) and \( H_z \).

\paragraph{Propagation in one direction}

Consider the undamped propagation along the z direction of \( E_x(z, t) \), which must satisfy
\begin{dmath}\label{eqn:uwavesLecture2:400}
\spacegrad^2 E_x = \epsilon \mu \PDSq{t}{E_x},
\end{dmath}

or
\begin{dmath}\label{eqn:uwavesLecture2:420}
\PDSq{z}{E_x} = \epsilon \mu \PDSq{t}{E_x}.
\end{dmath}

This is analogous to the voltage transmission line equation
\begin{dmath}\label{eqn:uwavesLecture2:440}
\PDSq{z}{V} = L C \PDSq{t}{V},
\end{dmath}

With solution 

\begin{equation}\label{eqn:uwavesLecture2:280}
V(z, t) = 
V_0^{+} f(z - v_\phi t)
+
V_0^{-} f(z + v_\phi t),
\end{equation}

where \( v_\phi = \ifrac{1}{\sqrt{L C}} \) is the phase velocity, \( f(z - v_\phi t)\) is the forward wave, and \( f(z + v_\phi t) \) is the reflected wave.  By analogy the solution to the 1D wave equation is

\begin{equation}\label{eqn:uwavesLecture2:300}
E_x(z, t) = 
E_0^{+} f(z - v_\phi t)
+
E_0^{-} f(z + v_\phi t),
\end{equation}

where \( v_\phi = \ifrac{1}{\sqrt{\epsilon \mu}} \) is the phase velocity.

For myself, the transmission line equation is something that I only encountered in this class, so this analogy isn't a great one.  That said, wave equation solutions are very familiar, so not much motivation for the structure of the solution is really required.

\paragraph{More rigorous derivation of the 1D solution}

Suppose that we assume the form of the solution is

\begin{equation}\label{eqn:uwavesLecture2:320}
\BE = \BE_0 f(z - v_\phi t ),
\end{equation}

where \( \BE_0 \) is a constant vector, so this describes a wave propagation in the z-direction, but without a-priori knowledge of the direction of this vector in space.

For source free media, we have

\begin{dmath}\label{eqn:uwavesLecture2:340}
\spacegrad \cdot ( \BE_0 f ) 
= \spacegrad f \cdot \BE_0 + f \cancel{ \spacegrad \cdot \BE }
= (\zcap f') \cdot \BE_0.
\end{dmath}

Because \( \BE_0 \cdot \zcap = 0 \), the vector \( \BE_0 \) is perpendicular to the direction of propagation ( z ).  The next task is to find the magnetic field that couples to this electric field solution.  Suppose the coordinate axis are picked so that 

\begin{equation}\label{eqn:uwavesLecture2:360}
\BE(z, t) = \BE_0 E_x( z \pm v_\phi t).
\end{equation}

Using Faraday's law \( \spacegrad \cross \BE = -\mu \PDi{t}{\BH} \) gives
\begin{dmath}\label{eqn:uwavesLecture2:380}
\spacegrad \cross (\BE_0 E_x)
=
(\spacegrad \cross \BE_0) E_x
-\BE_0 \cross \spacegrad E_x
=
-\BE_0 \cross \spacegrad E_x
=
-\BE_0 \cross \zcap \PD{z}{E_x},
\end{dmath}

so

\begin{dmath}\label{eqn:uwavesLecture2:460}
-\mu \PD{t}{\BH} = 
\lr{ \zcap \cross \BE_0 } \PD{z}{E_x}.
\end{dmath}

Because of the \( E_x( z \pm v_\phi t ) \) dependence on \( z, t \), we must have

\begin{dmath}\label{eqn:uwavesLecture2:480}
\PD{z}{E_x} = \pm \inv{v_\phi} \PD{t}{E_x},
\end{dmath}

or
\begin{dmath}\label{eqn:uwavesLecture2:500}
-\mu \PD{t}{\BH} = 
\lr{ \zcap \cross \BE_0 } 
= \pm \inv{v_\phi} \PD{t}{E_x}.
\end{dmath}

Integrating with respect to \( t \) and assuming a zero integration constant (physical justification for that?), we have

\begin{dmath}\label{eqn:uwavesLecture2:520}
\BH = \mp E_x(z, t) \frac{\zcap \cross \BE_0}{v_\phi \mu}.
\end{dmath}

The \( v_\phi \mu \) product is the intrinsic wave impedance of the medium

\begin{dmath}\label{eqn:uwavesLecture2:540}
v_\phi \mu = \frac{\mu}{\sqrt{\epsilon\mu}} = \sqrt{\frac{\mu}{\epsilon}} = \eta.
\end{dmath}

The magnetic field can be written in terms of the propagation direction \( \ncap = \mp \zcap \) has

\begin{equation}\label{eqn:uwavesLecture2:560}
\BH 
= \mp \frac{\zcap \cross \BE}{\eta}
= \frac{\ncap \cross \BE}{\eta}.
\end{equation}

Note that in free space we have \( \eta \approx 377 \Omega \).

\paragraph{summary}

For 1D wave propagation along the z axis we have

\begin{equation}\label{eqn:uwavesLecture2:580}
\begin{aligned}
\spacegrad^2 \BE &= \epsilon \mu \PDSq{t}{\BE} \\
\spacegrad^2 \BH &= \epsilon \mu \PDSq{t}{\BH} \\
\end{aligned}
\end{equation}

where 

\begin{equation}\label{eqn:uwavesLecture2:600}
\begin{aligned}
\BE &= \BE_0 f( z \mp v_\phi t ) \\
0 &= \BE_0 \cdot \zcap \\
\BH &= \pm \frac{\zcap \cross \BE_0}{\eta} f( z \mp v_\phi t )
\end{aligned}
\end{equation}

Note that \( \pm \zcap \) is the direction of propagation, with \( +z \) being the forward wave, and \( -z \) the backwards wave.

The phase velocity is

\begin{equation}\label{eqn:uwavesLecture2:620}
v_\phi = \frac{1}{\sqrt{\mu\epsilon}},
\end{equation}

and the intrinsic wave impedance is 

\begin{equation}\label{eqn:uwavesLecture2:640}
\eta = \sqrt{\frac{\mu}{\epsilon}}.
\end{equation}

Notes:

\begin{itemize}
\item \( \BE, \BH \) are perpendicular to the direction of propagation \( \zcap \).  These are transverse waves.
\item \( \Abs{\BE}/\Abs{\BH} = \eta \).  An analogy with transmission lines is
\begin{equation}\label{eqn:uwavesLecture2:660}
\begin{aligned}
V &\leftrightarrow E \\
I &\leftrightarrow H \\
Z_0 &\leftrightarrow \eta
\end{aligned}
\end{equation}
\item \( \BE, \BH \) are orthogonal to each other

F1: B.5
\end{itemize}

\section{Time harmonic fields}

Assume that the fields var sinusoidally and use phasor notation

\begin{equation}\label{eqn:uwavesLecture2:680}
\BE = \Real\lr{ \BE(\Br) e^{j \omega t} },
\end{equation}

where \( \BE(\Br) \) is a (complex) vector-valued phasor.  Notes:

\begin{enumerate}[(i)]
\item Physicists use a \( e^{-j \omega t} \) notation for the time dependence.
\item Sometimes RMS values are used instead of peak values, so that
\begin{equation}\label{eqn:uwavesLecture2:700}
\BE = \sqrt{2} \Real\lr{ \BE(\Br) e^{j \omega t} },
\end{equation}

and \( \BE(\Br) \) is an RMS phasor.
\item The time derivative \( \PDi{t}{} \rightarrow j\omega \) in Maxwell's equations.
\item The \( e^{j \omega t} \) variation is dropped for convienience.
\end{enumerate}

The \( e^{j \omega t} \) time harmonic dependence is not restrictive because one can synthesize any function of t using a Fourier transform

\begin{equation}\label{eqn:uwavesLecture2:720}
\calF \lr{ f(t) } = \tilde{f}(\omega) = \int_{-\infty}^\infty f(t) e^{-j \omega t} dt,
\end{equation}

where \( \tilde{f}(\omega) \) is the spectrum.  If the spectrum is known then the inverse Fourier transform is

\begin{equation}\label{eqn:uwavesLecture2:740}
f(t) = \inv{2 \pi} \int_{-\infty}^\infty \tilde{f}(\omega) e^{j \omega t} dt,
\end{equation}

If the responce to a time harmonic signal \( e^{j \omega t} \) is known then the response to an arbitrary waveform \( f(t) \) can be synthesized, as shown in F2 where the waveform input is \( f(t) = \inv{2 \pi} \int_{-\infty}^\infty \tilde{f}(\omega) e^{j \omega t} dt \) and the waveform output is \( \inv{2 \pi} \int_{-\infty}^\infty \tilde{f}(\omega) G(\omega) e^{j \omega t} dt \)

F2: B.7 (just the boxes and inputs)

If the fields are known in phasor form \( \BE(\Br, \omega ) \) then the repsonse to an arbitrary signal with spectrum \( \tilde{f}(\omega) \) is

\begin{equation}\label{eqn:uwavesLecture2:760}
\BE(\Br, t ) = \inv{2 \pi} \int_{-\infty}^\infty \BE(\Br, \omega) \tilde{f}(\omega) e^{j \omega t} dt.
\end{equation}

FIXME: unsure what \( \tilde{f}(\omega) \) is here?  Isn't \( \BE(\Br, \omega) \) the spectrum of the field?  Is this supposed to be the field response to the system \( G(\omega) \) ?

\paragraph{Maxwell's equations in phasor form}

Faraday's law, the Ampere-Maxwell law, the continuity equation, and the (linear, homogenous, isotropic) constituative relations respectively are
\begin{subequations}
\label{eqn:uwavesLecture2:780}
\begin{equation}\label{eqn:uwavesLecture2:800}
\spacegrad \cross \BE = - j \omega \BB
\end{equation}
\begin{equation}\label{eqn:uwavesLecture2:820}
\spacegrad \cross \BH = \BJ + j \omega \BD
\end{equation}
\begin{equation}\label{eqn:uwavesLecture2:840}
\spacegrad \cdot \BJ = - j \omega \rho
\end{equation}
\begin{equation}\label{eqn:uwavesLecture2:860}
\begin{aligned}
\BB &= \mu \BH \\
\BD &= \epsilon \BE \\
\end{aligned}
\end{equation}
\end{subequations}

Note that \( \BD(\Br, \omega) = \epsilon(\omega) \BE(\Br, \omega) \) becomes a convolution in the time domain

\begin{equation}\label{eqn:uwavesLecture2:880}
\BD(\Br, t) = \int \BE(\Br, t - \tau) \epsilon(\tau) d\tau.
\end{equation}

This describes the delay between the \( \BE \) field and the \( \BD \) field, the polarization.

\paragraph{1D propagation, and sinusoidal waves}

\EndArticle
%\EndNoBibArticle
