% 
% 
% 
% Copyright © 2012 Peeter Joot
% All Rights Reserved
% 
% This file may be reproduced and distributed in whole or in part, without fee, subject to the following conditions:
% 
% o The copyright notice above and this permission notice must be preserved complete on all complete or partial copies.
% 
% o Any translation or derived work must be approved by the author in writing before distribution.
% 
% o If you distribute this work in part, instructions for obtaining the complete version of this file must be included, and a means for obtaining a complete version provided.
% 
% 
% Exceptions to these rules may be granted for academic purposes: Write to the author and ask.
% 
% 
% 
\chapter*{Preface}\normalsize
  \addcontentsline{toc}{chapter}{Preface}

These are my personal lecture notes for the Spring 2012, University of Toronto Advanced Classical Mechanics course (PHY354H1S).  This class was taught by Prof.\ Erich Poppitz.  I audited the wednesday lectures since the timing was convienent.  I took occasional notes, and did some of the problems.

The official description of this course was:

Symmetry and conservation laws, stability and instability, generalized co-ordinates, Hamilton's principle, Hamilton's equations, phase space, Liouville's theorem, canonical transformations, Poisson brackets, Noether's theorem.

The text for the course is \citep{landau1960classical}, but I don't actually own that one (I've got \cite{goldstein1951cm}).

This document contains a few things

\begin{itemize}
\item A collection of miscellanious notes and problems for my personal (independent) classical mechanics studies.

\item My notes for some of the lectures I attended (some of the earlier lectures I didn't actively take notes for).

Typos and errors are probably mine (Peeter), and no claim nor attempt of spelling or grammar correctness will be made.  

These notes, when I took them, likely track along with the Professor's hand written notes very closely, since his lectures follow his notes very closely.

\item Some assigned problems from the phy354 course, ungraded (not submitted since I didn't actually take the course).

%\item Some worked problems associated with exam preparation.
\end{itemize}

Peeter Joot  \quad peeter.joot@gmail.com 
