% 
% 
% 
% Copyright © 2012 Peeter Joot
% All Rights Reserved
% 
% This file may be reproduced and distributed in whole or in part, without fee, subject to the following conditions:
% 
% o The copyright notice above and this permission notice must be preserved complete on all complete or partial copies.
% 
% o Any translation or derived work must be approved by the author in writing before distribution.
% 
% o If you distribute this work in part, instructions for obtaining the complete version of this file must be included, and a means for obtaining a complete version provided.
% 
% 
% Exceptions to these rules may be granted for academic purposes: Write to the author and ask.
% 
% 
% 
%\chapter{Preface}
% this suppresses an explicit chapter number for the preface.
\chapter*{Preface}%\normalsize
  \addcontentsline{toc}{chapter}{Preface}

This document is based on my lecture notes for the Winter 2012, University of Toronto Continuum Mechanics course (PHY454H1S), taught by Prof.\ Kausik S. Das.  

My thanks to Professor Das for teaching this course.  It covered the fundamentals of fluid dynamics in a sensible and logical fashion, providing a great base for further learning.

Official course description:

The theory of continuous matter, including solid and fluid mechanics.  Topics include the continuum approximation, dimensional analysis, stress, strain, the Euler and Navier-Stokes equations, vorticity, waves, instabilities, convection and turbulence.

What you will find in this document:

\begin{itemize}
\item My lecture notes, reworked after completion of the course into what is hopefully a more coherent and compact form.

\item Problem sets and midterm solutions. These have been incorporated into the lecture material as chapter end problems with solutions.

\item Some worked problems attempted for fun or for exam preparation.

\item Links to Mathematica workbooks associated with course content.

\end{itemize}

Peeter Joot  \quad peeter.joot@gmail.com 
