%\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{A different derivation of the adiabatic perturbation coefficient equation}
\label{chap:adiabaticPertTakeII}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2011/adiabaticPertTakeII.pdf}
\date{Oct 27, 2011}
\revisionInfo{adiabaticPertTakeII.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Motivation.}

Professor Sipe's adiabatic perturbation and that of the text \cite{desai2009quantum} in \S 17.5.1 and \S 17.5.2 use different notation for $\gamma_m$ and take a slightly different approach.  We can find Prof Sipe's final result with a bit less work, if a hybrid of the two methods is used.

\section{Guts}

Our starting point is the same, we have a time dependent slowly varying Hamiltonian 

\begin{equation}\label{eqn:adiabaticPertTakeII:10}
H = H(t),
\end{equation}

where our perturbation starts at some specific time from a given initial state

\begin{equation}\label{eqn:adiabaticPertTakeII:30}
H(t) = H_0, \qquad t \le 0.
\end{equation}

We assume that instantaneous eigenkets can be found, satisfying

\begin{equation}\label{eqn:adiabaticPertTakeII:50}
H(t) \ket{n(t)} = E_n(t) \ket{n(t)}
\end{equation}

Here I'll use $\ket{n} \equiv \ket{n(t)}$ instead of the $\ket{\hat{\psi}_n(t)}$ that we used in class because its easier to write.

Now suppose that we have some arbitrary state, expressed in terms of the instantaneous basis kets $\ket{n}$

\begin{equation}\label{eqn:adiabaticPertTakeII:70}
\ket{\psi} = \sum_n \bar{b}_n(t) e^{-i\alpha_n + i \gamma_n} \ket{n},
\end{equation}

where

\begin{equation}\label{eqn:adiabaticPertTakeII:90}
\alpha_n(t) = \inv{\hbar} \int_0^t dt' E_n(t'),
\end{equation}

and $\gamma_n$ (using the notation in the text, not in class) is to be determined.

For this state, we have at the time just before the perturbation

\begin{equation}\label{eqn:adiabaticPertTakeII:110}
\ket{\psi(0)} = \sum_n \bar{b}_n(0) e^{-i\alpha_n(0) + i \gamma_n(0)} \ket{n(0)}.
\end{equation}

The question to answer is: How does this particular state evolve?

Another question, for those that don't like sneaky bastard derivations, is where did that magic factor of $e^{-i\alpha_n}$ come from in our superposition state? We will see after we start taking derivatives that this is what we need to cancel the $H(t)\ket{n}$ in Schr\"{o}dinger's equation.

Proceeding to plug into the evolution identity we have

\begin{align*}
0 &=
\bra{m} \left( i \hbar \ddt{} - H(t) \right) \ket{\psi} \\
&=
\bra{m} \left(
\sum_n 
e^{-i \alpha_n + i \gamma_n}
(i \hbar) \left(
\ddt{\bar{b}_n}
+ \bar{b}_n \left(-i \cancel{\frac{E_n}{\hbar}} + i \dot{\gamma}_m \right)
\right) \ket{n}
+ i \hbar \bar{b}_n \ddt{} \ket{n}
- \cancel{E_n \bar{b}_n \ket{n}} 
\right)
\\
&=
e^{-i \alpha_m + i \gamma_m}
(i \hbar) 
\ddt{\bar{b}_m}
+
e^{-i \alpha_m + i \gamma_m}
(i \hbar) 
i \dot{\gamma}_m \bar{b}_m
+ i \hbar \sum_n \bar{b}_n \bra{m} \ddt{} \ket{n}
e^{-i \alpha_n + i \gamma_n} \\
&\sim
\ddt{\bar{b}_m}
+
i \dot{\gamma}_m \bar{b}_m
+ 
\sum_n 
e^{-i \alpha_n + i \gamma_n}
e^{i \alpha_m - i \gamma_m}
\bar{b}_n \bra{m} \ddt{} \ket{n} \\
&=
\ddt{\bar{b}_m}
+
i \dot{\gamma}_m \bar{b}_m
+ 
\bar{b}_m \bra{m} \ddt{} \ket{m}
+
\sum_{n \ne m} 
e^{-i \alpha_n + i \gamma_n}
e^{i \alpha_m - i \gamma_m}
\bar{b}_n \bra{m} \ddt{} \ket{n}
\end{align*}

We are free to pick $\gamma_m$ to kill the second and third terms

\begin{equation}\label{eqn:adiabaticPertTakeII:130}
0 =
i \dot{\gamma}_m \bar{b}_m
+ 
\bar{b}_m \bra{m} \ddt{} \ket{m},
\end{equation}

or

\begin{equation}\label{eqn:adiabaticPertTakeII:150}
\dot{\gamma}_m 
= 
i \bra{m} \ddt{} \ket{m},
\end{equation}

which after integration is

\begin{equation}\label{eqn:adiabaticPertTakeII:170}
\gamma_m(t)
= 
i \int_0^t dt' \bra{m(t')} \frac{d}{dt'} \ket{m(t)},
\end{equation}

as in class we can observe that this is a purely real function.  We are left with

\begin{equation}\label{eqn:adiabaticPertTakeII:190}
\ddt{\bar{b}_m}
=
-
\sum_{n \ne m} \bar{b}_n 
e^{-i \alpha_{nm} + i \gamma_{nm}}
\bra{m} \ddt{} \ket{n} 
,
\end{equation}

where

\begin{align}\label{eqn:adiabaticPertTakeII:210}
\alpha_{nm} &= \alpha_{n} -\alpha_m \\
\gamma_{nm} &= \gamma_{n} -\gamma_m 
\end{align}

The task is now to find solutions for these $\bar{b}_m$ coefficients, and we can refer to the class notes for that without change.

\EndArticle
