% mp134: mon/wed/fri: 11-12
\documentclass[12pt,leqno]{book}

\usepackage{amsmath,amssymb,amsfonts} % Typical maths resource packages
\usepackage{graphicx}
\usepackage{color}                   % For creating coloured text and background
\usepackage{txfonts} 
\usepackage{listings}
\usepackage[bookmarks=true,plainpages=false]{hyperref}

\parindent 1cm
\parskip 0.2cm
\topmargin 0.2cm
\oddsidemargin 1cm
\evensidemargin 0.5cm
\textwidth 15cm
\textheight 21cm

\input{../peeters_macros.tex}

\newcommand{\citep}[1]{\cite{#1}}
\newcommand{\citet}[1]{\cite{#1}}
\newcommand{\chapcite}[1]{\ref{chap:#1}}

%-----------------------------------------
%
% stubs for article class.
%
\newcommand{\blogpage}[1]{}
\newcommand{\email}[1]{}
\newcommand{\beginArtWithToc}[0]{}
\newcommand{\beginArtNoToc}[0]{}
\newcommand{\EndArticle}[0]{}
\newcommand{\EndNoBibArticle}[0]{}
\newcommand{\revisionInfo}[1]{}
%-----------------------------------------
\DeclareMathOperator{\Atan}{atan}

%\include{new_commands}

%\makeindex

\begin{document}
\pagenumbering{alph}

\title{Course notes and problems from\\University of Toronto PHY456H1F\\Quantum Physics II.}
\author{Peeter Joot \quad peeter.joot@gmail.com}

\maketitle

\clearpage\pagenumbering{roman}
\tableofcontents

\clearpage\pagenumbering{arabic}

\pagestyle{plain}

\include{preface}

\part{Readings}

\chapter{Reading and problem status}

\begin{itemize}
\item \S 24.2 Variational method.  -- reading done.  problems 3-6 done.  problem 7 would be good to do.
\item \S 16.1 - \S 16.2 Time independent pertubation, theory and HO example.  Reading done.
\item \S 16.4 stark effect.  Not covered in class.  I'm having trouble confirming 16.72 using 16.66?  See 'desai attempt to verify section 16.3.nb'.  Revisit after course.
\item \S 16.5 degeneracy.  Excellent description (helpful to also read \S 13.1.1 - \S 13.1.2 on two level problems.)  Revisit degeneracy problem set question having read this, and derive the pertubation expansion for two simple cases: 1 two fold degneracy + 1 non-degenerate state.  1 two fold degeneracy + 2 independent non-degenerate states.  Then do the general derivation.
\item \S 16.6: Problems: All look like good ones.  Haven't done any yet.
\item \S 3.3 Reread.  Prof Sipe's coverage was much clearer.

\item \S 17.1 Time dependent perturbation theory.  Reading: done.
\item \S 18.3 Coloumb excitation.  Given as an example in lecture 6 when starting time dependent pertubation.  Rather than working with the Taylor expansion and identifying the first term as the electric field the text works with the operators directly, and goes a lot deeper than we did in class (class goes to about 18.53 and then skips the rest of the section).  This would be good to revisit after the course.
\item \S 17.5.1 Adiabatic perturbation.  Reading: done.
\item \S 17.5.2 Berry phase. Reading: todo.
\item \S 17.2 Fermi's golden rule. Reading: todo.
\item SKIP: \S 17.3 - \S 17.4 scattering cross section, resonance and decay.  Not yet covered in class.  Skip for now, and revisit after or later in course.
\item \S 17.6 Problems.  Most look appropriate.
\end{itemize}
%-------------------------------------------------------

\part{Lecture Notes.}
\include{qmTwoL1}
\include{qmTwoL2}
\include{qmTwoL3}
\include{qmTwoL4}
\include{qmTwoL5}
\include{qmTwoL6}
\include{qmTwoL7}
\include{qmTwoL8}
\include{qmTwoL9}
\include{qmTwoL10}
\part{Notes and Problems.}
\include{phy456ps1}
\include{variationalHelium}
\include{variationHarmonicOscillator}
\include{entangledSimpleExample}
\include{qmTwoPs4}
%\part{More worked problems.}

%\part{Cronology}
%\include{cronology}
\part{Bibliography}

% END INCLUDES.
%-------------------------------------------------------

\bibliography{myrefs}
\bibliographystyle{unsrturl}
  \addcontentsline{toc}{chapter}{Bibliography}

\end{document}
%\include{qmTwoL11}
