\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}
%\newcommand{\LL}[0]{\mathcal{L}}
%\newcommand{\grad}[0]{\nabla}
%\newcommand{\pdot}[0]{\dot{p}}
%\newcommand{\pddot}[0]{\ddot{p}}

\usepackage[bookmarks=true]{hyperref}

\title{ Lorentz force interaction } % Declares the document's title.
\author{Peeter Joot \quad peeter.joot@gmail.com}
\date{ October 17, 2008.  Last Revision: $Date: 2009/04/15 03:40:22 $ }

\begin{document}

\maketitle{}
\tableofcontents

\section{ Motivation.}

Explore the relationships between proper velocity and current density.

\cite{goldstein1951cm} has a combined Lagrangian for the field and Lorentz force:

\begin{align*}
\LL = \int \left\{ \frac{E^2 - B^2}{8 \pi} - \sum_i q_i \delta(\Br-\Br_i)\left(\phi - \Bv_i/c \cdot \BA\right)\right\} dV + \inv{2} m_i \Bv_i^2
\end{align*}

Where alternate variation of the field, or coordinates produces the field equations or Lorentz equation respectively.

Compare to the covariant Lagrangian for the field and interaction (for metric $+---$)

\begin{align}\label{eqn:maxlag}
\LL_{\text{field}} &= -\frac{\epsilon_0}{2} (\grad \wedge A)^2 + A \cdot J/c \\
\LL_{\text{interaction}} &= \inv{2} m v^2 + q A \cdot (v/c)
\end{align}

(SI units this time).

These can be shown to produce Maxwell's equation and the Lorentz force equation respectively
\begin{align}
\grad F &= J/c\epsilon_0 \\
\pdot &= q F \cdot (v/c)
\end{align}

Derivations of these can be found in \cite{PJFieldLagrangian}, and \cite{PJSrLorentzForce} respectively.

The final aim is
find how to relate the $A\cdot J$, and $q A \cdot v$ interaction quantities in covariant field and
Lorentz Lagrangians.  Intuition tells me that considering the Lorentz transformation of $J$ and $A$ may help
shed some light on this.

Some four potentials discussion perhaps useful as background can be found in \cite{PJFourPotential}.

Also interesting is that in
\cite{pauli2000wm} the Lorentz force equation (in non-covariant form) is
derived as a limiting classical case via calculation of the expectation
value of the Hamiltonian

\begin{align}\label{eqn:interactionConjMomHamiltonian}
H = \inv{2m}\sum_{k=1}^3 \left(p_k - \frac{e}{c} A_k \right)^2 + V
\end{align}

This has a much different looking structure than the $\LL_{\text{interaction}}$ above, so reconciliation of the two
is justifiable.

\section{ Lorentz force Lagrangian with conjugate momentum. }

Let's start with the interaction in terms of the conjugate momentum.  Is there a relativistic form for the interaction
Lagrangian with a structure similar to \ref{eqn:interactionConjMomHamiltonian}?

Let's try

\begin{align*}
\LL
&= \inv{2 m}\left( m v - \kappa A \right)^2 \\
&= \inv{2 m}\left( m^2 v^2 - 2 m \kappa A \cdot v + \kappa^2 A^2 \right)^2 \\
&= \inv{2 m} \left(
m^2 \xdot^\alpha \xdot_\alpha
- 2 m \kappa A_\alpha \xdot^\alpha
+ \kappa^2 A_\alpha A^\alpha
\right) \\
\end{align*}

where $\kappa$ is to be determined.

For this Lagrangian the Euler-Lagrange calculation for variation of $S = \int d^4 x \LL$ is

\begin{align*}
\PD{x^\mu}{\LL}
&= - \kappa (\partial_\mu A_\alpha) \xdot^\alpha + \inv{m} \kappa^2 (\partial_\mu A_\alpha) A^\alpha
\frac{d}{d\tau} \PD{\xdot^\mu}{\LL} \\
&= \frac{d}{d\tau} \left(m \xdot_\mu - \kappa A_\mu \right)
\end{align*}

Assembling and shuffling we have
\begin{align*}
m \xddot_\mu
&= \kappa (\partial_\alpha A_\mu) \xdot^\alpha
- \kappa (\partial_\mu A_\alpha) \xdot^\alpha + \inv{m} \kappa^2 (\partial_\mu A_\alpha) A^\alpha \\
&= \kappa (\partial_\alpha A_\mu - \partial_\mu A_\alpha) \xdot^\alpha + \inv{m} \kappa^2 (\partial_\mu A_\alpha) A^\alpha \\
&= \kappa F_{\alpha\mu} \xdot^\alpha + \inv{m} \kappa^2 (\partial_\mu A_\alpha) A^\alpha \\
\end{align*}

Comparing to the Lorentz force equation (again for a $+---$ metric)

\begin{align*}
m v_\mu = \frac{q}{c} F_{\mu\nu} v^\nu
\end{align*}

We see that we need $\kappa = -q/c$, but we have an extra factor that doesn't look familiar.  In vector form this Lagrangian would give us the
equation of motion

\begin{align*}
m v = \frac{q}{c} F \cdot v + \frac{q^2}{m c^2} (\grad A_\mu) A^\mu
\end{align*}

Assuming that this extra term has no place in the Lorentz force equation we need to adjust the original Lagrangian as follows

\begin{align}
\LL
&= \inv{2 m}\left( m v + \frac{q}{c} A \right)^2 - \frac{q^2}{ 2 m c^2} A^2
\end{align}

Having put the Lagrangian in this form, expansion of the first square is just

\begin{align*}
\LL &= \inv{2} m v^2 + \frac{q}{c} A \cdot v 
\end{align*}

which is exactly the original Lorentz force Lagrangian.  Expressing this energy density in terms of the canonical momentum is 
somewhat interesting, and puts some extra structure into the equation, but in the end it doesn't be any fundamentally different.

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
