\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}
\newcommand{\EE}[0]{\boldsymbol{\mathcal{E}}}
\newcommand{\HH}[0]{\boldsymbol{\mathcal{H}}}
%\newcommand{\PDSq}[2]{\frac{\partial^2 {#2}}{\partial {#1}^2}}

\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage

\title{}
\author{Peeter Joot}
\date{ Feb dd, 2009.  Last Revision: $Date: 2009/02/07 15:20:07 $ }

\begin{document}

\maketitle{}

\tableofcontents

\section{}

\section{ Misc.  Doesn't fit yet. }

Taken from \cite{PJFourierVacuum}

\subsection{ Electric and magnetic field components. }

Given $F$ we can compute the electric and magnetic field components using a spacetime observer split

\begin{align*}
\EE &= \inv{2}(F - \gamma_0 F \gamma_0 ) \\
\HH &= \inv{2i}(F + \gamma_0 F \gamma_0 ) \\
\end{align*}

An alternative is a split by components, which also implicitly utilizes
a specific observer time basis vector.  That would be
\begin{align*}
F = E^m \sigma_m + i H^m \sigma_m
\end{align*}

This provides the electric and magnetic field components:
\begin{align*}
E^m &= F \cdot \sigma_m \\
H^m &= (-i F) \cdot \sigma_m
\end{align*}

We are used to separate electric and magnetic fields, as measurable and experiencable quantities.  So reexpressing the previous results 
in terms of separate fields should take some of the abstraction out of the picture.  Additionally, in the Fourier components we expect, or at least ought to be able to show, a $\Bk$ dependence between $\EE$ and $\HH$.  Let's try this

FIXME:TODO...

\subsection{ Separate electric and magnetic fields with the boundary conditions unspecified. }

Fixing the boundary value in terms of the initial field at $t=0$ is not the only option.  We see similar things in classical mechanics in constant acceleration problems where one can use an initial position and velocity, or positions at two different times, and so forth.  Bohm leaves his equivalents to the integration constants $C_{\Bk}$ unspecified.  If we do so too we have equation \ref{eqn:undetermined}.  How does that look with separate fields?

FIXME:TODO:...  

Cut and Paste Note.  That referenced equation was:
\begin{align}\label{eqn:undetermined}
F(\Bx,t) = \sum_{\Bk} 
\exp\left(i \Bk c t \right) 
\exp\left(-i \Bk \cdot \Bx \right) 
C_{\Bk} 
\end{align}

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
