%
% Copyright © 2016 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%{
\input{../blogpost.tex}
\renewcommand{\basename}{curlAndDivergenceVectorUniqueness}
\renewcommand{\dirname}{notes/phy1520/}
%\newcommand{\dateintitle}{}
%\newcommand{\keywords}{}

\input{../peeter_prologue_print2.tex}

\usepackage{peeters_layout_exercise}
\usepackage{peeters_braket}
\usepackage{peeters_figures}
\usepackage{siunitx}

\beginArtNoToc

\generatetitle{Does the divergence and curl uniquely determine the vector?}
%\chapter{Does the divergence and curl uniquely determine the vector?}
%\label{chap:curlAndDivergenceVectorUniqueness}

%Here's a homework question from ece1228 that screams for solution using Geometric Algebra techniques.  If I submitted such a solution, then my Prof probably wouldn't know what I was doing, so I'll probably also have to try to solve it another way too.

A problem posed in the ece1228 problem set was the following

\paragraph{Helmholtz theorem.}
Prove the first Helmholtz's theorem, i.e. if vector \(\BM\) is defined by its divergence

\begin{dmath}\label{eqn:emtProblemSet1Problem5:20}
\spacegrad \cdot \BM = s
\end{dmath}

and its curl
\begin{dmath}\label{eqn:emtProblemSet1Problem5:40}
\spacegrad \cross \BM = \BC
\end{dmath}

within a region and its normal component \( \BM_{\txtn} \) over the boundary, then \( \BM \) is
uniquely specified.

\paragraph{Solution.}

\input{../ece1228-electromagnetic-theory/gaPs1Problem5Attempt1.tex}

\paragraph{Commentary}

I attempted this problem in three different ways.  My first approach (above) assembled the divergence and curl relations above into a single (Geometric Algebra) multivector gradient equation and applied the vector valued Green's function for the gradient to invert that equation.  That approach logically led from the differential equation for \( \BM \) to the solution for \( \BM \) in terms of \( s \) and \( \BC \).  However, this strategy introduced some complexities
that make me doubt the correctness of the associated boundary analysis.

Even if the details of the boundary handling in my multivector approach is not correct, I thought that approach was interesting enough to share.  
%It is accompanied with a primer on Geometric Algebra that is hopefully enough to allow the reader to grasp the basic ideas of the approach, but is probably not sufficient to understand all the details without further study.

%}
\EndArticle
