\chapter*{Preface}\normalsize
  \addcontentsline{toc}{chapter}{Preface}

This document contains a few things

\begin{itemize}
\item My personal lecture notes for the Spring 2011, University of Relativistic Electrodynamics course (PHY450H1S).  This class was taught by Prof.\ Erich Poppitz, with Simon Freedman handling tutorials (which were excellent lecture style lessons).

Typos and errors, if any, are probably mine (Peeter), and no claim nor attempt of spelling or grammar correctness will be made.  

The text for the course is \cite{landau1980classical}.

These notes track along with the Professor's hand written notes very closely, since his lectures follow his notes very closely.  While I used the note taking exersize as a way to verify that I understood all the day's lecture materials, the Professor's notes are in many instances a much better study resource, since there are details in his notes that were left for us to read, and not neccessarily covered in the lectures.  On the other hand, there are details in these notes that I've added when I didn't find his approach simplistic enough for me to grasp, or I failed to follow the details in class.

\item Some notes from reading of the text.  One of the earlier of these was due to unfortunate use of an ancient edition of the text borrowed from the library, since mine had been lost in shipping.  That version didn't use the upper and lower index quantities that I'd expected, so I tried to puzzle out some of what myself from what I knew.

\item Some assigned problems, at least the parts of them that I did not hand write.  I've corrected some the errors after recieving grading feedback, and where I haven't done so I at least recorded some of the grading comments as a reference.  Not all the problems were graded, so I make no guarentees of correctness.

%  This includes observations, notes on what seem like errors, and some solved problems.  
%
%\item Some worked problems associated with exam preparation.
\end{itemize}

Peeter Joot  \quad peeter.joot@gmail.com 
