\documentclass[]{eliblog}

\title{Schr\"{o}dinger Equation as a Relativistic wave equation approximation.}
\author{Peeter Joot}
\date{June 18, 2009}
\blogpage{http://behindtheguesses.blogspot.com/2009/06/blah.html}
\email{peeter.joot@gmail.com}

\newcommand{\Abs}[1]{{\left\lvert{#1}\right\rvert}}
\newcommand{\BB}[0]{\mathbf{B}}
\newcommand{\BE}[0]{\mathbf{E}}
\newcommand{\BS}[0]{\mathbf{S}}
\newcommand{\Bk}[0]{\mathbf{k}}
\newcommand{\Bp}[0]{\mathbf{p}}
\newcommand{\Ba}[0]{\mathbf{a}}
\newcommand{\Bv}[0]{\mathbf{v}}
\newcommand{\Bx}[0]{\mathbf{x}}
\newcommand{\LL}[0]{\mathcal{L}}
\newcommand{\cross}[0]{\times}
\newcommand{\spacegrad}[0]{\boldsymbol{\nabla}}
\newcommand{\delambertian}[0]{\square}
\newcommand{\inv}[1]{\frac{1}{#1}}
\newcommand{\FF}[0]{\mathcal{F}}
\newcommand{\grad}[0]{\nabla}
\newcommand{\kcap}[0]{\hat{\Bk}}

\begin{document}
\maketitle

\section{Alternative approaches to motivate Schr\"{o}dinger's equation.}

It does not take particularily fancy mathematics to motivate the Schr\"{o}dinger's equation.  In
French and Taylor (\cite{french1998iqp}) the specialization of Schr\"{o}dinger's equation for a 
particle in a one dimensional potential

\begin{align}\label{eqn:oneDimSch}
-\frac{\hbar^2}{2m} \frac{\partial^2 \psi}{\partial x^2} + V\psi = i \hbar \frac{\partial \psi}{\partial t}
\end{align}

is presented from start to finish in only seven pages with no mathematics or physics unfamiliar to second year 
engineering undergrads.  However this does include a disclaimer upfront that
``we ... simply try to make the form of the equation plausible''.  The plausibility arguments
found in this text are not uncommon, and can be found for example (without the surrounding discussion of the text) 
in the ``Short heuristic derivation'' of the Wikipedia's Schr\"{o}dinger's equation article (\cite{wikiSchH}).

Calling these derivations is justifably debatable and is perhaps the origin of statements like
``Schr\"{o}dinger's equation cannot be derived'' (\cite{hyperphysicsSch})
Considerable care 
is required to construct logically consistent arguments that motivate the quantum wave
equation in a fashion that is not simply playing the equation in reverse.  Bohm's Quantum Theory
(\cite{bohm1989qt}) contains such a carefully crafted treatment.  The cost 
of this presentation
is that it takes nine chapters and two hundred pages to get to
the starting point of many other introductory Quantum texts.

Perhaps agreeing with a it cannot be derived opinion, the text of Liboff (\cite{liboff2003iqm}) appears to take a 
``let's calculate approach''.  There is little attempt to motivate the equation.  Instead one is left to learn
its characteristics by using it.  This engineering approach has some merits but must also contribute to much of the mystery
and confusion surrounding the subject.

Observe that the form of Schr\"{o}dinger's equation given in (\ref{eqn:oneDimSch}) or its three dimensional generalization is
explicitly non-relativisitic.  Time and space are not treated uniformly.  The order of the derivatives are not
even the same!

Given this apparent non-relativistic nature it is perhaps somewhat suprising that Pauli's Wave Mechanics (\cite{pauli2000wm}),
starts not with the Bohr model, nor black bodies, nor or any of the usual starting points for a Quantum text,
but instead with a lighting fast review of special relativity.
By assuming that the solution of the matter wave equation will take the same functional form as the solution for the classical wave equations
for light, and
by utilizing the
DeBroglie hypothesis that an energy momentum four vector for matter can be quantized, Pauli 
outlines a procedure to obtain a small momentum approximation of a relativistic matter wave equation (ie: of the Klien-Gordon equation).

Comments in
the wikipedia Klien-Gordon article (\cite{wikiKG}), suggest that Pauli's approach is probably closer historically to
the reasoning used initially
by Schr\"{o}dinger.  It is also interesting to note that the thesis of DeBroglie (\cite{AFkracklauerDeBroglie}) which
originally hypothosized extension of the notion of quantized freqencies from light to particles 
also heavily relied on special relativity arguments.

The following discussion follows the brief treatment of Pauli's text, but attempts to fill in some of the in between steps.

\section{Prerequiste concepts.  Wave equation, and Lorentz invariance.}

An unforced mechanical wave described by a function $\psi(t,\Bx)$, propagating undamped and unforced with
velocity $v$ is described by the familiar equation

\begin{align}\label{eqn:waveEquationV}
\inv{v^2} \frac{\partial^2 \psi}{\partial t^2} - \spacegrad^2 \psi = 0
\end{align}

For the purposes of this discussion, a relativistic wave is described by ($\ref{eqn:waveEquationV}$) with two
additional conditions.  The first is that the wave speed is
$v = c$, the speed of light.  The second condition required for the label relativistic 
is a restriction on the allowed coordinate transformations.  These are the linear transformations
of space time coordinates
$(t,x,y,z) \rightarrow (t', x', y', z')$ for which the wave equation retains precisely this form

\begin{align}\label{eqn:waveEquationInv}
\inv{c^2} \frac{\partial^2 \psi}{\partial {t}^2} 
- \frac{\partial^2 \psi}{\partial {x}^2} 
- \frac{\partial^2 \psi}{\partial {y}^2} 
- \frac{\partial^2 \psi}{\partial {z}^2} 
=
\inv{c^2} \frac{\partial^2 \psi}{\partial {t'}^2} 
- \frac{\partial^2 \psi}{\partial {x'}^2} 
- \frac{\partial^2 \psi}{\partial {y'}^2} 
- \frac{\partial^2 \psi}{\partial {z'}^2} 
\end{align}

Such transformations, the Lorentz transformations,
are those that introduce no cross term such as $\partial^2 \psi/\partial x' \partial y'$,
and do not change the wave velocity.  One can show that spatial rotations such as

\begin{align}
\begin{bmatrix}
ct' \\
x' \\
y' \\
z' \\
\end{bmatrix}
=
\begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & \cos\theta & \sin\theta \\
0 & 0 & -\sin\theta & \cos\theta \\
\end{bmatrix}
\begin{bmatrix}
ct \\
x \\
y \\
z \\
\end{bmatrix}
\end{align}

Or Lorentz boosts such as
\begin{align}
\begin{bmatrix}
ct' \\
x' \\
y' \\
z' \\
\end{bmatrix}
=
\begin{bmatrix}
\cosh\alpha & -\sinh\alpha & 0 & 0 \\
-\sinh\alpha & \cosh\alpha & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
\end{bmatrix}
\begin{bmatrix}
ct \\
x \\
y \\
z \\
\end{bmatrix}
\end{align}

Or any composition of such transformations meet this requirement.

In the linear transformations above the space and time coordinates were merged into a single vector representation,
the particle worldline vector, often written with shorthand such as

\begin{align}
X \equiv (ct, \Bx)
\end{align}

For such a vector, a Lorentz length can be defined

\begin{align}
X^2 \equiv c^2 t^2 - \Bx^2 \equiv c^2 t^2 - \Bx \cdot \Bx
\end{align}

Without specific discussion of the wave equation, 
a more usual but equivalent definition of Lorentz transformations, are those that leave this
Lorentz length unchanged, as in

\begin{align}
c^2 t^2 - \Bx^2 = c^2 {t'}^2 - {\Bx'}^2 
\end{align}

If we introduce a vector space time derivative operator

\begin{align}
\grad \equiv \left(\inv{c}\frac{\partial}{\partial t}, \spacegrad \right)
\end{align}

The Lorentz invariant length of this vector operator is in fact our wave equation operator

\begin{align}
\square \equiv \grad^2 = \inv{c^2}\frac{\partial^2}{\partial t^2} - \spacegrad \cdot \spacegrad
\end{align}

It is clear that the original requirement for wave equation invariance (\ref{eqn:waveEquationInv}) is also contained within 
this definition of Lorentz invariant length.

Unit vectors with respect to Lorentz length are neccessarily Lorentz invariant.  Considering for example the time rate of change of
a particle worldline we have

\begin{align}
\left(\frac{d}{dt}(ct, \Bx) \right)^2 = c^2 - \Bv^2
\end{align}

which implies that the Lorentz length of
\begin{align}
\frac{1}{\sqrt{1 - \Bv^2/c^2}}(1, \Bv/c)
\end{align}

is just one.  For the purposes of this 
Schr\"{o}dinger equation
discussion a scaling of this four vector so that it has dimensions of energy is required

\begin{align}
\frac{1}{\sqrt{1 - \Bv^2/c^2}}(m c^2, m\Bv c)
\end{align}

Algebraically, the Lorentz length of this four vector can easily be confirmed to be $(m c^2)^2$.  This quantity 
we will identify as an energy-momentum four vector as follows

\begin{align}\label{eqn:energyMomentumFourVec}
P = (E/c, \Bp)
\end{align}

With energy defined as
\begin{align}
E \equiv \frac{m c^2}{\sqrt{1 - \Bv^2/c^2}}
\end{align}

and spatial momentum defined as
\begin{align}
\Bp \equiv \frac{m \Bv}{\sqrt{1 - \Bv^2/c^2}}
\end{align}

It is beyond the scope of these notes to provide a good justification for this identification.
\footnote{This is a dodge, and having to make a statement like this shows that it is beyond the scope of the author's understanding to coherently justify this identification.  In the spirit of my engineering education I can at least work with it.}

When all is said and done, whether it has been adequately justified or not, the remainder
of the 
Schr\"{o}dinger equation
discussion to come really requires just one statement from special relativity.  That is
the Lorentz length of this energy momentum four vector.
From (\ref{eqn:energyMomentumFourVec}) that Lorentz length is

\begin{align}
P^2 = E^2/c^2 - \Bp^2 = m^2 c^2
\end{align}

\section{Wave equation form of Maxwell's equations}

In SI units and traditional vector notation, Maxwell's equations in the absence of charge and currents are

\begin{align}\label{eqn:divE}
\spacegrad \cdot \BE = 0
\end{align}
\begin{align}\label{eqn:divB}
\spacegrad \cdot \BB = 0
\end{align}
\begin{align}\label{eqn:curlB}
\spacegrad \cross \BB = \inv{c^2} \frac{\partial \BE}{\partial t}
\end{align}
\begin{align}\label{eqn:curlE}
\spacegrad \cross \BE = -\frac{\partial \BB}{\partial t}
\end{align}

The last two curl equations can be decoupled by once more calculating the curl.
Illustrating by example

\begin{align}\label{eqn:curlCurlE}
\spacegrad \cross (\spacegrad \cross \BE) = -\frac{\partial }{\partial t} \spacegrad \cross \BB = -\inv{c^2} \frac{\partial^2 \BE}{\partial t^2}
\end{align}

Digging out vector identities and utilizing the zero divergence we have

\begin{align}\label{eqn:identForcurlCurlE}
\spacegrad \cross (\spacegrad \cross \BE) = \spacegrad (\spacegrad \cdot \BE) - \spacegrad^2 \BE = -\spacegrad^2 \BE
\end{align}

Putting \ref{eqn:curlCurlE}, and \ref{eqn:identForcurlCurlE} together provides a wave equation for the electric field vector

\begin{align}\label{eqn:waveE}
\inv{c^2} \frac{\partial^2 \BE}{\partial t^2} - \spacegrad^2 \BE = 0
\end{align}

Operating with curl on the remainding Maxwell equation similarily produces a wave equation for the magnetic field vector

\begin{align}\label{eqn:waveB}
\inv{c^2} \frac{\partial^2 \BB}{\partial t^2} - \spacegrad^2 \BB = 0
\end{align}

This is really six wave equations, one for each of the field coordinates.

\section{Solution of the relativistic wave equation.}

Despite the fact that the six equations of (\ref{eqn:waveE}), and (\ref{eqn:waveB}) have the appearance of independence,
any solution must still meet the constraints of the original Maxwell equations.  For example for a linearly polarized
plane wave one can have $\BB = \kcap \cross \BE$, so the triplet of $\BE$, $\BB$, and $\kcap$ (the propagation direction) form a set 
of mutually perpendicular vectors.

Additional details on the dependence of the electromagnetic wave equations
can be found for example in Jackson (\cite{jackson1975cewWave}).  Fortunately, tackling this
in full detail is not required.  

There have been two specific aims discussing Maxwell wave equations.
One has been to provide a real world example of something that propagates with behaviour
governed by the relativistic wave equation.  The other is that we have an understanding 
of light quantization due to the photoelectric effect.  Together these 
will lead eventually to a relativistic matter wave equation and
the non-relativistic approximation that we know of as the 
Schr\"{o}dinger equation.

We can now move on to tackle the mathematical problem of finding solutions to our 
homogeneous wave equation for an arbitrary function $\psi(t,\Bx)$

\begin{align}
\delambertian \psi(t,\Bx) = \left( \frac{1}{c^2}\frac{\partial^2}{{\partial t}^2} - \spacegrad^2 \right) \psi(t,\Bx) = 0
\end{align}

Assuming preexisting knowledge of wave equation solutions here is not unreasonable, and nobody who has read this
far 
will likely be suprised that functions of the form $f(ct \pm \kcap \cdot \Bx)$
are solutions to $\delambertian f(t,\Bx) = 0$.  Superposition
of such functions would also be solutions, but general fitting to boundary conditions without additional constraints 
on the functions could be tricky.

An alternate approach is the application of more systematic methods.  In particular, the application 
of Fourier transformations to the wave equation is an effective technique.  As with Fourier solutions of the heat equation
(\cite{osgoodFourier}), performing this sort of transformation on the wave equation will result
in a much simpler differential equation in the wave number domain.

We will apply 3D spatial Fourier transforms, and write our transform pair in the symmetric form

\begin{align}
\FF(f(\Bx)) = \hat{f}(\Bk) = \inv{(\sqrt{2\pi})^3} \int f(\Bx) \exp\left( -i \Bk \cdot \Bx \right) d^3 x 
\end{align}
\begin{align}
\FF^{-1}({f}(\Bk)) = f(\Bx) = \inv{(\sqrt{2\pi})^3} \int \hat{f}(\Bk) \exp\left( i \Bk \cdot \Bx \right) d^3 k
\end{align}







\section{first try.}
Continuing the thread of Schr\"{o}dinger equation motivation, a peek at the relativistic origins for the quantum wave equation will be made.

The work of DeBroglie (\cite{AFkracklauerDeBroglie}) used by Schr\"{o}dinger to construct the quantum wave equation as we now know it relied heavily on special relativity arguments.  
Schr\"{o}dinger's initial attempt at the wave equation was in fact relativistic, but he ended up publishing a non-relativistic 
approximation.
According to the wikipedia Klien Gordon page (\cite{wikiKG}), 



%%In particular, one can derive the potential free Schr\"{o}dinger equation by employing a small momentum approximation to a solution of the relativistic wave equation.
%
% blah
%.  The following notes follow Pauli's treatment closely.
%Schr\"{o}dinger equation as an relativistic approximation in his Wave Mechanics text 
%for the quantum wave equation will be made.
%motivation, a peek at the relativistic 
%Here the relativistic origins for the quantum wave equation how do we end up with something so 
%
%It is interesting to note that Pauli, a master of relativity, having
%written the definitive encyclopedic article
%\cite{pauli1981tr}
%on special and general relativity
%as a teenager, introduces the
%Schr\"{o}dinger equation as an relativistic approximation in his Wave Mechanics text \cite{pauli2000wm}.  The following notes follow Pauli's treatment closely.
%%however
%%Pauli's preference for using
%%$h$ instead of $\hbar$ is avoided.
%%.  The second is an attempt to
%%avoid some of the terseness of Pauli's presentation.

\subsection{Invariant length for an energy momentum four vector}

The Lorentz invariant length of an energy momentum four vector

\begin{align}
P = (E/c, \Bp)
\end{align}

is given by

\begin{align}\label{eqn:Psquared}
P^2 \equiv \frac{E^2}{c^2} - \Abs{\Bp}^2 = \frac{(m c^2)^2}{c^2}
\end{align}

This invariance has meaning for particles, such as electrons, where $m \ne 0$, but also for photons where $m = 0$.
Einstein's explanation of the photoelectric effect introduced the concept that light comes in
quantized packets of energy (photons), with photon energy proportional to frequency

\begin{align}
E = h \nu = \hbar \omega
\end{align}

The relativistic invariance relation (\ref{eqn:Psquared}) implies that there is a corresponding photon momentum, also
quantized in magnitude, and this is typically written

\begin{align}
\Bp = \hbar \Bk
\end{align}

The magnitudes of the vector $\Bk$ and the frequency $\omega$ are not independent, since by (\ref{eqn:Psquared}) we have

\begin{align}\label{eqn:omegaKsquared}
c^2 \Abs{\Bk}^2 = \omega^2
\end{align}

%It is also worth noting that we have the concepts of electromagnetic energy and momentum in the continuous theory of
%electromagnetic radiation (i.e.: Maxwell's equations).  For an electromagnetic wave in vacuum the
%field energy density and momentum density also form a four vector
%
%\begin{align}
%%P = \epsilon_0 \left(\frac{\Abs{\BE}^2}{2} + \frac{c^2 \Abs{\BB}^2}{2}, c \BE \cross \BB \right)
%\end{align}
%
%Because $\Abs{\BE}^2 = c^2 \Abs{\BB}^2$, and $\BE \cdot \BB = 0$ for an electromagnetic wave one can show with some vector manipulation that in this continuous representation of light we also have $P^2 = 0$.
%
%\cite{bohm1989qt}
%\cite{feynman}
\subsection{Wave equation solutions to Maxwell's equations}

The coupling between wave number and angular frequency can 
%\begin{align}\label{eqn:omegaKsquared}
The 
In classical electrodynamics there are the field energy and momentum densities can also be are also coupled as in 
The 
%\begin{align}\label{eqn:omegaKsquared}

The equation for a wave $\psi(t,\Bx)$ propagating at the speed of light is

\begin{align}
\delambertian \psi = \left( \frac{1}{c^2}\frac{\partial^2}{{\partial t}^2} - \spacegrad^2 \right) \psi = 0
\end{align}

This is in fact the form that Maxwell's equations for both the electric and magnetic field take in the absence of charge or currents.
For example we have for the electric field

\begin{align}\label{eqn:eField}
\delambertian \BE = \left( \frac{1}{c^2}\frac{\partial^2}{{\partial t}^2} - \spacegrad^2 \right) \BE = 0
\end{align}

If we form the vector valued function

\begin{align}\label{eqn:testSolutionE}
\BE = \BE_0 e^{ i( \Bk \cdot \Bx - \omega t ) }
\end{align}

then application of the wave equation operator produces

\begin{align}\label{eqn:boxApplied}
\delambertian \BE = -\left( \frac{\omega^2}{c^2} - \Abs{\Bk}^2 \right) \BE
\end{align}

Therefore we have a solution of (\ref{eqn:eField}) provided

\begin{align}
\omega^2 = c^2 \Abs{\Bk}^2
\end{align}

This is exactly the relativistic invariance condition (\ref{eqn:omegaKsquared}) for a photon of energy $\hbar \omega$.

\subsection{Matter Wave equation}

Suppose that one has a function

\begin{align}\label{eqn:waveFunction}
\psi = e^{ i( \Bk \cdot \Bx - \omega t) }
\end{align}

that in some way represents a matter wave for a particle with energy $E = \hbar \omega$, and momentum
 $\Bp = \hbar \Bk$.  The multiplicative factor has been omitted for clarity.
From (\ref{eqn:boxApplied}), and
(\ref{eqn:Psquared})
we see that such a function will satisfy the relativistic wave equation

\begin{align}
\delambertian \psi = \left( \frac{1}{c^2}\frac{\partial^2}{{\partial t}^2} - \spacegrad^2 \right) \psi = - \frac{m^2 c^2}{\hbar^2} \psi
\end{align}

We have now reached the point that a non relativistic approximation can be made.
Again from (\ref{eqn:Psquared}) we have

\begin{align}
\omega = \frac{m c^2}{\hbar} \sqrt{ 1 + \frac{\hbar^2 \Abs{\Bk}^2}{m^2 c^2}}
\end{align}

If $(\hbar^2 \Abs{\Bk}^2)/(m^2 c^2)$ is small enough, then a first order Taylor gives

\begin{align}
\omega \approx \frac{m c^2}{\hbar} + \frac{\hbar \Abs{\Bk}^2}{2 m}
\end{align}

This allows for a factorization of the wave function (\ref{eqn:waveFunction})

\begin{align}\label{eqn:intermed}
\psi =
\exp\left( i \frac{m c^2}{\hbar} t \right)
\exp\left( i \left(
\Bk \cdot \Bx
-\frac{\hbar \Abs{\Bk}^2}{2 m}t
\right) \right)
\end{align}

The infinite plane wave solution of (\ref{eqn:waveFunction}) was only one possible solution to the original wave equation.
A general solution could be any superposition of such solutions, such as a weighted integral over wave numbers (i.e.: an
inverse Fourier transform from a wave number representation).  For this reason it is not unreasonable to drop the specific plane wave nature that generated (\ref{eqn:intermed}) instead writing $\psi(t,\Bx)$ in terms of an unknown function $\psi'$ as follows

\begin{align}
\psi = \exp\left( i \frac{m c^2}{\hbar} t \right) \psi'(t, \Bx)
\end{align}

We are now set to apply the relativistic wave equation operator to the wave function that we have now first order Taylor expanded
around the rest frequency $m c^2/\hbar$

\begin{align}
\delambertian \psi
=
\left(
\frac{1}{c^2}
\left(
\left( \frac{i m c^2}{\hbar} \right)^2 \psi' - 2 i \frac{m c^2}{\hbar} \frac{\partial \psi'}{\partial t} +
\frac{\partial^2 \psi'}{\partial t^2}
\right)
%\exp\left( i \frac{m c^2}{\hbar} t \right)
-
\spacegrad^2 \psi'
\right)
\exp\left( i \frac{m c^2}{\hbar} t \right)
= -
\exp\left( i \frac{m c^2}{\hbar} t \right)
\frac{m^2 c^2}{\hbar^2} \psi'
\end{align}

Rearranging we have
%\begin{align}
%\frac{1}{c^2}
%\left(
%-2 i \frac{m c^2}{\hbar} \frac{\partial \psi'}{\partial t} +
%\frac{\partial^2 \psi'}{\partial t^2}
%\right)
%=
%\spacegrad^2 \psi'
%\end{align}

\begin{align}
-\frac{\hbar^2}{2m} \spacegrad^2 \psi' = i \hbar \frac{\partial \psi'}{\partial t} -\frac{\hbar^2}{2m c^2} \frac{\partial^2 \psi'}{\partial t^2}
\end{align}

Now for the mass of an electron $\hbar^2/(m_e c^2) \approx 10^{-55} m^2 kg$, which is pretty small seeming.

Assuming this last term can be dropped we have the potential free Schr\"{o}dinger's equation.

\begin{align}
-\frac{\hbar^2}{2m} \spacegrad^2 \psi' = i \hbar \frac{\partial \psi'}{\partial t}
\end{align}

Observe that
this isn't the only wave equation we could construct in this fashion.  For example, we could also have started with $\psi \propto e^{i(\omega t - \Bx \cdot \Bk)}$.  Provided $\omega^2 = c^2 \Abs{\Bk}^2$ as before, this is still a solution to the wave equation, and this procedure would produce a wave equation with a negated sign on the time derivative.

\bibliography{blog}
\bibliographystyle{unsrturl}

\end{document}
