%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%
%
%\chapter{Preface}
% this suppresses an explicit chapter number for the preface.
\chapter*{Preface}%\normalsize
  \addcontentsline{toc}{chapter}{Preface}

This document is based on my lecture notes for the Fall 2012, University of Toronto Advanced Classical Optics course (PHY485H1F), taught by Prof.\ Joseph H. Thywissen.

My thanks to Professor Thywissen for teaching this course.  He knows his subject well, and I learned a lot.

Official course description:

This course builds on a student's knowledge of basic electromagnetic theory by focusing attention on light including elementary aspects of the propagation of optical beams and their interaction with matter.  We examine light polarization, coherence, interference and diffraction as we move towards a description of lasers within a semiclassical picture in which the fields are treated classically and matter is treated quantum mechanically.  In between we discuss Gaussian beam modes and their relation to optical resonators as well as fibre and slab waveguides.

In the course of the lectures the books \citep{hecht1998hecht}, \citep{fowles1989introduction}, and \citep{born1980principles} are referred to, but none were required texts for the course.

This document contains a few things

\begin{itemize}
\item Plain old lecture notes.   These mirror what was covered in class.  I've made no attempt to go and incorporate the much more extensive hand written supplementary notes supplied with the lectures.  In some cases I've likely filled in some of the same details that those hand written notes covered when working through my lecture notes trying to make sense of things.

\item Personal notes exploring details that were not clear to me from the lectures, or from the texts associated with the lecture material.

\item Assigned problems.  Like anything else take these as is.  I may or may not have gone back and corrected errors, and I definitely made some.

\item Some worked problems attempted as course prep, for fun, or for exam preparation, or post exam review.

\item Links to Mathematica workbooks associated with course content or these notes.

\end{itemize}

Peeter Joot  \quad peeterjoot@protonmail.com
