%
% Copyright © 2013 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\makeproblem{Transition from near- to far-field}{modernOptics:problemSet2:4}{
Consider the intensity distribution behind an illuminated  single slit whose width is \(A = 10 \lambda\). Immediately behind the slit, the intensity matches the transmission profile. Farther away, there will be diffraction patterns. Assume the illumination is plane-wave, with intensity \(I_0\), and monochromatic. Make a {\bf numerical calculation} of the intensity pattern \(I = | \Psi |^2 \) using the exact diffraction integral,
\(\) \Psi = \frac{\Psi_0}{i \lambda} \int_{-A/2}^{A/2} dx' \frac{e^{i k R}}{R} \(\)
where \(R = |\Br - \Br'|\) as in the notes, but we have restricted the problem to the plane (\(y=0\)).

When making plots, plot all lengths in units of \(\lambda\)  (ie, plot intensity \(I/I_0\) versus \(x/\lambda\)). Ignore the \(y\) direction. For each observation distance, comment on the validity of the approximations. Each plot should have an exact line, a Fresnel line, and a Fraunhofer line.

For each of the following observation distances, {\bf compare} your calculation to the Fresnel and Fraunhofer approximations:
\makesubproblem{
\(z=\lambda\)
}{modernOptics:problemSet2:4a}
\makesubproblem{
\(z=A=10 \lambda\)
}{modernOptics:problemSet2:4b}
\makesubproblem{
\(z=10 A = 100 \lambda\)
}{modernOptics:problemSet2:4c}
}

\makeanswer{modernOptics:problemSet2:4}{
\paragraph{Part \ref{modernOptics:problemSet2:4a}.  }
TODO.
\paragraph{Part \ref{modernOptics:problemSet2:4b}.  }
TODO.
\paragraph{Part \ref{modernOptics:problemSet2:4c}.  }
TODO.
%\makesubanswer{TODO.}{modernOptics:problemSet2:4a}
%\makesubanswer{TODO.}{modernOptics:problemSet2:4b}
%\makesubanswer{TODO.}{modernOptics:problemSet2:4c}
} % makeanswer
