% 
% 
% 
% Copyright © 2012 Peeter Joot
% All Rights Reserved
% 
% This file may be reproduced and distributed in whole or in part, without fee, subject to the following conditions:
% 
% o The copyright notice above and this permission notice must be preserved complete on all complete or partial copies.
% 
% o Any translation or derived work must be approved by the author in writing before distribution.
% 
% o If you distribute this work in part, instructions for obtaining the complete version of this file must be included, and a means for obtaining a complete version provided.
% 
% 
% Exceptions to these rules may be granted for academic purposes: Write to the author and ask.
% 
% 
% 
These are my personal lecture notes for the Fall 2012, University of Toronto Modern Optics course (PHY485H1F), taught by Prof.\ Henry M. van Driel.

The official description of this course is

This course builds on a student's knowledge of basic electromagnetic theory by focusing attention on light including elementary aspects of the propagation of optical beams and their interaction with matter.  We examine light polarization, coherence, interference and diffraction as we move towards a description of lasers within a semiclassical picture in which the fields are treated classically and matter is treated quantum mechanically.  In between we discuss Gaussian beam modes and their relation to optical resonators as well as fibre and slab waveguides.

The books \cite{hecht2002optics} and \cite{saleh1991fundamentals} are listed as useful references, but are not required texts for the course. 

This document will probably contain a few things

\begin{itemize}
\item My lecture notes.

Typos, if any, are probably mine (Peeter), and no claim nor attempt of spelling or grammar correctness will be made.

\item Some personal notes exploring details that weren't clear to me from the lectures, or from the texts associated with the lecture material.

\item Some worked problems.

\item Links to mathematica workbooks associated with course content.

\end{itemize}

Peeter Joot  \quad peeter.joot@gmail.com 

