%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

% 
% 
%\chapter{Preface}
% this suppresses an explicit chapter number for the preface.
\chapter*{Preface}%\normalsize
  \addcontentsline{toc}{chapter}{Preface}

These are my personal lecture notes for the Fall 2010, University of Toronto Quantum mechanics I course (PHY356H1F), taught by Prof.\ Vatche Deyirmenjian.

The official description of this course was:

The general structure of wave mechanics; eigenfunctions and eigenvalues; operators; orbital angular momentum; spherical harmonics; central potential; separation of variables, hydrogen atom; Dirac notation; operator methods; harmonic oscillator and spin.

This document contains a few things

\begin{itemize}
\item My lecture notes.

Typos, if any, are probably mine (Peeter), and no claim nor attempt of spelling or grammar correctness will be made.  The first four lectures I had chosen not to take notes for since they followed the text \citep{desai2009quantum} very closely.

\item Notes from reading of the text.  This includes observations, notes on what seem like errors, and some solved problems.  None of these problems have been graded.  Note that my informal errata sheet \citep{desaiTypos} for the text has been separated out from this document.

\item Some assigned problems.  I have corrected some the errors after receiving grading feedback, and where I have not done so I at least recorded some of the grading comments as a reference.

\item Some worked problems associated with exam preparation.

\end{itemize}

Peeter Joot  \quad peeterjoot@protonmail.com 
