\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}

%\usepackage{listings}
%\usepackage{txfonts} % for ointctr... (also appears to make "prettier" \int and \sum's)
\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage


\title{ Relativistic acceleration. }
\author{Peeter Joot \quad peeter.joot@gmail.com }
\date{ April 10, 2009.  Last Revision: $Date: 2009/04/11 01:00:47 $ }

\begin{document}

\maketitle{}
\tableofcontents
\section{ Motivation. }

Continuing on with reading of \cite{pauli1981tr}, having 
clarified aspects of the four vector velocity in \cite{PJrelativityFourVectorVelocity}, it is now
time to move on to acceleration.

Do the chain rule calculations for the acceleration four vector equation given in equation (193).

\section{ Compute it. }

Compute the spatial and timelike components of the acceleration

\begin{align*}
B^\mu 
&= \frac{d^2 x^\mu}{d\tau^2} \\
&= \frac{d }{d\tau} \left( \frac{d x^\mu }{d\tau} \right) \\
&= \frac{d }{d\tau} \left( \frac{d x^\mu }{dt} \frac{dt}{d\tau} \right) \\
&= \left( \frac{d }{d\tau} \frac{d x^\mu }{dt} \right) \frac{dt}{d\tau} + \frac{d x^\mu }{dt} \frac{d^2t}{d\tau^2} \\
&= \frac{d^2 x^\mu }{dt^2} \left( \frac{dt}{d\tau} \right)^2 + \frac{d x^\mu }{dt} \frac{d^2t}{d\tau^2} \\
\end{align*}

For $\mu \in \{1,2,3\}$, the ${d^2 x^\mu }/{dt^2}$ terms are the regular old spatial acceleration components.
, and $dx^4/dt = c$.  Writing $\Bu^2 = \sum_{k=1}^3 (dx^k/dt)^2$, and $\beta^2 = \Bu^2/c^2$, we have

\begin{align*}
B^k &= \frac{d^2 x^k }{dt^2} \inv{1-\beta^2} + \frac{d x^k }{dt} \frac{d^2t}{d\tau^2} \\
B^4 &= 0 + c \frac{d^2t}{d\tau^2} \\
\end{align*}

In both of these is the $d^2t/d\tau^2$ term.  Let's expand that.

\begin{align*}
\frac{d^2t}{d\tau^2} 
&= \frac{d}{d\tau} \left( \inv{\sqrt{ 1 - \Bu^2/c^2 }} \right) \\
&= \frac{-1}{c^2} \frac{(-1/2) }{({ 1 - \Bu^2/c^2 })^{3/2}} \frac{d\Bu^2}{d\tau} \\
&= \frac{1}{c^2} \frac{(1/2) }{({ 1 - \Bu^2/c^2 })^{3/2}} 2 \Bu \cdot \frac{d\Bu}{d\tau} \\
&= \frac{1}{c^2} \frac{1}{({ 1 - \Bu^2/c^2 })^{3/2}} \Bu \cdot \frac{d\Bu}{dt} \frac{dt}{d\tau} \\
&= \frac{1}{c^2} \frac{1}{({ 1 - \Bu^2/c^2 })^{2}} \Bu \cdot \frac{d\Bu}{dt} \\
\end{align*}

In vector form, with $\Ba = d\Bu/dt$, we now have the following 

\begin{align*}
\BB &= \Ba \inv{1-\beta^2} + \Bu (\Bu \cdot \Ba) \inv{c^2} \inv{ (1-\beta^2)^2} \\
B^4 &= \inv{c} (\Bu \cdot \Ba) \inv{ (1-\beta^2)^2}
\end{align*}

This reproduces the equation from the Pauli text (except for the imaginary factor $i$ due to the Minkowski notation).  Except for $\gamma$ factors this calculation has a similar final form to that of the decomposition of acceleration in terms of radial components.

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
