\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}

%\usepackage{listings}
\usepackage{txfonts} % for ointctr... (also appears to make "prettier" \int and \sum's)
\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage


\title{ Integer binomial theorem induction, the easy dumb way. }
\author{Peeter Joot \quad peeter.joot@gmail.com }
\date{ March 26, 2009.  Last Revision: $Date: 2009/03/27 00:59:32 $ }

\begin{document}
\maketitle{}

%\tableofcontents

\section{ Motivation. }

While working a problem with an induction requirement similar to but more 
complicated than the binomial theorem, I steped back and thought I'd try
this as an easier first step.  Had some trouble doing it, until I tried it
explicitly with for the power of three case.  Ironically, working it
out for an explicit index takes the abstraction out of the problem, and
generalizing further really only requires a search and replace.

\section{ Do it. }

Want to prove

\begin{align}
(t + x)^k = \sum_{m=0}^k \binom{k}{m} t^m x^{k-m}
\end{align}

where

\begin{align}
\binom{k}{m} = \frac{k!}{(k-m)!m!}
\end{align}

In particular want to prove this for the $k+1$ case, given $k$.

\subsection{ step with k+1 = 3 }

Three isn't actually the best way to start since it is almost too trivial, but
one gets the idea easily by doing it.

\begin{align*}
(t + x)^3 
&= (t + x)(t + x)^2  \\
&= (t + x)\sum_{m=0}^2 \binom{2}{m} t^m x^{2-m} \\
&= 
\sum_{m=0}^2 \binom{2}{m} t^{m+1} x^{2-m} 
+ \sum_{m=0}^2 \binom{2}{m} t^{m} x^{2-m + 1} \\
&= 
\sum_{m=1}^{2 + 1} \binom{2}{m-1} t^{m} x^{2 - m + 1} 
+ \sum_{m=0}^2 \binom{2}{m} t^{m} x^{2-m + 1} \\
\end{align*}

Now pull the lowest and highest order terms out of the sums, and group the
remaining bits.

\begin{align*}
(t + x)^3 
&= 
 \binom{2}{0} t^{0} x^{2 - 0 + 1} 
+ \sum_{m=1}^{2} \left( \binom{2}{m} + \binom{2}{m-1} \right) t^{m} x^{2 - m + 1} 
+ \binom{2}{2 + 1 -1} t^{2 + 1} x^{2 - (2 + 1) + 1} 
\\
\end{align*}

Now, observe that in all the steps above everywhere if all places that $2$, a nice easy to think with and concrete number, we could have used some
abstract index.

\subsection{ step with k+1 }

A straight text search and replace on $2$ with $k$ gives

\begin{align*}
(t + x)^{k+1}
&= (t + x)(t + x)^k  \\
&= (t + x)\sum_{m=0}^k \binom{k}{m} t^m x^{k-m} \\
&= 
\sum_{m=0}^k \binom{k}{m} t^{m+1} x^{k-m} 
+ \sum_{m=0}^k \binom{k}{m} t^{m} x^{k-m + 1} \\
&= 
\sum_{m=1}^{k + 1} \binom{k}{m-1} t^{m} x^{k - m + 1} 
+ \sum_{m=0}^k \binom{k}{m} t^{m} x^{k-m + 1} \\
&= 
 \binom{k}{0} t^{0} x^{k - 0 + 1} 
+ \sum_{m=1}^{k} \left( \binom{k}{m} + \binom{k}{m-1} \right) t^{m} x^{k - m + 1} 
+ \binom{k}{k + 1 -1} t^{k + 1} x^{k - (k + 1) + 1} \\
\end{align*}

This is EXACTLY the same as above with $k=2$ but it sure looks more complicated with an an abstract index.

To finish off we need a couple observations, that 
$\binom{k}{0} = 1 = \binom{k+1}{0}$, and $\binom{k}{k} = 1 = \binom{k+1}{k+1}$.  This leaves us with

\begin{align*}
(t + x)^{k+1}
= \binom{k+1}{0} t^{0} x^{k + 1} 
+ \sum_{m=1}^{k} \left( \binom{k}{m} + \binom{k}{m-1} \right) t^{m} x^{k - m + 1} 
+ \binom{k + 1}{k + 1} t^{k + 1} x^{0} \\
\end{align*}

So if we can show
\begin{align*}
\binom{k}{m} + \binom{k}{m-1} = \binom{k+1}{m}
\end{align*}

then we would have

\begin{align*}
(t + x)^{k+1}
&= \sum_{m=0}^{k+1} \binom{k+1}{m} t^{m} x^{k + 1 - m} 
\end{align*}

which is what was desired.

\subsection{ That last little piece. }

To prove that last little piece, let's do it again the dumb way, and let a regular expression $s/2/k/g$ in vim do the hard work.

\begin{align*}
\binom{2}{m} + \binom{2}{m-1} 
&=
\frac{2!}{(2-m)!m!}
+ \frac{2!}{(2-m + 1)!(m-1)!} \\
&=
\frac{2!}{(2-m)!m(m-1)!}
+ \frac{2!}{(3-m)!(m-1)!} \\
&=
\frac{2!}{(2-m)!m(m-1)!}
+ \frac{2!}{(3-m)(2-m)!(m-1)!} \\
&=
\frac{2!}{(2-m)!(m-1)!} \left( \inv{m} + \inv{3-m}\right) \\
&=
\frac{2!}{(2-m)!(m-1)!} \frac{3 -m + m }{m(3-m)} \\
&=
\frac{3!}{(3-m)!(m)!} \\
\end{align*}

So, generalizing the easy way with $s/3/k+1/g$, we have 

\begin{align*}
\binom{k}{m} + \binom{k}{m-1} 
&=
\frac{k!}{(k-m)!m!}
+ \frac{k!}{(k-m + 1)!(m-1)!} \\
&=
\frac{k!}{(k-m)!m(m-1)!}
+ \frac{k!}{((k+1)-m)!(m-1)!} \\
&=
\frac{k!}{(k-m)!m(m-1)!}
+ \frac{k!}{((k+1)-m)(k-m)!(m-1)!} \\
&=
\frac{k!}{(k-m)!(m-1)!} \left( \inv{m} + \inv{(k+1)-m}\right) \\
&=
\frac{k!}{(k-m)!(m-1)!} \frac{(k+1) -m + m }{m((k+1)-m)} \\
&=
\frac{(k+1)!}{((k+1)-m)!(m)!} \\
\end{align*}

Every step is EXACTLY the same as with $k=2$, the only differences were straight text substitution.  That leaves us with

\begin{align}
\binom{k}{m} + \binom{k}{m-1} 
&=
\binom{k+1}{m}
\end{align}

That's all we needed to complete the proof.

I think this is a superior way to do inductive proofs.  Just do the absolute easiest case and do it with a number that is easy to think with.  Search and replace in an editor does all the bits that would make you look clever if you were to leave off the fact that you were really only doing the easy version!

%\bibliographystyle{plainnat}
%\bibliography{myrefs}

\end{document}
