\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}

\DeclareMathOperator{\sgn}{sgn}
\newcommand{\PDSq}[2]{\frac{\partial^2 {#2}}{\partial {#1}^2}}
\newcommand{\PDN}[3]{\frac{\partial^{#3} {#2}}{\partial {#1}^{#3}}}
\DeclareMathOperator{\sinc}{sinc}
\DeclareMathOperator{\PV}{PV}
\newcommand{\FF}[0]{\mathcal{F}}
\newcommand{\Sw}[0]{\mathcal{S}}
\newcommand{\IIinf}[0]{ \int_{-\infty}^\infty }
\newcommand{\FM}[0]{\inv{\sqrt{2\pi\hbar}}}
\newcommand{\expectation}[1]{\langle{#1}\rangle}

%\usepackage{listings}
%\usepackage{txfonts} % for ointctr... (also appears to make "prettier" \int and \sum's)
% makes \grad look funny though (almost like spacegrad, but narrower)
\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage


\title{ QM notes and problems for Bohm, chapter 11. }
\author{Peeter Joot \quad peeter.joot@gmail.com }
\date{ May 8, 2009.  Last Revision: $Date: 2009/05/09 17:07:21 $ }

\begin{document}

\maketitle{}
\tableofcontents
\section{ Bohm Chapter 11 problems. }

Problems and additional details from reading of \cite{bohm1989qt}, chapter 11.

\subsection{ Problem 1.  Probability currents for step potential. }

\subsubsection{ Setup. }

This problem and the associated text has a step potential $V$ for $x>0$.  The 
idea is that we have a left to right stream of particles with an associated
wave function, with reflected and transmitted coefficients.

Solutions to the stationary equation are sought in each of the intervals

\begin{align}
-\frac{\hbar^2}{2m}\psi'' + (V-E)\psi = 0
\end{align}

That is
\begin{align}
\psi'' = - \frac{2m}{\hbar^2} (E-V)\psi 
\end{align}

With an assumption of exponential solutions on the left of the barrier
(ie: no decay and associated hyperbolic solutions in the $V=0$ interval),
we must have $E>0$.

So, the solution can be written as the sum

\begin{align*}
\psi_1 = \sum A_{\pm} \exp\left( \pm i \sqrt{2mE} x / \hbar \right)
\end{align*}

Similarily for $x>0$, non-hyperbolic solutions are

\begin{align*}
\psi_2 = \sum B_{\pm} \exp\left( \pm i \sqrt{2m(E-V)} x / \hbar \right)
\end{align*}

Mathematically, there isn't anything that prevents picking $E-V <0$ solutions

\begin{align*}
\psi_2 = \sum B_{\pm}' \exp\left( \pm \sqrt{-2m(E-V)} x / \hbar \right)
\end{align*}

TODO: Come back to this and think it through.  Where will follow through using
wave function continuity and derivative continuity with these solutions
in the $x>0$ interval?

Continuity at $x=0$ ($\psi_1(0) = \psi_2(0)$) requires

\begin{align*}
A_{+} + A_{-} = B_{+} + B_{-}
\end{align*}

Whereas derivative continuity, with $p_1 = \sqrt{2mE}$, and $p_2 = \sqrt{2m (E-V)}$ requires

\begin{align*}
\frac{i p_1}{\hbar} (A_{+} - A_{-}) = \frac{i p_2}{\hbar} (B_{+} - B_{-}) 
\end{align*}

This is

\begin{align*}
\begin{bmatrix}
1 & 1 \\
p_1 & -p_1
\end{bmatrix}
\begin{bmatrix}
A_{+} \\
A_{-} 
\end{bmatrix}
=
\begin{bmatrix}
1 & 1 \\
p_2 & -p_2
\end{bmatrix}
\begin{bmatrix}
B_{+} \\
B_{-} 
\end{bmatrix}
\end{align*}

Matrix inversion gives us the $B$ coefficients in terms of $A$

\begin{align*}
\begin{bmatrix}
B_{+} \\
B_{-} 
\end{bmatrix}
&=
\inv{-2p_2}
\begin{bmatrix}
-p_2 & -1 \\
-p_2 & 1
\end{bmatrix}
\begin{bmatrix}
1 & 1 \\
p_1 & -p_1
\end{bmatrix}
\begin{bmatrix}
A_{+} \\
A_{-} 
\end{bmatrix} \\
&=
\inv{2p_2}
\begin{bmatrix}
p_2 & 1 \\
p_2 & -1
\end{bmatrix}
\begin{bmatrix}
1 & 1 \\
p_1 & -p_1
\end{bmatrix}
\begin{bmatrix}
A_{+} \\
A_{-} 
\end{bmatrix} \\
&=
\inv{2p_2}
\begin{bmatrix}
(p_1 + p_2) & (p_2 - p_1) \\
(p_2 - p_1) & (p_1 + p_2)
\end{bmatrix}
\begin{bmatrix}
A_{+} \\
A_{-} 
\end{bmatrix} \\
\end{align*}

Or equivalently

\begin{align*}
\begin{bmatrix}
A_{+} \\
A_{-} 
\end{bmatrix}
&=
\inv{2p_1}
\begin{bmatrix}
(p_1 + p_2) & (p_1 - p_2) \\
(p_1 - p_2) & (p_1 + p_2)
\end{bmatrix}
\begin{bmatrix}
B_{+} \\
B_{-} 
\end{bmatrix} \\
\end{align*}

Using this last, the 
assumption of no further barriers in the $x>0$ interval (ie: no reflection at 
$x = \infty$), allows the physical situation to dictate $B_{-} = 0$.  Then we
have

\begin{align*}
\begin{bmatrix}
A_{+} \\
A_{-} 
\end{bmatrix}
&=
\frac{B_{+} }{2p_1}
\begin{bmatrix}
p_1 + p_2 \\
p_1 - p_2 
\end{bmatrix}
\end{align*}

The first is 
\begin{align*}
A_{+} 
&=
\frac{B_{+} }{2p_1} (p_1 + p_2)
\end{align*}

Or
\begin{align*}
B_{+} 
&=
\frac{2 p_1 A_{+} }{p_1 + p_2}
\end{align*}

and the second is
\begin{align*}
A_{-} 
&=
\frac{B_{+} }{2p_1} (p_1 - p_2) \\
&=
\frac{2 p_1 A_{+} }{p_1 + p_2}
\frac{1}{2p_1} (p_1 - p_2) \\
&=
A_{+} \left(
\frac{ p_1 - p_2 }{p_1 + p_2} \right)
\end{align*}

This reduces the free parameters in the wave functions to the single 
amplitude

\begin{align}\label{eqn:wavefunctions}
\psi_1 &= 
A_{+} e^{ i p_1 x/\hbar }
+A_{+} \left(
\frac{ p_1 - p_2 }{p_1 + p_2} \right)
e^{ -i p_1 x/\hbar } \\
\psi_2 &= A_{+} \frac{2 p_1 }{p_1 + p_2} e^{ i p_2 x/\hbar }
\end{align}

Inspection provides a check that these do in fact satisfy the desired
continuity requirements at $x=0$ (both the wave function and it's derivative).

Also observe that a sign error typo in the text is implicitly fixed above (
sign factor of $p_2$ in $\psi_2$).  That typo is corrected for right afterwards when $A,B,C$ is calculated, since it would otherwise result in a negative sign in the resulting linear equations there too.

\subsubsection{ Probability currents. }

From \ref{eqn:wavefunctions}
the probability currents can be calculated.  The probability 
conservation relation follows by taking time derivatives of $\psi^\conj\psi$, as done in \cite{PJprobCurrent}.  We have, for the $x$ component of the current

\begin{align*}
\PD{t}{\psi\psi^\conj} + \grad \cdot \left(
\frac{\hbar}{2 m i} 
\left( 
\psi^\conj \grad \psi
-\psi \grad \psi^\conj
\right)
\right) = 0
\end{align*}

Or in short

\begin{align*}
\PD{t}{\rho} + \grad \cdot \BJ = 0
\end{align*}

For $\psi_2$ we have

\begin{align*}
J_2 
&= 
\frac{\hbar}{2 m i} 
\left( 
B_{+}^\conj e^{ -i p_2 x/\hbar }
B_{+} (i p_2 /\hbar)e^{ i p_2 x/\hbar }
-B_{+} e^{ -i p_2 x/\hbar }
B_{+}^\conj (-i p_2 /\hbar)e^{ -i p_2 x/\hbar }
\right) \\
&= \frac{ p_2 \Abs{ B_{+} }^2}{m } \\
\end{align*}

In terms of $A_{+}$, this is

\begin{align*}
J_2 &= \frac{ 4 p_1^2 p_2 \Abs{ A_{+} }^2}{m (p_1 + p_2)^2}
\end{align*}

For $\psi_1$ the $x$ component of the current is

\begin{align*}
J_1 
&= 
\frac{\hbar}{2 m i} 
\left( 
(A_{+}^\conj e^{ -i p_1 x/\hbar }
+A_{-}^\conj e^{ i p_1 x/\hbar })
(A_{+} (i p_1 /\hbar)e^{ i p_1 x/\hbar }
+A_{-} (-i p_1 /\hbar)e^{ -i p_1 x/\hbar }) 
\right)
\\
&
-
\frac{\hbar}{2 m i} \left(
(
A_{+} e^{ i p_1 x/\hbar } +A_{-} e^{ -i p_1 x/\hbar })
(A_{+}^\conj (-i p_1 /\hbar)e^{ -i p_1 x/\hbar }
+A_{-}^\conj (i p_1 /\hbar)e^{ i p_1 x/\hbar })
\right) \\
&=
\frac{p_1}{2m} 
\left(
2\Abs{A_{+} }^2
+2\Abs{A_{-} }^2
+
A_{-}^\conj A_{+} e^{ 2 i p_1 x/\hbar }
-A_{+}^\conj A_{-} e^{ -2 i p_1 x/\hbar }
-A_{+} A_{-}^\conj e^{ 2 i p_1 x/\hbar }
+A_{-} A_{+}^\conj e^{ -2 i p_1 x/\hbar }
\right) \\
&=
\frac{p_1}{m} 
\left(
\Abs{A_{+} }^2
+\Abs{A_{-} }^2
\right) \\
\end{align*}

Again in terms of $A_{+}$, we have

\begin{align*}
\Abs{A_{+} }^2 +\Abs{A_{-} }^2 
&=
\Abs{A_{+}}^2 \left( 1 + \left( \frac{ p_1 - p_2 }{p_1 + p_2} \right)^2 \right) \\
&=
\Abs{A_{+}}^2 
\frac{ 2 p_1^2 + 2 p_2^2 + 2 p_1 p_2 - 2 p_1 p_2 }{(p_1 + p_2)^2} \\
&=
2 \Abs{A_{+}}^2 \frac{ p_1^2 + p_2^2 }{(p_1 + p_2)^2} \\
\end{align*}

Which provides the probability current for the sum of the incident and reflected wave functions

\begin{align*}
J_1
&=
\Abs{A_{+}}^2 
\frac{2 p_1}{m} 
\frac{ (p_1^2 + p_2^2) }{(p_1 + p_2)^2} \\
\end{align*}

The individual incident and reflected currents can also be calculated.  These are

\begin{align*}
J_i &= \frac{ p_1}{m} \Abs{ A_{+} }^2 \\
J_r &= \frac{ p_1}{m} \Abs{ A_{-} }^2  \\
&= 
\Abs{ A_{+} }^2 
\frac{ p_1 }{m } 
\left( \frac{ p_1 - p_2 }{p_1 + p_2} \right)^2 
\end{align*}

Perhaps unsuprisingly, we see that the total probability current for $x<0$, is the sum of the incident and reflected
currents $J_1 = J_i + J_r$.

\subsubsection{ Transmission and reflection coefficients. }

As in the text we can calculate the reflection coefficent, the ratio of the magnitudes of the reflected and the incident currents,

\begin{align*}
R 
&= \frac{J_r}{J_i}  \\
&=
\frac{\Abs{ A_{+} }^2 
\frac{ p_1 }{m } 
\left( \frac{ p_1 - p_2 }{p_1 + p_2} \right)^2 }{
\frac{ p_1}{m} 
\Abs{ A_{+} }^2 } \\
&=
\left( \frac{ p_1 - p_2 }{p_1 + p_2} \right)^2 
\\
\end{align*}

and can calculate the transmission coefficient, the ratio of the transmitted current $J_2$, to the incident current $J_i$.  This is

\begin{align*}
T 
&= \frac{J_2}{J_i} \\
&= 
\frac{
\frac{ 4 p_1^2 p_2 \Abs{ A_{+} }^2}{m (p_1 + p_2)^2}
}
{
\frac{ p_1}{m} \Abs{ A_{+} }^2 
} \\
&=
\frac{ 4 p_1 p_2 }{ (p_1 + p_2)^2}
\end{align*}

Without first calculating the currents explicitly it wasn't clear to me how $T$ was calculated in the text, and doing this first
makes that a bit more sensible.

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
