\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}

%\usepackage{listings}
\usepackage{txfonts} % for ointctr... (also appears to make "prettier" \int and \sum's)
\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage


\title{ Worked calculus of variations problems from Byron and Fuller. }
\author{Peeter Joot \quad peeter.joot@gmail.com }
\date{ March 21, 2009.  Last Revision: $Date: 2009/03/22 05:32:43 $ }

\begin{document}

\maketitle{}
\tableofcontents

\section{ Worked calculus of variations problems. }

Select problems from chapter II of \cite{byron1992mca}.

\subsection{ Problem 1.  Shortest line between points in polar coordinates. }

Problem.  Variational calculus exersize to find shortest distance between two points using polar coordinates.
 
The line element is:
 
\begin{align*}
ds^2 = r^2 d\theta^2 + {r'}^2
\end{align*}
 
So, the integral to minimize is
 
\begin{align*}
I = \int \sqrt{ r^2 + {r'}^2 } d\theta
\end{align*}
 
Application of the Euler-Lagrange equations yields
 
\begin{align*}
0 
&= \left( \PD{r}{} - \frac{d}{d\theta} \PD{r'}{} \right) \sqrt{ r^2 + {r'}^2 }  \\
&= \frac{r}{\sqrt{{r'}^2 + r^2}} - \frac{d}{d\theta} \left( \frac{{r'}}{\sqrt{{r'}^2 + r^2}}\right) \\
\end{align*}
 
Dividing through by $r$ and writing $v = u' = {r'}/r$ this is
 
\begin{align*}
\frac{1}{\sqrt{v^2 + 1}}
&= \frac{d}{d\theta} \left( \frac{v}{\sqrt{v^2 + 1}}\right) \\
&= \frac{v'}{\sqrt{v^2 + 1}} -\frac{v^2 v'}{(\sqrt{v^2 + 1})^3} \\
\end{align*}
 
\begin{align*}
1 
&= v' \left( 1 -\frac{v^2 }{v^2 + 1} \right) \\
&= \frac{v'}{v^2 + 1} \\
\end{align*}

This is now separable, and can be integrated directly

\begin{align*}
\theta - \theta_0
&= \int \frac{dv}{v^2 + 1} \\
&= \arctan(v) \\
\end{align*}
 
\begin{align*}
\tan(\theta - \theta_0)
&= \frac{{r'}}{r} \\
&= \frac{d \ln(r) }{d\theta}
\end{align*}

That solves the first of the second order differential equations resulting from the Euler-Lagrange equations, and the 
last becomes
\begin{align*}
\ln(r) &= \int \tan(\theta - \theta_0) d\theta \\
&= -\ln(\cos(\theta - \theta_0)) + \ln(r_0)
\end{align*}

Finally a polar parametric equation is obtained
 
\begin{align*}
\frac{r}{r_0} \cos(\theta - \theta_0) = 1
\end{align*}
 
If all went right, this should be the equation for a straight line in polar form.

It doesn't look like one, but if the cosine is expanded 

\begin{align*}
\cos(\theta - \theta_0) 
&= \Re\left( e^{i\theta}e^{-i\theta_0} \right) \\
&= \Re\left( (\cos\theta + i\sin\theta)(\cos\theta_0 - i\sin\theta_0) \right) \\
&= \cos\theta\cos\theta_0 + \sin\theta\sin\theta_0 \\
\end{align*}

With $x = r\cos\theta$, and $y = r\sin\theta$ this gives

\begin{align*}
r_0 
&= r \left( \cos\theta\cos\theta_0 + \sin\theta\sin\theta_0 \right) \\
&= x \cos\theta_0 + y\sin\theta_0 \\
\end{align*}

So, sure enough, following the math gives an equation for a straight line in a recognizable form.

\subsection{ Problem 2. Shortest line, in 3D. }

Did this one in \cite{PJgoldch1}

\subsection{ Problem 3.  Spherical geodesics. }

First calculate the line element (this was given in the problem, but I feel like working it out).
The position vector with $i = \Be_1 \Be_2$, is given by

\begin{align*}
\Br = a ( \sin\theta \Be_1 e^{i\phi} + \Be_3 \cos\theta )
\end{align*}

So, the differential given constant radius $a$ is

\begin{align*}
d\Br 
&= 
a \thetadot ( \cos\theta \Be_1 e^{i\phi} - \Be_3 \sin\theta )
+ a \phidot ( \sin\theta \Be_1 \Be_1 \Be_2 e^{i\phi} ) \\
&= 
a \left( \thetadot \cos\theta \Be_1 + \phidot \sin\theta \Be_2 \right) e^{i\phi} - a \thetadot \Be_3 \sin\theta \\
\end{align*}

And the square is
\begin{align*}
d\Br^2
&= a^2 \left( \thetadot^2 \cos^2\theta + \phidot^2 \sin^2\theta + \thetadot^2 \sin^2\theta \right) \\
&= a^2 \left( \thetadot^2 + \phidot^2 \sin^2\theta \right) \\
\end{align*}

Here the derivatives are with respect to some implicit variable that parameterizes the differential displacement.
This can be taken to be $\theta$, which gives the distance along any two points on the sphere as

\begin{align*}
S &= a^2 \int d\theta \sqrt{1 + \left(\frac{d\phi}{d\theta}\right)^2 \sin^2\theta } \\
\end{align*}

Writing $f(\theta, \phi, \phidot) = \sqrt{1 + \phidot^2 \sin^2\theta}$, the Euler-Lagrange equations can be
applied

\begin{align*}
0 
&= \left( \PD{\phi}{} - \frac{d}{d\theta} \PD{\phidot}{} \right) f \\
&= 0 - \frac{d}{d\theta} \frac{(1/2)(2\phidot) \sin^2\theta}{\sqrt{1 + \phidot^2 \sin^2\theta}} \\
\end{align*}

Introducing an integration constant $\kappa$, this is

\begin{align*}
\phidot \sin^2\theta = \kappa \sqrt{1 + \phidot^2 \sin^2\theta}
\end{align*}

squaring
\begin{align*}
\phidot^2 \sin^4\theta &= \kappa^2 \left(1 + \phidot^2 \sin^2\theta \right) \\
\phidot^2 \sin^2\theta \left( \sin^2 \theta - \kappa^2 \right) &= \kappa^2 \\
\end{align*}

\begin{align*}
\phi - \phi_0 &= \kappa \int \frac{d\theta}{\sin\theta \sqrt{ \sin^2 \theta - \kappa^2 }} \\
\end{align*}

This doesn't look particularily nice to integrate.  Instead, let's try writing the arc length integral as

\begin{align*}
\frac{S}{a^2} &= \int d\phi \sqrt{{\frac{d\theta}{d\phi}}^2 + \sin^2\theta } \\
\end{align*}

\begin{align*}
0 
&= \left( \PD{\theta}{} - \frac{d}{d\phi} \PD{\thetadot}{} \right) \sqrt{\thetadot^2 + \sin^2\theta } \\
&= \frac{\sin\theta \cos\theta}{ \sqrt{\thetadot^2 + \sin^2\theta } } -
\frac{d}{d\phi} \frac{\thetadot} { \sqrt{\thetadot^2 + \sin^2\theta } } \\
\end{align*}

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
