\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}

\newcommand{\PDSq}[2]{\frac{\partial^2 {#2}}{\partial {#1}^2}}
\DeclareMathOperator{\sinc}{sinc}
\DeclareMathOperator{\PV}{PV}
\newcommand{\FF}[0]{\mathcal{F}}
\newcommand{\Sw}[0]{\mathcal{S}}
\newcommand{\IIinf}[0]{ \int_{-\infty}^\infty }

\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage


\title{ Applications of Fourier distribution theory to some PDEs. }
\author{Peeter Joot \quad peeter.joot@gmail.com }
\date{ March 4, 2009.  Last Revision: $Date: 2009/03/06 00:55:51 $ }

\begin{document}
\maketitle{}
\tableofcontents

\section{ Motivation. }

I recently listened to Prof Brad Osgood's lectures on distributions in Fourier
transform theory and read the associated lecture notes
\cite{osgoodFourier}.  Here is an attempt to apply these ideas to solution
of some of the common PDEs of physics (wave and Poisson equations).

Some of these were tackled recently using ``classical'' Fourier methods
in \cite{PJpoisson}. 
This requires ad-hoc PV associations of integrals with delta and step
functions.  Such solutions do not inspire confidence.  Without
validation of the solutions by substituion back into the generating PDE
or comparison to a known solution one is left wondering
if all the right fudges were actually performed to get the answer.

\section{ Conventions and definitions. }

\subsection{ Transform pairs. }

The form of the Fourier transform pairs used here will be

\begin{align*}
\hat{f}(\mathbf{k}) &= \frac{1}{(\sqrt{2\pi})^n} \int f(\mathbf{x}) e^{-i \mathbf{k} \cdot \mathbf{x} } d^n x \\
{f}(\mathbf{x}) &= \frac{1}{(\sqrt{2\pi})^n} \int \hat{f}(\mathbf{k}) e^{i \mathbf{k} \cdot \mathbf{x} } d^n k \\
\end{align*}

Of particular interest will be forward and inverse Fourier transforms
of functions in $\Sw$, the space of Schwarz functions.  These are defined
more exactly in Osgood's notes, but in short have all the most desirable
properties one can imagine in a Fourier transformation context.  Namely
continuous, infinitely differentiable, with the function and all 
derivatives rapidly decreasing (faster than any polynomial).

\section{ 1D first order homogeneous wave equation. }

This is the simplest PDE that I can think of that one should be able 
to apply Fourier techniques to.  We seek solutions $f(x,t)$ of

\begin{align}\label{eqn:firstOrderAdvancedWave}
\inv{v} \PD{t}{f} - \PD{x}{f} = 0
\end{align}

\subsection{ Setup. }

Taking Fourier transforms of equation \ref{eqn:firstOrderAdvancedWave}, and integration
by parts, we have

\begin{align*}
\FF\left(\inv{v} \PD{t}{f} - \PD{x}{f} \right) &=
\inv{v} \inv{\sqrt{2\pi}} \int \PD{t}{f} e^{-i k x} dx - \inv{\sqrt{2\pi}} \int \PD{x}{f} e^{-i k x } dx
\end{align*}

The second term can be integrated by parts

\begin{align*}
- \inv{\sqrt{2\pi}} \int \PD{x}{f} e^{-i k x } dx
=
- \inv{\sqrt{2\pi}} {\left. {f} e^{-i k x } \right\vert}_{x = -\infty}^{\infty}
+ \inv{\sqrt{2\pi}} \int {f} (-i k ) e^{-i k x } dx
\end{align*}

Already we can see we are in some trouble here since we require 
$f(x,t)$ vanish at $\pm \infty$.
For now I close my eyes, and ignore the fact that the final solution will be seen to be more 
general, and not neccessarily have such a restriction.

FIXME: continue reading osgood and figure out how to do this setup in the distribution formalism?
With a proper setup of this problem in the distribution formalism one should expect that this trouble will
go away, since the function $f$ will be coupled with the Schwartz test function
that vanishes satisfactorily at the boundaries.

Again closing eyes, if one assumes that the order of integration and
differentiation in the time 
derivative term above can be reversed, then
we have a nice single variable first order LDE to solve for $\hat{f}$

\begin{align}
\inv{v} \PD{t}{\hat{f}}(k,t) + (-ik)\hat{f}(k,t) = 0
\end{align}

Our solution in the wave number space is thus
\begin{align}\label{eqn:startingPtFirstOrder}
{\hat{f}}(k,t) = A(k) e^{i k v t} = \hat{f}(k,0) e^{i k v t}
\end{align}

This will be the common starting point for both the classical and the distribution approach below.
Because of the integration by parts restriction and the reversion of order of integration and time differentiation, this probably ought not
be the starting point for the distribution case.  Let's compromise and work with what's known so far one step at a time.

\subsubsection{ On reversal of the time derivative and integration order. }

If one introduces a time variation
in the integral bounds, but still have them go to infinity based on some
other limiting process, as in

\begin{align*}
\PD{t}{} \int_{a(R,t)}^{b(R,t)} f(x,t) dx
&=
{\left.\PD{t}{f}(x,t)\right\vert}_{x = b(R,t)} \PD{t}{b}(R,t)
-{\left.\PD{t}{f}(x,t)\right\vert}_{x = a(R,t)} \PD{t}{a}(R,t)
+ \int_{a(R,t)}^{b(R,t)} \PD{t}{f}(x,t) dx
\end{align*}

then
it appears the time derivative of $f$ must vanish faster than the time derivatives of the 
integral bounds go to infinity in order to fully justify the switch of 
the order of differentiation and integration above.  That's a rather loose statement, but probably a correct one.  I expect that
a complete formulation of the problem in terms of distributions would also avoid this issue, since the Schwarz functions and their 
derivatives have the desirable property of vanishing ``rapidly''.

\subsection{ Classical way. }

To complete the solution in terms of initial conditions we can inverse Fourier transform
equation \ref{eqn:startingPtFirstOrder}

\begin{align*}
{{f}}(x,t) = \inv{\sqrt{2\pi}} \left( \inv{\sqrt{2\pi}}\int f(x',0) e^{-i k x'} dx' e^{i k v t} \right) e^{ i k x } dk
\end{align*}

Here's where things get fishy.  Reversing the order of integration, and performing
the $k$ integration first we want to ``integrate''

\begin{align*}
I(x,t) = \inv{2\pi} \int e^{ i k ((x-x') + v t) } dk
\end{align*}

This we can fudge by saying taking the principle value integral (symmetric about the origin)

\begin{align*}
I(x,t) &= \lim_{R \rightarrow \infty} \inv{\pi} \frac{\sin( R (x-x' + v t) )}{ (x-x' + vt)}
\end{align*}

this is a nascent form of the delta function:

\begin{align*}
\delta(x) \sim \lim_{R \rightarrow \infty} \frac{\sinc( R x )}{\pi x}
\end{align*}

So we have
\begin{align*}
I(x,t) &= \delta(x - x' + v t) 
\end{align*}

\begin{align*}
{{f}}(x,t) 
&= \int f(x',0) \delta(x - x' + v t) dx' \\
&= f(x + v t,0) 
\end{align*}

The first order advanced wave equation is thus solved by any function $g(x + vt)$.  This
is familar from
\cite{french1971vaw} where it was stated (then demonstrated correct) that solutions of the second
order homogenous wave equation in one dimension can be expressed in terms of any functions
$g(x \pm vt)$.

By invoking just the right incantations we can pull the delta function out of the magic hat and 
produce this expected result.

\subsection{ Using distributions. }

Starting with \ref{eqn:startingPtFirstOrder} for now, we can form a functional pairing with a test function $\psi$ in the Schwartz function space $\Sw$ by integration.  That is
assume that our function $\hat{f}(k,t)$, or equivalently $f(x,t)$ can be found for all test functions $\psi(k) \in \Sw$ as follows

\begin{align*}
\int {\hat{f}}(k,t) \psi(k) dk = \int \hat{f}(k,0) e^{i k v t} \psi(k) dk
\end{align*}

In terms of $f$ this is

\begin{align*}
\int \left( \inv{2\pi} \int f(x,t) e^{-i k x} dx \right) \psi(k) dk = \int \left(\inv{2\pi} \int f(x,0) e^{-i k x} dx \right) e^{i k v t} \psi(k) dk
\end{align*}

and reversing order of integration we have

\begin{align*}
\int f(x,t) \left( \inv{2\pi} \int \psi(k) e^{-i k x} dk \right) dx = \int f(x,0) \left(\inv{2\pi} \int \psi(k) e^{-i k (x - v t)} dk \right) dx
\end{align*}

What we have left on both sides is a plain old fourier transform, which exists since $\psi(k) \in \Sw$. We can write

\begin{align*}
\phi(x) = \inv{2\pi} \int \psi(k) e^{-i k x} dk 
\end{align*}

and we have

\begin{align*}
\int f(x,t) \phi(x) dx = \int f(x,0) \phi(x - vt) dx
\end{align*}

With a change of variables $x' = x - vt$ on the right, and $x = x'$ on the left we have
% x = x' + vt

\begin{align*}
\int f(x',t) \phi(x') dx' = \int f(x' + vt,0) \phi(x') dx'
\end{align*}

Or, for all $\phi(x)$,

\begin{align*}
\int (f(x,t) - f(x + vt, 0) \phi(x) dx = 0
\end{align*}

Now that's pretty slick, and we end up with the same result $f(x,t) = f(x + vt, 0)$ for any function(al) f.
What was a delta function in the classical approach became nothing more than variable substitution and evaluation at a point.

The only place where
any sort of unjustified step was made was the reversal of the order of integration, and the rapid decrease property of the test function $\psi$ can probably be used
to make that more rigorous (FIXME: have to think that through).

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
