\documentclass{article}      % Specifies the document class

\input{../peeters_macros.tex}
\newcommand{\grad}[0]{\nabla}
\newcommand{\PD}[2]{ \frac{\partial{#1}}{\partial {#2}} }

%
% The real thing:
%

                             % The preamble begins here.
\title{} % Declares the document's title.
\author{Peeter Joot}         % Declares the author's name.
%\date{}        % Deleting this command produces today's date.

\begin{document}             % End of preamble and beginning of text.

\maketitle{}

\section{ Relativistic dynamics from Lagrangian. }

David Tong's problem 3 on Lagrangian's is to show that the following

\begin{equation}
S = \kappa \sqrt{1 - \Bv^2/c^2} - \varphi
\end{equation}

can be used to find the equation of motion.  The constant $\kappa$ is determined
by requiring correspondence with the classical limit for small velocities:

\begin{equation*}
\kappa \sqrt{1 - \Bv^2/c^2} \simeq \kappa \left( 1 - \inv{2} \Bv^2/c^2 \right) = \text{constant} + \inv{2} m \Bv^2 + \cdots
\end{equation*}

Thus by neglecting the higher order terms, we have the Newtonian kinetic energy for $\kappa = -mc^2$, and the action to minimize is:

\begin{equation}
S = -m c^2\sqrt{1 - \Bv^2/c^2} - \varphi \simeq \inv{2}m \Bv^2 - \left(\varphi + m c^2\right)
\end{equation}

\subsection{ Newtonian equations of motion. }
In the small speed limit the Lagrangian conditions for action $S = mv^2/2 - \varphi$ yield:

\begin{equation*}
\sum \Be_i \left(\PD{S}{x^i} - \frac{d}{dt}\PD{S}{\dot{x}^i}\right) = \sum \Be_i \left(-\PD{\varphi}{x^i} - \frac{d}{dt}\left(m \dot{x}^i\right)\right) = 0
\end{equation*}

This is out familiar Newtonian force law for a point particle in a conservative potential field:

\begin{equation}
F = \frac{d (m \Bv)}{dt} = -\grad \varphi.
\end{equation}

Observe that the gradient and velocity vectors as expressed here have no requirement that the basis vectors to be an
othonormal set.  The only requirement is that the coordinates $x^i$ and the basis vectors $\Be_i$ be reciprocal pairs, defined by the following relationships:

\begin{equation*}
\Bx = \sum \Be_i x^i
\end{equation*}
\begin{equation*}
x^i = \Bx \cdot \Be^i
\end{equation*}
\begin{equation*}
\Be^i \cdot \Be_j = {\delta^i}_j
\end{equation*}

The requirement for reciprocal pairs of coordinates and basis frame vectors is due to the 
summation $\Bv = \sum \Be_i \dot{x}^i$.  The Lagrange equations that minimize the action still generate equations of
motion that hold when the coordinate and basis vectors cannot be summed in this fashion.
In such a case, however, the ability to merge the generalized coordinate equations of motion into a single
vector relationship will not be possible.

\subsection{ Relativistic three vector equation of motion. }
\subsection{ Relativistic four vector equation of motion. }

\section{ OLDER FOUR VECTOR NOTES. }

I was summarizing for myself the various four-vectors of mechanics:

\begin{align*}
x &= ct + \mathbf{x} \\
V &= \frac{dx}{d\tau} = \gamma(c + \mathbf{v}) \\
P &= m V = E/c + \gamma\mathbf{p} \\
f &= m\frac{d^2 x}{d\tau^2} = m\frac{d V}{d\tau} \\
\end{align*}

where:

\begin{align*}
\gamma^{-2} &= 1 - {\lvert \mathbf{v}/c \rvert}^2 \\
ds = c d\tau &= {\left(\frac{dx}{d\lambda} \cdot \frac{dx}{d\lambda}\right)}^{1/2} d\lambda \\
x \cdot x = {\lvert x \rvert}^2 &= c^2t^2 - {\lvert \mathbf{x} \rvert}^2 \\
E &= \int f \cdot (c d\tau) \\
\mathbf{v} &= \frac{d\mathbf{x}}{dt} \\
\mathbf{p} &= m\mathbf{v} \\
\end{align*}

Invarients for the first three four vectors are:

\begin{align*}
{\lvert x \rvert}^2 &= c^2 t^2 - {\lvert \mathbf{x} \rvert}^2 = c^2 \tau^2 \\
{\lvert V \rvert}^2 &= \gamma^2 (c^2 - {\lvert \mathbf{v} \rvert}^2) = c^2 \\
{\lvert P \rvert}^2 &= m^2 {\lvert V \rvert}^2 = m^2 c^2 \\
\end{align*}

Is the minkowski norm of the four vector force:

\[
f = m\frac{d^2 x}{d\tau^2} 
\]

also an invarient?  I think it has to be.  Assuming that is the case, what would the value (and significance if any) of this be?

\end{document}               % End of document.
