\documentclass{article}      % Specifies the document class

\input{../peeters_macros.tex}
\newcommand{\grad}[0]{\nabla}

%
% The real thing:
%

\title{ Potential and Kinetic Energy.} 
\author{Peeter Joot}         
\date{}

\begin{document}             

\maketitle{}

\section{}

Attempting some Lagrangian calculation problems I found I got all the signs of my potential energy terms wrong.  Here's a quick step back to basics to clarify for myself what the definition of potential energy is, and thus implicitly determine the correct signs.

Starting with kinetic energy, expressed in vector form:

\begin{equation*}
K 
= \inv{2} m \Br' \cdot \Br' 
= \inv{2} \Bp \cdot \Br',
\end{equation*}

one can calculate the rate of change of that energy:

\begin{align*}
\frac{dK }{dt}
&= \inv{2} \left(\Bp' \cdot \Br' + \Bp \cdot \Br''\right) \\
&= \inv{2} \left(\Bp' \cdot \Br' + \Br' \cdot \Bp'\right) \\
&= \Bp' \cdot \Br'.
\end{align*}

Note that the mass has been assumed constant above.

Integrating this time rate of change of kinetic energy produces a force
line 
integral:

\begin{align*}
K_2 - K_1 
&= \int_{t1}^{t2} \frac{dK}{dt} dt \\
&= \int_{t1}^{t2} \Bp' \cdot \Br' dt \\
&= \int_{t1}^{t2} \Bp' \cdot \frac{d\Br'}{dt} dt \\
&= \int_{\Br_1}^{\Br_2} \BF \cdot d\Br
\end{align*}

For the path integral to depend on only the end points or the corresponding end times requires a conservative force that can be expressed as a gradient.
Let's say that $\BF = \grad f$, then integrating:

\begin{align*}
K_2 - K_1 
&= \int_{\Br_1}^{\Br_2} \BF \cdot d\Br \\
&= \int_{\Br_1}^{\Br_2} \grad f \cdot d\Br \\
&= {\text{limit}}_{\epsilon \rightarrow 0} \int_{\Br_1}^{\Br_1 + \epsilon\rcap}
   \left(\rcap \frac{f(\Br + \epsilon \rcap)}{\epsilon}\right)
      \cdot d\Br \\
&= \text{handwaving} \\
&= f(\Br_2) - f(\Br_1).
\end{align*}

Assembling the quantities for times $1$, and $2$, we have

\begin{equation}
K_2 -f(\Br_2) = K_1 - f(\Br_1) = \text{constant}.
\end{equation}

This constant is what we give the name Energy.  The quantities $-f(\Br_i)$ we label potential energy $V_i$, and finally write the total energy as the sum of the kinetic and potential energies for a particle at a point in time and space:

\begin{equation}
K_2 + V_2 = K_1 + V_1 = E
\end{equation}
\begin{equation}
\BF = -\grad V
\end{equation}

\subsection{ Work with a specific example.  Newtonian gravitational force.}

Take the gravitional force:

\begin{equation}
F = -\frac{GmM}{r^2} \rcap
\end{equation}

The rate of change of kinetic energy with respect to such a force (FIXME: think though signs ... with or against?), is:

\begin{align*}
\frac{dK}{dt} 
&= \Bp' \cdot \Br' \\
&= -\frac{GmM}{r^2} \rcap \cdot \frac{d\Br}{dt} \\
&= -\frac{GmM}{r^3} \Br \cdot \frac{d\Br}{dt}.
\end{align*}

The vector dot products above can be eliminated with the standard trick:

\begin{align*}
\frac{dr^2}{dt} 
&= \frac{\Br \cdot \Br}{dt} \\
&= 2 \frac{d\Br}{dt} \cdot \Br.
\end{align*}

Thus,
\begin{align*}
\frac{dK}{dt} 
&= -\frac{GmM }{2r^3} \frac{dr^2}{dt} \\
&= -\frac{GmM }{r^2} \frac{dr}{dt} \\
&= \frac{d}{dt} \left( \frac{GmM }{r} \right).
\end{align*}

This can be integrated to find the kinetic energy difference associated with a change of position in a gravitational field:

\begin{align*}
K_2 - K_1 
&= \int_{t_1}^{t_2} \frac{d}{dt} \left( \frac{GmM }{r} \right) dt \\
&= GmM \left( \inv{r_2} - \inv{r_1} \right).
\end{align*}

Or, 

\begin{align*}
K_2 - \frac{GmM}{r_2} = K_1 - \frac{GmM}{r_1} = E.
\end{align*}

Taking gradients of this negative term:

\begin{align*}
\grad \left( - \frac{GmM}{r} \right)
&= \rcap \frac{\partial}{\partial r} \left( - \frac{GmM}{r} \right) \\
&= \rcap \frac{GmM}{r^2},
\end{align*}

returns the negation of the original force, so if we write $V = -Gmm/r^2$, it implies the force is:

\begin{equation}
\BF = -\grad V.
\end{equation}

By this example we see how one arrives at the negative sign convention for the potential energy, and also how we end up with strictly positive terms in the Lagrangian associated with a gravitational force:

\begin{equation}
L = \inv{2} m \Bv^2 + \frac{GmM}{r}.
\end{equation}

\end{document}               
