%\documentclass{article}

%\input{../peeters_macros.tex}
%\input{../peeters_macros2.tex}

%\usepackage[bookmarks=true]{hyperref}

%\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
%\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage


\chapter{Some rapidity angle notes. }
\label{chap:rapidity}
%\author{Peeter Joot \quad peeter.joot@gmail.com}
\date{ Dec 18, 2008.  $RCSfile: rapidity.tex,v $ Last $Revision: 1.8 $ $Date: 2009/06/14 23:51:45 $ }

%\begin{document}

%\maketitle{}
%\tableofcontents

\section{Motivation. }

Lut writes, "setting up a little calculation, I'm writing a 4-velocity as"
 
\begin{align*}
( \gamma, \gamma \beta_x, \gamma \beta_y, \gamma \beta_z ),
\end{align*}
 
which has length $-1$ if $\gamma^{-2} = 1 - (\beta_x)^2+(\beta_y)^2+(\beta_z)^2$.
 
Can you write this in terms of the 3 rapidities $a_1, a_2, a_3$ ?

I wasn't able to answer this right away so it is worth an examination
of rapidity angles to ensure that I understand the ideas.

\section{Stuff. }

Putting back in the $c$ factors, and switching to the $+---$ signature I'm used to, the position
vector is

\begin{align*}
x &= x^\mu \gamma_\mu = ct \gamma_0 + x^i \gamma_i,
\end{align*}

for which the corresponding proper velocity is
\begin{align*}
v &= \frac{dx}{d\tau} = c \frac{dt}{d\tau} \gamma_0 + \frac{dx^i}{dt} \frac{dt}{d\tau} \gamma_i
\end{align*}

Writing $\gamma = dt/d\tau$, and squaring the proper velocity we have

\begin{align*}
\frac{v^2}{c^2}
&= 1 \\
&= \gamma^2 \left(\gamma_0 + \inv{c}\frac{dx^i}{dt} \gamma_i\right)^2 \\
&= \gamma^2 \left(1 - \sum_i \inv{c^2} \left(\frac{dx^i}{dt}\right)^2 \right) \\
\end{align*}

So we have 

\begin{align*}
\gamma 
&= \inv{\sqrt{1 - \sum_i \inv{c^2} \left(\frac{dx^i}{dt}\right)^2 }} \\
\end{align*}

Observe that $\gamma$ ranges from $1$ to infinity, and can thus be described by the $[0,\infty]$ range of the hyperbolic cosine function.  With
the relative velocity $\Bv = \sum_i (dx^i/dt) \sigma_i$, this is

\begin{align*}
\frac{dt}{d\tau} &= \gamma  \\
&= \cosh\alpha \\
&= \inv{\sqrt{1 - (\Bv/c)^2}}
\end{align*}

In terms of the hyperbolic cosine for $\gamma$ our proper velocity then becomes

\begin{align*}
v/c &= \cosh\alpha \left(1 + \frac{\Bv}{c} \right) \gamma_0
\end{align*}

Taking the hint from the Lorentz transform where we have both $\sinh$ and $\cosh$ factors can one write

\begin{align*}
\gamma \frac{\Bv}{c} = \sinh\alpha
\end{align*}

This gives

\begin{align*}
\frac{\Bv}{c} = \tanh\alpha
\end{align*}

so we need $\alpha$ to be a spacetime relative vector.  With $\cosh$ being an even function $\cosh{\alpha} = \cosh{\Abs{\alpha}}$, so this
is still a 
scalar as desired.  Inverting the relationship for $\alpha$ we have

\begin{align*}
\alpha = \tanh^{-1} (\Bv/c) = \vcap \tanh^{-1} (\Abs{\Bv/c})
\end{align*}

The unit vector $\vcap$ can be factored out of the inverse hyperbolic tangent function since it is odd (consider the Taylor series expansion of $\tanh^{-1}$ to see why one can do this).

Finally, we have by 
dotting with the spatial basis vectors $\sigma_i$ three quantities in terms of spacetime vector rapidity angle

\begin{align*}
\alpha_i = (\vcap \cdot \sigma_i) \tanh^{-1} (\Abs{\Bv/c}).
\end{align*}

The $\vcap \cdot \sigma_i$ parts are direction cosines, so the three rapidities Lut was asking about all appear to be weighted direction cosines.

%\bibliographystyle{plainnat}
%\bibliography{myrefs}

%\end{document}
