%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%
%
\chapter{DC Power consumption formula for resistive load}
\label{chap:dcPower}
%\date{Jan 06, 2009.  dcPower.tex}

\section{Motivation}

Despite a lot of recent study of electrodynamics, faced with a simple electrical problem:

``What capacity generator would be required for an arc welder on a 30 Amp breaker using a 220 volt circuit''.

I could not think of how to answer this off the top of my head.  Back in school without hesitation I would have
been able to plug into \(P = I V\) to get a capacity estimation for the generator.

Having forgotten the formula, let us see how we get that \(P = I V\) relationship from Maxwell's equations.

\section{}

Having just derived the Poynting energy momentum density relationship from Maxwell's equations, let that be the starting
point

\begin{equation}\label{eqn:dcPower:20}
\begin{aligned}
\frac{d}{dt}\left(\frac{\epsilon_0}{2}\left(\BE^2 + c^2\BB^2 \right) \right) = - \inv{\mu_0} \left(\BE \cross \BB \right) - \BE \cdot \Bj
\end{aligned}
\end{equation}

The left hand side is the energy density time variation, which is power per unit volume, so we can integrate this
over a volume to determine the power associated with a change in the field.

\begin{equation}\label{eqn:dcPower:40}
\begin{aligned}
P = -\int dV \left( \inv{\mu_0} \left(\BE \cross \BB \right) + \BE \cdot \Bj \right)
\end{aligned}
\end{equation}

As a reminder, let us write the magnetic and electric fields in terms of potentials.

In terms of the ``native'' four potential our field is

\begin{equation}\label{eqn:dcPower:60}
\begin{aligned}
F
&= \BE + ic \BB \\
&= \grad \wedge A \\
&= \gamma^0 \gamma_k \partial_0 A^k + \gamma^j \gamma_0 \partial_j A^0 + \gamma^m \wedge \gamma_n \partial_m A^n \\
\end{aligned}
\end{equation}

The electric field is

\begin{equation}\label{eqn:dcPower:80}
\begin{aligned}
\BE &= \sum_k (\grad \wedge A) \cdot (\gamma^0 \gamma^k) \gamma_k \gamma_0 \\
\end{aligned}
\end{equation}

From this, with \(\phi = A^0\), and \(\BA = \sigma_k A^k\) we have
\begin{equation}\label{eqn:dcPower:100}
\begin{aligned}
\BE &= -\inv{c} \PD{t}{\BA} - \grad \phi \\
i\BB &= \spacegrad \wedge \BA
\end{aligned}
\end{equation}

Now, the arc welder is (I think) a DC device, and to
get a rough idea of what it requires lets just assume that its a rectifier that outputs RMS DC.
So if we make this simplification, and assume that we have a
purely resistive load (ie: no inductance and therefore no magnetic fields) and a DC supply and constant current, then
we eliminate the vector potential terms.

This wipes out the \(\BB\) and the Poynting vector, and leaves our electric field specified in terms
of the potential difference across the load \(\BE = -\spacegrad \phi\).

That is
\begin{equation}\label{eqn:dcPower:120}
\begin{aligned}
P &= \int dV (\spacegrad \phi) \cdot \Bj
\end{aligned}
\end{equation}

Suppose we are integrating over the length of a uniformly resistive load with some fixed cross sectional area.  \(\Bj dV\) is then the magnitude of the current directed along the wire for its length.  This basically leaves us with a line integral over the length of the wire that we are measuring our potential drop over so we are left with just

\begin{equation}\label{eqn:dcPower:140}
\begin{aligned}
P &= (\delta \phi) I
\end{aligned}
\end{equation}

This \(\delta \phi\) is just our voltage drop \(V\), and this gives us the desired \(P = I V\) equation.
Now, I also recall from school
now that I think about it that \(P = I V\) also applied to inductive loads, but it required that \(I\) and \(V\) be phasors that
represented the sinusoidal currents and sources.  A good followup exercise would be to show from Maxwell's equations
that this is in fact valid.  Eventually I will know the origin of all the formulas from my old engineering courses.
