
\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}

\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage

\title{ Field and wave energy and momentum. }
\author{Peeter Joot}
\date{ Jan 03, 2009.  Last Revision: $Date: 2009/01/04 16:59:26 $ }

\begin{document}

\maketitle{}

%\tableofcontents
\section{ Motivation. }

The concept of energy in the electric and magnetic fields I am getting closer to understanding, but there’s a few ways that I would like to approach it.

I've now explored the Poynting vector energy conservation relationships in 
\cite{PJpoynting}, and
\cite{PJemstresstensor}
, but hadn't understood fully where the energy expressions in the electro and magneto statics cases came from
separately.  I also don't yet know where the $F \gamma_k F$ terms of the stress tensor fit in the big picture?  I suspect that they can be obtained by Lorentz transforming the rest frame expression $F \gamma_0 F$ (the energy density, Poynting momentum density four vector).

It also ought to be possible to relate the field energies to a Lagrangian and Hamiltonian, but I haven't had success doing so.

The last thing that I'd like to understand is how the energy and momentum of a wave can be expressed, both in terms of the abstract conguate field momentum concept and with a concrete example such as the one dimensional oscillating rod that can be treated in a limiting coupled oscillator approach as in 
\cite{goldstein1951cm}.

Once I've got all these down I think I'll be ready to revisit Bohm's Rayleigh-Jeans law treatment in \cite{bohm1989qt}.  Unfortunately, each time I try persuing some 
interesting aspect of QM I find that I end up back studying electrodynamics, and suspect that needs to be my focus for the forseeable future (perhaps working throughly through Feynman's
volume II).

\section{ Electrostatic energy in a field. } 

Feynman’s treatment in 
\cite{feynman1963flp}
of the energy $\frac{\epsilon_0}{2}\BE^2$ associated with the electrostatic $\BE$ field is very easy to understand.  Here is a write up of this myself without looking at the book to see if I really understood the ideas.

The first step is consideration of the force times distance for two charges gives you the energy required (or gained) by moving one of those charges from infinity to some given separation

\begin{align*}
W &= \frac{1}{4\pi\epsilon_0} \int_{\infty}^{r} \frac{q_1 q_2}{x^2} \Be_1 \cdot (-\Be_1 dx) \\
&= \frac{q_1 q_2}{4 \pi \epsilon_0 r}
\end{align*}

This provides a quantization for an energy in a field concept.  A distribution of charge requires energy to assemble and it is possible to enumerate that energy separately by considering all the charges, or alternatively, by not looking at the final charge distribution, but only considering the net field associated with this charge distribution.  This is a pretty powerful, but somewhat abstract seeming idea.

The generalization to continuous charge distribution from there was pretty straightforward, requiring a double integration over all space

\begin{align*}
W &= \frac{1}{2} \int \frac{1}{4\pi\epsilon_0} \frac{\rho_1 dV_1 \rho_2 dV_2}{r_{12}}  \\
&= \frac{1}{2} \int \rho_1 \phi_2 dV_1
\end{align*}

The $1/2$ factor was due to double counting all "pairs" of charge elements.  The next step was to rewrite the charge density by using Maxwell's equations.  In terms of the four vector potential Maxwell's equation (with the $\grad \cdot A = 0$ gauge) is

\begin{align*}
\grad^2 A = \inv{\epsilon_0 c}( c \rho \gamma_0 + J^k \gamma_k)
\end{align*}

So, to write $\rho$ in terms of potential $A^0 = \phi$, we have

\begin{align*}
\left(\inv{c^2}\PDsq{t}{} - \spacegrad^2\right) \phi = \inv{\epsilon_0} \rho 
\end{align*}

In the statics case, where $\PD{t}{\phi} = 0$, we can thus write the charge density in terms of the potential

\begin{align*}
\rho = -\epsilon_0 \spacegrad^2 \phi
\end{align*}

and substitute back into the energy summation

\begin{align*}
W 
&= \frac{1}{2} \int \rho \phi dV \\
&= \frac{-\epsilon_0}{2} \int \phi \spacegrad^2 \phi dV \\
\end{align*}

Now, Feynman's last step was a bit sneaky, which was to convert the $\phi \spacegrad^2 \phi$ term into a divergence integral.  Working backwards to derive the identity that he used

\begin{align*}
\spacegrad \cdot (\phi \spacegrad \phi) 
&= \gpgradezero{ \spacegrad (\phi \spacegrad \phi) } \\
&= \gpgradezero{ (\spacegrad \phi) \spacegrad \phi + \phi \spacegrad (\spacegrad \phi) } \\
&= (\spacegrad \phi)^2 + \phi \spacegrad^2 \phi \\
\end{align*}

This can then be used with Stokes theorem in its dual form to convert our $\phi \spacegrad^2 \phi$ the into plain volume and surface integral
\begin{align*}
W 
&= \frac{\epsilon_0}{2} \int \left( (\spacegrad \phi)^2 -\spacegrad \cdot (\phi \spacegrad \phi) \right) dV \\
&= \frac{\epsilon_0}{2} \int (\spacegrad \phi)^2 dV - \frac{\epsilon_0}{2} \int_{\partial V} (\phi \spacegrad \phi) \cdot \ncap dA \\
\end{align*}

Letting the surface go to infinity and employing a limiting argument on the magnitudes of the $\phi$ and $\spacegrad \phi$ terms was enough to produce the final electrostatics 
energy result

\begin{align*}
W 
&= \frac{\epsilon_0}{2} \int (\spacegrad \phi)^2 dV \\
&= \frac{\epsilon_0}{2} \int \BE^2 dV 
\end{align*}

\section{ Magnetostatic field energy. } 

Feynman's energy discussion of the magnetic field for a constant current loop (magnetostatics), is not so easy to follow.  He considers the dipole moment of a small loop, obtained by comparision to previous electrostatic results (that I'd have to go back and read or re-derive) and some
subtle seeming arguments about the mechanical vs. total energy of the system.

As an attempt to understand all this, let's break it up into pieces.  First, is calculation of the field for a current loop.
Going back to
Maxwell's equation (with the $\grad \cdot A$ gauge again) will show us how this Biot-Savart calculation is expressed in the STA bivector form

\begin{align*}
\grad^2 A^\mu = J^\mu/\epsilon_0 c
\end{align*}

For a constant current situation, we have $\PDi{t}{A^\mu} = 0$, and our vector potential equations are

\begin{align*}
\spacegrad^2 A^k = -J^k/\epsilon_0 c
\end{align*}

Recall that the solution of $A^k$ can be expressed as the impulse response of a function of the following form

\begin{align*}
A^k = C\inv{r}
\end{align*}

and that $\spacegrad \cdot (\spacegrad (1/r))$ is zero for all $r \ne 0$.  Performing a volume integral of the expected Laplacian we
can integrate over an infinitesimal spherical volume of radius $R$

\begin{align*}
\int \spacegrad^2 A^k dV 
&= C \int \spacegrad \cdot \spacegrad \inv{r} dV \\
&= C \int \spacegrad \cdot \left( -\rcap \inv{r^2} \right) dV \\
&= -C \int_{\partial_V} \rcap \inv{r^2} \cdot \rcap dA \\
&= -C \inv{R^2} 4 \pi R^2 \\
&= - 4 \pi C \\
\end{align*}

Equating we can solve for $C$

\begin{align*}
- 4 \pi C &= -J^k/\epsilon_0 c \\
C &= \inv{4 \pi \epsilon_0 c} J^k
\end{align*}

Note that this is cheating slightly since C was kind of treated as a constant, whereas this equality makes it a function.  It works because
the infinitesimal volume can be made small enough that $J^k$ can be treated as a constant.  This therefore provides our 
potential function in terms of this impulse response

\begin{align*}
A^k &= \int dV \inv{4 \pi \epsilon_0 c r} J^k
\end{align*}

Now, this could have all been done with a comparison to the electrostatic result.  Regardless, it now leaves us in the position to
calcuate the field bivector

\begin{align*}
F 
&= \grad \wedge A \\
&= (\gamma^\mu \wedge \gamma_k) \partial_\mu A^k \\
&= -(\gamma_m \wedge \gamma_k) \partial_m A^k \\
&= (\sigma_m \wedge \sigma_k) \partial_m A^k \\
\end{align*}

Now, 
\begin{align*}
\partial_m A^k
&= \inv{4 \pi \epsilon_0 c} \int dV \partial_m \frac{J^k}{r} \\
&= \inv{4 \pi \epsilon_0 c} \int dV \left( J^k \partial_m \inv{r} + \inv{r} \partial_m {J^k} \right) \\
&= \inv{4 \pi \epsilon_0 c} \int dV \left( J^k \partial_m \left(\sum_j ((x^j)^2)^{-1/2}\right) + \inv{r} \partial_m {J^k} \right) \\
&= \inv{4 \pi \epsilon_0 c} \int dV \left( J^k \left(-\inv{2}\right) 2 x^m \inv{r^3} + \inv{r} \partial_m {J^k} \right) \\
&= \inv{4 \pi \epsilon_0 c} \int \inv{r^3} dV \left( -x^m J^k + r^2 \partial_m {J^k} \right) \\
\end{align*}

So with $\Bj = J^k \sigma_k$ we have

\begin{align*}
F 
&= \inv{4 \pi \epsilon_0 c} \int \inv{r^3} dV \left( -\Br \wedge \Bj + r^2 (\spacegrad \wedge \Bj) \right) \\
&= \inv{4 \pi \epsilon_0 c} \int dV \left( \frac{\Bj \wedge \rcap}{r^2} + \inv{r}(\spacegrad \wedge \Bj) \right) \\
\end{align*}

The first term here is essentially the Biot Savart law once the current density is converted to current $\int \Bj dV = I \int \jcap dl$, so we expect the second term to be zero.  How to show this?

\section{ Complete field energy. } 

TODO: Can a integral of the Lorentz force coupled with Maxwell's equations in their entirety produce the energy expression $\frac{\epsilon_0}{2}\left(\BE^2 + c^2\BB^2\right)$?  It seems like cheating to add these arbitrarily and then follow the Poynting derivation by taking derivatives.  That shows this quantity is a conserved quantity, but does it really show that it is the
energy?  One could imagine that there could be other terms in a total energy expression such as $\BE \cdot \BB$.

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
