%\documentclass{article}

%\input{../peeters_macros.tex}
%\input{../peeters_macros2.tex}
%\input{../peeters_macros3.tex}

%\usepackage{listings}
%\usepackage{txfonts} % for ointctr... (also appears to make "prettier" \int and \sum's)
% makes \grad look funny though (almost like spacegrad, but narrower)
%\usepackage[bookmarks=true]{hyperref}

%\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
%\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage


\chapter{Wave equation form of Maxwell's equations}\label{chap:emVacWave}
%\blogpage{http://sites.google.com/site/peeterjoot/math2009/emVacWave.pdf}
%\author{Peeter Joot \quad peeter.joot@gmail.com }
\date{ June 21, 2009.  $RCSfile: emVacWave.tex,v $ Last $Revision: 1.4 $ $Date: 2009/07/09 03:04:06 $ }

%\begin{document}

%\maketitle{}
%\tableofcontents

\section{Motivation.}

In \cite{jackson1975cew}, on plane waves, he writes "we find easily..." to show that the wave equation for each of the components of $\BE$, and $\BB$ in the absence of current and charge satisfy the wave equation.  Do this calculation.

\section{Vacuum case.}

Avoiding the non-vacuum medium temporarily, Maxwell's vacuum equations (in SI units) are

\begin{align}\label{eqn:emVacWave:divE}
\spacegrad \cdot \BE = 0
\end{align}
\begin{align}\label{eqn:emVacWave:divB}
\spacegrad \cdot \BB = 0
\end{align}
\begin{align}\label{eqn:emVacWave:curlB}
\spacegrad \cross \BB = \inv{c^2} \frac{\partial \BE}{\partial t}
\end{align}
\begin{align}\label{eqn:emVacWave:curlE}
\spacegrad \cross \BE = -\frac{\partial \BB}{\partial t}
\end{align}

The last two curl equations can be decoupled by once more calculating the curl.  Illustrating by example

\begin{align}\label{eqn:emVacWave:curlCurlE}
\spacegrad \cross (\spacegrad \cross \BE) = -\frac{\partial }{\partial t} \spacegrad \cross \BB = -\inv{c^2} \frac{\partial^2 \BE}{\partial t^2}
\end{align}

Digging out vector identities and utilizing the zero divergence we have

\begin{align}\label{eqn:emVacWave:identForcurlCurlE}
\spacegrad \cross (\spacegrad \cross \BE) = \spacegrad (\spacegrad \cdot \BE) - \spacegrad^2 \BE = -\spacegrad^2 \BE
\end{align}

Putting \ref{eqn:emVacWave:curlCurlE}, and \ref{eqn:emVacWave:identForcurlCurlE} together provides a wave equation for the electric field vector

\begin{align}\label{eqn:emVacWave:waveE}
\inv{c^2} \frac{\partial^2 \BE}{\partial t^2} - \spacegrad^2 \BE = 0
\end{align}

Operating with curl on the remaining Maxwell equation similarly produces a wave equation for the magnetic field vector

\begin{align}\label{eqn:emVacWave:waveB}
\inv{c^2} \frac{\partial^2 \BB}{\partial t^2} - \spacegrad^2 \BB = 0
\end{align}

This is really six wave equations, one for each of the field coordinates.

\section{With Geometric Algebra.}

Arriving at (\ref{eqn:emVacWave:waveE}), and (\ref{eqn:emVacWave:waveB}) is much easier using the GA formalism of (\cite{doran2003gap}).

Pre or post multiplication of the gradient with the observer frame time basis unit vector $\gamma_0$ has a conjugate like
action

\begin{align*}
\grad \gamma_0
&=
\gamma^0 \gamma_0 \partial_0 + \gamma^k \gamma_0 \partial_k \\
&=
\partial_0 - \spacegrad \\
\end{align*}

(where as usual our spatial basis is $\sigma_k = \gamma_k \gamma_0$).

Similarly
\begin{align*}
\gamma_0 \grad 
&=
\partial_0 + \spacegrad \\
\end{align*}

For the vacuum Maxwell's equation is just
\begin{align*}
\grad F = \grad (\BE + I c \BB) = 0
\end{align*}

With nothing more than an algebraic operation we have

\begin{align*}
0 
&= \grad \gamma_0 \gamma_0 \grad F \\
&=
( \partial_0 - \spacegrad ) ( \partial_0 + \spacegrad ) (\BE + I c \BB) \\
&=
\left( \inv{c^2} \frac{\partial^2}{\partial t^2} - \spacegrad^2 \right) (\BE + I c \BB) \\
\end{align*}

This equality is true independently for each of the components of $\BE$ and $\BB$, so we have as before

These wave equations are still subject to the constraints of the original Maxwell equations.  

\begin{align*}
0 &= \gamma_0 \grad F \\
&= (\partial_0 + \spacegrad) (\BE + I c \BB) \\
&= 
  \spacegrad \cdot \BE 
+ (\partial_0 \BE - c \spacegrad \cross \BB)
+ I ( c \partial_0 \BB + \spacegrad \cross \BE )
+ I c \spacegrad \cdot \BB 
\\
\end{align*}

\section{Tensor approach?}

In both the traditional vector and the GA form one can derive the wave equation relations of (\ref{eqn:emVacWave:waveE}), \ref{eqn:emVacWave:waveB}).  One can obviously summarize these in tensor form as

\begin{align}\label{eqn:emVacWave:waveFaraday}
\partial_\mu\partial^\mu F^{\alpha\beta} = 0
\end{align}

working backwards from the vector or GA result.  In this notation, the coupling constraint would be that the field variables $F^{\alpha\beta}$ are subject to the Maxwell divergence equation (name?)

\begin{align}\label{eqn:emVacWave:divergenceFaraday}
\partial_\mu F^{\mu\nu} = 0
\end{align}

and also the dual tensor relation

\begin{align}\label{eqn:emVacWave:dualFaraday}
\epsilon^{\sigma\mu\alpha\beta} \partial_\mu F_{\alpha\beta} = 0
\end{align}

I cannot seem to figure out how to derive (\ref{eqn:emVacWave:waveFaraday}) starting from these tensor relations?

This probably has something to do with the fact that we require both the divergence and the dual relations 
(
\ref{eqn:emVacWave:divergenceFaraday}
,
\ref{eqn:emVacWave:dualFaraday}
)
expressed together to do this.

\section{Electromagnetic waves in media.}

Jackson lists the Macroscopic Maxwell equations in (6.70) as 

\begin{align*}
\spacegrad \cdot \BB &= 0 \\
\spacegrad \cdot \BD &= 4 \pi \rho \\
\spacegrad \cross \BE + \inv{c}\PD{t}{\BB} &= 0 \\
\spacegrad \cross \BH - \inv{c}\PD{t}{\BD} &= \frac{4 \pi}{c} \BJ  \\
\end{align*}

(for this note this means unfortunately a switch from SI to CGS midstream)

For vacuum ($\rho =0$, and $\BJ =0$), plus $\BB = \mu \BH$, and $\BD = \epsilon \BE$, we can assemble these into his (7.1) equations

\begin{align*}
\spacegrad \cdot \BB &= 0 \\
\spacegrad \cdot \BE &= 0 \\
\spacegrad \cross \BE + \inv{c}\PD{t}{\BB} &= 0 \\
\spacegrad \cross \BB - \frac{\epsilon \mu}{c}\PD{t}{\BE} &= 0  \\
\end{align*}

In this macroscopic form, we cannot assemble the equations into a nice tidy GA form.  The best we can do is

\begin{align}
\spacegrad \BE + \partial_0 (I\BB) &= 0 \\
\spacegrad (I \BB) + \epsilon \mu \partial_0 \BE &= 0 
\end{align}

Although not as pretty, we can at least derive the wave equations from these.  For example for $\BE$, we apply one additional spatial gradient

\begin{align*}
0 
&= \spacegrad^2 \BE + \partial_0 (\spacegrad I \BB) \\
&= \spacegrad^2 \BE + \partial_0 ( -\epsilon \mu \partial_0 \BE ) \\
\end{align*}

For $\BB$ we get the same, and have two wave equations

\begin{align}
\frac{\mu \epsilon}{c^2} \frac{\partial^2 \BE}{\partial t^2} - \spacegrad^2 \BE &= 0 \\
\frac{\mu \epsilon}{c^2} \frac{\partial^2 \BB}{\partial t^2} - \spacegrad^2 \BB &= 0
\end{align}

The wave velocity is thus not $c$, but instead the reduced speed of $c/{\sqrt{\mu\epsilon}}$.

%\bibliographystyle{plainnat}
%\bibliography{myrefs}

%\end{document}
