\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}

\newcommand{\FF}[0]{\mathcal{F}}
\newcommand{\IIinf}[0]{ \int_{-\infty}^\infty }
\DeclareMathOperator{\PV}{PV}
\DeclareMathOperator{\sinc}{sinc}

\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage


\title{ A cheatsheet for Fourier transform conventions. }
\author{Peeter Joot}
\date{ Jan 21, 2009.  Last Revision: $Date: 2009/01/22 00:20:33 $ }

\begin{document}

\maketitle{}

%\tableofcontents

\section{ A cheatsheet for different Fourier integral notations. }

Damn.  There's too many different notations for the Fourier transform.  Examples are:

\begin{align*}
\tilde{f}(k) &= \IIinf f(x) \exp\left( - 2 \pi i k x \right) dx \\
\tilde{f}(k) &= \inv{\sqrt{2\pi}} \IIinf f(x) \exp\left( -i k x \right) dx \\
\tilde{f}(p) &= \sqrt{\frac{1}{2\pi\hbar}} \IIinf f(x) \exp\left( \frac{-i p x}{\hbar} \right) dx \\
\end{align*}

There's probably many more, with other variations such as using hats over things instead of twiddles, and so forth.

Unfortunately each of these have different numeric factors for the inverse transform.
Having just been bitten by rogue factors of $2 \pi$ after inoccently switching notations, it seems
worthwhile
to express the Fourier transform with a general fudge factor in the exponential.  Then it can be
seen at a 
glance what constants are required in the inverse transform given anybody's particular choice of the transform
definition.

Where to put all the factors can actually be seen from the QM formulation since one is free to treat $\hbar$
as an arbitrary constant, but let's do it from scratch in a mechanical fashion without having to think
back to QM as a fundamental.

Suppose we define the Fourier transform as 

\begin{align*}
\tilde{f}(s) = \kappa \IIinf f(x) \exp\left( - i \alpha s x \right) dx \\
{f}(x) = \kappa' \IIinf \tilde{f}(s) \exp\left( i \alpha x s \right) ds \\
\end{align*}

Now, what factor do we need in the inverse transform to make things work out right?  With the Rigor
Police on holiday, let's expand the inverse transform integral in terms of the original transform
and see what these numeric factors must then be to make this work out.

Omitting temporarily the $\kappa$ factors to be determined we have 

\begin{align*}
f(x) 
&\propto \IIinf \left( \IIinf f(u) \exp\left( - i \alpha s u \right) du \right) \exp\left( i \alpha x s \right) ds \\
&= \IIinf f(u) du \IIinf \exp\left( i \alpha s (x-u) \right) ds \\
&= \IIinf f(u) du \lim_{R\rightarrow \infty} 2 \pi \inv{ \pi \alpha(x-u) }\sin\left( \alpha R (x-u) \right) \\
&= \IIinf f(u) du 2 \pi \delta\left( \alpha (x-u) \right) \\
&= \inv{\alpha} \IIinf f(v/\alpha) dv 2 \pi \delta\left( \alpha x - v \right) \\
&= \frac{2 \pi}{\alpha} f((\alpha x)/\alpha) \\
&= \frac{2 \pi}{\alpha} f(x)
\end{align*}

Note that to get the result above, after switching order of integration, and assuming that we can take the principle value of the integrals,
the usual ad-hoc $\sinc$ and exponential integral identification of the delta function was made

\begin{align*}
\PV \inv{2\pi} \IIinf \exp\left( i  s  x \right) ds 
&= \lim_{R \rightarrow \infty} \inv{2\pi} \int_{-R}^R \exp\left( i  s  x \right) ds \\
&= \lim_{R\rightarrow \infty} \frac{\sin\left(  R  x \right)}{ \pi  x } \\
&\equiv \delta\left( x \right) \\
\end{align*}

The end result is that we will need to fix

\begin{align*}
\kappa \kappa' = \frac{\alpha}{2\pi}
\end{align*}

to have the transform pair produce the desired result.  Our transform pair is therefore

\begin{align*}
\tilde{f}(s) = \kappa \IIinf f(x) \exp\left( - i \alpha s x \right) dx \Leftrightarrow f(x) = {\frac{ \alpha} {2 \pi \kappa} } \IIinf \tilde{f}(s) \exp\left( i \alpha s x \right) ds
\end{align*}

Given this we can write the required numeric factors that accompany all the various forward transforms.

This was the form used in my old systems and signals course
\begin{align*}
\tilde{f}(s) &= \IIinf f(x) \exp\left( - 2 \pi i s x \right) dx \\
%\Leftrightarrow  \\
f(x) &= \IIinf \tilde{f}(s) \exp\left( 2 \pi i s x \right) ds \\
\end{align*}

See this one used by mathematicians

\begin{align*}
\tilde{f}(s) &= \inv{\sqrt{2\pi}} \IIinf f(x) \exp\left( - i  s x \right) dx \\
%\Leftrightarrow \\
f(x) &= \inv{\sqrt{2 \pi}} \IIinf \tilde{f}(s) \exp\left( i  s x \right) ds \\
\end{align*}

Remember this one from school too, probably in one of my circuits courses

\begin{align*}
\tilde{f}(s) &= \IIinf f(x) \exp\left( - i  s x \right) dx  \\
%\Leftrightarrow \\
f(x) &= \inv{{2 \pi}} \IIinf \tilde{f}(s) \exp\left( i  s x \right) ds \\
\end{align*}

and finally, from \cite{mcmahon2005qmd}

\begin{align*}
\tilde{f}(p) &= \inv{\sqrt{2\pi \hbar}} \IIinf f(x) \exp\left( - \frac{i  p x}{\hbar} \right) dx \\
%\Leftrightarrow \\
f(x) &= \inv{\sqrt{2 \pi \hbar}} \IIinf \tilde{f}(p) \exp\left( \frac{i  p x}{\hbar} \right) dp \\
\end{align*}

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
