%\input{../peeter_prologue.tex}

\chapter{Relativistic Doppler formula.}
\label{chap:frequencyTx}
\blogpage{http://sites.google.com/site/peeterjoot/math2009/frequencyTx.pdf}
%\date{June 27, 2009}
%\revisionInfo{$RCSfile: frequencyTx.tex,v $ Last $Revision: 1.5 $ $Date: 2009/07/11 05:49:02 $}

\date{June 27, 2009 $RCSfile: frequencyTx.tex,v $ Last $Revision: 1.5 $ $Date: 2009/07/11 05:49:02 $}

\beginArtWithToc

\section{Transform of angular velocity four vector.}

It was possible to derive the Lorentz boost matrix by requiring that the wave equation operator

\begin{align}
\grad^2 = \inv{c^2}\frac{\partial^2}{\partial t^2} - \spacegrad^2
\end{align}

retain its form under linear transformation (\cite{PJLorentzWave}).  Applying spatial Fourier transforms (\cite{PJwaveFourier}), one finds that solutions to the wave equation 

\begin{align}
\grad^2 \psi(t,\Bx) = 0
\end{align}

Have the form

\begin{align}
\psi(t, \Bx) = \int A(\Bk) e^{i(\Bk \cdot \Bx - \omega t)} d^3 k
\end{align}

Provided that $\omega = \pm c \Abs{\Bk}$.  Wave equation solutions can therefore be thought of as continuously weighted superpositions of constrained fundamental solutions

\begin{align}
\psi &= e^{i(\Bk \cdot \Bx - \omega t)} \\
c^2 \Bk^2 &= \omega^2
\end{align}

The constraint on frequency and wave number has the look of a Lorentz square

\begin{align}
\omega^2 - c^2 \Bk^2 = 0
\end{align}

Which suggests that in additional to the spacetime vector

\begin{align}
X = (ct, \Bx) = x^\mu \gamma_\mu
\end{align}

evident in the wave equation fundamental solution, we also have a frequency-wavenumber four vector

\begin{align}
K = (\omega/c, \Bk) = k^\mu \gamma_\mu
\end{align}

The pair of four vectors above allow the fundamental solutions to be put explicitly into covariant form

\begin{align}
K \cdot X = \omega t - \Bk \cdot \Bx = k_\mu x^\mu
\end{align}

\begin{align}
\psi = e^{-i K \cdot X}
\end{align}

Let's also examine the transformation properties of this fundamental solution, and see as a side effect that $K$
has transforms appropriately as a four vector.

\begin{align*}
0 &= \grad^2 \psi(t,\Bx) \\
&= {\grad'}^2 \psi(t',\Bx') \\
&= {\grad'}^2 e^{i(\Bx' \cdot \Bk' - \omega' t')} \\
&= -\left(\frac{{\omega'}^2}{c^2} - {\Bk'}^2 \right) e^{i(\Bx' \cdot \Bk' - \omega' t')} \\
\end{align*}

We therefore have the same form of frequency wave number constraint in the transformed frame (if we require that
the wave function for light is unchanged under transformation)

\begin{align}
{\omega'}^2 = c^2 {\Bk'}^2 
\end{align}

Writing this as

\begin{align}
0 = {\omega}^2 - c^2 {\Bk}^2 = {\omega'}^2 - c^2 {\Bk'}^2 
\end{align}

singles out the Lorentz invariant nature of the $(\omega, \Bk)$ pairing, and we conclude that this pairing 
does indeed transform as a four vector.

\section{Application of one dimensional boost.}

Having attempted to justify the four vector nature of the wave number vector $K$, now move on to application of a boost along the x-axis to this vector.

\begin{align*}
\begin{bmatrix}
\omega' \\
c k' \\
\end{bmatrix}
&=
\gamma
\begin{bmatrix}
1 & -\beta \\
-\beta& 1 \\
\end{bmatrix}
\begin{bmatrix}
\omega \\
c k \\
\end{bmatrix} 
\\
&=
\begin{bmatrix}
\omega - v k \\
c k - \beta \omega
\end{bmatrix} 
\end{align*}

We can take ratios of the frequencies if we make use of the dependency between $\omega$ and $k$.  Namely, $\omega = \pm c k$.  We then have

\begin{align*}
\frac{\omega'}{\omega}
%&= \omega - (v/c) (\pm \omega)
&= \gamma(1 \mp \beta) \\
&= \frac{1 \mp \beta}{\sqrt{1 - \beta^2}} \\
&= \frac{1 \mp \beta}{\sqrt{1 - \beta}\sqrt{1 + \beta}} \\
\end{align*}

For the positive angular frequency this is

\begin{align*}
\frac{\omega'}{\omega}
&= \frac{\sqrt{1 - \beta}}{\sqrt{1 + \beta}} 
\\
\end{align*}

and for the negative frequency the reciprocal.

Deriving this with a Lorentz boost is much simpler than the time dilation argument in wikipedia doppler article (\cite{wiki:relDoppler}).  EDIT: Later found exactly the above boost argument in the wiki k-vector article (\cite{wiki:kvector}).

What's missing here is putting this in a physical context properly with source and reciever frequencies spelled out.  That would make this more than just math.

\EndArticle
