\documentclass{article}      % Specifies the document class

\input{../peeters_macros.tex}

%
% The real thing:
%

                             % The preamble begins here.
\title{Inertia Tensor} % Declares the document's title.
\author{Peeter Joot}         % Declares the author's name.
%\date{}        % Deleting this command produces today's date.

\begin{document}             % End of preamble and beginning of text.

\maketitle{}

\section{}

GAFP derives the angular momentum for rotational motion in the following form

\[
L = R \left( \int \Bx \wedge (\Bx \cdot \Omega_B) dm \right) R^\dagger
\]

and calls the integral part, the inertia tensor

\[
\emph{I}(B) = \int \Bx \wedge (\Bx \cdot \Omega_B) dm
\]

which is a linear mapping from bivectors to bivectors.  To understand the
form of this I found it helpful to expanding the wedge product
part of this explicitly for the \R{3} case.

Ignoring the sum in this expansion write

\[
f(B) = \Bx \wedge (\Bx \cdot B)
\]

And writing $\Be_{ij} = \Be_i \Be_j$ introduce a basis

\[
b = \{ \Be_1 I, \Be_2 I, \Be_3 I \} = \{ \Be_{23}, \Be_{31}, \Be_{12} \}
\]

for the \R{3} bivector product space.

Now calculate $f(B)$ for each of the basis vectors

\begin{align*}
f(\Be_1 I) 
&= \Bx \wedge (\Bx \cdot \Be_{23}) \\
&= ( x_1 \Be_1 + x_2 \Be_2 + x_3 \Be_3) \wedge (x_2 \Be_3 - x_3 \Be_2) \\
\end{align*}

Completing this calculation for each of the unit basic bivectors, we have
%%+ (           + x_2 \Be_2            ) \wedge (x_2 \Be_3            ) \\
%%+ (                       + x_3 \Be_3) \wedge (          - x_3 \Be_2) \\
%+ ( x_1 \Be_1                        ) \wedge (x_2 \Be_3            ) \\
%+ ( x_1 \Be_1                        ) \wedge (          - x_3 \Be_2) \\
%%+ (           + x_2 \Be_2            ) \wedge (          - x_3 \Be_2) \\ %% = 0
%%+ (                       + x_3 \Be_3) \wedge (x_2 \Be_3            ) \\ %% = 0
\begin{align*}
f(\Be_1 I) &= (x_2^2 + x_3^2) \Be_{23} - (x_1 x_2) \Be_{31} - (x_1 x_3) \Be_{12} \\
f(\Be_2 I) &= -(x_1 x_2) \Be_{23} + (x_1^2 + x_3^2) \Be_{31} - (x_2 x_3) \Be_{12} \\
f(\Be_3 I) &= -(x_1 x_3) \Be_{23} - (x_2 x_3) \Be_{31} - (x_1^2 x_2^2) \Be_{12} \\
\end{align*}

Observe that taking dot products with $(\Be_i I)^\dagger$ will select just the $\Be_i I$ term of the result, so one can
form the matrix of this linear transformation that maps bivectors in basis $b$ to image vectors also in basis $b$ as follows

\[
{\begin{bmatrix}
\emph{I}(B)
\end{bmatrix}}_{b}^{b}
=
{\begin{bmatrix}
\emph{I}(\Be_i I) \cdot (\Be_j I)^\dagger
\end{bmatrix}}_{ij}
=
\int {\begin{bmatrix}
(x_2^2 + x_3^2)  & - (x_1 x_2)  & - (x_1 x_3)  \\
-(x_1 x_2)  & + (x_1^2 + x_3^2)  & - (x_2 x_3)  \\
-(x_1 x_3)  & - (x_2 x_3)  & - (x_1^2 x_2^2)  \\
\end{bmatrix}} dm
\]

Observe that this can also be written in a more typical tensor notation

\[
I_{ij} 
= \emph{I}(\Be_i I) \cdot (\Be_j I)^\dagger 
= \int (\delta_{ij} \Bx^2 - x_i x_j) dm
\]

Where, as usual for tensors, the meaning of the indexes and whether summation is required is implied.  In this case
the coordinate transformation matrix for this linear transformation has components $I_{ij}$ (and no summation).

\subsection{ coordinate transformation matrix for a couple other linear transformations }

Seeing a function of a bivector for the first time is kind of intriging.  We can form the matrix of such a linear transformation
from a basis of the bivector space to the space spanned by function.  For fun, let's calculate that matrix for the basis $b$ above
for the following function:

\[
f(B) = \Be_1 \wedge (\Be_2 \cdot B)
\]

For this function operating on \R{3} bivectors we have:

\begin{align*}
f(\Be_{23}) &= \Be_1 \wedge (\Be_2 \cdot \Be_{23}) = -\Be_{31} \\
f(\Be_{31}) &= \Be_1 \wedge (\Be_2 \cdot \Be_{31}) = 0 \\
f(\Be_{12}) &= \Be_1 \wedge (\Be_2 \cdot \Be_{12}) = 0 \\
\end{align*}

So

\[
{\begin{bmatrix}
f
\end{bmatrix}}_b^b
= 
\begin{bmatrix}
-1 & 0 & 0
\end{bmatrix}
\]

For \R{4} one orthonormal basis is

\[
b = \{ 
\Be_{12},
\Be_{13},
\Be_{14},
\Be_{23},
\Be_{24},
\Be_{34}
\}
\]

A basis for the span of $f$ is $b' = \{
\Be_{13},
\Be_{14}
\}$.  Like any other coordinate transformation associated with a linear transformation we can write the matrix of the transformation that
takes a coordinate vector in one basis into a coordinate vector for the basis for the image:

\[
{\begin{bmatrix}
f(x)
\end{bmatrix}}_{b'}
=
{\begin{bmatrix}
f
\end{bmatrix}}_b^{b'}
{\begin{bmatrix}
x
\end{bmatrix}}_{b}
\] 

For this function $f$ and these pair of basis bivectors we have:

\[
{\begin{bmatrix}
f
\end{bmatrix}}_b^{b'}
= 
\begin{bmatrix}
0 & 0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 \\
\end{bmatrix}
\]

\subsection{ Equation 3.126 details. }

This statement from GAFP deserves expansion (or at least an exersize):

\[
A \cdot ( \Bx \wedge (\Bx \cdot B) )
= \langle{A \Bx (\Bx \cdot B)}\rangle
= \langle (A \cdot \Bx) \Bx B \rangle
= B \cdot ( \Bx \wedge (\Bx \cdot A) )
\]

Perhaps this is obvious to the author, but wasn't to me.  To clarify this observe the following product

\[
\Bx ( \Bx \cdot B ) = \Bx \cdot ( \Bx \cdot B ) + \Bx \wedge ( \Bx \cdot B ) 
\]

By writing $B = \Bb \Bc = \Bb \wedge \Bc$ we can show that the dot product part of this product is zero:

\begin{align*}
\Bx \cdot ( \Bx \cdot B ) 
&= \Bx \cdot ( (\Bx \cdot \Bb) \Bc - (\Bx \cdot \Bc) \Bb ) \\
&= (\Bx \cdot \Bc) (\Bx \cdot \Bb) - (\Bx \cdot \Bb) (\Bx \cdot \Bc)) \\
&= 0
\end{align*}

This provides the justification for the wedge product removal in the text, since
one can write

\begin{equation}
\Bx \wedge ( \Bx \cdot B ) = \Bx ( \Bx \cdot B )
\end{equation}\label{eqn:inertia_wedge_to_product}

Although it wasn't stated in the text (\ref{eqn:inertia_wedge_to_product}), can
be used to put this inertia product in a pure dot product form

\begin{align*}
A^\dagger \cdot (\Bx \wedge (\Bx \cdot B) )
&= -\langle {A \Bx (\Bx \cdot B)} \rangle \\
&= -\langle (A \cdot \Bx + A \wedge \Bx)(\Bx \cdot B) \rangle \\
\end{align*}

The trivector vector product has only vector and trivector components
\[
(A \wedge \Bx)(\Bx \cdot B) = \langle{ (A \wedge \Bx)(\Bx \cdot B)}\rangle_1 + \langle{(A \wedge \Bx)(\Bx \cdot B)}\rangle_3
\]

So $\langle{(A \wedge \Bx)(\Bx \cdot B)}\rangle_0 = 0$, and one can write

\begin{align*}
A^\dagger \cdot (\Bx \wedge (\Bx \cdot B) )
&= - (A \cdot \Bx ) \cdot (\Bx \cdot B) \\
&= (\Bx \cdot A) \cdot (\Bx \cdot B) \\
\end{align*}

As pointed out in the text this is symetric.  That can't be more clear than above.

\end{document}               % End of document.
