\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}

%\usepackage{listings}
%\usepackage{txfonts} % for ointctr... (also appears to make "prettier" \int and \sum's)
\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage


\title{ Lorentz force Lagrangian with conjugate momentum. }
\author{Peeter Joot \quad peeter.joot@gmail.com }
\date{ April 15, 2009.  Last Revision: $Date: 2009/04/16 03:43:05 $ }

\begin{document}

\maketitle{}
%\tableofcontents

\section{ Motivation }

The covariant Lorentz force Lagrangian (for metric $+---$)

\begin{align}
\LL &= \inv{2} m v^2 + q A \cdot (v/c)
\end{align}

Can be used to find the Lorentz force equation (here in four vector form)

\begin{align}
m\vdot &= q F \cdot (v/c)
\end{align}

A derivation of this can be found in \cite{PJSrLorentzForce}.

However, in
\cite{pauli2000wm} the Lorentz force equation (in non-covariant form) is
derived as a limiting classical case via calculation of the expectation
value of the Hamiltonian

\begin{align}\label{eqn:interactionConjMomHamiltonian}
H = \inv{2m}\sum_{k=1}^3 \left(p_k - \frac{e}{c} A_k \right)^2 + V
\end{align}

This has a much different looking structure than $\LL$ above, so reconciliation of the two
is justifiable.

\section{ Lorentz force Lagrangian with conjugate momentum. }

Is there a relativistic form for the interaction Lagrangian with a structure similar to \ref{eqn:interactionConjMomHamiltonian}?

Let's try

\begin{align*}
\LL
&= \inv{2 m}\left( m v - \kappa A \right)^2 \\
&= \inv{2 m}\left( m^2 v^2 - 2 m \kappa A \cdot v + \kappa^2 A^2 \right)^2 \\
&= \inv{2 m} \left(
m^2 \xdot^\alpha \xdot_\alpha
- 2 m \kappa A_\alpha \xdot^\alpha
+ \kappa^2 A_\alpha A^\alpha
\right) \\
\end{align*}

where $\kappa$ is to be determined.

For this Lagrangian the Euler-Lagrange calculation for variation of $S = \int d^4 x \LL$ is

\begin{align*}
\PD{x^\mu}{\LL}
&= - \kappa (\partial_\mu A_\alpha) \xdot^\alpha + \inv{m} \kappa^2 (\partial_\mu A_\alpha) A^\alpha \\
\frac{d}{d\tau} \PD{\xdot^\mu}{\LL} 
&= \frac{d}{d\tau} \left(m \xdot_\mu - \kappa A_\mu \right)
\end{align*}

Assembling and shuffling we have
\begin{align*}
m \xddot_\mu
&= \kappa (\partial_\alpha A_\mu) \xdot^\alpha
- \kappa (\partial_\mu A_\alpha) \xdot^\alpha + \inv{m} \kappa^2 (\partial_\mu A_\alpha) A^\alpha \\
&= \kappa (\partial_\alpha A_\mu - \partial_\mu A_\alpha) \xdot^\alpha + \inv{m} \kappa^2 (\partial_\mu A_\alpha) A^\alpha \\
&= \kappa F_{\alpha\mu} \xdot^\alpha + \inv{m} \kappa^2 (\partial_\mu A_\alpha) A^\alpha \\
\end{align*}

Comparing to the Lorentz force equation (again for a $+---$ metric)

\begin{align*}
m v_\mu = \frac{q}{c} F_{\mu\nu} v^\nu
\end{align*}

We see that we need $\kappa = -q/c$, but we have an extra factor that doesn't look familiar.  In vector form this Lagrangian would give us the
equation of motion

\begin{align*}
m v = \frac{q}{c} F \cdot v + \frac{q^2}{m c^2} (\grad A_\mu) A^\mu
\end{align*}

Assuming that this extra term has no place in the Lorentz force equation we need to adjust the original Lagrangian as follows

\begin{align}
\LL
&= \inv{2 m}\left( m v + \frac{q}{c} A \right)^2 - \frac{q^2}{ 2 m c^2} A^2
\end{align}

Expressing this energy density in terms of the canonical momentum is 
somewhat interesting.  It provides some extra structure, allowing for a loose identification of the two terms as

\begin{align*}
\LL &= K - V
\end{align*}

(ie: $K = p^2/2m$, where $p$ is the sum of the (proper) mechanical momentum and electromagnetic momentum).

However, that said, observe that expanding the square gives

\begin{align*}
\LL &= \inv{2} m v^2 + \frac{q}{c} A \cdot v 
\end{align*}

which is exactly the original Lorentz force Lagrangian, so in the end this works out to only differ from the original cosmetically.

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
