\documentclass[12pt,leqno]{book}

\usepackage{amsmath,amssymb,amsfonts} % Typical maths resource packages
\usepackage{graphicx}
\usepackage{color}                   % For creating coloured text and background
\usepackage{txfonts} 
\usepackage{listings}

\usepackage[
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% I like the colour instead of boxes better.  Keep this one:
% 
colorlinks=true,
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
plainpages=false,
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
bookmarks=true,
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% this is unfriendly, and generates viewer popup:
%
%pdfpagemode=FullScreen,
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% More screen real estate by default:
%
pdftoolbar=false,
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Avoid control-2 in acrobat?
%
pdffitwindow=true,
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Blech:
%
%pdfpagelayout=TwoColumnLeft,
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Try this:
%
%pdfwindowui=false,
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% NOTE: backref takes multiple build passes to take effect:
%
backref 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
]{hyperref}


\parindent 1cm
\parskip 0.2cm
\topmargin 0.2cm
\oddsidemargin 1cm
\evensidemargin 0.5cm
\textwidth 15cm
\textheight 21cm

\input{../peeters_macros.tex}

\newcommand{\Dslash}[0]{ \not\!D }

%-----------------------------------------
%
% stubs for article class.
%
\newcommand{\blogpage}[1]{}
\newcommand{\email}[1]{}
\newcommand{\beginArtWithToc}[0]{}
\newcommand{\beginArtNoToc}[0]{}
\newcommand{\EndArticle}[0]{}
\newcommand{\EndNoBibArticle}[0]{}
\newcommand{\revisionInfo}[1]{}
\newcommand{\chapcite}[1]{\ref{chap:#1}}
%-----------------------------------------

%\makeindex

\date{ Last $Revision: 1.73 $ }
\begin{document}
\pagenumbering{alph}

\title{Exploring physics with Geometric Algebra.}
\author{Peeter Joot  \quad peeter.joot@gmail.com \\
{\small\em \copyright \  Draft date \today }}

% FIXME: this is something that ought to be configurable with \author
\hypersetup{%
pdfauthor={Peeter Joot, peeter.joot@gmail.com},
pdftitle={Exploring physics with Geometric Algebra.},
pdfsubject={Geometric Algebra, Physics},
pdfkeywords={Clifford Algebra, Geometric Algebra, Physics, Mechanics, Electrodynamics, Quantum Theory}
}



\maketitle

\clearpage\pagenumbering{roman}
\addcontentsline{toc}{chapter}{Contents}
\tableofcontents
\listoffigures
\listoftables

\clearpage\pagenumbering{arabic}

\pagestyle{plain}

\include{preface}

%-------------------------------------------------------

\part{Basics and Geometry.}
\include{introGa}
\include{gaGradeDotWedge}
\include{gaWiki}
\include{gaWikiCramers}
\include{gaWikiTorque}
\include{gaWikiUnitDerivative}
\include{radialVectorDerivatives}
\include{angularVelocity}
\include{bivector}
\include{trivector}
\include{scalarCommutes}
\include{bladegradereduction}
\include{plane}
\include{quaternion}
\include{cauchyGradient}
\include{legendre}
\include{levi}
\include{nfcmCh2}
\include{outermorphismDet}
\part{Projection.}
\include{reciprocalFrame}
\include{projectionWithMatrixComparison}
\include{obliqueProj}
\include{projectionAndMoorePenroseVectorInverse}
\include{angleBetweenLineAndPlane}
\include{orthodecomp}
\include{matrixOfLinearTx}
\part{Rotation.}
\include{rotor}
\include{eulerangle}
\include{sphericalPolar}
\include{slerp}
\include{kvectorExponential}
\include{rotationGenerator}
\include{sphericalPolarUnit}
\part{Calculus.}
\include{multivectorTaylors}
\include{gradientAndForms}
\include{vectorIntegralRelations}
\include{stokesRevisited}
\include{nvolume}
\include{vectorDifferentialIdentities}
\include{stokesGradeTwo}
\include{stokesNoTensor}
\part{General Physics.}
\include{angularAcc}
\include{angularAccCross}
\include{keRotation}
\include{radial}
\include{tensor}
\include{inertialTensor}
\include{locateSatellite}
\include{laplace}
\include{complex}
\part{Relativity.}
\include{lorentz}
\include{massVaryLagrangian}
\include{velocityTx}
\include{fourvecDotinvariance}
\include{spacetimegrad}
\include{gem}
\include{frequencyTx}
\include{poincareTx}
\part{Electrodynamics.}
\include{maxwellsGa}
\include{gaMaxwell}
\include{macroscopicMaxwell}
\include{waveEqn}
\include{gaussianSurface}
\include{emWave}
\include{sgMx41}
\include{chargeArcElement}
\include{chargeLineElement}
\include{biotSavart}
\include{vectorMaxwellsProjection}
\include{stokesMaxwellApplication}
\include{maxwellToTensor}
\include{emPotential}
\include{emBivectorMetricDependencies}
\include{emVacWave}
\include{maxwellVacuum}
\include{transverseField}
\include{transverseWave}
\include{covariantMedia}
\include{jackson12Dash1Gauge}
\part{Lorentz Force.}
\include{lorentzForceTx}
\include{gafpLorentz}
\include{lorentzRotation}
\include{electronRotor}
\include{radiationGeometry}
\include{nuclearInteraction}
\part{Electrodynamics Stress Energy.}
\include{poynting}
\include{poyntingRate}
\include{electricFieldEnergy}
\include{energyMomentumTensor}
\include{stressEnergyLorentz}
\include{enMTensor}
\include{dcPower}
\include{rayleighJeans}
\part{Lagrangian Topics}
\include{newtonianLagrangianAndGradient}  % Aug 9/08
\include{srLagrangian}                      % Aug 21/08
\include{canonicalMomentum}                 % Sep 1/08
\include{lagrangianFieldDensity}           % Sep 8/08
\include{lorentzForce}                      % Oct 8/08
\include{fieldLagrangian}                   % Oct 10/08
\include{maxwellTensorLagrangian}          % Oct 12/08
\include{eulerLagrange}                     % Oct 13/08
\include{boostMaxwellLagrangian}           % Oct 19/08
\include{noethersLorentzForce}             % Oct 22/08
\include{noethersField}                     % Oct 29/08
\include{lorentzForcePQA}                 % Apr 15/09
\include{maxwellTensorFromLagrangian}     % Apr 20/09
\include{stressEnergyNoethers}
\include{lForceLag2}
\include{rotationCurrents}
\include{jackson12Dash9}
\include{sPolarMultiPendulum}
\include{multiPendulumSpherical2}
\part{Quantum Mechanics.}
\include{bohr}
\include{schCurrent}
\include{diracLagrangian}
\include{pauliMatrix}
\include{gamma}
\include{qmAngularMom}
\include{L1Associated}
\include{bivectorSelect}
\part{Fourier treatments}
\include{heatFourier}
\include{poisson}
\include{waveFourier}
\include{fourierMaxwell}
\include{firstorderFourierMaxwell}
\include{4dFourier}
\include{fourierSeriesMaxwell}
\include{planewave}
\include{potentialFourier}
\part{Appendix.}
\include{notationTable}
\include{fourierTx}
\include{fourierNotation}
\include{projGeneralizedDotProd}
\include{tongMf1}
\part{Cronology}
\include{cronology}
\part{Bibliography and Readings.}
\include{gabookmark}

% END INCLUDES.
%-------------------------------------------------------

% from the template:

%\begin{thebibliography}{99}
%  \addcontentsline{toc}{chapter}{Bibliography}
%\bibitem{lamport} L. Lamport. {\bf \LaTeX \ A Document Preparation System}
%Addison-Wesley, California 1986.
%
%\end{thebibliography}

%\bibliographystyle{plainnat}
\bibliographystyle{unsrturl}
  \addcontentsline{toc}{chapter}{Bibliography}
\bibliography{myrefs}

%Note the tag used to make an index entry. You may need to consult Lamport's
%book~\cite{lamport} for details of the procedure to make the index input
%file; \LaTeX \ will create a pre-index by listing all the tagged
%items in the file {\tt bookex.idx} then you edit this into
%a {\tt theindex} environment, as {\tt index.tex}.

%\include{index}
%  \addcontentsline{toc}{chapter}{Index}

\end{document}
