\documentclass{article}      % Specifies the document class

\input{../peeters_macros.tex}
\newcommand{\LL}[0]{\mathcal{L}}
\newcommand{\grad}[0]{\nabla}
\newcommand{\PD}[2]{\frac{\partial {#2}}{\partial {#1}}}
\newcommand{\xdot}[0]{\dot{x}}
\newcommand{\vdot}[0]{\dot{v}}
\newcommand{\mdot}[0]{\dot{m}}
\newcommand{\xddot}[0]{\ddot{x}}

%
% The real thing:
%

                             % The preamble begins here.
\title{ Equations of motion given mass variation with spacetime position. } % Declares the document's title.
\author{Peeter Joot}         % Declares the author's name.
\date{August 28, 2008}        % Deleting this command produces today's date.

\begin{document}             % End of preamble and beginning of text.

\maketitle{}

\section{}

Let
\begin{align*}
x &= \sum \gamma_{\mu} {x}^{\mu} \\
v &= \frac{dx}{d\tau} = \sum \gamma_{\mu} \xdot^{\mu}
\end{align*}

Where whatever spacetime basis you pick has a corresponding reciprocal frame defined implicitly by:

\begin{equation*}
\gamma^{\mu} \cdot \gamma_{\nu} = {\delta^{\mu}}_{\nu}
\end{equation*}

You could for example pick these so that these are othornormal with:

\begin{align*}
\gamma_{i}^2 &= \gamma_i \cdot \gamma_i = -1 \\
\gamma^{i} &= -\gamma_{i} \\
\gamma^{0} &= \gamma_{0} \\
\gamma_{0}^2 &= 1 \\
\gamma_{i} \cdot \gamma_0 &= 0
\end{align*}

ie: the frame vectors define the metric tensor implicitly:

\begin{equation*}
g_{\mu\nu} = \gamma_{\mu} \cdot \gamma_{\nu} =
\begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & -1 & 0 & 0 \\
0 & 0 & -1 & 0 \\
0 & 0 & 0 & -1 \\
\end{bmatrix}
\end{equation*}

Now, my assumption is that given a Lagrangian of the form:

\begin{equation*}
\LL = \inv{2} m v^2 - \phi
\end{equation*}

That the equations of motion follow by computation of:

\begin{equation*}
\PD{x^{\mu}}{\LL} = \frac{d}{d\tau} \PD{\xdot^{\mu}}{\LL}
\end{equation*}

I don't have any proof of this (I don't yet know any calculus of variations, and this is a guess based on intuition).  It does however work out to get the covariant form of the Lorentz force law, so I think it is right.

To get the EOM we need the squared proper velocity.  This is just $c^2$.  Example: for an orthonormal spacetime frame one has:

\begin{align*}
v^2 &= 
\left(\gamma^0 c dt/d\tau + \sum \gamma_i dx/d\tau\right)^2  \\
&= \gamma \left(\gamma_0 c + \sum \gamma_i dx/dt\right)^2 \\
&= \gamma^2 \left(c^2 - \Bv^2\right) = c^2
\end{align*}

but if we leave this expressed in terms of coordinates (also don't have to assume the diagonal metric tensor, since we can use non-orthonormal basis vectors if desired) we have:

\begin{align*}
v^2 
&= \left(\sum \gamma_{\mu} \xdot^{\mu}\right) \cdot \left(\sum \gamma_{\nu} \xdot^{\nu}\right) \\
&= \sum \gamma_{\mu} \cdot \gamma_{\nu} \xdot^{\mu} \xdot^{\nu} \\
&= \sum g_{\mu\nu} \xdot^{\mu} \xdot^{\nu}
\end{align*}

Therefore the Lagrangian to minimize is:

\begin{equation*}
\LL = \inv{2} m \sum g_{\mu\nu} \xdot^{\mu} \xdot^{\nu} - \phi.
\end{equation*}

Performing the calculations for the EOM, and in this case, also allowing mass to be a function of space or time position ($m = m(x^{\mu})$)

\begin{align*}
\PD{x^{\mu}}{\LL} &= \frac{d}{d\tau} \PD{\xdot^{\mu}}{\LL} \\
- \PD{x^{\mu}}{\phi} + \inv{2} \PD{x^{\mu}}{m} \sum g_{\alpha\beta} \xdot^{\alpha} \xdot^{\beta} &= \\
- \PD{x^{\mu}}{\phi} + \inv{2} \PD{x^{\mu}}{m} v^2 &= \\
%- \PD{x^{\mu}}{\phi} + \inv{2} \PD{x^{\mu}}{m} c^2 &= \\
&= \inv{2} \frac{d}{d\tau} m \sum g_{\alpha\beta} \PD{x^{\mu}}{}\left(\xdot^{\alpha} \xdot^{\beta}\right) \\
&= \inv{2} \frac{d}{d\tau} m \sum g_{\alpha\beta} \left(\delta^{\mu\alpha} \xdot^{\beta} + \xdot^{\alpha} \delta^{\mu\beta}\right) \\
&= \frac{d}{d\tau} m \sum g_{\alpha\mu} \xdot^{\alpha} \\
&= \sum \PD{x^{\beta}}{m} \xdot^{\beta} g_{\alpha\mu} \xdot^{\alpha} + m g_{\alpha\mu} \xddot^{\alpha} \\
\end{align*}

Now, the metric tensor values can be removed by summing since they can be used to switch upper and lower indexes of the frame vectors:

\begin{align*}
\gamma_{\mu} &= \sum a^{\nu} \gamma^{\nu} \\
\gamma_{\mu} \cdot \gamma_{\beta}
&= \sum a^{\nu} \gamma^{\nu} \cdot \gamma_{\beta} \\
&= \sum a^{\nu} {\delta^{\nu}}_{\beta} \\
&= a^{\beta} \\
\implies \\
\gamma_{\mu}
&= \sum \gamma_{\mu} \cdot \gamma_{\nu} \gamma^{\nu} \\
&= \sum g_{\mu\nu} \gamma^{\nu} \\
\end{align*}

If you are already familiar with tensors then this may be obvious to you (but wasn't to me with only vector background).

Multiplying throughout by $\gamma^{\mu}$, and summing over $\mu$ one has:

\begin{align*}
\sum \gamma^{\mu} \left( - \PD{x^{\mu}}{\phi} + \inv{2} \PD{x^{\mu}}{m} v^2 \right) 
&= \sum \gamma^{\mu} \left(\PD{x^{\beta}}{m} \xdot^{\beta} g_{\alpha\mu} \xdot^{\alpha} + m g_{\alpha\mu} \xddot^{\alpha} \right) \\
- \left(\sum \gamma^{\mu} \PD{x^{\mu}}{}\right) \phi + \inv{2} v^2 \left(\sum \gamma^{\mu} \PD{x^{\mu}}{}\right) m &= \\
&= \sum \PD{x^{\beta}}{m} \xdot^{\beta} \gamma^{\mu} \gamma_{\alpha} \cdot \gamma_{\mu} \xdot^{\alpha} + m \gamma^{\mu} \gamma_{\alpha} \cdot \gamma_{\mu} \xddot^{\alpha}  \\
&= \sum \PD{x^{\beta}}{m} \xdot^{\beta} \gamma_{\alpha} \xdot^{\alpha} + m \gamma_{\alpha} \xddot^{\alpha}  \\
\end{align*}

Writing:
\begin{equation*}
\nabla = \sum \gamma^{\mu} \frac{\partial}{\partial x^{\mu}}
\end{equation*}

This is:
\begin{equation*}
- \grad \phi + \inv{2} v^2 \grad m = v \sum \PD{x^{\beta}}{m} \xdot^{\beta} + m \vdot 
\end{equation*}

However, 
\begin{align*}
(\grad m) \cdot v 
&= 
\left(\sum \gamma^{\mu} \PD{x^{\mu}}{m}\right) \cdot \left( \sum \gamma_{\nu} \xdot^{\nu} \right) \\
&= \sum \gamma^{\mu} \cdot \gamma_{\nu} \PD{x^{\mu}}{m} \xdot^{\nu} \\
&= \sum {\delta^{\mu}}_{\nu} \PD{x^{\mu}}{m} \xdot^{\nu} \\
&= \sum \PD{x^{\mu}}{m} \xdot^{\mu} = \frac{dm}{d\tau}
\end{align*}

That allows for expressing the EOM in strict vector form:
\begin{equation*}
- \grad \phi + \inv{2} v^2 \grad m = v \grad m \cdot v + m \vdot.
\end{equation*}

However, there is still an asymmetry here, as one would expect a $\mdot v$ term.  Regrouping slightly, and using some algebraic vector
manipulation we have:

\begin{align*}
m \vdot + v \grad m \cdot v - \inv{2} v^2 \grad m &= - \grad \phi \\
m \vdot + \inv{2} v ( \underbrace{2 \grad m \cdot v - v \grad m}_{2 a \cdot b - b a = a b}) &= \\
m \vdot + \inv{2} v (\grad m) v &= \\
m \vdot + \inv{2} (v \grad m) v &= \\
m \vdot + \inv{2} (2 v \cdot \grad m - \grad m v) v &= \\
m \vdot + (v \cdot \grad m) v - \inv{2} (\grad m v) v &= \\
m \vdot + \mdot v - \inv{2} \grad m (v v) &= \\
\implies \\
\frac{d (m v)}{d\tau} = m \vdot + \mdot v
&= \inv{2} \grad m c^2 -\grad \phi \\
&= -\grad \left(\phi - \inv{2} m c^2 \right) \\
&= -\grad \left(\phi - \inv{2} m v^2 \right) \\
\end{align*}

So, after a whole wack of algebra, the end result is to show the proper time varient of the Lagrangian equations imply that our 
proper force can be expressed as a (spacetime) gradient.

The cavaet is that if the mass is allowed to vary, it also needs to be
included in the generalized potential associated with the equation of motion.

\subsection{ Summarizing. }

We took this Lagrangian with kinetic energy and non-velocity dependent potential terms, where the
mass in the kinetic energy term is allowed to vary with position or time.  That plus the
presumed proper-time Lagrange equations:

\begin{align*}
\LL &= \inv{2} m v^2 - \phi \\
\PD{x^{\mu}}{\LL} &= \frac{d}{d\tau} \PD{\xdot^{\mu}}{\LL},
\end{align*}

when followed to their algebraic conclusion together imply that the equation of motion is:

\begin{equation*}
\frac{d (m v)}{d\tau} = \grad \LL,
\end{equation*}

\end{document}               % End of document.
