%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

% 
% 
\chapter{Tensor derivation of Maxwell equation (non-dual part) from Lagrangian}
\label{chap:maxwellTensorFromLagrangian}
\date{April 20, 2009.  maxwellTensorFromLagrangian.tex}

\section{Motivation. }

Looking through my notes for a purely tensor derivation of Maxwell's equation, and not finding one.  Have done this on
paper a number of times, but writing it up once for reference to refer to for signs will be useful.

\section{Lagrangian. }

Notes containing derivations of Maxwell's equation

\begin{align}\label{eqn:maxTenLag:maxwell}
\grad F = J/\epsilon_0 c
\end{align}

From the Lagrangian

\begin{align}\label{eqn:maxTenLag:maxlag}
\LL &= -\frac{\epsilon_0}{2} (\grad \wedge A)^2 + \frac{J}{c} \cdot A
\end{align}

can be found in \chapcite{PJFieldLagrangian}, and the earlier \chapcite{PJMaxwellLagrangian}.

We will work from the scalar part of this Lagrangian, expressed strictly in tensor form

\begin{align}
\LL &= \frac{\epsilon_0}{4} F_{\mu\nu}F^{\mu\nu} + \inv{c} J_\mu \cdot A^\mu
\end{align}

\section{Calculation. }

\subsection{Preparation. }

In preparation, an expansion of the Faraday tensor in terms of potentials is desirable

\begin{align*}
F_{\mu\nu}F^{\mu\nu}
&=
(\partial_\mu A_\nu - \partial_\nu A_\mu) (\partial^\mu A^\nu - \partial^\nu A^\mu) \\
&=
\partial_\mu A_\nu \partial^\mu A^\nu 
-\partial_\mu A_\nu \partial^\nu A^\mu
- \partial_\nu A_\mu \partial^\mu A^\nu 
+ \partial_\nu A_\mu \partial^\nu A^\mu \\
&=
2 (\partial_\mu A_\nu \partial^\mu A^\nu - \partial_\mu A_\nu \partial^\nu A^\mu)
\\
\end{align*}

So we have
\begin{align*}
\LL &= \frac{\epsilon_0}{2} \partial_\mu A_\nu (\partial^\mu A^\nu - \partial^\nu A^\mu)
+ \inv{c} J_\mu \cdot A^\mu
\end{align*}

\subsection{Derivatives. }

We want to compute

\begin{align}
\PD{A_\alpha}{\LL} = \sum \partial_\beta \PD{(\partial_\beta A_\alpha)}{\LL}
\end{align}

Starting with the LHS we have

\begin{align*}
\PD{A_\alpha}{\LL} = \inv{c} J^\alpha 
\end{align*}

and for the RHS

\begin{align*}
\PD{(\partial_\beta A_\alpha)}{\LL}
&=
\frac{\epsilon_0}{2} 
\PD{(\partial_\beta A_\alpha)}{}
\partial_\mu A_\nu (\partial^\mu A^\nu - \partial^\nu A^\mu) \\
&=
\frac{\epsilon_0}{2} 
\left(
F^{\beta\alpha}
+
\partial^\mu A^\nu 
\PD{(\partial_\beta A_\alpha)}{}
(\partial_\mu A_\nu - \partial_\nu A_\mu) 
\right) \\
&=
\frac{\epsilon_0}{2} 
\left(
F^{\beta\alpha}
+ \partial^\beta A^\alpha 
-\partial^\alpha A^\beta 
\right) \\
&=
{\epsilon_0} F^{\beta\alpha} \\
\end{align*}

Taking the $\beta$ derivatives and combining the results for the LHS and RHS this is

\begin{align}\label{eqn:maxTenLag:maxwellTensor}
\partial_\beta F^{\beta\alpha} = \inv{\epsilon_0 c} J^\alpha 
\end{align}

\subsection{Compare to STA form. }

To verify that no sign errors have been made during the index manipulations above, this result should also match
the STA Maxwell equation of \ref{eqn:maxTenLag:maxwell}, the vector part of which is

\begin{align*}
\grad \cdot F = J/\epsilon_0 c
\end{align*}

Dotting the LHS with $\gamma^\alpha$ we have

\begin{align*}
(\grad \cdot F) \cdot \gamma^\alpha 
&=
((\gamma^\mu \partial_\mu) \cdot (\inv{2} F^{\beta\sigma} (\gamma^\beta \wedge \gamma_\sigma))) \cdot \gamma^\alpha \\
&=
\inv{2} \partial_\mu F^{\beta\sigma} 
(\gamma^\mu \cdot (\gamma_\beta \wedge \gamma_\sigma)) \cdot \gamma^\alpha \\
&=
\inv{2} 
\left( \partial_\beta F^{\beta\sigma} \gamma_\sigma -\partial_\sigma F^{\beta\sigma} \gamma_\beta \right) \cdot \gamma^\alpha \\
&=
\inv{2} 
\left( \partial_\beta F^{\beta\alpha} -\partial_\sigma F^{\alpha\sigma} \right) \\
&=
\partial_\beta F^{\beta\alpha} 
\end{align*}

This gives us

\begin{align}
\partial_\beta F^{\beta\alpha} = J^\alpha/\epsilon_0 c
\end{align}

In agreement with \ref{eqn:maxTenLag:maxwellTensor}.
