\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}

\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage



\title{ Some NFCM exersize solutions and notes. }
\author{Peeter Joot}
\date{ Nov 27, 2008.  Last Revision: $Date: 2008/11/28 03:02:16 $ }

\begin{document}

\maketitle{}

%\tableofcontents
\section{ Chapter 2 }

I recall that some of the problems from this chapter of
\cite{hestenes1999nfc}
were fairly tricky.  Did I end up doing them all?  I intended to
revisit these and make sure I understood it all.  As I do so, write up
solutions, starting with $1.3$, a question on the Geometric Algebra group.

Another thing I recall from the text is that I was fairly confused about
all the mass of identities by the time I got through it, and it wasn't clear
to me which were the fundamental ones.
Eventually I figured out that it is
really grade selection that is the fundamental operation, and the
definitions of the generalized dot

\begin{align*}
A_r \cdot B_s = \gpgrade{A B}{{\Abs{r -s}}}
\end{align*}

and wedge product

\begin{align*}
A_r \wedge B_s = \gpgrade{A B}{r + s}
\end{align*}

and most everything else
follows from that.
I also found it was really easy to go around in circles attempting the various proofs, and intended to revisit all of these
and prove them all for myself making sure I didn't invoke any circular arguments and used only things already proven.

\subsection{ Exersize 1.3 }

Solve for $x$

\begin{align*}
\alpha x + a x \cdot b = c
\end{align*}

where $\alpha$ is a scalar and all the rest are vectors.

\subsubsection{ Solution. }

Can dot or wedge the entire equation with the constant vectors.  In particular

\begin{align*}
c \cdot b &= (\alpha x + a x \cdot b) \cdot b \\ 
&= (\alpha + a \cdot b) x \cdot b 
\end{align*}
\begin{align*}
\implies
x \cdot b &= \frac{c \cdot b}{\alpha + a \cdot b} \\
\end{align*}

and
\begin{align*}
c \wedge a &= (\alpha x + a x \cdot b) \wedge a \\ 
&= \alpha (x \wedge a) + \underbrace{(a \wedge a)}_{=0} (x \cdot b) \wedge a \\
\end{align*}
\begin{align*}
\implies
x \wedge a &= \inv{\alpha} (c \wedge a) \\
\end{align*}

This last can be reduced by dotting with $b$, and then substitute the 
result for $x \cdot b$ from above

\begin{align*}
(x \wedge a) \cdot b
&= x (a \cdot b) - (x \cdot b) a \\
&= x (a \cdot b) - \frac{c \cdot b}{\alpha + a \cdot b} a \\
\end{align*}

Thus the final solution is

\begin{align*}
x = \inv{a \cdot b}\left(
\frac{c \cdot b}{\alpha + a \cdot b} a 
+ \inv{\alpha} (c \wedge a) \cdot b
\right)
\end{align*}

Question: was there a geometric or physical motivation for this question.  I can't recall one?

\section{ Sequential proofs of required identities. }

\subsection{ bivector dot with vector reduction. }

In the $1.3$ solution above the identity

\begin{align}
(a \wedge b) \cdot c &= a (b \cdot c) - (a \cdot c) b \\
\end{align}

was used.  Let's prove this, taking the vector dot and wedge identities
as given (FIXME: should do that first for completeness).

\begin{align*}
(a \wedge b) \cdot c
&= \gpgradeone{(a \wedge b) c} \\
\implies \\
2 (a \wedge b) \cdot c
&= \gpgradeone{a b c - b a c} \\
&= \gpgradeone{a b c - b (- c a + 2 a \cdot c )} \\
&= \gpgradeone{a b c + b c a} - 2 b (a \cdot c ) \\
&= \gpgradeone{a (b \cdot c + b \wedge c) + (b \cdot c + b \wedge c) a} - 2 b (a \cdot c ) \\
&= 2 a (b \cdot c) + a \cdot (b \wedge c) + (b \wedge c) \cdot a - 2 b (a \cdot c ) \\
\end{align*}

To complete the proof we need $a \cdot B = -B \cdot a$ (FIXME: do this one).  Given that one is left with the desired identity after dividing through
by $2$.

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
