\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}
\input{../peeters_macros3.tex}


%\usepackage{listings}
%\usepackage{txfonts} % for ointctr... (also appears to make "prettier" \int and \sum's)
% makes \grad look funny though (almost like spacegrad, but narrower)
\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage

\title{Poincare transformation symmetries}
\author{Peeter Joot \quad peeter.joot@gmail.com }
\date{ June 1, 2009.  $RCSfile: poincareTx.tex,v $ Last $Revision: 1.1 $ $Date: 2009/07/01 14:33:59 $ }

\begin{document}

\maketitle{}
\tableofcontents
\section{Motivation}

In \cite{montesinos2006sem} a Poincare transformation is used to 
develop the symmetric stress energy tensor directly, in contrast to the
non-symmetric canonical stress energy tensor that results from 
spacetime translation.

Attempt to decode some bits of this article.

\section{Guts}

Equation (11) in the article, is labelled an infinitesimal Poincare
transformation

\begin{align}\label{eqn:txComponents}
{x'}^\mu
&=
{x'}^\mu
+ {{\epsilon}^\mu}_\nu x^\nu
+ {\epsilon}^\mu
\end{align}

It is stated that an antisymmetrization conditon $\epsilon_{\mu\nu} = \epsilon_{\nu\mu}$.  This is somewhat confusing 
since the infinitesimal transformation is given by a mixed upper and lower index tensor.   Due to the antisymmetry
perhaps this all a coordinate statement of the following vector to vector linear transformation

\begin{align}\label{eqn:blah}
x' = x + \epsilon + A \cdot x
\end{align}

This transformation is less restricted than a plain old spacetime transformation, as it also contains a projective term, where $x$ is projected onto the spacetime (or spatial) plane $A$ (a bivector), plus a rotation in that plane.

Writing as usual

\begin{align*}
x = \gamma_\mu x^\mu
\end{align*}

So that components are recovered by taking dot products, as in
\begin{align*}
x^\mu = x \cdot \gamma^\mu
\end{align*}

For the bivector term, write

\begin{align*}
A = c \wedge d = c^\alpha d^\beta (\gamma_\alpha \wedge \gamma_\beta)
\end{align*}

For
\begin{align*}
(A \cdot x ) \cdot \gamma^\mu 
&=
c^\alpha d^\beta x_\sigma ((\gamma_\alpha \wedge \gamma_\beta) \cdot \gamma^\sigma) \cdot \gamma^\mu \\
&=
c^\alpha d^\beta x_\sigma ( {\delta_\alpha}^\mu {\delta_\beta}^\sigma -{\delta_\beta}^\mu {\delta_\alpha}^\sigma ) \\
&=
(c^\mu d^\sigma -c^\sigma d^\mu ) x_\sigma 
\end{align*}

This allows for an identification $\epsilon^\mu\sigma = c^\mu d^\sigma -c^\sigma d^\mu$ which is antisymmetric as required.
With that identification we can write (\ref{eqn:txComponents}) via the equivalent vector relation (\ref{eqn:blah}) if
we write

\begin{align*}
{\epsilon^\mu}_\sigma x^\sigma = (c^\mu d_\sigma -c_\sigma d^\mu ) x^\sigma 
\end{align*}

Where ${\epsilon^\mu}_\sigma$ is defined implicitly in terms of components of the bivector $A = c \wedge d$.

Is this what a Poincare transformation is?  The \href{http://mathworld.wolfram.com/PoincareTransformation.html}{Poincare Transformation} article suggests not.  This article suggests that the Poincare transformation is a spacetime translation plus
a Lorentz tranformation (composition of boosts and rotations).  That Lorentz transformation will not be antisymmetric
however, so how can these be reconciled?  The key is probably the fact that this was an infinistesimal Poincare transformation
so lets consider a Taylor expansion of the Lorentz boost or rotation rotor.

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
