\input{../peeter_prologue.tex}

\chapter{Bivector form of quantum angular momentum operator}
\label{chap:qmAngularMom}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2009/qmAngularMom.pdf}
%\date{July 27, 2009}
%\revisionInfo{$RCSfile: qmAngularMom.tex,v $ Last $Revision: 1.6 $ $Date: 2009/08/26 12:14:43 $}

\date{July 27, 2009.  $RCSfile: qmAngularMom.tex,v $ Last $Revision: 1.6 $ $Date: 2009/08/26 12:14:43 $}

\beginArtWithToc

\section{Spatial bivector representation of the angular momentum operator.}

Reading (\cite{bohm1989qt}) on the angular momentum operator, the form of the operator is suggested by analogy where components of $\Bx \cross \Bp$ with 
the position representation $\Bp \sim -i \hbar \spacegrad$ used to expand the coordinate representation of the operator.

The result is the following coordinate representation of the operator

%\begin{align*}
%L_x &= -i \hbar( y \partial_z - z \partial_y ) \\
%L_y &= -i \hbar( z \partial_x - x \partial_z ) \\
%L_z &= -i \hbar( x \partial_y - y \partial_x ) \\
%\end{align*}
\begin{align*}
L_1 &= -i \hbar( x_2 \partial_3 - x_3 \partial_2 ) \\
L_2 &= -i \hbar( x_3 \partial_1 - x_1 \partial_3 ) \\
L_3 &= -i \hbar( x_1 \partial_2 - x_2 \partial_1 ) \\
\end{align*}

It is interesting to put these in vector form, and then employ the freedom to use for $i = \sigma_1 \sigma_2 \sigma_3$ the spatial pseudoscalar.

\begin{align*}
\BL 
&= 
-\sigma_1 (\sigma_1 \sigma_2 \sigma_3) \hbar( x_2 \partial_3 - x_3 \partial_2 ) 
-\sigma_2 (\sigma_2 \sigma_3 \sigma_1) \hbar( x_3 \partial_1 - x_1 \partial_3 ) 
-\sigma_3 (\sigma_3 \sigma_1 \sigma_2) \hbar( x_1 \partial_2 - x_2 \partial_1 ) \\
&= 
-\sigma_2 \sigma_3 \hbar( x_2 \partial_3 - x_3 \partial_2 ) 
-\sigma_3 \sigma_1 \hbar( x_3 \partial_1 - x_1 \partial_3 ) 
-\sigma_1 \sigma_2 \hbar( x_1 \partial_2 - x_2 \partial_1 ) \\
&=
-\hbar ( \sigma_1 x_1 +\sigma_2 x_2 +\sigma_3 x_3 ) \wedge ( \sigma_1 \partial_1 +\sigma_2 \partial_2 +\sigma_3 \partial_3 ) \\
\end{align*}

The choice to use the pseudoscalar for this imaginary seems a logical one and the end result is a pure bivector representation of angular momentum operator

\begin{align}\label{eqn:qmAngularMom:ang}
\BL &= - \hbar \Bx \wedge \spacegrad
\end{align}

The choice to represent angular momentum as a bivector $\Bx \wedge \Bp$ is also natural in classical mechanics (encoding the orientation of the plane and the magnitude of the momentum in the bivector), although its dual form the axial vector $\Bx \cross \Bp$ is more common, at least in introductory mechanics.  Observe that there is no longer any explicit imaginary in (\ref{eqn:qmAngularMom:ang}), since the bivector itself has an implicit complex structure.

\section{Factoring the gradient and Laplacian.}

The form of (\ref{eqn:qmAngularMom:ang}) suggests a more direct way to extract the angular momentum operator from the Hamiltonian (i.e. from the Laplacian).  Bohm uses the spherical polar representation of the Laplacian as the starting point.  Instead let's project the gradient itself in a specific constant direction $\Ba$, much as we can do to find the polar form angular velocity and acceleration components.

Write 

\begin{align*}
\spacegrad 
&=
\inv{\Ba} \Ba \spacegrad \\
&=
\inv{\Ba} (\Ba \cdot \spacegrad + \Ba \wedge \spacegrad) \\
\end{align*}

Or
\begin{align*}
\spacegrad 
&=
\spacegrad \Ba \inv{\Ba} \\
&=
(\spacegrad \cdot \Ba + \spacegrad \wedge \Ba) \inv{\Ba} \\
&=
(\Ba \cdot \spacegrad - \Ba \wedge \spacegrad) \inv{\Ba} \\
\end{align*}

The Laplacian is therefore

\begin{align*}
\spacegrad^2 
&=
\gpgradezero{ \spacegrad^2 } \\
&=
\gpgradezero{ 
(\Ba \cdot \spacegrad - \Ba \wedge \spacegrad) \inv{\Ba} \inv{\Ba} (\Ba \cdot \spacegrad + \Ba \wedge \spacegrad) 
} \\
&=
\inv{\Ba^2} \gpgradezero{ 
(\Ba \cdot \spacegrad - \Ba \wedge \spacegrad) (\Ba \cdot \spacegrad + \Ba \wedge \spacegrad) 
} \\
&=
\inv{\Ba^2} ((\Ba \cdot \spacegrad)^2 - (\Ba \wedge \spacegrad)^2 ) \\
\end{align*}

So we have for the Laplacian a representation in terms of projection and rejection components 

\begin{align*}
\spacegrad^2
&=
(\acap \cdot \spacegrad)^2 - \inv{\Ba^2} (\Ba \wedge \spacegrad)^2 \\
&=
(\acap \cdot \spacegrad)^2 - (\acap \wedge \spacegrad)^2 \\
\end{align*}

The vector $\Ba$ was arbitrary, and just needed to be constant with respect to the factorization operations.  Setting $\Ba = \Bx$, the radial position from the origin one may guess that we have

\begin{align}\label{eqn:qmAngularMom:wrong}
\spacegrad^2 &= \frac{\partial^2 }{\partial r^2} - \inv{\Bx^2} (\Bx \wedge \spacegrad)^2 
\end{align}

however, with the switch to a non-constant position vector $\Bx$, this cannot possibly be right.

\section{The Coriolis term}

The radial factorization of the gradient relied on the direction vector $\Ba$ being constant.  If we evaluate (\ref{eqn:qmAngularMom:wrong}), then there should be a non-zero remainder compared to the Laplacian.  Evaluation by coordinate expansion is one way to verify this, and should produce the difference.  Let's do this in two parts, starting with $(x \wedge \grad)^2$.  Summation will be implied by mixed indexes, and for generality a general basis and associated reciprocal frame will be used.

\begin{align*}
(x \wedge \grad)^2 f 
&=
((x^\mu \gamma_\mu) \wedge (\gamma_\nu \partial^\nu)) \cdot 
((x_\alpha \gamma^\alpha) \wedge (\gamma^\beta \partial_\beta)) \\
&=
(\gamma_\mu \wedge \gamma_\nu) \cdot (\gamma^\alpha \wedge \gamma^\beta) x^\mu \partial^\nu (x_\alpha \partial_\beta) f \\
&=
({\delta_\mu}^\beta {\delta_\nu}^\alpha -{\delta_\mu}^\alpha {\delta_\nu}^\beta) x^\mu \partial^\nu (x_\alpha \partial_\beta) f \\
&=
x^\mu \partial^\nu ((x_\nu \partial_\mu) - x_\mu \partial_\nu) f \\
&=
x^\mu (\partial^\nu x_\nu) \partial_\mu f - x^\mu (\partial^\nu x_\mu) \partial_\nu f \\
&+x^\mu x_\nu \partial^\nu \partial_\mu f - x^\mu x_\mu \partial^\nu \partial_\nu f \\
&=
(n-1) x \cdot \grad f +x^\mu x_\nu \partial^\nu \partial_\mu f - x^2 \grad^2 f \\
\end{align*}

For the dot product we have
\begin{align*}
(x \cdot \grad)^2 f 
&=
x^\mu \partial_\mu( x^\nu \partial_\nu ) f \\
&=
x^\mu (\partial_\mu x^\nu) \partial_\nu  f + x^\mu x^\nu \partial_\mu \partial_\nu f \\
&=
x^\mu \partial_\mu f + x^\mu x_\nu \partial^\nu \partial_\mu f \\
&=
x \cdot \grad f + x^\mu x_\nu \partial^\nu \partial_\mu f \\
\end{align*}

So, forming the difference we have

\begin{align*}
(x \cdot \grad)^2 f - (x \wedge \grad)^2 f &=
-(n - 2) x \cdot \grad f + x^2 \grad^2 f \\
\end{align*}

Or
\begin{align}
\grad^2 &= \inv{x^2} (x \cdot \grad)^2 - \inv{x^2} (x \wedge \grad)^2 + (n - 2) \inv{x} \cdot \grad 
\end{align}

Going back to the quantum Hamiltonian we do still have the angular momentum operator as one of the distinct factors of the Laplacian.  As operators we have something akin to the projection of the gradient onto the radial direction, as well as terms that project the gradient onto the tangential plane to the sphere at the radial point

\begin{align}
-\frac{\hbar^2}{2m} \spacegrad^2 + V
&=
-\frac{\hbar^2}{2m} \left( \inv{\Bx^2} (\Bx \cdot \spacegrad)^2 - \inv{\Bx^2} (\Bx \wedge \spacegrad)^2 + \inv{\Bx} \cdot \spacegrad \right) + V
\end{align}

\section{Correspondence with explicit radial form.}

We've seen above that we can factor the 3D Laplacian as 

\begin{align}\label{eqn:qmAngularMom:3dLaplacian}
\spacegrad^2 \psi = \inv{\Bx^2}( (\Bx \cdot \spacegrad)^2 + \Bx \cdot \spacegrad - (\Bx \wedge \spacegrad)^2) \psi
\end{align}

Contrast this to the explicit $r,\theta,\phi$ form as given in (Bohm's \cite{bohm1989qt}, 14.2)

\begin{align}\label{eqn:qmAngularMom:LaplacianRTP}
\spacegrad^2 \psi = \inv{r} \frac{\partial^2}{\partial r^2} (r\psi) + \inv{r^2} \left(
\inv{\sin\theta} \partial_\theta \sin\theta \partial_\theta + \inv{\sin^2\theta} + \partial_{\phi \phi} \right) \psi
\end{align}

Let's expand out the non-angular momentum operator terms explicitly as a partial verification of this factorization.  The radial term in Bohm's Laplacian formula expands out to

\begin{align*}
\inv{r} \frac{\partial^2}{\partial r^2} (r\psi) 
&=
\inv{r} \partial_r (\partial_r r \psi) \\
&=
\inv{r} \partial_r (\psi + r\partial_r \psi) \\
&=
\inv{r} \partial_r \psi + \inv{r}( \partial_r \psi + r \partial_{rr} \psi) \\
&=
\frac{2}{r} \partial_r \psi + \partial_{rr} \psi \\
\end{align*}

On the other hand, with $\Bx = r\rcap$, what we expect to correspond to the radial term in the vector factorization is

\begin{align*}
\inv{\Bx^2}( (\Bx \cdot \spacegrad)^2 + \Bx \cdot \spacegrad ) \psi
&=
\inv{r^2}( (r \rcap \cdot \spacegrad)^2 + r \rcap \cdot \spacegrad  ) \psi \\
&=
\inv{r^2}( (r \partial_r )^2 + r \partial_r  ) \psi \\
&=
\inv{r^2}( r \partial_r \psi + r^2 \partial_{rr} \psi + r \partial_r \psi ) \\
&=
\frac{2}{r} \partial_r \psi + \partial_{rr} \psi 
\end{align*}

Okay, good.  It's a brute force way to verify things, but it works.  With $\Bx \wedge \spacegrad = I (\Bx \cross \spacegrad)$ we can eliminate the wedge product from the factorization expression (\ref{eqn:qmAngularMom:3dLaplacian}) and express things completely in quantities that can be understood without any resort to Geometric Algebra.  That is

\begin{align}\label{eqn:qmAngularMom:LaplacianRadialCross}
\spacegrad^2 \psi = \inv{r} \frac{\partial^2}{\partial r^2} (r\psi) + \inv{r^2} (\Bx \cross \spacegrad)^2 \psi
\end{align}

Bohm resorts to analogy and an operatorization of $L_c = \epsilon_{abc} (x_a p_b - x_b p_a)$, then later a spherical polar change of coordinates to match exactly the $L^2$ expression with (\ref{eqn:qmAngularMom:LaplacianRTP}).  With the GA formalism we see this a bit more directly, although it is not the least bit obvious that the operator $\Bx \cross \spacegrad$ has no radial dependence.  Without resorting to a comparison with the explicit $r,\theta,\phi$ form that wouldn't be so easy to see.

\section{Raising and Lowering operators in GA form.}

Having seen in (\chapcite{L1Associated}) that we have a natural GA form for the $l=1$ spherical harmonic eigenfunctions $\psi_1^{m}$, and that we have the vector angular momentum operator $\Bx \cross \spacegrad$ showing up directly in a sort-of-radial factorization of the Laplacian, it is natural to wonder what the GA form of the raising and lowering operators are.  At least for the $l=1$ harmonics use of $i = I \Be_3$ (unit bivector for the $x-y$ plane) for the imaginary ended up providing a nice geometric interpretation.

Let's see what that provides for the raising and lowering operators.  First we need to express $L_x$ and $L_y$ in terms of our bivector angular momentum operator.  Let's switch notations and drop the $-i \hbar$ factor from (\ref{eqn:qmAngularMom:ang}) writing just

\begin{align}\label{eqn:qmAngularMom:Ang}
\BL &= \Bx \wedge \spacegrad
\end{align}

We can now write this in terms of components with respect to the basis bivectors $I \Be_k$.  That is

\begin{align}\label{eqn:qmAngularMom:BLprojected}
\BL = \sum_k \left( (\Bx \wedge \spacegrad) \cdot \inv{I \Be_k}\right) I \Be_k
\end{align}

These scalar product results are expected to match the $L_x$, $L_y$, and $L_z$ components at least up to a sign.  Let's check, picking $L_z$ as representative

\begin{align*}
(\Bx \wedge \spacegrad) \cdot \inv{I \Be_3}
&=
(\sigma_m \wedge \sigma^k) \cdot {-\sigma_1 \sigma_2 \sigma_3 \sigma_3} x^m \partial_k \\
&=
(\sigma_m \wedge \sigma^k) \cdot {-\sigma_1 \sigma_2} x^m \partial_k \\
&=
-( x^2 \partial_1 - x^1 \partial_2 )
\end{align*}

With the $-i\hbar$ factors dropped this is $L_z = L_3 = x^1 \partial_2 - x^2 \partial_1$, the projection of $\BL$ onto the $x-y$ plane $I \Be_k$.  So, now how about the raising and lowering operators

\begin{align*}
L_x \pm i L_y
&=
L_x \pm I \Be_3 L_y \\
&=
\BL \cdot \inv{I\Be_1} \pm I \Be_3 \BL \cdot \inv{I\Be_2} \\
&=
-\Be_1 I \left( I \Be_1 \BL \cdot \inv{I\Be_1} \pm I \Be_2 \BL \cdot \inv{I\Be_2} \right) \\
\end{align*}

Or
\begin{align}\label{eqn:qmAngularMom:raisingLoweringBivector}
(I \Be_1) L_x \pm i L_y &= I \Be_1 \BL \cdot \inv{I\Be_1} \pm I \Be_2 \BL \cdot \inv{I\Be_2} 
\end{align}

Compare this to the projective split of $\BL$ (\ref{eqn:qmAngularMom:BLprojected}).  We have projections of the bivector angular momentum operator onto the bivector directions $I\Be_1$ and $I\Be_2$ (really the bivectors for the planes perpendicular to the $\xcap$ and $\ycap$ directions).

We have the Laplacian in explicit vector form and have a clue how to vectorize (really bivectorize) the raising and lowering operators.  We have also seen how to geometrize the first spherical harmonics.  The next logical step is to try to apply this vector form of the raising and lowering operators to the vector form of the spherical harmonics.  I don't know yet how that will work out, and it's a game for a different night.

\section{Explicit expansion of the angular momentum operator.}

There's a couple of things to explore before going forward.  One is an explicit verification that $\Bx \wedge \spacegrad$ has no radial dependence (something not obvious).  Another is that we should be able to compare the $\Bx^{-2} (\Bx \wedge \spacegrad)^2$ (as done for the $\Bx \cdot \spacegrad$ terms) the explicit $r,\theta,\phi$ expression for the Laplacian to verify consistency and correctness.

For the spherical polar rotation we use the rotor

\begin{align}\label{eqn:qmAngularMom:goo1}
R = e^{\Be_{31}\theta/2} e^{\Be_{12}\phi/2}
\end{align}

Our position vector in spherical polar coordinates is then

\begin{align}\label{eqn:qmAngularMom:goo2}
\Bx &= r \tilde{R} \Be_3 R 
\end{align}

\begin{align}\label{eqn:qmAngularMom:goo3}
\spacegrad &= \rcap \partial_r + \thetacap \inv{r} \partial_\theta + \phicap \inv{r \sin\theta} \partial_\phi
\end{align}

and our unit vectors translate from the standard basis as

\begin{align}\label{eqn:qmAngularMom:goo4}
\begin{pmatrix}
\rcap \\
\thetacap \\
\phicap \\
\end{pmatrix}
=
\tilde{R}
\begin{pmatrix}
\Be_3 \\
\Be_1 \\
\Be_2 \\
\end{pmatrix}
R
\end{align}

This last mapping can be used to express the gradient unit vectors in terms of the standard basis, as we did for the position vector $\Bx$.  That is

\begin{align}\label{eqn:qmAngularMom:goo5}
\spacegrad &= \tilde{R} \left( \Be_3 R \partial_r + \Be_1 R \inv{r} \partial_\theta + \Be_2 R \inv{r \sin\theta} \partial_\phi \right)
\end{align}

Okay, we've now got all the pieces collected, ready to evaluate $\Bx \wedge \spacegrad$

\begin{align*}
\Bx \wedge \spacegrad 
&=
r \gpgradetwo{\tilde{R} \Be_3 R 
\tilde{R} \left( \Be_3 R \partial_r + \Be_1 R \inv{r} \partial_\theta + \Be_2 R \inv{r \sin\theta} \partial_\phi \right) } \\
&=
r \gpgradetwo{\tilde{R} \left( R \partial_r + \Be_3 \Be_1 R \inv{r} \partial_\theta + \Be_3 \Be_2 R \inv{r \sin\theta} \partial_\phi \right) } \\
\end{align*}

Observe that the ${\Be_3}^2$ contribution is only a scalar, so bivector selection of that is zero.  In the remainder we have cancellation of $r/r$ factors, leaving just

\begin{align}\label{eqn:qmAngularMom:goo6}
\Bx \wedge \spacegrad 
&=
\tilde{R} \left( \Be_3 \Be_1 R \partial_\theta + \Be_3 \Be_2 R \inv{\sin\theta} \partial_\phi \right) 
\end{align}

Using (\ref{eqn:qmAngularMom:goo4}) this is

\begin{align}\label{eqn:qmAngularMom:goo7}
\Bx \wedge \spacegrad 
&=
\rcap \left( \thetacap \partial_\theta + \phicap \inv{\sin\theta} \partial_\phi \right) 
\end{align}

As hoped, there is no explicit radial dependence here, taking care of the first of the desired verifications.

Next we want to square this operator.  It should be noted that in the original derivation where we ``factored'' the Laplacian operator with respect to the reference vector $\Bx$ we really used $(\Bx \wedge \spacegrad)^2 \equiv \gpgradezero{(\Bx \wedge \spacegrad)^2}$.  That's worth noting since a regular bivector would square to a negative constant, whereas the operator factors of the vectors in this expression do not intrinsically commute.

An additional complication for evaluating the square of $\Bx \wedge \spacegrad$ using the result of (\ref{eqn:qmAngularMom:goo7}) is that $\thetacap$ and $\rcap$ are functions of $\theta$ and $\phi$, so we'd have to operate on those too.  Without that operator subtlety we get the wrong answer

\begin{align*}
-(\Bx \wedge \spacegrad)^2
&=
\gpgradezero{ 
\tilde{R} \left( \Be_1 R \partial_\theta + \frac{\Be_2 R}{\sin\theta}\partial_\phi \right)
\tilde{R} \left( \Be_1 R \partial_\theta + \frac{\Be_2 R}{\sin\theta}\partial_\phi \right)
 } \\
&\ne
\partial_{\theta\theta} + \inv{\sin^2\theta \partial_{\phi\phi}}
\end{align*}

Equality above would only be if the unit vectors were fixed.  By comparison we also see that this is missing a $\cot\theta \partial_\theta$ term.  That must come from the variation of the unit vectors with position in the second application of $\Bx \wedge \spacegrad$.

\EndArticle
