\documentclass{article}

\input{../peeters_macros.tex}
\input{../peeters_macros2.tex}
\input{../peeters_macros3.tex}


%\usepackage{listings}
%\usepackage{txfonts} % for ointctr... (also appears to make "prettier" \int and \sum's)
% makes \grad look funny though (almost like spacegrad, but narrower)
\usepackage[bookmarks=true]{hyperref}

\usepackage{color,cite,graphicx}
   % use colour in the document, put your citations as [1-4]
   % rather than [1,2,3,4] (it looks nicer, and the extended LaTeX2e
   % graphics package. 
\usepackage{latexsym,amssymb,epsf} % don't remember if these are
   % needed, but their inclusion can't do any damage


\title{Stress Energy Tensor from Lagrangian}\label{chap:stress_energy_noethers}
\author{Peeter Joot \quad peeter.joot@gmail.com }
\date{ June 5, 2009.  Last Revision: $Date: 2009/06/06 23:04:17 $ }

\begin{document}

\maketitle{}
\tableofcontents
\section{ Summary and direction. }

In 
\cite{PJstressEnergyLorentz} 
%\ref{chap:PJstressEnergyLorentz} 
we saw that it was
possible to express the Lorentz force equation for the charge per unit
volume in terms of the stress energy tensor.

Repeating 
%from \ref{eqn:seLorentz:lorentzForceT} and
%\ref{eqn:seLorentz:seTensorVecForm}

\begin{align}
\grad \cdot T(\gamma_\mu) &= \inv{c} \gpgradezero{ F \gamma_\mu J } \\
T(a) &= \frac{\epsilon_0}{2} F a \tilde{F}
\end{align}

While these
may not appear too much like the Lorentz force equation as we are used
to seeing it, with some manipulation we found 
%\ref{eqn:seLorentz:lorentzForcePair}

\begin{align}
\inv{c} \gpgradezero{ F \gamma_0 J } &= -\Bj \cdot \BE \\
\inv{c} \gpgradezero{ F \gamma_k J } &= (\rho \BE + \Bj \cross \BB) \cdot \sigma_k
\end{align}

where we now have an energy momentum pair of equations, the second
of which if integrated over a volume is the Lorentz force for the charge
in that volume.

We've also seen in
%\ref{eqn:seLorentz:lorentzForceGA}
that we can express the Lorentz force equation in GA form
\begin{align}
m \ddot{x} &= q F \cdot \dot{x}/c
\end{align}

In 
%\ref{eqn:seLorentz:lorentzForceTensor}
this was expressed in tensor form, toggling indexes that was
\begin{align}
m \ddot{x}_\mu &= {q} F_{\mu\alpha} \dot{x}^\alpha
\end{align}

We then saw in
%\cite{PJenMtensor}
\cite{PJenMtensor}
%\ref{chap:en_m_tensor} in equations \ref{eqn:stressEnTen:miracle} and
%\ref{eqn:stressEnTen:covariantTensor}
, that the
the covariant form of the stress energy tensor relation was

\begin{align}
T^{\mu\nu} &= {\epsilon_0} \left( F^{\alpha\mu} {F^{\nu}}_{\alpha} + \inv{4} F^{\alpha\beta} F_{\alpha\beta} \eta^{\mu\nu} \right) \\
\partial_\nu T^{\mu\nu} &= F^{\alpha\mu} J_\alpha/c
\end{align}

this has identical structure (FIXME: sign error here?) to the
covariant Lorentz force equation.

Now the stress energy conservation equations above did not require
the Lorentz force equations at all for their derivation, nor have we
used the Lorentz force interaction Lagrangian to arrive at them.  
With Maxwell's equation and the Lorentz force equation together (
or the equivalent field and interaction Lagrangians) we have
the complete specificiation of classical electrodynamics.  Curiously
it
appears that we have most of the structure of the Lorentz force equation
(except for the association with mass) all in embedded in Maxwell's equation
or the Maxwell field Lagrangian.  

Now, a proper treatment of the field and charged mass interaction
likely requires the Dirac Lagrangian, and hiding in there if one
could extract it, is probably everything that could be said on the
topic.  It will be a long journey to get to that point, but how
much can we do considering just the field Lagranian?

For these reasons it seems desirable to
understand the background behind the stress energy tensor much better.
In particular, it is natural to then expect that these conservation
relations may also be found as a 
consequence of a symmetry and an associated Noether current
(see 
\cite{PJNoethersField}
%\ref{chap:PJNoethersField}
).  What
is that symmetry?  That symmetry should leave the field equations as 
calculated by the field Euler-Lagrange equations (see
\cite{PJFieldLagrangian}
%\ref{chap:PJFieldLagrangian}
) unchanged.
Given that symmetry how would one go about
actually showing that this is the case?   These are the questions
to tackle here.

\section{ Symmetry due to total derivative addition to the Lagrangian }

In \cite{doran2003gap} the stress energy tensor is treated 
by considering spacetime translation, but I've unfortunately
not understood much more than vague direction in that treatment.

In \cite{srednicki2007qft} it is also stated that the stress energy tensor
is the result of a Lagrangian spacetime translation, but I didn't find
details there.

There is also fairly extensive treatment
of the stress energy tensor (in the simpler non-GA tensor form) in
\cite{jackson1975cew}.  However, that treatment relies on analogy with
mechanical form of Noether's theorem, and I'd rather see it developed
explicitly.

In an unexpected place, the 
clue required to understand the details of 
how this spacetime translation results in the stress energy tensor
was found in
\cite{TongQFT}.

In Tong's treatment it is pointed out there is a symmetry for the Lagrangian
if it is altered by a divergence.

\begin{align*}
\LL \rightarrow \LL + \partial_\mu F^\mu
\end{align*}

It took me a while to figure out how this was a symmetry, but after a
nice refreshing motorcycle ride, the answer suddenly surfaced.  One
can add a derivative to a mechanical Lagrangian and not change
the resulting equations of motion.  While tackling problem 5 of 
Tong's mechanics
in 
\cite{PJTongMf1}
%\ref{chap:PJTongMf1}
, such an invariance was considered in
detail
%\ref{chap:PJTongMf1:addDerivative}
.

FOLLOWUP: No Noether current was 
derived for this sort of mechanical symmetry.  From
what I've read, such a Noether's current calculation is 
likely to provide a demonstration of linear momentum conservation.  Worth
verifying.

If one has altered the Lagrangian by adding an arbitrary function $f$
to it.

\begin{align*}
\LL' = \LL + f
\end{align*}

Assuming to start a Lagrangian that is a function of a single field variable
$\LL = \LL(\phi, \partial_\mu \phi)$, then the 
variation of the Lagrangian for the field equations yields

\begin{align*}
\frac{\delta \LL'}{\delta \phi}
&=
\frac{\partial \LL'}{\partial \phi}
- 
\partial_\sigma \frac{\partial \LL'}{\partial (\partial_\sigma \phi)} \\
&=
\underbrace{\frac{\partial \LL}{\partial \phi}
- 
\partial_\sigma \frac{\partial \LL}{\partial (\partial_\sigma \phi)} }_{=0}
+\frac{\partial f}{\partial \phi}
- 
\partial_\sigma \frac{\partial f}{\partial (\partial_\sigma \phi)} \\
\end{align*}

So, if this transformed Lagrangian is a symmetry, it is sufficient to
find the conditions for the variation of additional part to be zero

\begin{align}\label{eqn:requirement}
\frac{\delta f}{\delta \phi} = 0
\end{align}

\subsection{ Some examples adding a divergence. }

To validate the fact that we can add a divergence to the Lagrangian without
changing the field equations lets work out a few concrete examples 
of \ref{eqn:requirement} of for Lagrangian alterations by a divergence
$f = \partial_\mu F^\mu$.

Each of these examples will be for a single field variable Lagrangian 
with generalized coordinates $x^1 = x$, and $x^1 = y$.
 
\subsubsection{ Simplest case.  No partials. }

Let 

\begin{align*}
F^1 &= \phi \\
F^2 &= 0 \\
\end{align*}

With this the divergence is 

\begin{align*}
f 
&= \partial_x F^x + \partial_y F^y  \\
&=
\PD{x}{\phi}
\end{align*}

Now the variation is
\begin{align*}
\frac{\delta f}{\delta \phi} 
&= 
\left( \PD{\phi}{} 
- \PD{x}{}\PD{(\PDi{x}{\phi})}{} 
- \PD{y}{}\PD{(\PDi{y}{\phi})}{} 
\right) \PD{x}{\phi} \\
&= 
\PD{x}{}\PD{\phi}{\phi} - \PD{x}{1} \\
&= 0
\end{align*}

Okay, so far so good.

\subsubsection{ One partial. }

Now, let 

\begin{align*}
F^1 &= \PD{x}{\phi} \\
F^2 &= 0 \\
\end{align*}

With this the divergence is 

\begin{align*}
f 
&= \partial_x F^x + \partial_y F^y  \\
&=
\PD{x}{}\PD{x}{\phi}
\end{align*}

And the variation is
\begin{align*}
\frac{\delta f}{\delta \phi} 
&= 
\left( \PD{\phi}{} 
- \PD{x}{}\PD{(\PDi{x}{\phi})}{} 
- \PD{y}{}\PD{(\PDi{y}{\phi})}{} 
\right) \PD{x}{}\PD{x}{\phi} \\
&= 
\PD{\phi}{} \PD{x}{}\PD{x}{\phi} \\
&= 
\PD{x}{} \PD{x}{}\PD{\phi}{\phi} \\
&= 
\PD{x}{} \PD{x}{1} \\
&= 0
\end{align*}

Again, assuming I'm okay to switch the differentiation order, we have zero.

\subsubsection{ Another partial. }

For the last concrete example before going on to the general case, try

\begin{align*}
F^1 &= \PD{y}{\phi} \\
F^2 &= 0 \\
\end{align*}

The divergence is 

\begin{align*}
f 
&= \partial_x F^x + \partial_y F^y  \\
&=
\PD{x}{}\PD{y}{\phi}
\end{align*}

And the variation is
\begin{align*}
\frac{\delta f}{\delta \phi} 
&= 
\left( \PD{\phi}{} 
- \PD{x}{}\PD{(\PDi{x}{\phi})}{} 
- \PD{y}{}\PD{(\PDi{y}{\phi})}{} 
\right) \PD{x}{}\PD{y}{\phi} \\
&= 
- \PD{y}{}\PD{x}{1} \\
&= 0
\end{align*}

\subsubsection{ The general case. }

Because of linearity we've now seen that we can construct functions with
any linear combinations of first and second derivatives

\begin{align*}
F^\mu = a^\mu \phi + \sum_\sigma {b_\sigma}^\mu \PD{x^\sigma}{\phi}
\end{align*}

and for such a function we will have

\begin{align*}
\frac{\delta (\partial_\mu F^\mu)}{\delta \phi}  = 0
\end{align*}

How general can the function $F^\mu = F^\mu(\phi, \partial_\sigma \phi)$ be 
made and still yield a zero variational derivative?

To answer this, let's compute the derivative for a general divergence 
added to a single field variable Lagrangian.  This is

\begin{align*}
\frac{\delta (\partial_\mu F^\mu)}{\delta \phi} 
&= 
\sum_\mu \left( \PD{\phi}{} 
- \sum_\sigma \PD{x^\sigma}{}\PD{(\PDi{x^\sigma}{\phi})}{} 
\right) \PD{x^\mu}{F^\mu} \\
&= 
\sum_\mu \PD{x^\mu}{} \PD{\phi}{F^\mu} 
- \sum_{\mu,\sigma} \PD{x^\sigma}{}\PD{(\PDi{x^\sigma}{\phi})}{} 
\left(
\PD{\phi}{F^\mu} \PD{x^\mu}{\phi}
+ \sum_\alpha \PD{(\PDi{x^\alpha}{\phi})}{F^\mu} \PD{x^\mu}{(\PDi{x^\alpha}{\phi})} \right) \\
&= 
\partial_\mu \PD{\phi}{F^\mu} 
- \partial_\sigma \PD{(\partial_\sigma \phi)}{} 
\left(
\PD{\phi}{F^\mu} \partial_\mu {\phi}
+ \PD{(\partial_\alpha \phi)}{F^\mu} \partial_{\mu\alpha} \phi \right) \\
\end{align*}

For tractability in this last line the shorthand for the partials has been injected.  
Sums over $\alpha$, $\mu$, and $\sigma$ are also now implied (this was made explicit prior to 
this in all cases where upper and lower indexes were matched).

Treating these two last derivatives separately, we have for the first 

\begin{align*}
\partial_\sigma 
\PD{(\partial_\sigma \phi)}{} 
\PD{\phi}{F^\mu} \partial_\mu {\phi}
&=
\partial_\sigma 
\left(\PD{(\partial_\sigma \phi)}{} \PD{\phi}{F^\mu} \right) \partial_\mu {\phi}
+
\partial_\sigma 
\PD{\phi}{F^\mu} \PD{(\partial_\sigma \phi)}{} \partial_\mu {\phi} \\
&=
\partial_\sigma 
\left(\PD{(\partial_\sigma \phi)}{} \PD{\phi}{F^\mu} \right) \partial_\mu {\phi}
+
\partial_\mu \PD{\phi}{F^\mu}
\end{align*}

So our $\PDi{\phi}{F^\mu}$'s cancel out, and we are left with

\begin{align*}
\frac{\delta (\partial_\mu F^\mu)}{\delta \phi} 
&=
-\partial_\sigma 
\left(
\left(\PD{(\partial_\sigma \phi)}{} \PD{\phi}{F^\mu} \right) \partial_\mu {\phi}
+ 
\PD{(\partial_\sigma \phi)}{} 
\left(
\PD{(\partial_\alpha \phi)}{F^\mu} \partial_{\mu\alpha} \phi 
\right) 
\right) 
\\
&=
-\partial_\sigma 
\left(
\partial_\mu {\phi}
\left(\PD{(\partial_\sigma \phi)}{} \PD{\phi}{F^\mu} \right) 
+ 
\partial_{\mu\alpha} \phi 
\PD{(\partial_\sigma \phi)}{} 
\left(
\PD{(\partial_\alpha \phi)}{F^\mu} 
\right) 
\right) 
\\
&=
-\partial_\sigma 
\left(
(\partial_\mu {\phi})
\PD{\phi}{} 
\PD{(\partial_\sigma \phi)}{}
F^\mu
+ 
\left(\partial_{\mu} 
\PD{x^\alpha}{\phi}\right)
\PD{(\partial_\alpha \phi)}{}
\PD{(\partial_\sigma \phi)}{} 
{F^\mu} 
\right) 
\\
\end{align*}

Now there's a lot of indexes and derivatives floating around.  Writing $g^\mu = \PDi{(\partial_\sigma \phi)}{F^\mu}$, we have something a bit easier to look at

\begin{align*}
\frac{\delta (\partial_\mu F^\mu)}{\delta \phi} 
&=
-\partial_\sigma 
\left(
(\partial_\mu {\phi})
\PD{\phi}{ g^\mu }
+ 
\left(\partial_{\mu} 
\PD{x^\alpha}{\phi}\right)
\PD{(\partial_\alpha \phi)}{ g^\mu }
\right) 
\\
\end{align*}

But this is a chain rule expansion of the 
derivative $\partial_\mu g^\mu$

\begin{align*}
\PD{x^\mu}{g^\mu} &=
\PD{x^\mu}{\phi}\PD{\phi}{g^\mu} 
+ \PD{x^\mu}{\partial_\beta \phi}\PD{\partial_\beta \phi}{g^\mu} 
\end{align*}

So, we finally have

\begin{align*}
\frac{\delta (\partial_\mu F^\mu)}{\delta \phi} 
&=
-\partial_{\sigma \mu} g^\mu \\
&=
\end{align*}

This is
\begin{align}\label{eqn:requiredZeroForSymmetry}
\frac{\delta (\partial_\mu F^\mu)}{\delta \phi} 
&=
-\partial_{\sigma \mu} \PD{(\partial_\sigma \phi)}{F^\mu} 
\end{align}

I don't think we have any right asserting that this is zero for arbitrary $F^\mu$.  However
if the Taylor expansion of $F^\mu$ with respect to variables $\phi$, and $\partial_\sigma \phi$ has
no higher than first order terms in the field variables $\partial_\sigma \phi$, we will
certainly have a zero variational derivative and a corresponding symmetry.

\subsubsection{ More examples to confirm the symmetry requirements. }

As a confirmation that a zero in \ref{eqn:requiredZeroForSymmetry} requires linear field derivatives, 
lets lets try two more example calculations.  

First with non-linear
powers of $\phi$ to show that we have more freedom to construct the function first powers.  Let

\begin{align*}
F^1 &= \phi^2 \\
F^2 &= 0
\end{align*}

We have

\begin{align*}
\frac{\delta (\partial_\mu F^\mu)}{\delta \phi} 
&=
\left(\PD{\phi}{} - \partial_\sigma \PD{(\partial_\sigma \phi)}{}\right) 2 \phi \phi_x \\
&=
2 \phi_x - \partial_x (2 \phi) \\
&= 0
\end{align*}

Zero as expected.  Generalizing the function to include arbitrary polynomial powers is no harder.

Let
\begin{align*}
F^1 &= \phi^k \\
F^2 &= 0 \\
\partial_\mu F^\mu &= k \phi^{k-1} \phi_x \\
\end{align*}

So we have
\begin{align*}
\frac{\delta (\partial_\mu F^\mu)}{\delta \phi}  
&= 
k (k-1) \phi^{k-2} \phi_x - \partial_x (k \phi^{k-1})  \\
&= 0
\end{align*}

Okay, now moving on to the derivatives.  Picking a divergence that should not will not generate a
symmetry, something with a non-linear derivative should do the trick.  Let's Try

\begin{align*}
F^1 &= (\phi_x)^2 \\
F^2 &= 0
\end{align*}

\begin{align*}
\frac{\delta (\partial_\mu F^\mu)}{\delta \phi} 
&=
\left(\PD{\phi}{} - \partial_\sigma \PD{(\partial_\sigma \phi)}{}\right) 2 \phi_x \phi_{xx} \\
&=
- 2 \partial_x \phi_{xx} \\
&=
- 2 \phi_{xxx} \\
\end{align*}

So, sure enough, unless additional conditions can be imposed on $\phi$, such a transformation
will not be a symmetry.

\section{ Symmetry for Wave equation under spacetime translation. }


The Lagrangian for a one dimensional wave equation is

\begin{align}\label{eqn:oneDimWave}
\LL = 
\inv{2 v^2} \left(\PD{t}{\phi}\right)^2 - \inv{2} \left(\PD{x}{\phi}\right)^2
\end{align}

Under a transformation of variables 

\begin{align*}
x &\rightarrow x' = x + a \\
t &\rightarrow t' = t + \tau 
\end{align*}

Employing a multivariable Taylor expansion
(see 
\cite{PJmultiTaylors}
%\ref{chap:PJmultiTaylors}
)
for our Lagrangian having no explicit dependence on $t$ and $x$, we have

\begin{align*}
\LL' &= \LL + \underbrace{(a \partial_x + \tau \partial_t)\LL}_{(*)} + \cdots
\end{align*}

That first order term of the Taylor expansion $(*)$, 
can be written as a divergence $\partial_\mu F^\mu$, with $F^1 = a \LL$, and $F^2 = \tau\LL$, however
both of these are quadratic in $\phi_x$, and $\phi_t$, which isn't linear.
That linearity in the derivatives was required for \ref{eqn:requiredZeroForSymmetry} to be 
definitively zero for the transformation to be a symmetry.  So after all that goofing around
with derivatives and algebra it is defeated by the
simplest field Lagrangian.

Now, if we continue we find that we do in fact still have a symmetry by introducing a linearized spacetime translation.
This follows from direct expansion

\begin{align*}
(*) 
&= (a \partial_x + \tau \partial_t) \LL \\
&= 
a 
\left(
\inv{v^2} \phi_t \partial_x \phi_t
- \phi_x \partial_x \phi_x
\right)
+ \tau 
\left(
\inv{v^2} \phi_t \partial_t \phi_t
- \phi_x \partial_t \phi_x
\right) \\
\end{align*}

Next, calculation of the variational derivative we have

\begin{align*}
\frac{\delta(*)}{\delta \phi}
&=
\left( \PD{\phi}{} - \partial_x \PD{\phi_x}{} - \partial_t \PD{\phi_t}{} \right) (*) \\
&=
-\partial_x \left( -a \partial_{xx} \phi - \tau \partial_{tx}\phi \right) 
-\inv{v^2} \partial_t \left( a \partial_{xt} \phi + \tau \partial_{tt}\phi \right) \\
&=
a \left( 
\partial_x \left(
\phi_{xx} - \inv{v^2} \phi_{tt}
\right)
\right)
+ \tau \left( 
\partial_t \left(
\phi_{xx} - \inv{v^2} \phi_{tt}
\right)
\right) \\
\end{align*}

Since we have $\phi_{xx} = \inv{v^2} \phi_{tt}$ by variation of \ref{eqn:oneDimWave}.  So we do in fact have a symmetry from the 
linearized spacetime translation for any shift $(t,x) \rightarrow (t+\tau, x+a)$.

\section{ Symmetry condition for arbitrary linearized spacetime translation. }

If we want to be able to alter the Lagragian with a linearized vector translation of the generalized coordinates by some arbitrary 
shift, since we do not have the linear derivatives for many Lagrangians of interest (wave equations, Maxwell equation, ...)
then can we find a general condition that is responsible for the translation symmetry that we've observed must exist for
the simple wave equation.

For a general Lagrangian $\LL = \LL(\phi(x), \partial_\mu \phi(x))$ under shift by some vector $a$

\begin{align*}
x \rightarrow x' = x + a
\end{align*}

we have

\begin{align*}
\LL' = e^{a \cdot \grad} \LL = \LL + (a \cdot \grad)\LL + \cdots
\end{align*}

Compuating the variational derivative for the first order term we have

\begin{align*}
\frac{\delta (a \cdot \grad) \LL}{\delta \phi} 
&=
\left(\PD{\phi}{} - \partial_\sigma \PD{\phi_\sigma}{} \right) \inv{2}(a \grad + \grad a) \LL \\
&=
\inv{2} \left( a \grad \PD{\phi}{\LL} - \grad a \partial_\sigma \PD{\phi_\sigma}{\LL} \right) \\
\end{align*}

But $\PDi{\phi}{\LL} = \partial_\sigma \PDi{\phi_\sigma}{\LL}$, so we have

\begin{align*}
\frac{\delta (a \cdot \grad) \LL}{\delta \phi} 
&=
\inv{2} \left( a \grad \PD{\phi}{\LL} - \grad a \PD{\phi}{\LL} \right) \\
\end{align*}

This is
\begin{align}
\frac{\delta (a \cdot \grad) \LL}{\delta \phi} 
&= (a \wedge \grad) \PD{\phi}{\LL}
\end{align}

So for the wave equation, which has only derivatives and no explicit dependence on $\phi$ itself, this 
shows why we will have a symmetry (not yet calculated) under linearized translation for the wave equation despite the quadratic derivative terms.

\section{ Noether current. }

\bibliographystyle{plainnat}
\bibliography{myrefs}

\end{document}
