\input{../peeter_prologue.tex}

\chapter{transverseWave}
\label{chap:transverseWave}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2009/transverseWave.pdf}
\date{August 6, 2009}
\revisionInfo{$RCSfile: transverseWave.tex,v $ Last $Revision: 1.2 $ $Date: 2009/08/09 00:33:08 $}

\beginArtWithToc

\section{Motivation}

In (\cite{maxwellVacuum}) a phasor like form of the transverse wave equation was derived by considering Fourier solutions of the Maxwell equation.  Curiously no perpendicularity condition for $\BE$ and $\BB$ seemed to be required there.  Why would that be the case?  In Jackson's treatment, which employed the traditional dot and cross product form of Maxwell's equations, this followed by back substituting the assumed phasor solution back into the equations.  This back substitution wasn't done in (\cite{maxwellVacuum}), and if we attempt this we should find the same sort of additional mutual perpendicularity constraints on the fields.

Here we start with the equations from Jackson (\cite{jackson1975cew}, ch7), expressed in GA form.  Using the same assumed phasor form we should get the same resuls using GA.  Anything else indicates a misunderstanding.  After using a more traditional phasor form (where one would have to take real parts) we revisit the ``phasor'' like form found in (\cite{maxwellVacuum}), and find that when all is said and done it ends up being equivalent to the phasor form where the imaginary $i$ has been replaced by the spatial pseudoscalar $I = \sigma_1\sigma_2\sigma_3$.

\section{With assumed phasor field}

Maxwell's equations in abscence of charge and current ((7.1) of Jackson) can be summarized by 

\begin{align}\label{eqn:foo1}
0 &= (\grad + \sqrt{\mu\epsilon}\partial_0) F 
\end{align}

The $F$ above is a composite electric and magnetic field merged into a single multivector.  In the spatial basic the electric field component $\BE$ is a vector, and the magnetic component $I\BB$ is a bivector (in the Dirac basis both are bivectors).

\begin{align}\label{eqn:foo1}
F &= \BE + I \BB/\sqrt{\mu\epsilon}) 
\end{align}

With an assumed phasor form

\begin{align}\label{eqn:foo2}
F = \FF e^{ i(\Bk \cdot \Bx - \omega t) } 
\end{align}

Although there are many geometric multivectors that square to -1, we do not assume here that the imaginary $i$ has any specific geometric meaning, and in fact commutes with all multivectors.  Because of this we have to take the real parts later when done.

Operating on $F$ with Maxwell's equation we have 

\begin{align}\label{eqn:foo3}
0 = (\grad + \sqrt{\mu\epsilon}\partial_0) F = i \left( \Bk - \sqrt{\mu\epsilon}\frac{\omega}{c} \right) F 
\end{align}

Similarily, left multiplication of Maxwell's equation by the conjugate operator $\grad - \sqrt{\mu\epsilon}\partial_0$, we have the wave equation

\begin{align}\label{eqn:foo4}
0 &= \left(\grad^2 - \frac{\mu\epsilon}{c^2}\frac{\partial^2}{\partial_t^2}\right) F \\
\end{align}

and substitution of the assumed phasor solution gives us

\begin{align}\label{eqn:foo7}
0 = (\grad^2 - {\mu\epsilon}\partial_{00}) F = -\left( \Bk^2 - {\mu\epsilon}\frac{\omega^2}{c^2} \right) F 
\end{align}

This provides the relation between the magnitude of $\Bk$ and $\omega$, namely

\begin{align}\label{eqn:foo8}
\Abs{\Bk} = \pm \sqrt{\mu\epsilon}\frac{\omega}{c} 
\end{align}

Without any real loss of generality we can pick the positive root, so the result of the Maxwell equation operator on the phasor is

\begin{align}\label{eqn:foo5}
0 = (\grad + \sqrt{\mu\epsilon}\partial_0) F = i \sqrt{\mu\epsilon}\frac{\omega}{c} \left( \kcap - 1\right) F 
\end{align}

Rearranging we have the curious property that the field $F$ can ``swallow'' a left multiplication by the propagation direction unit vector 

\begin{align}\label{eqn:foo9}
\kcap F = F 
\end{align}

\begin{align}\label{eqn:fooA}
\kcap F = F 
\end{align}

Selection of the scalar and pseudoscalar grades of this equation shows that the electric and magnetic fields $\BE$ and $\BB$ are both completely transverse to the propagation direction $\kcap$.  For the scalar grades we have

\begin{align*}
0 &= \gpgradezero{\kcap F - F} \\
  &= \kcap \cdot \BE
\end{align*}

and for the pseudoscalar
\begin{align*}
0 &= \gpgradethree{\kcap F - F} \\
  &= I \kcap \cdot \BB
\end{align*}

From this we have $\kcap \cdot \BB = \kcap \cdot \BB = 0$.  Because of this transverse property we see that the $\kcap$ multiplication of $F$ in (\ref{eqn:fooA}) serves to map electric field (vector) components into bivectors, and the magnetic bivector components into vectors.  For the result to be the same means we must have an additional coupling between the field components.  Writing out (\ref{eqn:fooA}) in terms of the field components we have

\begin{align*}
\BE + I\BB/\sqrt{\mu\epsilon} 
&= \kcap (\BE + I\BB/\sqrt{\mu\epsilon} ) \\
&= \kcap \wedge \BE + I (\kcap \wedge \BB)/\sqrt{\mu\epsilon} 
&= I \kcap \cross \BE + I^2 (\kcap \cross \BB)/\sqrt{\mu\epsilon} 
\end{align*}

Equating left and right hand grades we have

\begin{align}\label{eqn:fooD}
\BE &= -(\kcap \cross \BB)/\sqrt{\mu\epsilon} \\
\BB &= \sqrt{\mu\epsilon} \kcap \cross \BE 
\end{align}

\EndArticle
