%
% Copyright © 2013 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
These are collected notes of believed typos from my reading of \citep{desai2009quantum}.  Some were confirmed by Prof.\ Vatche Deyirmenjian who taught the PHY356 course I took at UofT.  He also pointed out still more (and pointed out where I had identified the wrong source for some typos).

Some but not all of the typos found in the second third of the book have been confirmed by Prof.\ John Sipe, who teaches the PHY456 course, also at UofT.

Any of the typos found in the last third of the text (Dirac content) have not been confirmed, and are based only on private study.

\section{Chapter 1}

\begin{itemize}
\item Page 1.  Prof.\ Deyirmenjian: The Hermitian, not complex conjugate, of \(\ket{}\) is \(\bra{}\).
\item Page 5-6.  Prof.\ Deyirmenjian: Change the \(\conj\) in (1.26), (1.31), and (1.33) to a dagger.
\item Page 7.  Text before (1.43).  \(\alpha\) instead of \(a\) used.
\item Page 19.  Equation (1.122).  \(\dagger\)s omitted after first equality.
\end{itemize}

\section{Chapter 2}
\begin{itemize}
\item Page 40.  Text before (2.137).  Reference to equation (2.133) should be (2.135)
\item Page 53.  Is the "Also show that" here correct?  I get a different answer.
\end{itemize}

\section{Chapter 3}
\begin{itemize}
\item Page 61.  Equation (3.51).  \(1/\Hbar\) missing.
\item Page 62.  Equation (3.58).  Prof.\ Deyirmenjian: Remove the \(U_I\) operators from Eq. (3.58)
\item Page 66.  Equation (3.92).  \(-(d/dt \bra{\alpha}) \ket{\alpha}\) should be \(+\ket{\alpha} d/dt \bra{\alpha}\).
\item Page 66.  Equation (3.93).  \(H\) on wrong side of \(\bra{\alpha}\) 
\item Page 74,76.  Prof.\ Deyirmenjian: remove the extra brackets from Eq (4.9) and (4.21).
\item Page 79.  Prof.\ Deyirmenjian: "The probability of finding this particle" should read "The probability density for this state at point x is"
\end{itemize}

\section{Chapter 4}
\begin{itemize}
\item Page 81.  Equation (4.52).  Should be \(-2\alpha\) in the exponent.
\item Page 82.  Equation (4.65).  Prof.\ Deyirmenjian: a \(1/\sqrt{2\pi}\) is missing before the integral.  Note that without this (4.67) appears incorrect (off by a factor of \(\sqrt{2\pi}\), but the error is really just in (4.65).
\item Page 82.  Equation (4.67).  Prof.\ Deyirmenjian: the negative sign should appear inside the large square brackets.
%%% NO:  See above.
%\item Page 82.  Equation (4.67).  \(2\alpha\) in the denominator of the normalization should be \(\alpha'/\pi\).
\item Page 83.  Equation (4.74).  A normalized wave function is not required for the discussion, but if that was intended, a \(1/\sqrt{2\pi}\) factor is missing.
\item Page 83-84.  Prof.\ Deyirmenjian: In (4.67) and (4.77), the derivative should be evaluated at \(k=k_0\).
\item Page 86.  Equation (4.99).  Extra brace in the exponent.
\item Page 87.  Equation (4.106).  Extra brace in the exponent.
\item Page 89.  Equation (4.124-4.130).  Prof.\ Deyirmenjian: \(C e^{\pm \sqrt{\mu}\phi}\) is not a solution to (4.122).  This should be \(Q(\phi) = C e^{i \sqrt{\mu} \phi}\) and (4.126) should be \(\sqrt{\mu} = m\).  This fixes the apparent error in sign in equations 4.129 and 4.130 which are correct as is.
\item Page 92.  Equation (4.158).  Prof.\ Deyirmenjian: should read \(P_l(1) = 1\).
\item Page 93.  Equation (4.169).  conjugation missing for \(Y_{lm}\).  \(Y_{l'm'}\) is missing prime on the \(l\) index.
\item Page 95.  Second line of text.  Language choice?  "We now implement".  perhaps utilize would be better?
\item Page 95.  Text before (4.193).  \(i\) is in bold.
\item Page 96.  Text before (4.196).  \(i\) is in bold.
\item Page 97.  (4.205).  \(i\) is in bold.
\item Page 97.  (4.207-209).  \(\Bi\), and \(\Bj\)s are not in bold like \(\Bk\)
\item Page 101.  (4.245).  The right side should read \(Y_{l,m+1}\)
\item Page 101.  (4.239-240).  The approach here is unclear.  FIXME: incorporate lecture notes from class that did this using braket notation.
\item Page 102.  (4.248-249).  Commas missing to separate \(l\), and \(m\pm 1\) in the kets.
\end{itemize}

\section{Chapter 5}
\begin{itemize}
\item Page 109.  (5.49). minor: Remove bold font in right hand side state \(\ket{\chi_{n+}}\).
\item Page 113.  (5.86). One \(\sigma\) is not in bold.
\item Page 114.  (5.100). \(\chi\) is in bold.
\item Page 115.  Text before (5.106). \(\alpha\) in bold.
\item Page 118.  Switch of notation in problem 5 for ensemble averages.  \([S_i]\) used instead of \(\expectation{S_i}_{\text{av}}\).
\end{itemize}

\section{Chapter 6}
\begin{itemize}
\item Page 120.  \(\phi\) in bold.  \(A\) not in bold.
\item Page 123.  (6.26).  \(1/i \Hbar\) factor missing on RHS.
\item Page 124.  Text before (6.37).  You say canonical momenta \(P_k\), but call these mechanical momenta on prev page.
\item Page 125.  (6.41).  Some \(\psi\)s are in bold.
\item Page 126.  (6.49).  There is no mention that \(\BB\) is constant, leaving it unclear how the gauge condition and how the curl of \(\BA\) reproduces \(\BB\).  This would also help clarify how you are able to write \(\Bmu \cdot \BB = \BB \cdot \Bmu\).
\item Page 128.  (6.65).  \(\Bmu \cdot \BL\) should be \(\Bmu \cdot \BB\).
\item Page 129.  (6.75).  \(\Bmu \cdot \BL\) should be \(\Bmu \cdot \BB\).
\item Page 130.  (6.80).  integral looks like it should be \(\int_{\Br' = \Br_0}^\Br \frac{e}{c \Hbar} \BA(\Br') \cdot d\Br'\).  ie: Clarify bounds, and add a factor of \(c\) in the denominator which is required for the cancellation of (6.82).
\item Page 131.  (6.81,6.86).  Factors of \(c\)s should be with each of the \(\Hbar\)s.
\item Page 131.  Problem 1.  bold missing on \(\BE\).
\end{itemize}

\section{Chapter 8}
\begin{itemize}
\item Page 143.  (8.58).  \(\beta\) should be negated.
\item Page 159.  (8.6.3).  Two references to Chapter 2 should be Chapter 4.
\item Page 160.  (8.199).  Want \(\Hbar^2\) not \(\Hbar\) in expression for \(k\).
\item Page 162.  (Fig 8.9).  Figure is backwards compared to text (a bump instead of a well).
\item Page 165.  (8.235).  Extra \(R_l\) factor inside parens.
\end{itemize}

\section{Chapter 9}
\begin{itemize}
\item Page 174.  (9.5).  Have \(\Hbar/2m\omega\) instead of \(\Hbar m \omega/2\) in expression for \(P\).
\item Page 181.  (9.57).  Factor of two missing.  Want \(\frac{\alpha}{2 \sqrt{\pi}}\).
\item Page 186.  (Problem 10).  Sequencing the text and problems is off.  The green's function technique is not introduced until chapter 10.
\end{itemize}

\section{Chapter 10}
\begin{itemize}
\item Page 189.  (10.22).  It would be nice to have a reference to the appendix (ie: 10.100) for the chapter so that this identity is not pulled out of a magic hat.
\item Page 192.  (10.44, 10.45).  \(2 \alpha {\alpha^\conj}'\) should be \(\alpha {\alpha^\conj}' + \alpha' \alpha^\conj\)
\item Page 193.  (10.51).  Application (slowly, step by step explicitly) of 10.100 to expand the \(e^{\frac{i}{\Hbar}(p_0 X - x_0 P)}\) in the braket gives

\begin{equation}\label{eqn:desaiTyposInnards:20}
\begin{aligned}
\bra{x} e^{\frac{i}{\Hbar}(p_0 X - x_0 P)} \ket{0}
&=
\bra{x} e^{\frac{i}{\Hbar}p_0 X }
e^{-\frac{i}{\Hbar}x_0 P}
e^{-\frac{i}{2\Hbar}x_0 p_0 \antisymmetric{X}{P}}
\ket{0} \\
&=
\bra{x} e^{\frac{i}{\Hbar}p_0 X }
e^{-\frac{i}{\Hbar}x_0 P}
e^{\frac{x_0 p_0}{2} }
\ket{0} \\
&=
e^{\frac{x_0 p_0}{2} }
\bra{x} e^{\frac{i}{\Hbar}p_0 X} 
e^{-\frac{i}{\Hbar}x_0 P}
\ket{0} \\
&=
e^{\frac{x_0 p_0}{2} }
\left(
\bra{0} 
e^{\frac{i}{\Hbar}x_0 P}
e^{-\frac{i}{\Hbar}p_0 X} 
\ket{x}\right)^\conj \\
&=
e^{\frac{x_0 p_0}{2} }
\left(
\bra{0} 
e^{\frac{i}{\Hbar}x_0 P}
\ket{x}
e^{-\frac{i}{\Hbar}p_0 x} 
\right)^\conj \\
&=
e^{\frac{x_0 p_0}{2} } e^{\frac{i}{\Hbar}p_0 x} 
\left(
\bra{0} 
e^{\frac{i}{\Hbar}x_0 P}
\ket{x}
\right)^\conj \\
&=
e^{\frac{x_0 p_0}{2} } e^{\frac{i}{\Hbar}p_0 x} 
\left(
\braket{0}{x - x_0}
\right)^\conj \\
&=
e^{\frac{x_0 p_0}{2} } e^{\frac{i}{\Hbar}p_0 x} 
\braket{x - x_0}{0} \\
&=
e^{\frac{x_0 p_0}{2} } e^{\frac{i}{\Hbar}p_0 x} 
\psi_0(x - x_0, 0)
\end{aligned}
\end{equation}

This is the same as (10.51) with the exception of a real scalar constant \(e^{ x_0 p_0/2}\) multiplying the wave function.  Because of this I think that (10.51) should be a proportionality statement, and not an equality as in

\begin{equation}\label{eqn:desaiTyposInnards:40}
\begin{aligned}
\bra{x} e^{\frac{i}{\Hbar}(p_0 X - x_0 P)} \ket{0} \propto
e^{\frac{i}{\Hbar}p_0 x} \psi_0(x - x_0, 0)
\end{aligned}
\end{equation}

(ie: building this additional factor into the wave function normalization instead).
\item Page 196.  (text after 10.76).  Looks like reference to Chapter 9, should be Chapter 9 problem 5.

\item Page 197.  (text after 10.85).  Reference to Chapter 1 should be Chapter 2.
\end{itemize}

\section{Chapter 16}
\begin{itemize}
\item Page 278.  minor: On the line ``Pulling out'', \(\ket{\psi_s^0}\) should be \(\ket{\psi_s^{(0)}}\).
\item Page 286.  (16.73).  \(a\) should be \(a_0\)
\end{itemize}

\section{Chapter 17}
\begin{itemize}
\item Page 297.  (17.38).   appears to be off by a factor of \(2\) since \((\sin^2 2x)' = 2 \sin x \cos x = \sin(2 x)\).
\item Page 311.  (17.134).  \(d/dt'\) missing on the \(H_{ss}'\) term in the integral.
\item Page 311.  (17.136).  First term (non-integral part) should be negated.
\item Page 312.  (17.144).  Sign on \(\lambda\) before sum positive instead of negative.
\item Page 313.  (17.149).  \(1/\Hbar\) missing.
\item Page 313.  (17.152,17.154).  extra bra around the bra.
\item Page 313.  (17.153).  bra missing on \(\phi_n\)
\end{itemize}

\section{Chapter 19}
\begin{itemize}
\item Page 345.  fig (19.2).  Infinite potential is positioned at \(a\) instead of \(0\) (no reference to \(a\) in \S 19.2).
\item Page 346.  (19.40).  \(1/\Abs{A}^2\) missing from both RHS terms.
\item Page 347.  (19.50).  minor: misplaced comma.
\item Page 349.  (19.66).  \(1/2\) missing on RHS.
\item Page 349.  (19.67).  \(1/2\) missing on RHS.
\end{itemize}

\section{Chapter 24}
\begin{itemize}
\item Page 450.  (24.6).  \(k^2(x)\) should be \(k^2(x) u\).
\item Page 452.  (24.18).  In the \(E < V\) case should be \(1/\sqrt{\kappa}\) instead of \(1/\sqrt{k}\) (although what is in the text is strictly still correct since it only changes the phase of the wavefunction).
% wrong: think that we have a sqrt squared because of the product of two
% wavefunctions.
\item Page 455.  (24.40).  RHD should be multiplied by \(\Hbar\).
%\item Page 459.  (24.68).  should be square root around \(Z^3/\pi a_0^3\).
\item Page 460.  (24.71).  \(\psi\) should be \(\phi\).
\item Page 460.  Third paragraph.  \((\Br_1 - \Br)\) should be \((\Br_1 - \Br_2)\).
\item Page 460.  (24.76).  The integral should be \(1/3\), not \(5 \pi/32\).  This messes up some of the subsequent stuff, unless there is also another compensating error.  Note that one can check this easily since the derivative of \(-1/(3 (1+x)^3)\) is \((1 + x)^{-4}\).
\end{itemize}

\section{Chapter 25}
\begin{itemize}
\item Page 471.  (25.18).  \(i\) subscripts missing on \(\LL\) and \(\dot{y}^2\).
\end{itemize}

\section{Chapter 26}
\begin{itemize}
\item Page 486.  (26.60).  \(\Bn \cross \Br \cdot \spacegrad\) ought to have braces and read \((\Bn \cross \Br) \cdot \spacegrad\).
\item Page 487.  (26.67).  \(0\) in the \(3,3\) position should be \(1\).
\item Page 489.  before (26.84).  For rotations about the \textunderline{imaginary} axis was probably meant to be the i'th axis.
\item Page 495.  (26.149,26.150).  Looks like \(\Hbar\)'s are missing (esp. compared to 26.144-145).
\item Page 495.  (26.150).  \(J_y\) off by \(-1\). (\(J_x + iJ_y \ne J_{+}\)) as is.
\item Page 496.  (26.154).  An extra \(Y_{l'm}\) in the integral, in between \(Y_{l' m'}\) and the \((\theta, \phi)\).
\item Page 498.  (26.175).  \(e^{i\phi}\) should be \(e^{-i\phi}\) in the first line.
\item Page 498.  (26.178).  An \(\Hbar\) factor has been lost in either (26.178) or (26.179).
\item Page 499.  (26.190).  minor: \(\Bj\) should be \(\BJ\).
\item Page 450.  (26.192).  minor: \(\Bj\) should be \(\BJ\), and \(R\) should be \(\Hbar\).
\end{itemize}

\section{Chapter 27}
\begin{itemize}
\item Page 503.  (27.8).  minor: bold \(\chi\).  \(R(\theta, \phi)\) probably meant to be \(R(\chi)\).
\item Page 504.  (27.20).  Missing \(\Hbar E_n\) factor on LHS.
\item Page 507.  before (27.53).  minor: Velocity \(v\) missing bold.
\item Page 510.  before (27.78).  minor: periods in the two kets should be commas.
\item Page 510.  (27.80).  \(\sigma_y\) and \(\sigma_z\) should be interchanged (if \(\alpha\) is the polar angle then \(\ncap = \ycap\) for that rotation, and \(\ncap = \zcap\) for the rotation in the \(x,y\) plane).   The \(\Hbar\)'s here should also be dropped.
\item Page 511.  (27.81).  Same as 27.80.
\item Page 511.  (27.83).  \(\Hbar\)'s should be dropped.
\item Page 514.  (27.109).  minor: dot instead of cdot.
\item Page 515.  (27.117).  Same error as in (26.149-150).  \(\Hbar\)'s missing, and wrong sign on \(J_y\).
\end{itemize}

\section{Chapter 28}
This chapter written as if \(\Hbar = 1\), without a statement that this is being done.
\begin{itemize}
\item Page 518.  (28.4).  \(\Hbar\) missing.  Also in text following, eigenvalue should be \(m \Hbar = \Hbar (m_1 + m_2)\).
\item Page 519.  (28.9).  \(\Hbar\) missing LHS.
\item Page 519.  (28.11).  \(\Hbar\) missing (after each equality).  Text following \(m\) and \(j(j+1)\) eigenvalues should be multipled by \(\Hbar\) and \(\Hbar^2\) respectively.
\item Page 519.  following (28.14).  \(\Hbar\) missing in \(J_{-}\) equality.
\item Page 520.  (28.15).  \(\Hbar\) missing for two factors after last \(=\).
\item Page 520.  (28.16).  \(\Hbar\) missing LHS.
\item Page 520.  (28.21).  \(\downarrow \downarrow\) should be \(\uparrow \downarrow\).
\item Page 521.  following (28.25).  Chapter 2 should read Chapter 5.
\item Page 522.  (28.31).  Notational inconsistency.  \(\ket{j_1 j_2 jm}\) should read \(\ket{j_1 j_2, j m}\)
\item Page 522.  (28.31).  Extra \(\vert\) between braket and ket.
\item Page 523.  (28.36).  Notational inconsistency.  \(\braket{m_1, m_2}{jm -1}\) should read \(\braket{m_1, m_2}{j,m -1}\) 
\item Page 525.  following (28.52).  \(l(l+1)\), \(s(s+1)\), \(j(j+1)\) eigenvalues all missing \(\Hbar^2\).
\item Page 525.  (28.53).  LHS missing \(\Hbar^2\).
\item Page 525.  (28.54).  LHS missing \(\Hbar^2\).
\item Page 525.  (28.57).  RHS missing \(\Hbar\).
\item Page 525.  (28.58).  RHS missing \(\Hbar\).  \(m \pm \inv{2}\) should be \(m \pm 1\).
\item Page 526.  (28.60).  \(\Hbar^2\) missing from both terms.
\item Page 526.  (28.61).  In first term \(\sqrt{l + m_1 + 1}\) should be \(\sqrt{(l + m_1 + 1)(l - m_1)}\).
\end{itemize}

\section{Chapter 29}
This chapter written as if \(\Hbar = 1\), without a statement that this is being done.
\begin{itemize}
\item Page 531.  (29.23).  \(\Hbar\) missing from second two lines.
\item Page 533.  (29.25).  \(\Hbar\) should multiply all.
\item Page 533.  (29.26).  \(\Hbar\) should multiply all (RHS).
\item Page 533.  (29.29).  \(\Hbar\) should multiply all (RHS).
\item Page 533.  (29.30).  \(\Hbar\) should multiply all (RHS).
\item Page 533.  (29.31).  \(\Hbar\) should multiply all (RHS).
\item Page 536.  (29.59).  \(\Hbar\) should multiply RHS.
\item Page 536.  (29.60).  \(\Hbar\) should multiply RHS.  \(\ket{j m + 1}\) should be \(\ket{j, m+1}\).
\item Page 536.  (29.61).  \(\Hbar\) should multiply RHS.  \(\ket{j'm' - 1}\) should be \(\ket{j', m'-1}\).
\item Page 537.  (29.65).  \(\braket{j'm' - 1}{m, q}\) should be \(\braket{j', m'-1}{m, q}\).
\end{itemize}

\section{Chapter 31}
\begin{itemize}
\item Page 562.  (31.56).  \(T' = L T \tilde{M}\) is given for a mixed tensor representation.  This is \(T^\mu_{.\nu}\).  The other mixed representation \(T_{\mu}^{.\nu}\) transforms as \(T' = M T \tilde{L}\).
\end{itemize}

\section{Chapter 32}
\begin{itemize}
\item Page 575.  minor: \(E t - \Bp . \Br\) written instead of \(E t - \Bp \cdot \Br\)
\item Page 576.  minor: \(\spacegrad . \Bj\) instead of \(\spacegrad \cdot \Bj\)
\item Page 577.  (32.23).  \(\Bj\) accidentally includes the divergence.
\item Page 579.  (32.35).  Sign missing in exponential.  Should be \(e^{-i k \cdot x}\) not \(e^{i k \cdot x}\).
\item Page 579.  \(\Hbar \omega_k\) is the energy of the particle, not \(\omega_k\).  There is also an \(\Hbar\) missing in the expression for \(\omega_k\).  That is \(\omega_k = \sqrt{ c^2 \Bk^2 + m_0^2 c^4/\Hbar^2}\).
\item Page 580.  (32.40).  The factor of \(g\) presumed constant ought to be incorporated into \(\chi\) if this is to be consistent with the (32.45) that follows.
\item Page 583.  (32.70).  sign error.  negate integral.
\item Page 584.  (32.74).  sign error in the both the square root and subsequent approximation, which should be \(p_4 = \pm \sqrt{ \Bp^2 + m^2 - i \epsilon} \approx \pm (\omega_p - i \epsilon')\).  (I have added approximately equal for the second part since that was not specified which I found confusing).
\item Page 584.  (32.75-76).  there are multiple sign errors in these equations which should be

\begin{equation}\label{eqn:desaiTyposInnards:60}
\begin{aligned}
\inv{p^2 - m^2 + i\epsilon} 
&\approx \inv{
(p_4 - \omega_p + i\epsilon')
(p_4 + \omega_p - i\epsilon')
} \\
&\approx
\inv{2 \omega_p}\left(
\inv{p_4 - \omega_p + i\epsilon'}
-\inv{p_4 + \omega_p - i\epsilon'}
\right)
\end{aligned}
\end{equation}

Note that an attempt to confirm (32.76) yields

\begin{equation}\label{eqn:desaiTyposInnards:80}
\begin{aligned}
\inv{p_4 - \omega_p + i\epsilon'}
-\inv{p_4 + \omega_p - i\epsilon'}
=
\frac{2 \omega_p - 2 i \epsilon'}{ p^2 - m^2 + i \epsilon + \epsilon^2/4(\Bp^2 + m^2)}
\end{aligned}
\end{equation}

So we need approximations twice for the ``equality''.

\item Page 584.  (before 32.78).  minor: bold script used for \(\Bp \cdot \Br\) on second like of the change of variables.
\item Page 585.  (32.82, 32.83).  minor: \(p_n . x\) instead of \(p_n \cdot x\).
\item Page 585.  (32.82, 32.83).  wrong normalization?  would not we want \(1/\sqrt{2 \omega_{p_n}}\).
\item Page 585.  (32.84).  notation switch?  \(0n\) index whereas \(n0\) used above?  What are the definitions of \(\psi_{0n}\)? that allow the integral to be converted to a sum?
\item Page 586.  (32.88).  minor: \(i \Bk .\) instead of \(i \Bk \cdot\)
\item Page 586.  (32.88).  I calculate a negated \(G(\Bk)\) from (32.88).  Guessing that (32.88), and (32.93 on pg 587) where intended to be negated like done earlier (for example in (32.57)). 
\item Page 587.  (32.93).  minor: \(i \Bk .\) instead of \(i \Bk \cdot\)
\item Page 588-589.  (32.104-105).  This substitution does not appear to work?

\begin{equation}\label{eqn:desaiTyposInnards:100}
\begin{aligned}
&(\spacegrad^2 - \mu^2)(\phi' e^{-\mu r}) \\
&= \inv{r^2} \PD{r}{} \left( r^2 \PD{r}{} \left( \phi' e^{-\mu r} \right) \right) - \mu^2 \phi' e^{-\mu r} \\
&= \inv{r^2} \PD{r}{} \left( r^2 \left( \PD{r}{\phi'} e^{-\mu r} -\mu \phi' e^{-\mu r} \right) \right) - \mu^2 \phi' e^{-\mu r} \\
&= 
\PD{r}{} \left( \left( \PD{r}{\phi'} e^{-\mu r} -\mu \phi' e^{-\mu r} \right) \right) 
+ 2 \inv{r} \left( \PD{r}{\phi'} e^{-\mu r} -\mu \phi' e^{-\mu r} \right)  
- \mu^2 \phi' e^{-\mu r} \\
&= 
\PDSq{r}{\phi'} e^{-\mu r} -\mu \PD{r}{\phi'} e^{-\mu r} 
-\mu \left( \left( \PD{r}{\phi'} e^{-\mu r} - \cancel{\mu \phi' e^{-\mu r}} \right) \right) 
+ 2 \inv{r} \left( \PD{r}{\phi'} e^{-\mu r} -\mu \phi' e^{-\mu r} \right)  
- \cancel{\mu^2 \phi' e^{-\mu r}} \\
&=
\PDSq{r}{\phi'} e^{-\mu r} -\mu \PD{r}{\phi'} e^{-\mu r} 
-\mu \PD{r}{\phi'} e^{-\mu r} 
+ 2 \inv{r} \left( \PD{r}{\phi'} e^{-\mu r} -\mu \phi' e^{-\mu r} \right) 
\\
&=
\left( \PDSq{r}{\phi'} 
+ 2 \inv{r} \PD{r}{\phi'} \right) e^{-\mu r} 
-2 \mu \PD{r}{\phi'} e^{-\mu r} 
-2 \mu \inv{r} \phi' e^{-\mu r} 
\\
&=
(\spacegrad^2 \phi') e^{-\mu r}
- 2 \mu \left( \PD{r}{\phi'} + \inv{r} \phi' \right) e^{-\mu r} 
\\
&=
(\spacegrad^2 \phi') e^{-\mu r}
- 2 \frac{\mu}{r} \left( \PD{r}{(r \phi')} \right) e^{-\mu r} 
\\
\end{aligned}
\end{equation}

There is an extra term here that does not show up in (32.105) with this transformation.  Can that be argued away somehow?

\item Page 589.  (32.107).  minor: \(i \Bk_f .\) instead of \(i \Bk_f \cdot\)
\item Page 589.  (32.108).  minor: \(i \Bq .\) instead of \(i \Bq \cdot\)
\item Page 590.  (after 32.118).  minor: \(i \Bq .\) instead of \(i \Bq \cdot\)
\item Page 591.  (32.124).  minor: \(i \Bq .\) instead of \(i \Bq \cdot\)
\end{itemize}

NOTE: up to commit efc6cd3bfee91f43949016f8ba851de273e4fa8d of these notes emailed to Desai May 10, 2011.
 
\section{Chapter 33}

\begin{itemize}
\item Page 597.  (before 33.5).  Last paragraph references chapter 12.  Chapter 32 meant here?  (or chapter 4).
\end{itemize}

\section{Chapter 35}

\begin{itemize}
\item Page 635.  (35.46).  minor: Bold on gamma. 
\item Page 636.  (35.50).  minor: Bold on gamma. 
\item Page 636-638.  (35.51-35.58).  minor: incomplete notation switch.  This chapter uses \(E_p\) instead of \(\Abs{E}\), but many formulas on these pages continue to use the \(\Abs{E}\) notation from chapter 33, even mixing the two in some places.
\item Page 643.  (35.107).  \(e_\mu^{\nu}\) should be \(e_\mu^{.\nu}\).  There are also some missing positional indicators in (35.105) and (35.106).
\item Page 643.  (35.114).  \(\overbar{\psi}'(x')\psi(x')\) should be \(\overbar{\psi}'(x')\psi'(x')\).
\item Page 645.  (35.134).  Wrong sign on \(\gamma_5\).  It should be \(-i \gamma^1 \gamma^2 \gamma^3 \gamma^4\).
\end{itemize}

\section{Chapter 36}
\begin{itemize}
\item Page 647.  (36.7).  Sign errors on both terms in the second line.
\item Page 648.  (36.12).  \(+i e \Bsigma'\) should be negative.
\item Page 648.  (36.16-17).  \(\Hbar\)s should be omitted for consistency.
\item Page 648.  (36.16-17).  It appears that the \(- e \Bsigma'\) should be \(+ e \Bsigma'\).
\item Page 649.  (36.22).  \(+i e \Balpha \cdot \BE\) should be negative (although this whole term drops out later after the non-relativistic expectation is taken).
\item Page 649.  (36.25).  Same sign error on the \(\Balpha \cdot \BE\) term.
\item Page 650.  Equation reference to (36.32) should be (32.31)
\item Page 650.  (36.35).  \(1/2m\) in the denominator should be omitted.
\item Page 653.  (36.60,61).  \(\Bsigma\) should be \(\Bsigma'\)
\item Page 653.  (36.63).  Should read \(p_\mu \rightarrow p_\mu - e A_\mu\).
\item Page 653.  (36.64).  Should read \(D_\mu = \partial_\mu + i e A_\mu\).
\item Page 654.  (36.78).  Missing factor of two.  \(-e \sigma^{\mu\nu} F_{\mu\nu}\) should be \(-\frac{e}{2} \sigma^{\mu\nu} F_{\mu\nu}\).
\item Page 655.  (36.80).  \(\partial - e A\) should read \(\partial + i e A\).
\end{itemize}
