%\input{../peeter_prologue_print.tex}

%\label{chap:desaiTypos}
%\useCCL
%\blogpage{http://sites.google.com/site/peeterjoot/math2010/desaiTypos.pdf}
%\date{Oct 31, 2010}
%\revisionInfo{desaiTypos.tex}

\chapter{Believed to be typos in Desai's QM Text}

\beginArtNoToc

I have found some of the obvious stuff in my reading of \citep{desai2009quantum}.  Prof.\ Vatche Deyirmenjian who teaches our PHY356 course has pointed out still more (and pointed out where I'd identified the wrong source for some typos).

\section{Chapter 1.}

\begin{itemize}
\item Page 1.  Vatche: The Hermitian, not complex conjugate, of $\ket{}$ is $\bra{}$.
\item Page 5-6.  Vatche: Change the $\conj$ in (1.26), (1.31), and (1.33) to a dagger.
\item Page 7.  Text before (1.43).  $\alpha$ instead of $a$ used.
\item Page 19.  Equation (1.122).  $\dagger$s omitted after first equality.
\end{itemize}

\section{Chapter 2.}
\begin{itemize}
\item Page 40.  Text before (2.137).  Reference to equation (2.133) should be (2.135)
\item Page 53.  Is the "Also show that" here correct?  I get a different answer.
\end{itemize}

\section{Chapter 3.}
\begin{itemize}
\item Page 61.  Equation (3.51).  $1/\hbar$ missing.
\item Page 62.  Equation (3.58).  Vatche: Remove the $U_I$ operators from Eq. (3.58)
\item Page 66.  Equation (3.92).  $-(d/dt \bra{\alpha}) \ket{\alpha}$ should be $+\ket{\alpha} d/dt \bra{\alpha}$.
\item Page 66.  Equation (3.93).  $H$ on wrong side of $\bra{\alpha}$ 
\item Page 74,76.  Vatche: remove the extra brackets from Eq (4.9) and (4.21).
\item Page 79.  Vatche: "The probability of finding this particle" should read "The probability density for this state at point x is"
\end{itemize}

\section{Chapter 4.}
\begin{itemize}
\item Page 81.  Equation (4.52).  Should be $-2\alpha$ in the exponent.
\item Page 82.  Equation (4.65).  Vatche: a $1/\sqrt{2\pi}$ is missing before the integral.  Note that without this (4.67) appears incorrect (off by a factor of $\sqrt{2\pi}$, but the error is really just in (4.65).
\item Page 82.  Equation (4.67).  Vatche: the negative sign should appear inside the large square brackets.
%%% NO:  See above.
%\item Page 82.  Equation (4.67).  $2\alpha$ in the denominator of the normalization should be $\alpha'/\pi$.
\item Page 83.  Equation (4.74).  A normalized wave function isn't required for the discussion, but if that was intended, a $1/\sqrt{2\pi}$ factor is missing.
\item Page 83-84.  Vatche: In (4.67) and (4.77), the derivative should be evaluated at $k=k_0$.
\item Page 86.  Equation (4.99).  Extra brace in the exponent.
\item Page 87.  Equation (4.106).  Extra brace in the exponent.
\item Page 89.  Equation (4.124-4.130).  Vatche: $C e^{\pm \sqrt{\mu}\phi}$ is not a solution to (4.122).  This should be $Q(\phi) = C e^{i \sqrt{\mu} \phi}$ and (4.126) should be $\sqrt{\mu} = m$.  This fixes the apparent error in sign in equations 4.129 and 4.130 which are correct as is.
\item Page 92.  Equation (4.158).  Vatche: should read $P_l(1) = 1$.
\item Page 93.  Equation (4.169).  conjugation missing for $Y_{lm}$.  $Y_{l'm'}$ is missing prime on the $l$ index.
\item Page 95.  Second line of text.  Language choice?  "We now implement".  perhaps utilize would be better?
\item Page 95.  Text before (4.193).  $i$ is in bold.
\item Page 96.  Text before (4.196).  $i$ is in bold.
\item Page 97.  (4.205).  $i$ is in bold.
\item Page 97.  (4.207-209).  $\Bi$, and $\Bj$s aren't in bold like $\Bk$
\item Page 101.  (4.245).  The right side should read $Y_{l,m+1}$
\item Page 101.  (4.239-240).  The approach here is unclear.  FIXME: incorporate lecture notes from class that did this using braket notation.
\item Page 102.  (4.248-249).  Commas missing to separate $l$, and $m\pm 1$ in the kets.
\end{itemize}

\section{Chapter 5.}
\begin{itemize}
\item Page 109.  (5.49). Remove bold font in right hand side state $\ket{\chi_{n+}}$.
\item Page 113.  (5.86). One $\sigma$ isn't in bold.
\item Page 114.  (5.100). $\chi$ is in bold.
\item Page 115.  Text before (5.106). $\alpha$ in bold.
\item Page 118.  Switch of notation in problem 5 for ensemble averages.  $[S_i]$ used instead of $\expectation{S_i}_{\text{av}}$.
\end{itemize}

\section{Chapter 6.}
\begin{itemize}
\item Page 120.  $\phi$ in bold.  $A$ not in bold.
\item Page 123.  (6.26).  $1/i \hbar$ factor missing on RHS.
\item Page 124.  Text before (6.37).  You say canonical momenta $P_k$, but call these mechanical momenta on prev page.
\item Page 125.  (6.41).  Some $\psi$s are in bold.
\item Page 126.  (6.49).  There's no mention that $\BB$ is constant, leaving it unclear how the gauge condition and how the curl of $\BA$ reproduces $\BB$.  This would also help clarify how you are able to write $\Bmu \cdot \BB = \BB \cdot \Bmu$.
\item Page 128.  (6.65).  $\Bmu \cdot \BL$ should be $\Bmu \cdot \BB$.
\item Page 129.  (6.75).  $\Bmu \cdot \BL$ should be $\Bmu \cdot \BB$.
\item Page 130.  (6.80).  integral looks like it should be $\int_{\Br' = \Br_0}^\Br \frac{e}{c \hbar} \BA(\Br') \cdot d\Br'$.  ie: Clarify bounds, and add a factor of $c$ in the denominator which is required for the cancellation of (6.82).
\item Page 131.  (6.81,6.86).  Factors of $c$s should be with each of the $\hbar$s.
\item Page 131.  Problem 1.  bold missing on $\BE$.
\end{itemize}

\section{Chapter 8.}
\begin{itemize}
\item Page 143.  (8.58).  $\beta$ should be negated.
\item Page 159.  (8.6.3).  Two references to Chapter 2 should be Chapter 4.
\item Page 160.  (8.199).  Want $\hbar^2$ not $\hbar$ in expression for $k$.
\item Page 162.  (Fig 8.9).  Figure is backwards compared to text (a bump instead of a well).
\item Page 165.  (8.235).  Extra $R_l$ factor inside parens.
\end{itemize}

\section{Chapter 9.}
\begin{itemize}
\item Page 174.  (9.5).  Have $\hbar/2m\omega$ instead of $\hbar m \omega/2$ in expression for $P$.
\item Page 181.  (9.57).  Factor of two missing.  Want $\frac{\alpha}{2 \sqrt{\pi}}$.
\item Page 186.  (Problem 10).  Sequencing the text and problems is off.  The green's function technique isn't introduced until chapter 10.
\end{itemize}

\section{Chapter 10.}
\begin{itemize}
\item Page 189.  (10.22).  It would be nice to have a reference to the appendix (ie: 10.100) for the chapter so that this identity isn't pulled out of a magic hat.
\item Page 192.  (10.44, 10.45).  $2 \alpha {\alpha^\conj}'$ should be $\alpha {\alpha^\conj}' + \alpha' \alpha^\conj$
\item Page 193.  (10.51).  Application (slowly, step by step explicitly) of 10.100 to expand the $e^{\frac{i}{\hbar}(p_0 X - x_0 P)}$ in the braket gives

\begin{align*}
\bra{x} e^{\frac{i}{\hbar}(p_0 X - x_0 P)} \ket{0}
&=
\bra{x} e^{\frac{i}{\hbar}p_0 X }
e^{-\frac{i}{\hbar}x_0 P}
e^{-\frac{i}{2\hbar}x_0 p_0 \antisymmetric{X}{P}}
\ket{0} \\
&=
\bra{x} e^{\frac{i}{\hbar}p_0 X }
e^{-\frac{i}{\hbar}x_0 P}
e^{\frac{x_0 p_0}{2} }
\ket{0} \\
&=
e^{\frac{x_0 p_0}{2} }
\bra{x} e^{\frac{i}{\hbar}p_0 X} 
e^{-\frac{i}{\hbar}x_0 P}
\ket{0} \\
&=
e^{\frac{x_0 p_0}{2} }
\left(
\bra{0} 
e^{\frac{i}{\hbar}x_0 P}
e^{-\frac{i}{\hbar}p_0 X} 
\ket{x}\right)^\conj \\
&=
e^{\frac{x_0 p_0}{2} }
\left(
\bra{0} 
e^{\frac{i}{\hbar}x_0 P}
\ket{x}
e^{-\frac{i}{\hbar}p_0 x} 
\right)^\conj \\
&=
e^{\frac{x_0 p_0}{2} } e^{\frac{i}{\hbar}p_0 x} 
\left(
\bra{0} 
e^{\frac{i}{\hbar}x_0 P}
\ket{x}
\right)^\conj \\
&=
e^{\frac{x_0 p_0}{2} } e^{\frac{i}{\hbar}p_0 x} 
\left(
\braket{0}{x - x_0}
\right)^\conj \\
&=
e^{\frac{x_0 p_0}{2} } e^{\frac{i}{\hbar}p_0 x} 
\braket{x - x_0}{0} \\
&=
e^{\frac{x_0 p_0}{2} } e^{\frac{i}{\hbar}p_0 x} 
\psi_0(x - x_0, 0)
\end{align*}

This is the same as (10.51) with the exception of a real scalar constant $e^{ x_0 p_0/2}$ multiplying the wave function.  Because of this I think that (10.51) should be a proportionality statement, and not an equality as in

\begin{align*}
\bra{x} e^{\frac{i}{\hbar}(p_0 X - x_0 P)} \ket{0} \propto
e^{\frac{i}{\hbar}p_0 x} \psi_0(x - x_0, 0)
\end{align*}

(ie: building this additional factor into the wave function normalization instead).
\item Page 196.  (text after 10.76).  Looks like reference to Chapter 9, should be Chapter 9 problem 5.

\item Page 197.  (text after 10.85).  Reference to Chapter 1 should be Chapter 2.
\end{itemize}

\section{Chapter 26.}
\begin{itemize}
\item Page 486.  (26.60).  $\Bn \cross \Br \cdot \spacegrad$ ought to have braces and read $(\Bn \cross \Br) \cdot \spacegrad$.
\item Page 496.  (26.154).  Remove $Y_{l'm}(\theta, \phi)$ term from the integral.
\end{itemize}

\section{Chapter 31.}
\begin{itemize}
\item Page 562.  (31.56).  $T' = L T \tilde{M}$ is given for a mixed tensor representation.  This is $T^\mu_{.\nu}$.  The other mixed representation $T_{\mu}^{.\nu}$ transforms as $T' = M T \tilde{L}$.
\end{itemize}

\section{Chapter 32.}
\begin{itemize}
\item Page 575.  minor: $E t - \Bp . \Br$ written instead of $E t - \Bp \cdot \Br$
\item Page 576.  minor: $\spacegrad . \Bj$ instead of $\spacegrad \cdot \Bj$
\item Page 577.  (32.23).  $\Bj$ accidentally includes the divergence.
\item Page 579.  (32.35).  Sign missing in exponential.  Should be $e^{-i k \cdot x}$ not $e^{i k \cdot x}$.
\item Page 579.  $\hbar \omega_k$ is the energy of the particle, not $\omega_k$.  There's also an $\hbar$ missing in the expression for $\omega_k$.  That is $\omega_k = \sqrt{ c^2 \Bk^2 + m_0^2 c^4/\hbar^2}$.
\item Page 580.  (32.40).  The factor of $g$ presumed constant ought to be incorporated into $\chi$ if this is to be consistent with the (32.45) that follows.
\item Page 586.  (32.88).  minor: $i \Bk .$ instead of $i \Bk \cdot$
\item Page 587.  (32.93).  minor: $i \Bk .$ instead of $i \Bk \cdot$
\item Page 589.  (32.107).  minor: $i \Bk_f .$ instead of $i \Bk_f \cdot$
\item Page 589.  (32.108).  minor: $i \Bq .$ instead of $i \Bq \cdot$
\item Page 590.  (after 32.118).  minor: $i \Bq .$ instead of $i \Bq \cdot$
\item Page 591.  (32.124).  minor: $i \Bq .$ instead of $i \Bq \cdot$
\end{itemize}

\EndArticle
