%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%
%
%\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

%\chapter{PHY456H1F: Quantum Mechanics II.  Lecture 5 (Taught by Prof J.E. Sipe).  Pertubation theory and degeneracy.  Review of dynamics}
\index{pertubation theory}
\index{degeneracy}
\index{dynamics}
\label{chap:qmTwoL5}
\blogpage{http://sites.google.com/site/peeterjoot/math2011/qmTwoL5.pdf}
%\date{Sept 23, 2011}





\section{Issues concerning degeneracy}

\paragraph{When the perturbed state is non-degenerate}

Suppose the state of interest is non-degenerate but others are

FIXME: diagram.  states designated by dashes labeled \(n1\), \(n2\), \(n3\) degeneracy \(\alpha = 3\) for energy \(E_n^{(0)}\).

This is no problem except for notation, and if the analysis is repeated we find

\begin{align}\label{eqn:qmTwoL5:10}
E_s &= E_s^{(0)} + \lambda {H'}_{ss} + \lambda^2
\sum_{m \ne s, \alpha}
\frac{\Abs{{H'}_{m \alpha ; s}}^2 }
{ E_s^{(0)} - E_m^{(0)} }
+ \cdots
\\
\ket{\overbar{\psi}_s} &= \ket{{\psi_s}^{(0)}} + \lambda
\sum_{m \ne s, \alpha}
\frac{{H'}_{m \alpha ; s}}
{ E_s^{(0)} - E_m^{(0)} } \ket{{\psi_{m \alpha}}^{(0)}}
+ \cdots,
\end{align}

where

\begin{equation}\label{eqn:qmTwoL5:30}
{H'}_{m \alpha ; s} =
\bra{{\psi_{m \alpha}}^{(0)}} H' \ket{{\psi_{s \alpha}}^{(0)}}
\end{equation}

\paragraph{When the perturbed state is also degenerate}

FIXME: diagram.  states designated by dashes labeled \(n1\), \(n2\), \(n3\) degeneracy \(\alpha = 3\) for energy \(E_n^{(0)}\), and states designated by dashes labeled \(s1\), \(s2\), \(s3\) degeneracy \(\alpha = 3\) for energy \(E_s^{(0)}\).

If we just blindly repeat the derivation for the non-degenerate case we would obtain

\begin{align}\label{eqn:qmTwoL5:50}
E_s &= E_s^{(0)} + \lambda {H'}_{s1 ; s1}
+ \lambda^2
\sum_{m \ne s, \alpha}
\frac{\Abs{{H'}_{m \alpha ; s1}}^2 }
{ E_s^{(0)} - E_m^{(0)} }
+ \lambda^2
\sum_{\alpha \ne 1}
\frac{\Abs{{H'}_{s \alpha ; s1}}^2 }
{ E_s^{(0)} - \color{red}{E_s^{(0)}} }
+ \cdots
\\
\ket{\overbar{\psi}_s} &= \ket{{\psi_s}^{(0)}}
+ \lambda
\sum_{m \ne s, \alpha}
\frac{{H'}_{m \alpha ; s}}
{ E_s^{(0)} - E_m^{(0)} } \ket{{\psi_{m \alpha}}^{(0)}}
+ \lambda
\sum_{\alpha \ne s1}
\frac{{H'}_{s \alpha ; s1}}
{ E_s^{(0)} - \color{red}{E_s^{(0)}} } \ket{{\psi_{s \alpha}}^{(0)}}
+ \cdots,
\end{align}

where

\begin{equation}\label{eqn:qmTwoL5:70}
{H'}_{m \alpha ; s1} =
\bra{{\psi_{m \alpha}}^{(0)}} H' \ket{{\psi_{s1}}^{(0)}}
\end{equation}

Note that the \(E_s^{(0)} -\color{red}{E_s^{(0)}}\) is NOT a typo, and why we run into trouble.  There is one case where a perturbation approach is still possible.  That case is if we happen to have

\begin{equation}\label{eqn:qmTwoL5:90}
\bra{{\psi_{m \alpha}}^{(0)}} H' \ket{{\psi_{s1}}^{(0)}} = 0.
\end{equation}

That may not be obvious, but if one returns to the original derivation, the right terms cancel so that one will not end up with the \(0/0\) problem.

FIXME: performing this derivation outside of class (below), it was found that we do not need the matrix elements of \(H'\) to be diagonal, but just need

\begin{equation}\label{eqn:qmTwoL5:90b}
\bra{{\psi_{s \alpha}}^{(0)}} H' \ket{{\psi_{s \beta}}^{(0)}} = 0, \quad \mbox{for \(\beta \ne \alpha\)}.
\end{equation}

That is consistent with problem set III where we did not diagonalize \(H'\), but just the subset of it associated with the degenerate states.  I am unsure now if \eqnref{eqn:qmTwoL5:90} was copied in error or provided in error in class, but it definitely appears to be a more severe requirement than actually needed to deal with perturbation of a state found in a degenerate energy level.

\input{approx/qmTwoL5a}

\paragraph{Diagonalizing the perturbation Hamiltonian}

Suppose that we do not have this special zero condition that allows the perturbation treatment to remain valid.  What can we do.  It turns out that we can make use of the fact that the perturbation Hamiltonian is Hermitian, and diagonalize the matrix

\begin{equation}\label{eqn:qmTwoL5:110}
\bra{{\psi_{s \alpha}}^{(0)}} H' \ket{{\psi_{s \beta}}^{(0)}}
\end{equation}

In the example of a two fold degeneracy, this amounts to us choosing not to work with the states

\begin{equation}\label{eqn:qmTwoL5:130}
\ket{\psi_{s1}^{(0)}}, \ket{\psi_{s2}^{(0)}},
\end{equation}

both some linear combinations of the two
\begin{align}\label{eqn:qmTwoL5:150}
\ket{\psi_{sI}^{(0)}} &= a_1 \ket{\psi_{s1}^{(0)}} + b_1 \ket{\psi_{s2}^{(0)}} \\
\ket{\psi_{sII}^{(0)}} &= a_2 \ket{\psi_{s1}^{(0)}} + b_2 \ket{\psi_{s2}^{(0)}}
\end{align}

In this new basis, once found, we have

\begin{equation}\label{eqn:qmTwoL5:170}
\bra{{\psi_{s \alpha}}^{(0)}} H' \ket{{\psi_{s \beta}}^{(0)}} = \calH_\alpha \delta_{\alpha \beta}
\end{equation}

Utilizing this to fix the previous, one would get if the analysis was repeated correctly

\begin{align}\label{eqn:qmTwoL5:190}
E_{s\alpha} &= E_s^{(0)} + \lambda {H'}_{s\alpha ; s\alpha}
+ \lambda^2
\sum_{m \ne s, \beta}
\frac{\Abs{{H'}_{m \beta ; s \alpha}}^2 }
{ E_s^{(0)} - E_m^{(0)} }
+ \cdots
\\
\ket{\overbar{\psi}_{s \alpha}} &= \ket{{\psi_{s \alpha}}^{(0)}}
+ \lambda
\sum_{m \ne s, \beta}
\frac{{H'}_{m \beta ; s \alpha}}
{ E_s^{(0)} - E_m^{(0)} } \ket{{\psi_{m \beta}}^{(0)}}
+ \cdots
\end{align}

FIXME: why do we have second order in \(\lambda\) terms for the energy when we found those exactly by diagonalization?  We found there that the perturbed energy eigenvalues were multivalued with values \(E_{s\alpha} = E_s^{(0)} + \lambda {H'}_{s\beta ; s\beta}\) for all degeneracy indices \(\beta\).  Have to repeat the derivation for these more carefully to understand this apparent discrepancy.

We see that a degenerate state can be split by applying perturbation.

FIXME: diagram.  \(E_s^{(0)}\) as one energy level without perturbation, and as two distinct levels with perturbation.

\paragraph{guess} I will bet that this is the origin of the spectral line splitting, especially given that an atom like hydrogen has degenerate states.

