Copyright \copyright 2016 Peeter Joot
All Rights Reserved

This work is licenced under Creative Commons Attribution-NonCommercial-NoDerivs 3.0 (CC BY-NC-ND 3.0 \ccbyncnd) as described in \href{http://creativecommons.org/licenses/by-nc-nd/3.0/}{http://creativecommons.org/licenses/by-nc-nd/3.0/}

You are free \textbf{to Share}, to copy, distribute and transmit the work

\paragraph{Under the following conditions}:

\begin{itemize}
\item \ccAttribution Attribution.  You must attribute the work in the manner specified by the author or licensor (but not in any way that suggests that they endorse you or your use of the work).
\item \ccNonCommercial Noncommercial. You may not use this work for commercial purposes.
\item \ccNoDerivatives No Derivative Works. You may not alter, transform, or build upon this work.
\end{itemize}

With the understanding that:

\begin{itemize}
\item \textbf{Waiver}. Any of the above conditions can be waived if you get permission from the copyright holder.
\item \textbf{Public Domain}. Where the work or any of its elements is in the public domain under applicable law, that status is in no way affected by the license.
\item \textbf{Other Rights}. In no way are any of the following rights affected by the license:

\begin{itemize}
\item Your fair dealing or fair use rights, or other applicable copyright exceptions and limitations;
\item The author's moral rights;
\item Rights other persons may have either in the work itself or in how the work is used, such as publicity or privacy rights.
\end{itemize}

\item \textbf{Notice}. For any reuse or distribution, you must make clear to others the license terms of this work. The best way to do this is with a link to this web page (\href{http://creativecommons.org/licenses/by-nc-nd/3.0/}{http://creativecommons.org/licenses/by-nc-nd/3.0/})
\end{itemize}
