%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%
%
\chapter{Multimedia figures}

A table of animated figures in this document.  These require Adobe Reader (google's online pdf reader will not display these for example).  Some patience is required to view these (waiting for the underlying javascript to run before clicking the play button ... on my machine I can hear acrobat churning away on the CPU at these points.)

\begin{itemize}
\item Fluids with densities of mercury and water in the lower (1) and upper (2) layers respectively.  \cref{fig:continuumProblemSet2:continuumProblemSet2Animation1}
\item Fluids with densities of water and mercury in the lower (1) and upper (2) layers respectively.  \cref{fig:continuumProblemSet2:continuumProblemSet2Animation2}
\item Time evolution of channel flow velocity profile after turning on a constant pressure gradient.  \cref{fig:channelFlowWithStepPressureGradient:channelFlowWithStepPressureGradientFig3}
\item Early time evolution of channel flow velocity profile after turning on a constant pressure gradient.  \cref{fig:channelFlowWithStepPressureGradient:channelFlowWithStepPressureGradientFig4}
\item Animation of Couette flow, with continuous variation of outer angular velocity. \cref{fig:couetteFlow:couetteFlowFig6}.
\item Animating the two cylinder velocity field for a set of parameter values.  \cref{fig:twoCylinders:twoCylindersFig3}.
\end{itemize}

%-------------------------------------------------------
