%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

Simon (our TA) stated \cref{eqn:relativisticElectrodynamicsT1:130} without justification.  Here's a little justification for the frequency four vector.

We know some of it from the QM context, and if we have been reading ahead know a bit of this from our text \citep{landau1980classical} (the energy momentum four vector relationships).  Let us go back to the classical electromagnetism and recall what we know about the relation of frequency and wave numbers for continuous fields.  We want solutions to Maxwell's equation in vacuum and can show that such solution also implies that our fields obey a wave equation

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:131}
\inv{c^2} \frac{\partial^2 \Psi}{\partial t^2} - \spacegrad^2 \Psi = 0,
\end{equation}

where $\Psi$ is one of $\BE$ or $\BB$.  We have other constraints imposed on the solutions by Maxwell's equations, but require that they at least obey \eqnref{eqn:relativisticElectrodynamicsT1:131} in addition to these constraints.

With application of a spatial Fourier transformation of the wave equation, we find that our solution takes the form

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:132}
\Psi = (2 \pi)^{-3/2} \int \tilde{\Psi}(\Bk, 0) e^{i (\omega t \pm \Bk \cdot \Bx) } d^3 \Bk.
\end{equation}

If one takes this as a given and applies the wave equation operator to this as a test solution, one finds without doing the Fourier transform work that we also have a constraint.  That is

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:133}
\inv{c^2} (i \omega)^2 \Psi - (\pm i \Bk)^2 \Psi = 0.
\end{equation}

So even in the continuous field domain, we have a relationship between frequency and wave number.  We see that this also happens to have the form of a lightlike spacetime interval

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:134}
\frac{\omega^2}{c^2} - \Bk^2 = 0.
\end{equation}

Also recall that the photoelectric effect imposes an experimental constraint on photon energy, where we have

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:135}
E = h \nu = \frac{h}{2\pi} 2 \pi \nu = \Hbar \omega
\end{equation}

Therefore if we impose a mechanics like $P = (E/c, \Bp)$ relativistic energy-momentum relationship on light, it then makes sense to form a nilpotent (lightlike) four vector for our photon energy.  This combines our special relativistic expectations, with the constraints on the fields imposed by classical electromagnetism.  We can then write for the photon four momentum

\begin{equation}\label{eqn:relativisticElectrodynamicsT1:136}
P = \left( \frac{\Hbar \omega}{c}, \Hbar k \right)
\end{equation}

