%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%\chapter{Spacetime, events, worldlines, proper time, invariance}
\label{chap:relativisticElectrodynamicsL2}
%\blogpage{http://sites.google.com/site/peeterjoot/math2011/relativisticElectrodynamicsL2.pdf}
%\date{Jan 12, 2011}

\paragraph{Reading}

No reading from \citep{landau1980classical} appears to have been assigned, but relevant stuff can be found in chapter 1.

From \href{http://www.physics.utoronto.ca/~poppitz/epoppitz/PHY450_files/RelEM12-26.pdf}{Professor Poppitz's lecture notes}, we have reading: pp.12-26: spacetime, spacetime points, worldlines, interval (12-14); invariance of infinitesimal intervals (15-17).


\section{Einstein's relativity principle}

\begin{enumerate}

\item Replace Galilean transformations between coordinates in differential inertial frames with Lorentz transforms between $(\Bx, t)$.  Postulate that these constitute the symmetries of physics.  Recall that Galilean transformations are symmetries of the laws of non-relativistic physics.  

\item Speed of light $c$ is the same in all inertial frames.  Phrased in this form, relativity leads to ``relativity of simultaneity''.

PICTURE: Three people on a platform, at positions $1,3,2$, all with equidistant separation.  This stationary frame is labeled $O$.  1 and 2 flash light signals at the same time and in frame $O$ the reception of the light signal by 3 is observed as arriving at 3 simultaneously.

Now introduce a moving frame with origin $O'$ moving along the positive x axis.  To a stationary observer in $O'$ the three guys are seen to be moving in the $-x$ direction.  The middle guy (3) is eventually going to be seen to receive the light signal by this $O'$ observer, but less time is required for the light to get from 1 to 3, and more time is required for the light to get from 2 to 1 (3 is moving away from the light according to the $O'$ observer).  Because the speed of light is perceived as constant for all observers, the perception is then that the light must arrive at 3 at different times.

This is very non-intuitive since we are implicitly trained by our surroundings that Galilean transformations govern mechanical behavior.

In $O$, 1 and 2 send light signals simultaneously while in $O'$ 1 sends light later than 2.  The conclusion, rather surprisingly compared to intuition, is that simultaneity is relative.

\end{enumerate}

\makequestion{On symmetries.}{question:relativisticElectrodynamicsL2:1}{
\paragraph{Q:}
A comment made that the symmetries impose the dynamics, and the symmetries provided the form of the Lagrangian in classical physics.  My question to this comment was

``When we have transformations that leave the Lagrangian unchanged (a symmetry), we have a conserved current.  I have done various exercises to compute those currents for various types of transformations (translation, spacetime translation, rotation, boosts, ...), but can not think of a way that the Lagrangian itself is defined these sorts symmetries.  Can you elaborate on what you mean by this?''

\paragraph{A:}

Ah, you see, what I meant by that is the following. For a free particle $\LL$ should depend on $\Bx$, $\dot{\Bx}$ and $t$.  Homogeneity of space and time do not allow to have $x$ and $t$ dependence and isotropy of space only permits dependence on $\Abs{\dot{\Bx}}$. Finally, Gallilean relativity only allows $\LL = \dot{\Bx}^2$ (times a constant). (See \citep{landau1960classical} vol 1 or \href{http://www.physics.utoronto.ca/~poppitz/epoppitz/PHY354_files/CMpp13.1-27.pdf}{my notes on PHY354 website}, p. 23-27).

So what was used is:

\begin{itemize}
\item Having only dependence on $x$ and $dx/dt$.
\item Spacetime homogeneity/isotropy.
\item Gallilean relativity.
\end{itemize}

Similar story holds in relativity, as we will see. 


}


