%
%
%
% Copyright © 2012 Peeter Joot
% All Rights Reserved
%
% This file may be reproduced and distributed in whole or in part, without fee, subject to the following conditions:
%
% o The copyright notice above and this permission notice must be preserved complete on all complete or partial copies.
%
% o Any translation or derived work must be approved by the author in writing before distribution.
%
% o If you distribute this work in part, instructions for obtaining the complete version of this file must be included, and a means for obtaining a complete version provided.
%
%
% Exceptions to these rules may be granted for academic purposes: Write to the author and ask.
%
%
%
%\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

%\usepackage[english]{babel}
%\usepackage{media9}
\chapter{Surface for spinning bucket of water.}

\label{chap:constantSpinSurfaces}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/constantSpinSurfaces.pdf}
\date{Apr 29, 2012}
\gitRevisionInfo{constantSpinSurfaces}
\keywords{Navier-Stokes, Bernoulli's theorem, PHY454H1S, PHY454H1}

\beginArtWithToc
%\beginArtNoToc
%\wordpresscategory{}

\section{Motivation.}

Here's a problem from the 2009 phy1530 final that was appropriate for exam prep for this course too.  It also serves as a nice example of how to determine a surface as a function of pressure, something I want to do for the non-bottomless coffee problem to be attempted.

\section{Statement}

An undergraduate student is assigned a problem about an ideal fluid rotating at a constant angular velocity $\Omega$ under gravity $g$.  The velocity field is $\Bu = (-\Omega y, \Omega x, 0)$.  Here, $x$ and $y$ are horizontal and $z$ points up.  The student is supposed to find the surfaces of constant pressure, and hence the shape of the free surface of water in a rotating bucket.  The free surface corresponds to the surface for which $p = p_0$, where $p_0$ is the atmospheric pressure.  Surface tension is neglected.

\begin{enumerate}
\item On their homework assignment, the student writes:

``By Bernoulli's equation:

\begin{equation}\label{eqn:constantSpinSurfaces:10}
B = \frac{p}{\rho} + \inv{2}u^2 + g z
\end{equation}

where $B$ is a constant.  So the constant pressure surface at $p = p_0$ is 

\begin{equation}\label{eqn:constantSpinSurfaces:30}
z = \left( \frac{B}{g} - \frac{p_0}{\rho g}
\right)
- \frac{\Omega^2 }{2 g} \left( x^2 + y^2 \right).
\end{equation}
''

But this seems to show that the surface of the water in a rotating bucket is \textit{highest in the middle}!  What is wrong with the student's argument?

\item
Write down the Euler equations in component form and integrate them directly to find the pressure $p$, and hence obtain the correct parabolic shape for the free surface.
\end{enumerate}

\section{Solution.  Part 1.}

Let's recall how we derived Bernoulli's theorem.  We started with Navier-Stokes and used the identity

\begin{equation}\label{eqn:constantSpinSurfaces:50}
(\Bu \cdot \spacegrad ) \Bu = \spacegrad \inv{2} \Bu^2 + (\spacegrad \cross \Bu) \cross \Bu.
\end{equation}

Navier-Stokes for a steady state incompressible flow, with external body force per unit volume $\rho \Bg = -\rho \spacegrad \chi$ take the form

\begin{equation}\label{eqn:constantSpinSurfaces:70}
\spacegrad \inv{2} \Bu^2 + (\spacegrad \cross \Bu) \cross \Bu
= -\inv{\rho} \spacegrad p + \nu \spacegrad^2 \Bu - \spacegrad \chi.
\end{equation}

For the non-viscous (``dry-water'') case where we take $\mu = \nu \rho = 0$, and treat the density $\rho$ as a constant we find

\begin{equation}\label{eqn:constantSpinSurfaces:90}
\Bu \cross (\spacegrad \cross \Bu)
=
\spacegrad 
\left( 
\inv{2} \Bu^2 + \frac{p}{\rho} + \chi
\right).
\end{equation}

Observe that we only arrive at Bernoulli's theorem if the flow is also irrotational (as well as incompressible and non-viscous), as we require an irrotational flow where $\spacegrad \cdot \Bu = 0$ to claim that the gradient on the RHS is zero.  
%
%The most general claim that we can make, even for irrotational flows is that we have
%
%\begin{equation}\label{eqn:constantSpinSurfaces:110}
%0 = \Bu \cdot
%\spacegrad 
%\left( 
%\inv{2} \Bu^2 + \frac{p}{\rho} + \chi
%\right).
%\end{equation}
%
%That holds even for flows that are not irrotational, since $\Bu \cdot (\spacegrad \cross \Bu) = 0$.

In this problem we do not have an irrotational flow, which can be demonstrated by direct calculation.  We have

\begin{equation}\label{eqn:constantSpinSurfaces:130}
\begin{aligned}
\spacegrad \cross \Bu
&=
\Omega
\begin{vmatrix}
\xcap & \ycap & \zcap \\
\partial_x & \partial_y & 0 \\
-y & x & 0
\end{vmatrix} \\
&=
2 \zcap \Omega \\
&\ne 0
\end{aligned}
\end{equation}

In fact we have

\begin{equation}\label{eqn:constantSpinSurfaces:150}
\begin{aligned}
\Bu \cross (\spacegrad \cross \Bu)
&=
2 \Omega^2
\begin{vmatrix}
\xcap & \ycap & \zcap \\
-y & x & 0  \\
0 & 0 & 1
\end{vmatrix} \\
&=
2 \Omega^2 (\xcap + \ycap)
\end{aligned}
\end{equation}

The closest we can get to Bernoulli's theorem for this problem is

\begin{equation}\label{eqn:constantSpinSurfaces:210}
2 \Omega^2 (\xcap + \ycap)
= 
\spacegrad 
\left( 
\inv{2} \Bu^2 + \frac{p}{\rho} + g z
\right).
\end{equation}

We can say that the directional derivatives in directions perpendicular to $\xcap + \ycap$ are zero, and that 

\begin{equation}\label{eqn:constantSpinSurfaces:230}
\begin{aligned}
2 \Omega^2 
&= (\partial_x + \partial_y) 
\left( 
\inv{2} \Bu^2 + \frac{p}{\rho} + g z
\right) \\
&= (\partial_x + \partial_y) 
\left( 
\inv{2} \Bu^2 + \frac{p}{\rho} 
\right) \\
\end{aligned}
\end{equation}

Perhaps those could be used to solve for the surface, but we no longer have something that is obviously integrable.

Because $\Bu \cdot (\Bu \cross (\spacegrad \cross \Bu)) = 0$, we can also say that

\begin{equation}\label{eqn:constantSpinSurfaces:250}
\begin{aligned}
0 
&= \Bu \cdot \spacegrad
\left( 
\inv{2} \Bu^2 + \frac{p}{\rho} + g z
\right) \\
&= 
\Omega ( y \partial_x - x \partial_y )
\left( 
\inv{2} \Bu^2 + \frac{p}{\rho} 
\right).
\end{aligned}
\end{equation}

Perhaps this could also be used to find the surface?

\section{Solution.  Part 2.}

We want to write down the steady state, incompressible, non-viscous Navier-Stokes equations.  The first of these is trivially satisfied by our assumed solution

\begin{equation}\label{eqn:constantSpinSurfaces:270}
0 
= \spacegrad \cdot \Bu 
= \partial_x (-\Omega y) + \partial_y(\Omega x).
\end{equation}

For the inertial term we've got

\begin{equation}\label{eqn:constantSpinSurfaces:290}
\begin{aligned}
\Bu \cdot \spacegrad \Bu
&=
\Omega^2 (-y \partial_x + x \partial_y (-y, x, 0) \\
&=
\Omega^2 (-x, -y, 0),
\end{aligned}
\end{equation}

Leaving us with

\begin{equation}\label{eqn:constantSpinSurfaces:310}
\begin{aligned}
-\Omega^2 x &= -\inv{\rho} \partial_x p \\
-\Omega^2 y &= -\inv{\rho} \partial_y p \\
          0 &= -\inv{\rho} \partial_z p - g
\end{aligned}
\end{equation}

Integrating these, we seek seek simultaneous solutions to

\begin{equation}\label{eqn:constantSpinSurfaces:330}
\begin{aligned}
p &= \inv{2} \rho \Omega^2 x^2 + f(y,z) \\
p &= \inv{2} \rho \Omega^2 y^2 + g(x,z) \\
p &= h(x, y) - \rho g z.
\end{aligned}
\end{equation}

It's clear that one solution would be

\begin{equation}\label{eqn:constantSpinSurfaces:350}
p = p_0 + \inv{2} \rho \Omega^2 (x^2 + y^2) - \rho g z.
\end{equation}

where $p_0$ is some constant to be determined, dependent on where we set our origin.  Putting the origin of the coordinate system at the lowest point in the parabolic profile $(x, y, z) = (0, 0, 0)$, we have $p(0, 0, 0) = p_0$, which fixes $p_0$ as the atmospheric pressure.  If the radius of the bucket is $R$, the max height $h$ of the surface above that point is also found on this surface of constant pressure

\begin{equation}\label{eqn:constantSpinSurfaces:370}
p_0 = p_0 + \inv{2} \rho \Omega^2 R^2 - \rho g h,
\end{equation}

or 

\begin{equation}\label{eqn:constantSpinSurfaces:390}
h = \frac{\Omega^2 R^2 }{2 g}.
\end{equation}

\section{Appendix.  Proof of vector identities used.}

\begin{equation}\label{eqn:constantSpinSurfaces:170}
\begin{aligned}
\left( \spacegrad \inv{2} \Bu^2 + (\spacegrad \cross \Bu) \cross \Bu \right)_i
&=
\partial_i \inv{2} u_j u_j + \partial_a u_b \epsilon_{a b r} u_s \epsilon_{r s i} \\
&=
u_j \partial_i u_j + \partial_a u_b u_s \delta^{[a b]}_{s i} \\
&=
u_j \partial_i u_j 
+ u_s \partial_s u_i 
- u_s \partial_i u_s \\
&= (\Bu \cdot \spacegrad) \Bu)_i 
\end{aligned}
\end{equation}

Also observe that our claim that $\Bu \cdot (\Bu \cross (\spacegrad \Bu)) = 0$ follows easily after expansion in coordinates

\begin{equation}\label{eqn:constantSpinSurfaces:190}
\Bu \cdot (\Bu \cross (\spacegrad \cross \Bu) )
=
u_i u_s ( \partial_s u_i - \partial_i u_s ).
\end{equation}

We've got a symmetric and antisymmetric factor in the summation, so the end result is zero.

%\EndArticle
\EndNoBibArticle
