\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

%\chapter{Introduction and strain tensor.}
\chapter{PHY454H1S\\Continuum Mechanics.  Lecture 2.  Introduction and strain tensor.  Taught by Prof. K. Das.}
\label{chap:continuumL2}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumL2.pdf}
\date{Jan 13, 2012}
\revisionInfo{continuumL2.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Introduction.}

Mechanics could be defined as the study of effects of forces and displacements on a physical body.

FIXME: F1.

In continuum mechanics we have a physical body and we are interested in the internal motions in the object.  

FIXME: F2.

For the first time considering mechanics we have to introduce the concepts of fields to make progress tackling these problems.

We will have use of the following types of fields

\begin{itemize}
\item Scalar fields.  $3^0$ components.  Examples: density, Temperature, ...
\item Vector fields.  $3^1$ components.  Examples: Force, velocity.
\item Tensor fields.  $3^2$ components.  Examples: stress, strain.
\end{itemize}

We have to consider objects (a control volume) that is small enough that we can consider that we have a point in space limit for the quantities of density and velocity.  At the same time we cannot take this limiting process to the extreme, since if we use a control volume that is sufficiently small, quantum and inter-atomic effects would have to be considered.

FIXME: F3.

\subsection{Stress and Strain definitions.}

\begin{definition}
\emph{(Stress)}
\label{dfn:continuumL2:10}
Measure of the Internal force on the surfaces.
\end{definition}

\begin{definition}
\emph{(Strain)}
\label{dfn:continuumL2:30}
Measure of the deformation of the body.
\end{definition}

\section{Strain Tensor.}

This follows \cite{landau1960theory} \S 1 very closely.

Utilizing summation convention consider a set of small internal displacements $u_1, u_2, u_3$ to the $x, y, z$ coordinates so that the transformation $x_i \rightarrow x_i'$ is related by

FIXME: F4

\begin{align}\label{eqn:continuumL2:50}
u_i &= x_i' - x_i \\
x_i' &= g(x_i) \\
u_i &= f(x_i)
\end{align}

(ie: $x_i'$ is a function of all the initial coordinates, as are the displacements $u_i$).

\begin{equation}\label{eqn:continuumL2:70}
dx_i' = dx_i + du_i
\end{equation}

\begin{align}\label{eqn:continuumL2:90}
dl &= \sqrt{dx_k dx_k} \\
dl' &= \sqrt{d{x'}_k d{x'}_k}
\end{align}

or

\begin{equation}\label{eqn:continuumL2:110}
{dl'}^2 = 
(dx_k + du_k)
(dx_k + du_k)
= 
dl^2 + 2 dx_k du_k + du_k du_k
\end{equation}

with 

\begin{equation}\label{eqn:continuumL2:130}
du_i = \PD{x_k}{u_i} dx_k
\end{equation}

we have

\begin{equation}\label{eqn:continuumL2:150}
du_i^2 = 
\PD{x_k}{u_i} dx_k
\PD{x_l}{u_i} dx_l
\end{equation}

\begin{align*}
{dl'}^2 
&= 
dl^2 
+ 2 \PD{x_k}{u_i} dx_k dx_i 
+ \PD{x_i}{u_l} 
\PD{x_k}{u_l} 
dx_i dx_k \\
&= 
dl^2 
+ 
\left(
\PD{x_k}{u_i} 
+
\PD{x_i}{u_k} 
\right)
dx_k dx_i 
+ \PD{x_i}{u_l} 
\PD{x_k}{u_l} 
dx_i dx_k \\
&=
dl^2 
+ 
2 e_{ik} dx_i dx_k
\end{align*}

We write

\begin{equation}\label{eqn:continuumL2:170}
{dl'}^2 - dl^2 = 2 e_{ik} dx_i dx_k
\end{equation}

where we define the \emph{strain tensor} as

\begin{equation}\label{eqn:continuumL2:190}
e_{ik} = \inv{2} \left(
\left(
\PD{x_k}{u_i} 
+
\PD{x_i}{u_k} 
\right)
+ \PD{x_i}{u_l} 
\PD{x_k}{u_l} 
\right)
\end{equation}

Here $e_{ik}$ is a $3 \times 3$ matrix in Cartesian coordinates

\begin{equation}\label{eqn:continuumL2:210}
\begin{bmatrix}
e_{11} & e_{12} & e_{13} \\
e_{21} & e_{22} & e_{23} \\
e_{31} & e_{32} & e_{33} \\
\end{bmatrix}
\end{equation}

We see from \ref{eqn:continuumL2:190} that $e_{ik}$ is symmetric, so we have

\begin{align}\label{eqn:continuumL2:n}
e_{21} &= e_{12} \\
e_{31} &= e_{13} \\
e_{32} &= e_{23}
\end{align}

Because any real symmetric matrix can be diagonalized we can write in some coordinate system

\begin{equation}\label{eqn:continuumL2:n}
\bar{e}_{ik} = 
\begin{bmatrix}
\bar{e}_{11} & 0 & 0 \\
0 & \bar{e}_{22} & 0 \\
0 & 0 & \bar{e}_{33} \\
\end{bmatrix}
\end{equation}

\begin{align}\label{eqn:continuumL2:n}
{dx_1'}^2 &= (1 + 2 \bar{e}_{11}) dx_1^2 \\
{dx_2'}^2 &= (1 + 2 \bar{e}_{22}) dx_2^2 \\
{dx_3'}^2 &= (1 + 2 \bar{e}_{33}) dx_3^2
\end{align}

If our changes are small enough we can also write approximately, taking the first order term in the square root evaluation

\begin{align}\label{eqn:continuumL2:n}
dx_1' &\approx (1 + \bar{e}_{11}) dx_1 \\
dx_2' &\approx (1 + \bar{e}_{22}) dx_2 \\
dx_3' &\approx (1 + \bar{e}_{33}) dx_3
\end{align}

%FIXME: some hand waving here to think through.
We are also free to define a volume element

\begin{equation}\label{eqn:continuumL2:n}
dV' = 
dx_1'
dx_2'
dx_3'
\approx
(1 + e_{11})
(1 + e_{22})
(1 + e_{33})
dx_1 dx_2 dx_3
\end{equation}

\begin{equation}\label{eqn:continuumL2:n}
dV' = (1 + e_{11} +e_{22} +e_{33} ) dV
\end{equation}

So the change of volume is given by the trace

\begin{equation}\label{eqn:continuumL2:n}
dV' = ( 1 + e_{ii} )^2 dV
\end{equation}

\subsection{Strain Tensor in cylindrical coordinates.}

At the end of the section in the text, the formulas for the spherical and cylindrical versions (to first order) of the strain tensor is given without derivation.  Let's do that derivation for the cylindrical case, which is simpler.  It appears that use of explicit vector notation is helpful here, so we write

\begin{align}\label{eqn:continuumL2:n}
\Bx &= r \rcap + z \zcap \\
\Bu & u_r \rcap + u_\phi \phicap + u_z \zcap
\end{align}

where

\begin{align}\label{eqn:continuumL2:n}
\rcap &= \Be_1 e^{i\phi} \\
\phicap &= \Be_2 e^{i\phi} \\
i &= \Be_1 \Be_2
\end{align}

Since $\rcap$ and $\phicap$ are functions of position, we will need their differentials

\begin{align}\label{eqn:continuumL2:n}
d\rcap &= \Be_1 \Be_1 \Be_2 e^{i\phi} d\phi = \Be_2 e^{i \phi} d\phi \\
d\phicap &= \Be_2 \Be_1 \Be_2 e^{i\phi} d\phi = -\Be_2 e^{i \phi} d\phi,
\end{align}

but these are just scaled basis vectors

\begin{align}\label{eqn:continuumL2:n}
d\rcap &= \phicap d\phi \\
d\phicap &= -\rcap d\phi.
\end{align}

So for our $\Bx$ and $\Bu$ differentials we find

\begin{align*}
d\Bx 
&= dr \rcap + r d\rcap + dz \zcap \\
&= dr \rcap + r \phicap d\phi + dz \zcap,
\end{align*}

and
\begin{align*}
d\Bu 
&= du_r \rcap + du_\phi \phicap + du_z \zcap 
+ u_r \phicap d\phi - u_\phi \rcap d\phi \\
&= \rcap( du_r - u_\phi d\phi )
+ \phicap ( du_\phi + u_r d\phi )
+ \zcap ( du_z ).
\end{align*}

Putting these together we have

\begin{align*}
d\Bl' 
&= d\Bu + d\Bx 
\\
&= \rcap( du_r - u_\phi d\phi + dr )
+ \phicap ( du_\phi + u_r d\phi + r d\phi )
+ \zcap ( du_z + dz ).
\end{align*}

For the squared magnitude's difference from $d\Bx^2$ we have

\begin{align*}
(d\Bl')^2 - d\Bx^2
&= 
( du_r - u_\phi d\phi + dr )^2
+ ( du_\phi + u_r d\phi + r d\phi )^2
+ ( du_z + dz )^2
-dr^2 - r^2 d\phi^2 - dz^2 \\
&=
( du_r - u_\phi d\phi )^2 
+ 2 dr ( du_r - u_\phi d\phi )
+ ( du_\phi + u_r d\phi )^2
+ 2 r d\phi ( du_\phi + u_r d\phi )
+ du_z^2 + 2 du_z dz \\
\end{align*}

Expanding this out, but dropping all the terms that are quadratic in the components of $\Bu$ or its differentials, we have

\begin{align*}
(d\Bl')^2 - d\Bx^2
&\approx
  2 dr ( du_r - u_\phi d\phi )
+ 2 r d\phi ( du_\phi + u_r d\phi )
+ 2 du_z dz \\
&=
  2 dr du_r 
- 2 dr u_\phi d\phi 
+ 2 r d\phi du_\phi 
+ 2 r d\phi u_r d\phi 
+ 2 du_z dz 
\\
&=
  2 dr 
\left( 
\PD{r}{u_r} dr
+\PD{\phi}{u_r} d\phi
+\PD{z}{u_r} dz
\right) \\
&- 2 dr d\phi u_\phi  \\
&+ 2 r d\phi 
\left( 
\PD{r}{u_\phi} dr
+\PD{\phi}{u_\phi} d\phi
+\PD{z}{u_\phi} dz
\right) \\
&+ 2 r d\phi d\phi u_r \\
&+ 2 
dz 
\left( 
\PD{r}{u_z} dr
+\PD{\phi}{u_z} d\phi
+\PD{z}{u_z} dz
\right) \\
\end{align*}

Grouping all terms, with all the second order terms neglected, we have

\begin{equation}\label{eqn:continuumL2:n}
\begin{aligned}
(d\Bl')^2 - d\Bx^2
&=
2 dr dr \PD{r}{u_r} 
+ 2 r^2 d\phi d\phi \left( \inv{r} \PD{\phi}{u_\phi} +\inv{r} u_r \right)
+ 2 dz dz \PD{z}{u_z}  \\
&+ 2 dz dr \left( \PD{z}{u_r} + \PD{r}{u_z} \right)
+ 2 dr r d\phi \left( \PD{r}{u_\phi} - \inv{r} u_\phi + \inv{r} \PD{\phi}{u_r} \right)
+ 2 dz r d\phi \left( \PD{z}{u_\phi} +\inv{r} \PD{\phi}{u_z} \right).
\end{aligned}
\end{equation}

From this we can read off the result quoted in the text

\begin{align}\label{eqn:continuumL2:n}
2 e_{rr} &= \PD{r}{u_r}  \\
2 e_{\phi\phi} &= \inv{r} \PD{\phi}{u_\phi} +\inv{r} u_r  \\
2 e_{zz} &= \PD{z}{u_z}  \\
2 e_{zr} &= \PD{z}{u_r} + \PD{r}{u_z} \\
2 e_{r\phi} &= \PD{r}{u_\phi} - \inv{r} u_\phi + \inv{r} \PD{\phi}{u_r} \\
2 e_{\phi z} &= \PD{z}{u_\phi} +\inv{r} \PD{\phi}{u_z}.
\end{align}

Observe that we have to introduce factors of $r$ along with all the $d\phi$'s, when we factored out the tensor components.  That's an important looking detail, which isn't obvious unless one works through the derivation.

Note that in class we retained the second order terms.  If we do the same here (omitting the details), we get

\begin{equation}\label{eqn:continuumL2:n}
\begin{aligned}
(d\Bl')^2 - d\Bx^2 &= \\
\text{dr}^2 \left(\text{PDruphi}^2+2 \text{PDrur}+\text{PDrur}^2+\text{PDruz}^2\right)+\text{dr} \text{dz} (2 \text{PDruz}+2 \text{PDruphi} \text{PDzuphi}+2 \text{PDzur}+2 \text{PDrur} \text{PDzur}+2 \text{PDruz} \text{PDzuz})+\text{dz}^2 \left(1+\text{PDzuphi}^2+\text{PDzur}^2+2 \text{PDzuz}+\text{PDzuz}^2\right)+\text{dphi} \text{dr} (2 \text{PDphiur}+2 \text{PDphiuphi} \text{PDruphi}+2 \text{PDphiur} \text{PDrur}+2 \text{PDphiuz} \text{PDruz}+2 \text{PDruphi} r-2 \text{uphi}-2 \text{PDrur} \text{uphi}+2 \text{PDruphi} \text{ur})+\text{dphi} \text{dz} (2 \text{PDphiuz}+2 \text{PDphiuphi} \text{PDzuphi}+2 \text{PDphiur} \text{PDzur}+2 \text{PDphiuz} \text{PDzuz}+2 \text{PDzuphi} r-2 \text{PDzur} \text{uphi}+2 \text{PDzuphi} \text{ur})+\text{dphi}^2 \left(\text{PDphiuphi}^2+\text{PDphiur}^2+\text{PDphiuz}^2+2 \text{PDphiuphi} r-2 \text{PDphiur} \text{uphi}+\text{uphi}^2+2 \text{PDphiuphi} \text{ur}+2 r \text{ur}+\text{ur}^2\right)
\end{aligned}
\end{equation}

\EndArticle
