% 
% 
% 
% Copyright © 2012 Peeter Joot
% All Rights Reserved
% 
% This file may be reproduced and distributed in whole or in part, without fee, subject to the following conditions:
% 
% o The copyright notice above and this permission notice must be preserved complete on all complete or partial copies.
% 
% o Any translation or derived work must be approved by the author in writing before distribution.
% 
% o If you distribute this work in part, instructions for obtaining the complete version of this file must be included, and a means for obtaining a complete version provided.
% 
% 
% Exceptions to these rules may be granted for academic purposes: Write to the author and ask.
% 
% 
% 
%\input{../peeter_prologue_print.tex}
%\input{../peeter_prologue_widescreen.tex}

\chapter{Overview.}
%\chapter{Continuum mechanics.  Lecture I.  Overview.}
\label{chap:continuumL1}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot2/math2012/continuumL1.pdf}
\date{Jan 11, 2012}
\revisionInfo{continuumL1.tex}

\beginArtWithToc
%\beginArtNoToc

\section{Texts.}

The course notes that will be posted with each lecture are based on the following

\begin{itemize}
\item \href{http://www.amazon.com/Elementary-Dynamics-Applied-Mathematics-Computing/dp/0198596790/ref=sr_1_1?ie=UTF8&qid=1326302753&sr=8-1}{Elementary fluid dynamics} \cite{acheson1990elementary}.
\item \href{http://www.amazon.com/Theory-Elasticity-Third-Theoretical-Physics/dp/075062633X/ref=sr_1_6?s=books&ie=UTF8&qid=1326302957&sr=1-6}{Continuum mechanics} \cite{landau1960theory}
\end{itemize}

%http://www.biblio.com/search.php?author=acheson&title=elementary+fluid+dynamics&keyisbn=&format=

This course won't actually follow these texts specifically.  While they are not required reading, I found the Acheson text particularly helpful when looking for additional details about subjects covered in class.

\section{Nomenclature.}

Some terms we'll end up discussing as the course progresses

\begin{itemize}
\item rheological.  Study of the flow of matter, primarily in the liquid state.  \cite{wiki:rheology}.
\item Newtonian fluids: stress and strain are linear.
\item non-Newtonian fluids: stress and strain are non-linearly related.
\item constitutive relation: relations between stress and strain.
\item stress: A quantity constructed so that its divergence supplies the force per unit volume.
\item strain: A relation between stretch and position.
\item elastic waves
\item elastic/plastic
\item Navier stokes relation: equivalent of Newton's law for fluids.
\item nondimensionalisation: parameter substitutions that put the differential equations of interest into dimensionless form.
\item boundary layer theory: investigation of the region of fluid flow around a solid where viscous forces dominate.
\item stability analysis
\item nonlinearily 
\end{itemize}

\EndArticle
