% 
% 
% 
% Copyright © 2012 Peeter Joot
% All Rights Reserved
% 
% This file may be reproduced and distributed in whole or in part, without fee, subject to the following conditions:
% 
% o The copyright notice above and this permission notice must be preserved complete on all complete or partial copies.
% 
% o Any translation or derived work must be approved by the author in writing before distribution.
% 
% o If you distribute this work in part, instructions for obtaining the complete version of this file must be included, and a means for obtaining a complete version provided.
% 
% 
% Exceptions to these rules may be granted for academic purposes: Write to the author and ask.
% 
% 
% 
\chapter{Multimedia figures.}

A table of animated figures in this document.  These require Adobe Reader (google's online pdf reader will not display these for example).  Some patience is required to view these (waiting for the underlying javascript to run before clicking the play button ... on my machine I can hear acrobat churning away on the CPU at these points.)

\begin{itemize}
\item Fluids with densities of mercury and water in the lower (1) and upper (2) layers respectively.  \ref{fig:continuumProblemSet2:continuumProblemSet2Animation1}
\item Fluids with densities of water and mercury in the lower (1) and upper (2) layers respectively.  \ref{fig:continuumProblemSet2:continuumProblemSet2Animation2}
\item Time evolution of channel flow velocity profile after turning on a constant pressure gradient.  \ref{fig:channelFlowWithStepPressureGradient:channelFlowWithStepPressureGradientFig3}
\item Early time evolution of channel flow velocity profile after turning on a constant pressure gradient.  \ref{fig:channelFlowWithStepPressureGradient:channelFlowWithStepPressureGradientFig4}
\item Animation of Couette flow, with continuous variation of outer angular velocity. \ref{fig:couetteFlow:couetteFlowFig6}.
\end{itemize}

%-------------------------------------------------------
