%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%
%
\input{../peeter_prologue.tex}

\chapter{hamiltonian}
\label{chap:hamiltonian}
%\useCCL
\blogpage{http://sites.google.com/site/peeterjoot/math2009/hamiltonian.pdf}
\date{Nov 15, 2009}
\revisionInfo{$RCSfile: hr.tex,v $ Last $Revision: 1.6 $ $Date: 2009/12/03 03:24:40 $}

\beginArtWithToc
%\beginArtNoToc

\section{Relativistic force free Hamiltonian.}

Having considered a number of Newtonian mechanical systems, lets move on to the force free relativistic Lagrangian in preparation for considering electromagnetic forces in a Hamiltonian context.

\subsection{Covariant Lagrangian.  Force free equations of motion.}

We have two options for the relativistic Lagrangian (and perhaps more).  One of which is the covariant form with exactly the same structure as a purely kinetic Newtonian Lagrangian $m \Bv^2/2$, except for replacement of the velocity dot product with a four vector dot product of proper time velocities.  That is

\begin{align}\label{eqn:hamiltonian:qqq1}
\LL &= \inv{2} m \xdot^\mu \xdot_\mu \\
\xdot^\mu &= \frac{d x^\mu}{d \tau},
\end{align}

where the corresponding covariant action is an integral over proper time

\begin{align}\label{eqn:hamiltonian:qqq2}
S = \int d\tau \LL.
\end{align}

Varying the proper time action, and picking the upper index coordinates as our generalized coordinates, we get for the Euler-Lagrange equations 

\begin{align}\label{eqn:hamiltonian:qqq3}
\frac{d}{d \tau} \frac{d \LL}{d \xdot^\alpha} = \frac{d \LL}{d x^\alpha}.
\end{align}

This is the setup required to evaluate the Euler-Lagrange equations.  First for the left hand side we want the conjugate momenta

\begin{align*}
P_\mu 
&= \frac{d \LL}{d \xdot^\alpha} \\
&= m \xdot_\mu.
\end{align*}

Our equations of motion, without yet using the Hamiltonian equations are just

\begin{align}\label{eqn:hamiltonian:qqq2x}
\frac{d}{d\tau} \left( \xdot_\mu \right) = 0.
\end{align}

Provided the covariant Lagrangian has no explicit proper time dependence, following the same procedure as in the non-relativistic case, we find

\begin{align}\label{eqn:hamiltonian:foo2r}
\frac{d}{d\tau} \left( \dot{x}^\mu \PD{\dot{x}^\mu}{\LL} -\LL \right) = 0.
\end{align}

We therefore have a constant for the system associated with proper time translation

\begin{align}\label{eqn:hamiltonian:foo2q}
H = \dot{x}^\mu \PD{\dot{x}^\mu}{\LL} -\LL.
\end{align}

Taking partials we find easily the covariant form of the Hamiltonian equations

\begin{align}\label{eqn:hamiltonian:qqq30}
\PD{P_\mu}{H} &= \dot{x}^\mu \\
\PD{x^\mu}{H} &= -\dot{P}_\mu 
\end{align}

Returning to the specifics of our covariant Lagrangian, the Hamiltonian is unsurprisingly just the Lagrangian

\begin{align}\label{eqn:hamiltonian:foo2s}
H = \inv{2} m \dot{x}^\mu \dot{x}_\mu = \LL = \inv{2m} P^\mu P_\mu.
\end{align}

Writing this out explicitly in space time coordinates we have

\begin{align}\label{eqn:hamiltonian:qqq10}
H = \inv{2} m
\left( \frac{dt}{d\tau} \right)^2
(c^2 - \Bv^2)
\end{align}

In particular in the rest frame where $t = \tau$ and $\Bv = 0$, we have

\begin{align}\label{eqn:hamiltonian:qqq11}
H = \inv{2} m c^2.
\end{align}

An immediate consequence, without even evaluating the Hamiltonian equations, is that our $\gamma$ factor falls out

\begin{align}\label{eqn:hamiltonian:qqq12}
\frac{dt}{d\tau} = \inv{\sqrt{1 - \Bv^2/c^2}}.
\end{align}

This also allows the equations of motion \eqnref{eqn:hamiltonian:qqq2x} to be transformed from covariant to a spacetime form

%FIXME.
%m \frac{d}{d\tau}

\subsection{An aside.}

The fact that the Hamiltonian is a constant of motion also follows directly just by taking derivatives and applying the Euler-Lagrange equations.  That is

\begin{align*}
\frac{d}{dt} \left( \inv{2} m \dot{x}^\mu \dot{x}_\mu \right) 
=
\dot{x}^\mu \underbrace{\frac{d}{dt} \left( m \dot{x}_\mu \right)}_{=0} \\
\end{align*}

\subsection{Applying the Hamiltonian equations.}

Okay.  Let us apply the Hamiltonian equations \eqnref{eqn:hamiltonian:qqq30}.  For our force free Lagrangian where we have no $x^\mu$ dependence, we recover \eqnref{eqn:hamiltonian:qqq2x} immediately by evaluating

\begin{align}\label{eqn:hamiltonian:qqq30b}
\PD{x^\mu}{H} &= -\dot{P}_\mu,
\end{align}

since the left hand side is zero.

%\EndArticle
\EndNoBibArticle
