%
% Copyright © 2015 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

% 
%\chapter{Preface}
% this suppresses an explicit chapter number for the preface.
\chapter*{Preface}%\normalsize
  \addcontentsline{toc}{chapter}{Preface}

This document was produced while taking the Spring 2015, University of Toronto Graduate Quantum Mechanics course (PHY1520H), taught by Prof.\ Arun Paramekanti.

%\paragraph{Official course description:}
%
%``
%''

\begin{enumerate}
\item Basics - Postulates, Wavefunctions, Density matrices, Measurements
\item Time evolution - Schrodinger picture, Heisenberg picture, Interaction picture
\item Harmonic oscillator - Operator method, Wavefunctions, Coherent states
\item Particle in a magnetic field - Local gauge invariance, 2D Landau levels
\item Symmetries - Parity, Translations, Rotations, Time-reversal
\item Angular momentum, Spin, and Angular momentum addition
\item Time-independent perturbation theory
\item Time-dependent perturbation theory
\item Variation approach
\item Scattering theory
\item Dirac equation - one dimension
\item Path integrals
\end{enumerate}

Texts include
\begin{itemize}
	\item [Required Text] 
%\citep{balanis2005antenna} C.A. Balanis, ``Antenna Theory,'' Wiley, 3rd Edition
J. J. Sakurai, Modern Quantum Mechanics, 2nd edition. Addison Wesley 1993.
	\item [Optional Text] 
K. Gottfried and T.-M. Yan, Quantum Mechanics (Springer, New York, 2004), 2nd Edition.
%	\item [Reference books] 
%M. Le Bellac, Quantum Physics (Cambridge University Press, New York, 2006).
%C. Cohen-Tannoudji, B. Diu & F. Laloe, Quantum Mechanics, Vol. 1-2, Wiley-VCH.
%R. Shankar, Principles of Quantum Mechanics, 2nd edition, Plenum Press
%E. Merzbacher, Quantum Mechanics, 3rd edition, Wiley.
\end{itemize}

\paragraph{This document contains:}

\begin{itemize}

\item Personal notes exploring details that were not clear to me from the lectures, or from the texts associated with the lecture material.

\ifthenelse{\boolean{redacted}}%
{%
\item Assigned problems.  Like anything else take these as is.  I have attempted to either correct errors or mark them as such.%
}%
{\relax}

%\item Some worked problems attempted as course prep, for fun, or for test preparation, or post test reflection.
%\item Links to Matlab function implementations associated with the problem sets.
\end{itemize}

My thanks go to Professor Paramekanti for teaching this course.

Peeter Joot  \quad peeter.joot@gmail.com 
