%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%
%
%\chapter{Preface}
% this suppresses an explicit chapter number for the preface.
\chapter*{Preface}%\normalsize
  \addcontentsline{toc}{chapter}{Preface}

These are my personal lecture notes for the Spring 2011, University of Toronto, Relativistic Electrodynamics course (PHY450H1S).  This class was taught by Prof.\ Erich Poppitz, with Simon Freedman handling tutorials (which were excellent lecture style lessons).

Official course description:

Special Relativity, four-vector calculus and relativistic notation, the relativistic Maxwell's Equations, electromagnetic waves in vacuum and conducting and non-conducting materials, electromagnetic radiation from point charges and systems of charges.

The text for the course is \citep{landau1980classical}.

This document contains a few things

\begin{itemize}
\item My lecture notes.

Typos and errors are probably mine (Peeter), and no claim nor attempt of spelling or grammar correctness will be made.

These notes track along with the Professor's hand written notes very closely, since his lectures follow his notes very closely.  While I used the note taking exercise as a way to verify that I understood all the day's lecture materials, the Professor's notes are in many instances a much better study resource, since there are details in his notes that were left for us to read, and not necessarily covered in the lectures.  On the other hand, there are details in these notes that I have added when I did not find his approach simplistic enough for me to grasp, or I failed to follow the details in class.

\item Some notes from reading of the text.  One of the earlier of these was due to unfortunate use of an ancient edition of the text borrowed from the library, since mine had been lost in shipping.  That version did not use the upper and lower index quantities that I had expected, so I was left to try to reconcile that with what I knew.

\item Some assigned problems, at least the parts of them that I did not hand write.  I have corrected some the errors after receiving grading feedback, and where I have not done so I at least recorded some of the grading comments as a reference.  Not all the problems were graded, so I make no guarantees of correctness.
\end{itemize}

Peeter Joot  \quad peeterjoot@protonmail.com
