%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

\chapter{Integrating the equation of motion for a one dimensional problem}
\index{pendulum}
\label{chap:1dpotentialIntegral}
%\blogpage{http://sites.google.com/site/peeterjoot/math2010/1dpotentialIntegral.pdf}
%\date{Jan 1, 2010}

\section{Motivation}

While linear approximations, such as the small angle approximation for the pendulum, are often used to understand the dynamics of non-linear systems, exact solutions may be possible in some cases.  Walk through the setup for such an exact solution.

\section{Guts}

The equation to consider solutions of has the form

\begin{equation}
\label{eqn:1dpotentialIntegral:1}
\frac{d}{dt} \left( m \frac{dx}{dt} \right) = -\PD{x}{U(x)}.
\end{equation}

We have an unpleasant mix of space and time derivatives, preventing any sort of antidifferentiation.  Assuming constant mass \(m\), and employing the chain rule a refactoring in terms of velocities is possible.

\begin{equation}\label{eqn:1dpotentialIntegral:45}
\begin{aligned}
\frac{d}{dt} \left( \frac{dx}{dt} \right)
&=
\frac{dx}{dt} \frac{d}{dx} \left( \frac{dx}{dt} \right)  \\
&=
\inv{2} \frac{d}{dx} \left( \frac{dx}{dt} \right)^2  \\
\end{aligned}
\end{equation}

The one dimensional Newton's law \autoref{eqn:1dpotentialIntegral:1} now takes the form
\begin{equation}
\label{eqn:1dpotentialIntegral:2}
\frac{d}{dx} \left( \frac{dx}{dt} \right)^2 = -\frac{2}{m} \PD{x}{U(x)}.
\end{equation}

This can now be antidifferentiated for

\begin{equation}\label{eqn:1dpotentialIntegral:3}
\left( \frac{dx}{dt} \right)^2 = \frac{2}{m} (H - U(x)).
\end{equation}

What has now been accomplished is the removal of the second derivative.  Having done so, one can see that this was a particularly dumb approach, since \autoref{eqn:1dpotentialIntegral:3} is nothing more than the Hamiltonian for the system, something more obvious if rearranged slightly

\begin{equation}\label{eqn:1dpotentialIntegral:3b}
H = \inv{2} m \dot{x}^2 + U(x).
\end{equation}

We could have started with a physics principle instead of mechanically plugging through calculus manipulations and saved some work.  Regardless of the method used to get this far, one can now take roots and rearrange for

\begin{equation}\label{eqn:1dpotentialIntegral:4}
dt = \frac{dx}{\sqrt{ \frac{2}{m} (H - U(x)) } }.
\end{equation}

We now have a differential form implicitly relating time and position.  One can conceivably integrate this and invert to solve for position as a function of time, but substitution of a more specific potential is required to go further.

\begin{equation}\label{eqn:1dpotentialIntegral:5}
t(x) = t(x_0) + \int_{y=x_0}^{x} \frac{dy}{\sqrt{ \frac{2}{m} (H - U(y)) } }.
\end{equation}
