%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%
%
%\chapter{Preface}
% this suppresses an explicit chapter number for the preface.
\chapter*{Preface}%\normalsize
  \addcontentsline{toc}{chapter}{Preface}

This is a collection of notes on classical mechanics, and contains a few things

\begin{itemize}
\item A collection of miscellaneous notes and problems for my personal (independent) classical mechanics studies.  A fair amount of those notes were originally in my collection of Geometric (Clifford) Algebra related material so may assume some knowledge of that subject.

\item My notes for some of the PHY354 lectures I attended.  That class was taught by Prof.\ Erich Poppitz.  I audited some of the Wednesday lectures since the timing was convenient.  I took occasional notes, did the first problem set, and a subset of problem set 2.

These notes, when I took them, likely track along with the Professor's hand written notes very closely, since his lectures follow his notes very closely.

%The official description of this course was:
%
%Symmetry and conservation laws, stability and instability, generalized co-ordinates, Hamilton's principle, Hamilton's equations, phase space, Liouville's theorem, canonical transformations, Poisson brackets, Noether's theorem.
%
The text for PHY354 is \citep{landau1976mechanics}.  I'd done my independent study from \citep{goldstein1951cm}, also a great little book.

\item Some assigned problems from the PHY354 course, ungraded (not submitted since I did not actually take the course).  I ended up only doing the first problem set and two problems from the second problem set.

\item Miscellaneous worked problems from other sources.

\end{itemize}

Peeter Joot  \quad peeterjoot@protonmail.com
