
usepackage: \ce{}. chem.

\paragraph{Druid model}

In this section we will investigate the optical properties of free electrons, or what is commonly called free electron gas.

By free electron gas we mean electrons that do not experience the restoring force which we considered for bound garges in the case of Lorentz model.  In particular, the resonance frequency \( \omega_0 \) for free electrons is zero.

There are two typical cases of free electron systems 

\begin{itemize}[a]
\item Metals.
\item Doped (n or p type) semiconductors.
\end{itemize}

For the moment we consider the case of metals.

Free electrons are responsible for high reflectivity and good thermal conductivity of metals up to optical frequencies.  A model that can be used to describe the high reflectivity of metals is the Drude model.

\paragraph{Plasma:} A neutral gas of free eletrons and heavy ions is called plasma.  Examples of plasma are metals and doped semiconduction, since these materials are a compination of free electrons and heavy ions which are, in sum, electrically neutral.

\paragraph{Drude-Lorentz model}, (or Drude model for short): similar to the case of bound charges we already studied for free electron plasma, we can start with a harmonic oscillator model.  However, in this case, since electrons are free, there is no restoring force (i.e. \(\omega_0 = 0 \).  Recall that in the spring mass model \( \omega_0^2 = S/m \) where \( S \) was the spring tension coefficient.

With such a model the Lorentz model equation

\begin{dmath}\label{eqn:druid:20}
\frac{d^2 x}{dt^2} + \gamma \ddt{x} + \omega_0^2 x = \frac{Q E_0}{m} e^{j \omega t},
\end{dmath}

is reduced to

\begin{dmath}\label{eqn:druid:40}
\frac{d^2 x}{dt^2} + \gamma \ddt{x} = \frac{Q E_0}{m} e^{j \omega t},
\end{dmath}

Again, assuming a solution of the form \( x_p = x_0 e^{j \omega t} \) for the particular solution and substituting in \cref{eqn:druid:40}, we have

\begin{dmath}\label{eqn:druid:80}
x_0 \lr{ (j\omega)^2 + \gamma (j \omega)} = \frac{Q E_0}{m},
\end{dmath}

or
\begin{dmath}\label{eqn:druid:60}
x 
= 
\frac{Q E/m}{-\omega^2 + j \gamma \omega },
\end{dmath}

Once more assuming identical particles that are not coupled and a linear isotropic medium and using the fact that \( \BP = N \Bp = N Q \Bx \), and

\begin{dmath}\label{eqn:druid:100}
\chi_e = \frac{\Abs{\BP}}{\epsilon_0 \Abs{\BE} }, 
\end{dmath}

we have

\begin{dmath}\label{eqn:druid:120}
\chi_e 
=
\frac{Q^2 N/m \epsilon_0}{-\omega^2 + j \gamma \omega },
\end{dmath}

or with \( \omega_p^2 = Q^2 N/m\epsilon_0\),

\begin{dmath}\label{eqn:druid:140}
\epsilon_r 
= 1 + \chi_e 
= 
1+
\frac{\omega_p^2}{-\omega^2 + j \gamma \omega }.
\end{dmath}

Plasma frequency, \( \omega_p \), can be understood as the natural resonance frequency by which the free electron gas (plasma) collectively (not individulal electrons ) oscillates.

Note that if we neglect the last term, i.e., let \( \gamma = 0 \) then

\begin{dmath}\label{eqn:druid:160}
\epilson_r = 1 - \frac{\omega_p^2}{\omega^2}.
\end{dmath}

From this it is clear that when \( \omega < \omega_p \), we have \( \epsilon_r < 1 \) and \( n = \sqrt{\epsilon_r} \) is purely imaginary, and the wave attenuates inside the electron plasma.

This means that for \( \omega < \omega_p \) electromagnetic waves do not propagate a large distance inside of metal.  However, for \( \omega > \omega_p \) the electron plasma (e.g. metal) is transparent.  The latter is called ultraviolate transparency of metal, because for most metals \( \omega_p \) is in the ultraviolate part of the spectrum.  For example, 

\begin{itemize}
\item For \ce{Al}
\begin{dmath}\label{eqn:druid:180}
\frac{\omega_p}{2 \pi} = 3.82 \times 10^{15} \si{Hz} \implies \lambda_p = 79 [nm],
\end{dmath}
\item For \ce{Au}
\begin{dmath}\label{eqn:druid:200}
\frac{\omega_p}{2 \pi} = 5.9 \times 10^{15} \si{Hz} \implies \lambda_p = 138 [nm],
\end{dmath}
\end{itemize}

Using \cref{eqn:druid:160} one can calculate 

\begin{dmath}\label{eqn:druid:220}
\tilde{n} = \sqrt{\epsilon_r},
\end{dmath}

and plot the reflectivity \( R \) at normal incidence

\begin{dmath}\label{eqn:druid:240}
R = \Abs{ \frac{\tilde{n} - 1 }{\tilde{n} + 1} },0jA
\end{dmath}

which will have a shape similar to

F3

\paragraph{Conductivity}

%\ddt{} \lr{ m \ddt{x} } + \gamma \ddt{x} = Q E_0 e^{i \omega t}

\begin{dmath}\label{eqn:druid:260}
\spacegrad \cross \BE(\Br, \omega) 
= \sigma \BE(\Br, \omega) + j \omega \epsilon_0 \BE(\Br, \omega)
= j \omega \epsilon_0 \lr{ 1 + \frac{\sigma}{j \omega \epsilon_0} } \BE(\Br, \omega)
= j \omega \epsilon_0 \lr{ 1 - \frac{j \sigma}{\omega \epsilon_0} } \BE(\Br, \omega)
\end{dmath}

This complex factor is the relative permittivity

\begin{dmath}\label{eqn:druid:280}
\epsilon_r  
\end{dmath}
= 1 - \frac{j \sigma}{\omega \epsilon_0},
\end{dmath}

and is why we write

\begin{dmath}\label{eqn:druid:300}
\epsilon(\omega) = \epsilon'(\omega) - j \epsilon''(\omega) 
\end{dmath}
