%
% Copyright © 2016 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\input{../blogpost.tex}
\renewcommand{\basename}{emt8}
\renewcommand{\dirname}{notes/ece1228/}
\newcommand{\keywords}{ECE1228H}
\input{../peeter_prologue_print2.tex}

%\usepackage{ece1228}
\usepackage{peeters_braket}
%\usepackage{peeters_layout_exercise}
\usepackage{peeters_figures}
\usepackage{mathtools}
\usepackage{siunitx}

\beginArtNoToc
\generatetitle{ECE1228H Electromagnetic Theory.  Lecture 8: Waves.  Taught by Prof.\ M. Mojahedi}
%\chapter{Waves}
\label{chap:emt8}

%\paragraph{Disclaimer}
%
%Peeter's lecture notes from class.  These may be incoherent and rough.
%
%These are notes for the UofT course ECE1228H, Electromagnetic Theory, taught by Prof. M. Mojahedi, covering \textchapref{{1}} \citep{balanis1989advanced} content.

\paragraph{Cylindrical coorindate wave equation solutions}

Seek a function

\begin{dmath}\label{eqn:emtLecture8:20}
\BE = E_\rho \rhocap + E_\phi \phicap + E_z \zcap
\end{dmath}

solving 

\begin{dmath}\label{eqn:emtLecture8:40}
\spacegrad^2 \BE = -\beta^2 \BE.
\end{dmath}

One way to find the Laplacian in cylindrical coordinates is to use

\begin{dmath}\label{eqn:emtLecture8:60}
\spacegrad^2 \BE = 
\spacegrad \lr{ \spacegrad \cdot \BE }
-\spacegrad \cross \lr{ \spacegrad \cross \BE },
\end{dmath}

where 

\begin{dmath}\label{eqn:emtLecture8:80}
\spacegrad = \rhocap \PD{\rho}{} + \frac{\phicap}{\rho} \PD{\phi}{} + \zcap \PD{z}{}
\end{dmath}

Can be shown that:
\begin{dmath}\label{eqn:emtLecture8:100}
\spacegrad \cdot \BE = \inv{\rho} \PD{\rho}{} \lr{ \rho E_\rho } + \inv{\rho}\PD{\phi}{E_\phi} + \PD{z}{E_z}
\end{dmath}

and
\begin{dmath}\label{eqn:emtLecture8:120}
\spacegrad \cross \BE 
%= 
%\begin{vmatrix}
%\rhocap & \phicap & \zcap \\
%\partial_\rho & \inv{\rho}\partial_\phi & \partial_z \\
%E_\rho & \rho E_\phi & E_z
%\end{vmatrix}
=
\rhocap  \lr{ \inv{\rho} \partial_\phi E_z - \partial_z E_\phi }
+\phicap \lr{ \partial_z E_\rho - \partial_\rho E_z }
+\zcap   \lr{ \inv{\rho} \partial_\rho (\rho E_\phi) - \inv{\rho} \partial_\phi E_\rho }
\end{dmath}

This gives
\begin{dmath}\label{eqn:emtLecture8:200}
\spacegrad^2 \psi = 
\PDSq{\rho}{\psi}
+\inv{\rho} \PD{\rho}{\psi}
+\inv{\rho^2} \PDSq{\phi}{\psi}
+\PDSq{z}{\psi}.
\end{dmath}

and
\begin{dmath}\label{eqn:emtLecture8:220}
\begin{aligned}
\spacegrad^2 E_\rho &= \lr{ -\frac{E_\rho}{\rho^2} - \frac{2}{\rho^2} \PD{\phi}{E_\phi} } \\
\spacegrad^2 E_\phi &= \lr{ -\frac{E_\phi}{\rho^2} + \frac{2}{\rho^2} \PD{\phi}{E_\rho} } \\
\spacegrad^2 E_z    &= -\beta^2 E_\phi.
\end{aligned}
\end{dmath}

%Note that with \( i = \Be_1 \Be_2 \),
%
%\begin{dmath}\label{eqn:emtLecture8:140}
%\rhocap = \Be_1 e^{i \phi}
%\end{dmath}
%
%so
%\begin{equation}\label{eqn:emtLecture8:160}
%\PD{\phi}{\rhocap} = \Be_2 e^{i \phi} = \thetacap
%\end{equation}
%
%... the end result is
%
%\begin{dmath}\label{eqn:emtLecture8:180}
%\end{dmath}

\paragraph{TEM:} If we want to have a TEM mode it can be shown that we need an axial distribution mechanism, such as the core of a co-axial cable.

These are messy to solve in general, but we can solve the z-component without too much pain

\begin{dmath}\label{eqn:emtLecture8:240}
\PDSq{\rho}{E_z}
+\inv{\rho} \PD{\rho}{E_z}
+\inv{\rho^2} \PDSq{\phi}{E_z}
+\PDSq{z}{E_z}
=
-\beta^2 E_z
\end{dmath}

Solving this using separation of variables with

\begin{dmath}\label{eqn:emtLecture8:260}
E_z = R(\rho) P(\phi) Z(z)
\end{dmath}

\begin{dmath}\label{eqn:emtLecture8:280}
\inv{R}\lr{R'' + \inv{\rho} R'} + \inv{\rho^2 P} P'' + \frac{Z''}{Z} = -\beta^2
\end{dmath}

Assuming for some constant \( \beta_z \) that we have
\begin{dmath}\label{eqn:emtLecture8:300}
\frac{Z''}{Z} = -\beta_z^2,
\end{dmath}

then

\begin{dmath}\label{eqn:emtLecture8:320}
\inv{R}\lr{\rho^2 R'' + \rho R'} + \inv{P} P'' + \rho^2 \lr{\beta^2 - \beta_z^2} = 0
\end{dmath}

Now assume that
\begin{dmath}\label{eqn:emtLecture8:340}
\inv{P} P'' = -m^2,
\end{dmath}

and let \( \beta^2 - \beta_z^2 = \beta_\rho^2 \), which leaves

\begin{dmath}\label{eqn:emtLecture8:360}
\rho^2 R'' + \rho R' + \lr{ \rho^2 \beta_\rho^2 -m^2 } R = 0.
\end{dmath}

This is the Bessel differential equation, with travelling wave solution

\begin{dmath}\label{eqn:emtLecture8:380}
R(\rho) = 
A H_m^{(1)}(\beta_\rho \rho) 
+B H_m^{(2)}(\beta_\rho \rho),
\end{dmath}

and standing wave solutions
\begin{dmath}\label{eqn:emtLecture8:400}
R(\rho) = 
A J_m(\beta_\rho \rho)
+B Y_m(\beta_\rho \rho).
\end{dmath}

Here \( H_m^{(1)}, H_m^{(2)} \) are Hankel functions of the first and second kinds, and
\( J_m, Y_m \) are the Bessel functions of the first and second kinds.

For \( P(\phi) \) 
\begin{dmath}\label{eqn:emtLecture8:460}
P'' = -m^2 P
\end{dmath}

\paragraph{Quadropole potential}

In Jackson
\citep{jackson1975cew}
, the prove of which is scattered through chapter 3, is the following

\begin{dmath}\label{eqn:emtLecture8:420}
\inv{\Abs{\Bx - \Bx'}}
= 
4 \pi \sum_{l= 0}^\infty \sum_{m = -l}^l \inv{2 l + 1} \frac{(r')^l}{r^{l+1}} 
Y^\conj_{l,m}(\theta', \phi')
Y_{l,m}(\theta, \phi),
\end{dmath}

where \( Y_{l,m} \) are the spherical harmonics.  Plugging this into the potential we have

\begin{dmath}\label{eqn:emtLecture8:440}
\phi(\Bx) 
= \inv{4 \pi \epsilon_0} \int \frac{\rho(\Bx') d^3 x'}{\Abs{\Bx - \Bx'}}
= 
\inv{4 \pi \epsilon_0} \int \rho(\Bx') d^3 x' \lr{
4 \pi \sum_{l= 0}^\infty \sum_{m = -l}^l \inv{2 l + 1} \frac{(r')^l}{r^{l+1}} 
Y^\conj_{l,m}(\theta', \phi')
Y_{l,m}(\theta, \phi)
}
= 
\inv{\epsilon_0} 
\sum_{l= 0}^\infty \sum_{m = -l}^l \inv{2 l + 1} 
\int \rho(\Bx') d^3 x' \lr{
\frac{(r')^l}{r^{l+1}} 
Y^\conj_{l,m}(\theta', \phi')
Y_{l,m}(\theta, \phi)
}
= 
\inv{\epsilon_0} 
\sum_{l= 0}^\infty \sum_{m = -l}^l \inv{2 l + 1} 
\lr{ 
\int (r')^l \rho(\Bx') 
Y^\conj_{l,m}(\theta', \phi')
d^3 x' 
}
\frac{
Y_{l,m}(\theta, \phi)
}
{
r^{l+1}
}
\end{dmath}

The integral terms are called the coefficients of the multipole moments, denoted
\begin{dmath}\label{eqn:emtLecture8:480}
q_{l,m} = 
\int (r')^l \rho(\Bx') 
Y^\conj_{l,m}(\theta', \phi')
d^3 x',
\end{dmath}

The \( l = 0,1,2\) terms are, respectively, called the monopole, dipole, and quadropole terms of the potential
\begin{dmath}\label{eqn:emtLecture8:500}
\rho(\Bx) =
\inv{4 \pi \epsilon_0} 
\sum_{l= 0}^\infty \sum_{m = -l}^l \frac{4\pi} {2 l + 1} 
q_{l,m}
\frac{
Y_{l,m}(\theta, \phi)
}
{
r^{l+1}
}.
\end{dmath}

Note the power of this expansion.  Should we wish to compute the electric field, we have only to compute the qradient of  the last (\(Y_{l,m} r^{-l-1} \)) portion (since \( q_{l,m} \) is a constant).
 
\begin{dmath}\label{eqn:emtLecture8:520}
q_{1,1}
= 
-\int \sqrt{\frac{3}{8 \pi}} \sin\theta' e^{-i\phi'} r' \rho(\Bx') dV'
=
-\sqrt{\frac{3}{8 \pi}} \int \sin\theta' \lr{ \cos\phi' - i\sin\phi'} r' \rho(\Bx') dV'
=
-\sqrt{\frac{3}{8 \pi}} \lr{ 
\int x' \rho(\Bx') dV'
-i \int y' \rho(\Bx') dV'
}
=
-\sqrt{\frac{3}{8 \pi}} \lr{ 
p_x - i p_y
}.
\end{dmath}

Here we've used
\begin{dmath}\label{eqn:emtLecture8:540}
\begin{aligned}
x' &= r' \sin\theta' \cos\phi' \\
y' &= r' \sin\theta' \sin\phi' \\
z' &= r' \cos\theta'
\end{aligned}
\end{dmath}

and the \( Y_{11} \) representation

\begin{dmath}\label{eqn:emtLecture8:560}
\begin{aligned}
Y_{00} &= -\sqrt{\frac{1}{4 \pi}} \\
Y_{11} &= -\sqrt{\frac{3}{8 \pi}} \sin\theta e^{i\phi} \\
Y_{10} &=  \sqrt{\frac{3}{4 \pi}} \cos\theta  \\
Y_{22} &= -\inv{4} \sqrt{\frac{15}{2 \pi}} \sin^2\theta e^{2 i\phi} \\
Y_{21} &=  \inv{2} \sqrt{\frac{15}{2 \pi}} \sin\theta \cos\theta e^{i\phi} \\
Y_{20} &=  \inv{4} \sqrt{\frac{5}{\pi}} \lr{ 3 \cos^2\theta - 1 } \\
\end{aligned}
\end{dmath}

%HW: compute a few of the more tedious moment coeffients.  These have been exam questions in the past.

With the usual dipole moment expression

\begin{dmath}\label{eqn:emtLecture8:580}
\Bp = \int \Bx' \rho(\Bx') d^3 x',
\end{dmath}

and a quadropole moment defined as
\begin{dmath}\label{eqn:emtLecture8:600}
Q_{i,j} = \int \lr{ 3 x_i' x_j' - \delta_{ij} (r')^2 } \rho(\Bx') d^3 x',
\end{dmath}

the first order terms of the potential are now fully specified
\begin{dmath}\label{eqn:emtLecture8:620}
\phi(\Bx)
=
\inv{4 \pi \epsilon_0}
\lr{ 
q + \frac{\Bp \cdot \Bx}{r^3} + 
\inv{2} \sum_{ij} Q_{ij} \frac{x_i x_j}{r^5}
}.
\end{dmath}

\paragraph{Waves}

\begin{itemize}
\item The field is a modification of space-time
\item Mode is a particular field configuration for a given boundary value problem.  Many field configurations can satisfy Maxwell equations (wave equation).  These usually are referrred to as modes.  A mode is a self-consistent field distribution.
\item In a TEM mode, \( \BE \) and \( \BH \) are every point in space are constrained in a local plane, independent of time.  This plane is called the equiphase plane.  In general equiphase planes are not parallel at two different points along the trajectory of the wave.
\item If equiphase planes are parallel (i.e. the space orientation of the planes for TEM mode...
... next time.
\end{itemize}

%}
\EndArticle
%\EndNoBibArticle
