%
% Copyright © 2016 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\input{../blogpost.tex}
\renewcommand{\basename}{emt10}
\renewcommand{\dirname}{notes/ece1228/}
\newcommand{\keywords}{ECE1228H}
\input{../peeter_prologue_print2.tex}

%\usepackage{ece1228}
\usepackage{peeters_braket}
%\usepackage{peeters_layout_exercise}
\usepackage{peeters_figures}
\usepackage{mathtools}
\usepackage{siunitx}

\beginArtNoToc
\generatetitle{ECE1228H Electromagnetic Theory.  Lecture 10: Fresnel relations.  Taught by Prof.\ M. Mojahedi}
%\chapter{Fresnel relations}
\label{chap:emt10}

%\paragraph{Disclaimer}
%
%Peeter's lecture notes from class.  These may be incoherent and rough.
%
%These are notes for the UofT course ECE1228H, Electromagnetic Theory, taught by Prof. M. Mojahedi, covering \textchapref{{1}} \citep{balanis1989advanced} content.
%
\paragraph{Single interface TE mode.}

We considered a Fresnel reflection geometry last class with \( \BE \) parallel to the interface

F1

\begin{dmath}\label{eqn:emtLecture10:20}
   \BE_i = \ycap E_i e^{j \omega t - j \Bk_{1i} \cdot \Bx },
\end{dmath}

with an assumption that this field maintains it's polarization in both its reflected and transmitted components, so that

\begin{dmath}\label{eqn:emtLecture10:40}
   \BE_i = \ycap r E_i e^{j \omega t - j \Bk_{1r} \cdot \Bx },
\end{dmath}

and
\begin{dmath}\label{eqn:emtLecture10:60}
   \BE_t = \ycap t E_i e^{j \omega t - j \Bk_{2t} \cdot \Bx },
\end{dmath}

The end result is

\begin{subequations}
\label{eqn:emtLecture10:80}
\begin{dmath}\label{eqn:emtLecture10:100}
r = \frac{\mu_2 k_{1z} - \mu_1 k_{2z}}{\mu_2 k_{1z} + \mu_1 k_{2z}}
\end{dmath}
\begin{dmath}\label{eqn:emtLecture10:120}
t = \frac{2 \mu_2 k_{1z}}{\mu_2 k_{1z} + \mu_1 k_{2z}}
\end{dmath}
\end{subequations}

\paragraph{Two interface problem.}

It turns out that light passing through two interfaces ends up with reflected and transmitted components, but no mode that goes through the slab in a wave guide fashion.  Such a mode is possible when the light is incident in an end-fire configuration (i.e. a waveguide).  It's interesting that such a wave guide like mode isn't possible.

Consider a geometry with interfaces at \( z = 0, d \).  

The incident, and reflected wave components in the first medium are
\begin{dmath}\label{eqn:emtLecture10:140}
   A e^{-j k_{1z} z},
\end{dmath}
\begin{dmath}\label{eqn:emtLecture10:160}
   r A e^{j k_{1z} z},
\end{dmath}

in the second medium 

\begin{dmath}\label{eqn:emtLecture10:180}
   C e^{-j k_{1z} z},
\end{dmath}
\begin{dmath}\label{eqn:emtLecture10:200}
   D e^{j k_{1z} z},
\end{dmath}

and in the final medium
\begin{dmath}\label{eqn:emtLecture10:220}
   A t e^{-j k_{1z} (z-d)},
\end{dmath}

\begin{itemize}
   \item \( z = 0 \).
      % FIXME
   %A r_{12} = C + D
   \item \( z = d \).
      % FIXME
\end{itemize}
These relationships can be assembled into matrix form. The end result is

FIXME.

Something important to consider is that the index of refraction can be complex valued

\begin{dmath}\label{eqn:emtLecture10:240}
n(\omega) = n'(\omega) + j n''(\omega),
\end{dmath}

but all of these derivations are independent of this.

\paragraph{special case.}

If medium 1 and 3 are identical then

\begin{dmath}\label{eqn:emtLecture10:260}
r_{21} = r_{23} = -r_{12},
\end{dmath}

which gives

\begin{dmath}\label{eqn:emtLecture10:280}
   t^{\textrm{TE}} = \frac{t_{12} t_{21} e^{j \phi}}{1 - r_{21}^2 e^{2 j \phi}}
\end{dmath}
\begin{dmath}\label{eqn:emtLecture10:300}
   r^{\textrm{TE}} = r_{12} + \frac{t_{12} t_{21} r_{21} e^{2 j \phi}}{1 - r_{21}^2 e^{2 j \phi}}
\end{dmath}

where
\begin{dmath}\label{eqn:emtLecture10:320}
   \phi = -k_{2z} d = - \frac{\omega}{c} n_2 \cos\theta_2 d
   % \cos(\theta_2 d)?
\end{dmath}

It's possible to design the material so that there is no reflection from the slab, called the matched condition.

This requires

\begin{dmath}\label{eqn:emtLecture10:340}
r_{21} = \frac{-\mu_2 k_{1z} + \mu_1 k_{2z}}{\mu_2 k_{1z} + \mu_1 k_{2z}} = -r_{12}.
\end{dmath}

To force this to be zero, one can set
\begin{dmath}\label{eqn:emtLecture10:360}
\mu_2 k_{1z} = \mu_1 k_{2z},
\end{dmath}

or
\begin{dmath}\label{eqn:emtLecture10:380}
   \mu_2 \frac{\omega}{c} n_1 \cos \theta_1 =
   \mu_1 \frac{\omega}{c} n_2 \cos \theta_2
\end{dmath}

or
\begin{dmath}\label{eqn:emtLecture10:400}
   \mu_2 \sqrt{\epsilon_1 \mu_1} \cos \theta_1 = \mu_1 \sqrt{\epsilon_2 \mu_2} \cos \theta_2,
\end{dmath}

which is
\begin{dmath}\label{eqn:emtLecture10:420}
   \sqrt{\epsilon_1/\mu_1} \cos \theta_1 = \sqrt{\epsilon_2/\mu_2} \cos \theta_2,
\end{dmath}

or
\begin{dmath}\label{eqn:emtLecture10:440}
   \eta_1 \cos \theta_1 = \eta_2 \cos \theta_2.
\end{dmath}

At normal incidence, this reduces to

\begin{dmath}\label{eqn:emtLecture10:460}
   \eta_1 = \eta_2.
\end{dmath}

What happens to the transmission coefficient for such a slab.

One can find 

\begin{equation}\label{eqn:emtLecture10:480}
t_{12} = t_{21} = 1,
\end{equation}

under a matched condition, so

\begin{equation}\label{eqn:emtLecture10:500}
   t = t_{12} t_{21} e^{j \phi}.
\end{equation}

The matched slab only introduces a phase delay (at the specific frequency for which the slab is matched).

\paragraph{Group delay}

Under matched conditions where \( t^{\textrm{TE}} = e^{j \phi} = e^{j k_{2z} d} \), we can write

\begin{dmath}\label{eqn:emtLecture10:520}
- \PD{\phi}{\omega} 
= \PD{\omega}{} \lr{ k_{2z} d }
= d \PD{\omega}{k_{2z}}
= d /\PD{k_{2z}}{\omega}
= d / v_g,
\end{dmath}

so 

\begin{dmath}\label{eqn:emtLecture10:540}
v_g = -\frac{d}{\PD{\phi}{\omega}}.
\end{dmath}
%FIXME: last relation?

This is called the \textAndIndex{group delay}.

This is different than the phase velocity \( v_p = \omega/k \), as illustrated in the sketch of 

F2

If (FIXME: why?) 

\begin{dmath}\label{eqn:emtLecture10:n}
v_g
= \PD{k}{\omega}
= 1/\PD{\omega}{k},
\end{dmath}

and in an unbounded medium
\begin{dmath}\label{eqn:emtLecture10:560}
   k = \frac{\omega}{c} n(\omega),
\end{dmath}

so
\begin{dmath}\label{eqn:emtLecture10:580}
\PD{\omega}{k} 
= 
\inv{c} \lr{ n(\omega) + \omega \PD{\omega}{n} },
\end{dmath}

so
\begin{equation}\label{eqn:emtLecture10:600}
v_g 
= \frac{c}{n(\omega) + \omega \PD{\omega}{n} }
= \frac{c}{n_g}
\equiv
\frac{d}{\tau_g}.
\end{equation}

%}
%\EndArticle
\EndNoBibArticle
